### 2021-04-01 publication/manuscript.Rmd
inital commit

---
title: Regularized target encoding outperforms traditional methods in supervised machine learning with high cardinality features
author:
  - name: Florian Pargent
    email: florian.pargent@psy.lmu.de
    affiliation: LMU Munich 1
    footnote: 1
  - name: Florian Pfisterer
    #email: florian.pfisterer@stat.uni-muenchen.de
    affiliation: LMU Munich 2
  - name: Janek Thomas
    #email: janek.thomas@iis.fraunhofer.de
    affiliation: Fraunhofer
  - name: Bernd Bischl
    #email: bernd.bischl@stat.uni-muenchen.de
    affiliation: LMU Munich 2
address:
  - code: LMU Munich 1
    address: Department of Psychology, Psychological Methods and Assessment, LMU Munich, Leopoldstraße 13, 80802 Munich, Germany
  - code: LMU Munich 2
    address: Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstraße 33, 80539 Munich, Germany
  - code: Fraunhofer
    address: Fraunhofer Institute for Integrated Circuits, Hansastraße 32, 8068 Munich, Germany
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  Because most machine learning (ML) algorithms are designed for numerical inputs, efficiently encoding categorical variables is a crucial aspect during data analysis. An often encountered problem are high cardinality features, i.e. unordered categorical predictor variables with a high number of levels. We study techniques that yield numeric representations of categorical variables which can then be used in subsequent ML applications. We focus on the impact of those techniques on a subsequent algorithm's predictive performance, and -- if possible -- derive best practices on when to use which technique. We conducted a large-scale benchmark experiment, where we compared different encoding strategies together with five ML algorithms (lasso, random forest, gradient boosting, k-nearest neighbours, support vector machine) using datasets from regression, binary- and multiclass- classification settings. Throughout our study, regularized versions of target encoding (i.e. using target predictions based on the feature levels in the training set as a new numerical feature) consistently provided the best results. Traditional encodings often make unreasonable assumptions to map levels to integers (e.g. integer encoding) or reduce the number of levels (possibly based on target information, e.g. leaf encoding) before creating binary indicator variables (one-hot or dummy encoding) were not as effective.
keywords: supervised machine learning, benchmark, high-cardinality categorical features, target encoding, impact encoding, dummy encoding

journal: "Computational Statistics and Data Analysis"
date: "`r Sys.Date()`"
bibliography: literature.bib
#linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
output:
  #rticles::elsevier_article: default
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
#output: rticles::elsevier_article

header-includes:
   - \usepackage{algorithm}
   - \usepackage{array}
   - \usepackage[noend]{algpseudocode}
   - \usepackage{amsmath}
   - \usepackage{amssymb}
   - \usepackage{amsfonts}
   - \usepackage{mathtools}
   - \usepackage{bm}
   - \usepackage{tikz}
   - \usepackage{threeparttable}
   - \usetikzlibrary{trees,arrows, chains, fit, positioning, calc, shapes, shadows}
   - \setlength\parindent{0pt}
   - \usepackage{pdflscape}
---

<!--
Possible titles:
- Target encoding outperforms traditional methods in supervised machine learning with high cardinality features
- Experimental evaluation of categorical encoding techniques
- Encoding Categorical Variables -- An empirical study
- A Benchmark Experiment on How to Encode High Cardinality Features in Supervised Machine Learning
- Encoding Categorical Variables -- An empirical study
- Your algorithm can not handle categorical variables? 10 solutions you would have never thought of.
- You will not believe how these categorical variable encoding techniques perform in practice!
-->

```{r setup, include=FALSE}
options(scipen=3)
```

```{r analysis-preferences, include=FALSE}
# Seed for random number generation (use old rng to be identical with master thesis)
RNGkind(sample.kind = "Rounding")
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed, echo = FALSE)
options(knitr.kable.NA = " ")
# options(papaja.na_string = "")
```

```{r colour defaults, include=FALSE}
library(RColorBrewer)
box_colors = c("#F8766D", "#D89000", "#A3A500", "#39B600", "#00BF7D", "#00BFC4", "#00B0F6", "#9590FF", "#E76BF3", "#FF62BC", "#808080", "#000000")
#box_colors = c(brewer.pal(9, "Set1"), "#778877", "#808080", "#000000")
```

Introduction
============

While increasing sample size is usually considered the most important step to improve the predictive performance of a machine learning (ML) model, using effective feature engineering comes as a close second. One remaining challenge is how to handle high cardinality features -- categorical predictor variables with a high number of different levels but without any natural ordering. While categorical variables with only a small number of possible levels can often be efficiently dealt with using standard techniques such as one-hot encoding, this approach becomes inefficient as the number of levels increases. Although domain knowledge can sometimes be used to reduce the number of theoretically relevant levels, finding strategies that work well on a large variety of problems is highly important for many applications as well as in automated ML [@thomas_2018; @thornton_2013; @feurer_2015]. Optimally, strategies should be model-agnostic, as benchmarking encoding methods together with ML algorithms from different classes is often necessary for applications. While a variety of strategies exists, there are very few benchmarks that can be used to decide which technique is expected to yield good predictive performance. Furthermore, there has recently been increasing attention on scientific benchmark studies that compare different methods to provide a clearer picture in light of a large number of methods available to practitioners [@delgado; @bommert], as they can provide at least partial answers to such questions. The goal of this study is to provide an overview of existing approaches for encoding categorical variables and to study their effect on a model's predictive performance.

## Notation

We consider the classical setting of supervised learning from an $i.i.d$ tabular dataset $\mathcal{D}$ of size $N$ sampled from a joint distribution $\mathcal{P}(\bm{x}, y)$ of a set of features $\bm{x}$ and an associated target variable $y$.
Here, $\bm{x}$ consists of a mix of numeric (real-valued or integer-valued) features and categorical features, the latter of which we seek to transform feature-wise to numeric features using a categorical encoding technique.
Let $x$ be a single unordered categorical feature from a feature space $\mathcal{X}$ with cardinality $card(\mathcal{X}) \leq card(\mathbb{N})$.
It holds either $y \in \mathbb{R}$ (regression), $y \in \mathcal{C}$ from a finite class space $\mathcal{C} = \{c_1, ..., c_C\}$ with $C = 2$ (binary classification) or $C > 2$ (multiclass classification). We always assume to observe all $C$ classes in our training sample, however we might only observe a subset $\mathcal{L}^{train} \subseteq \mathcal{X}$ of a features' available $L$ levels, $\mathcal{L}^{train} = \{l_1, ..., l_L\}$ for categorical features. We denote the observed frequency of class $c$ in the training set with $N_c$ and the observed frequency of a level $l$ in the training set with $N_l$.
As most ML algorithms require numerical features, we use categorical encoding techniques to transform each nominal feature $x^{train}$ into numerical features  $\hat{x}^{train}$ which are then used for training. If clear from the context, we use $\hat{x}_l$ as the encoded value for observation with level $l$. Although datasets might contain multiple high cardinal features, we encode each feature separately but using the same strategy.

## Related Work

We broadly categorize feature encoding techniques into \textit{target-agnostic} methods and methods that incorporate target statistics [@micci_barreca_2001].
Figure \ref{fig:overview} contains a taxonomy of the encoding methods we consider in this work. \textit{Target-agnostic} methods do not rely on any information about the target variable and can therefore also be used in unsupervised settings. Simple strategies from this domain e.g. one-hot or dummy encoding are widely used to embed variables for classical ML algorithms as well as (deep) neural networks in the scientific literature [@kuhn_2019; @hancock] but also on  Kaggle\footnote{\url{https://www.kaggle.com/}}.
Such indicator methods map each level of a categorical variable to a set of dichotomous features encoding the presence or absence of a particular level.
An obvious drawback of indicator encoding is that it adds one additional feature per level of a categorical variable.
When the number of levels increases to a point where indicator encoding leads to an unreasonable number of features, levels are often simply mapped to integer values with random order (integer encoding). Alternatively, the "hashing trick" [@weinberger_2009] can be used to randomly collapse feature levels into a smaller number of indicator variables [@kuhn_2019], or levels can be encoded by using the observed frequency of a given level in the dataset (frequency encoding).

\tikzset{
  basic/.style  = {draw, text width=4cm, drop shadow, font=\sffamily \scriptsize, rectangle},
  root/.style   = {basic, rounded corners=2pt, thin, align=center, fill=white},
  level-2/.style = {basic, rounded corners=5pt, thin,align=center, fill=white, text width=2.3cm},
  level-3/.style = {basic, thin, align=center, fill=white, text width=1.8cm},
  level-3-focus/.style = {basic, thick, align=center, fill=white, text width=1.8cm}
}
\begin{figure}[ht]
  \centering
  \begin{tikzpicture}[
    level 1/.style={sibling distance=8em, level distance=3em},
    edge from parent/.style={->,solid,black,thick,draw},
    edge from parent path={(\tikzparentnode.south) -- (\tikzchildnode.north)},
    >=latex, node distance=0.8cm, edge from parent fork down]
    \node[root] {\textbf{Taxonomy of categorical encoding techniques}}
      child {node[level-2, xshift=-10pt] (c1) {\textbf{Target-agnostic}}}
      child {node[level-2, xshift=10pt] (c2) {\textbf{Target-based}}};

    \begin{scope}[every node/.style={level-3}]
      \node [below of = c1, xshift=5pt] (c11) {indicator: dummy, one-hot};
      \node [below of = c11] (c12) {integer};
      \node [below of = c12] (c13) {frequency};
      \node [below of = c13] (c14) {hash};

      \node [below of = c2, xshift=5pt](c21) {leaf};
      \node [below of = c21] (c22) {impact};
      \node [below of = c22] (c23) {regularized impact};
    \end{scope}

    \foreach \value in {1,2,3,4}
      \draw[->] (c1.192) |- (c1\value.west);

    \foreach \value in {1,...,3}
      \draw[->] (c2.192) |- (c2\value.west);
  \end{tikzpicture}
  \caption{Taxonomy of common categorical variable encoding techniques.}
  \label{fig:overview}
\end{figure}

Target-based methods try to incorporate information about the target values associated with a given level.
Early strategies aimed to reduce the number of levels by methods like hierarchical clustering or decision trees based on statistics of the target variable, although this has been rarely described in the scientific literature [for a short mention, see @micci_barreca_2001].
The basic idea of more advanced methods called target, impact, mean, or likelihood encoding is to use the training set to make a simple prediction of the target for each level of the categorical feature, and to use the prediction as to the numerical feature value $\hat{x}_l$ for the respective level.
One of the earliest formal descriptions of this strategy is @micci_barreca_2001.
A very similar approach is called *weights of evidence* in the credit scoring literature, see @hand_1997.
In basic target encoding for regression problems, the mean target value in the training set from all observations with a certain feature level is used as the numeric value to encode that level for all observations: $\hat{x}_l = \frac{\sum_{i:x^{train}_i = l}y^{train}_i}{N_l}$.
Simple target encoding often does not perform well with rare levels, where it tends to overfit to the training data and fail to generalize well for new observations.
In the extreme case of a categorical feature with unique values (e.g. some hashed ID variable) studied in @prokhorenkova_2018, the mean target for each level of this feature in simple target encoding is similar to the true target value of a single observation.
Based on the encoded feature, all observations can be predicted perfectly in the training set, even if the original variable did not contain any useful information.
ML models would place a high priority on such an encoded feature during training but would perform badly on test data.
To avoid this, practitioners often use regularized target encoding by adding a smoothing parameter $\epsilon$ which shrinks those effects towards the global mean [@micci_barreca_2001].
An alternative strategy is to combine target encoding with cross-validation (CV) techniques [@prokhorenkova_2018].

## Previous Encoding Benchmarks

Several small-scale studies that investigate aspects of our research questions have been previously conducted.
However, those works do not yield conclusive results due to narrower or different scopes or not considering high cardinality variables.
One benchmark ($6$ datasets) on encoding high cardinality features (cardinality between $103$ and $9095$) in combination with gradient boosting has been published on the Kaggle forums [@prokopev_2018].
Different versions of target encoding are compared with indicator, integer, and frequency encoding. They recommend combining the smoothed version of target encoding with 4 or 5-fold CV, never using simple target encoding and use indicator encoding only for small datasets.
Interestingly, frequency encoding did perform well in many cases.
@coors_2018 performed a benchmark ($12$ datasets) on encoding high cardinality features (maximum number of levels per dataset between 10 and 25847) when developing the automatic gradient boosting (autoxgboost) library [@thomas_2018]. They compared different variants of target encoding with integer and indicator encoding. In their benchmarks, target encoding only improved over target-agnostic methods on $2$ datasets, while it led to worse results on 4 datasets. For a smaller number of levels, indicator and integer encoding yielded similar results.
As those studies only consider gradient boosting and a limited amount of datasets, it is unclear whether results generalize to other datasets and ML algorithms.
Several other publications studied encoding text data based on similarity [@cerda_2018; @cerda_2020] and employed indicator or target encoding as baselines.
Another line of work studies variable encodings employed within specific ML models.
@wright_2019 study treatments for categorical variables in random forests together with dummy and integer encoding (18 datasets, cardinality between $3$ and $38$), concluding that indicator and integer encoding perform subpar in comparison to methods that re-order levels according to the target variables.
@prokhorenkova_2018 compared their new CatBoost variant of target encoding to smoothed target encoding with no CV, hold-out, and leave-one-out CV on $8$ datasets.
While the CatBoost method performed best, hold-out came second and target encoding without CV performed worst.
An overview of encoding techniques tailored to be used together with Neural Networks is provided in [@hancock].
Amongst others, they survey indicator-based methods as well as embedding-based methods but do not conduct a benchmark study.
One widely adopted approach, in particular, has been proposed by [@guo_2016] under the term \textit{entity embeddings}.
In contrast, our work focuses on high cardinality variables and techniques that are agnostic to the subsequent ML method, which has not been done before.
We furthermore study this problem on a sufficiently large variety of datasets and settings to allow for more compelling conclusions.
\newline

The main goal of our study is to assess the impact different categorical encoding techniques have on subsequent models predictive performance.
As the optimal encoding might differ depending on the ML algorithm, we consider various state-of-the-art algorithms, \textit{regularized linear models}, \textit{random forests}, \textit{gradient tree boosting}, \textit{k-nearest neighbors}, and \textit{support vector machines}.
To find default settings for high cardinality features, we analyse a variety of datasets with different characteristics including regression, binary classification and multiclass classification problems.
As our methods vary in runtime and complexity, we aim to learn whether more complex methods, in general, are to be preferred, or if simpler approaches suffice.
Furthermore, the relationship between a feature's cardinality and the choice of encoding technique has not been investigated, which we study by varying the minimum number of levels above which features are transformed.

## Contributions

We first survey various categorical encoding techniques and conduct a comprehensive benchmark study of a broad set of categorical encoding techniques with a focus on high cardinality features.
We study $7$ different encoding techniques in conjunction with $5$ commonly used ML algorithms across $24$ datasets, both from a classification and a regression regime. We give a detailed description of our study design to highlight important considerations for studying high cardinality features.
Our results provide an overview of the performance of various approaches heavily used in the literature. After a discussion of results concerning predictive performance, we provide further analyses seeking to inform practitioners which methods to apply. This includes an important discussion on runtimes.

<!--
how many different encoders do we count?
7? 10? 11?
-->

Encodings
=========
\label{sec:methods}

Pseudocode is presented in the main text (for non-standard encodings), or the Supplementary Material (for standard methods).
An important detail is how encoding techniques treat new levels during the prediction phase, as this is often not obvious.

## Integer Encoding

The simplest strategy for categorical features is integer encoding, which is also often referred to as ordinal encoding.
To make sure that the order of the observed levels does not affect our results, we randomly map the observed levels from the training set to the integers $1$ to $L$.
Although new levels could be mapped to $L+1$ or $0$, model predictions would be completely arbitrary as the order of the integers does not contain any information.
Thus, we encode new levels as missing values and use mode imputation to obtain the integer which matches the most frequent level in the training set.
Integer encoding should be an acceptable strategy for tree-based models, which can separate all original levels with repeated splits, but less so for other algorithms.

## Frequency Encoding

Frequency encoding simply maps each level to its observed frequency in the training set ($\hat{x}_l = N_l$).
On the one hand, this assumes a functional relationship between the frequency of a level and the target.
On the other hand, it implicitly reduces the number of levels, as the subsequent model can best differentiate between levels with dissimilar frequencies. This approach is typically used e.g. in natural language processing to encode token or n-gram counts.
We encode new levels with a frequency of $1$, although choosing the minimal observed frequency for any level during training might be another reasonable strategy.

## Indicator Encoding

We use indicator encoding as an umbrella term for the two most common strategies to encode categorical features with a small to moderate the number of levels: one-hot and dummy encoding.
One-hot encoding transforms the original feature into $L$ binary indicator columns, each representing one original level.
An observation is coded with $1$ for the indicator column representing its level ($x^{train}_i = l$) and $0$ for all other indicators.
Dummy encoding results in only $L-1$ indicator columns.
A reference feature level is chosen that is encoded with $0$ in all indicator columns.
For one-hot encoding, the zero vector can be used to encode new levels which were not observed during training.
For dummy encoding, this is not useful as it collapses new levels to the (often arbitrary) reference category.
In our study, the first level in alphabetical order is used as the reference category for dummy encoding.
We replace new levels in the prediction phase with the most frequent level in the training set.
In datasets with a high total number of feature levels, constructing all indicator variables can be detrimental to the predictive performance or at least increase the computational load.
As indicator encoding is practically infeasible for high cardinality variables (a feature with $10^5$ levels would add $10^5$ new columns), we limit the number of indicator coded features by collapsing rare levels beyond a varying threshold to a single \textit{other} category before encoding.
Closely related to dummy encoding in classical analysis of variance and regression are Helmert and Polynomial contrasts [@chambers_1992].
However, these encodings focus on ordinal categorical features, which we did not consider in our study.

## Hash Encoding

Hash Encoding can be used to compute indicator variables based on a hash function [@weinberger_2009].
The basic idea is to transform each feature level $l$ to an integer $hash(l) \in \mathbb{N}$, based on its label.
This integer is then transformed to an indicator representation, with $1$ in indicator column number $(hash(l) \mod hash.size) + 1$ and $0$ in all remaining columns [@kuhn_2019].
Some levels will be hashed to the same indicator representation.
The smaller the $hash.size$, the higher the number of collapsed levels.
The number of indicators is often effectively lower than $hash.size$, as some indicators can be constant in the training set (we remove those columns for both training and prediction).
Although hashing would allow us to jointly hash multiple features, we hash each feature separately to improve comparability with the other encoders.

## Leaf Encoding

In leaf encoding, a decision tree is fitted on the training set to predict the target based on the categorical feature.
Each level is encoded by the number of the terminal node, in which an observation with the respective level ends up.
In that way, leaf encoding combines feature levels with similar target values.
We use the rpart package in R [@therneau_2018] which grows CARTs with categorical feature support that can be pruned based on internal performance estimates from 10-fold CV.
Thus, our leaf encoder automatically uses an "optimal" number of new levels.
To speed up the computation for multiclass classification, our implementation uses the ordering approach presented in [@wright_2019].
New levels are encoded with the arbitrary number of the terminal node with most observations during training.
Encoded values are treated as a new categorical feature and encoded by one-hot encoding.
Our leaf encoder can be thought of as a simplification of the approach suggested by [@grabczewski_2003].

\begin{algorithm}
  \caption{Leaf Encoding}
  \label{alg:leaf}
  \begin{algorithmic}
    \State \underline{Training:} require number of cross-validation folds $K \in \mathbb{N}$
    \State fit CART tree on $\mathcal{D}^{train}$ with complexity pruning based on $K$-fold cross-validation
      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
        $\tilde{x}_i = t$ with $x^{train}_i$ in terminal node $t$
      \EndFor
    \State \underline{Prediction:}
    \For{$x^{new}$}
        \If{$x^{new} \in \mathcal{L}^{train}$}\ {$\tilde{x}^{new} = t$ with $x^{new}$ in terminal node $t$}
        \Else\ {$\tilde{x}^{new} = b$ where $b$ indicates the biggest terminal node}
        \EndIf
    \EndFor
  \end{algorithmic}
\end{algorithm}

## Impact Encoding

The first formal description of a so-called impact or target encoder was provided by @micci_barreca_2001.
The basic idea is to encode each feature level with the conditional target mean (regression) or the conditional relative frequency of one or more target classes (classification).
The impact encoder for classification uses a logit link and transforms the original feature into $C$ numeric features, each representing one target class.
A smoothing parameter $\epsilon$ is introduced to avoid division by zero.
This parameter could be further used to regularize towards the unconditional mean.
To allow comparing regularized (introduced below) with simple target encoding, we always choose a small $\epsilon = 0.0001$ for the impact encoder.
Similar techniques have been proposed in the literature:
Weights of evidence encoding from the credit scoring classification literature [@hand_1997] are almost identical to impact encoding, but does not
consider any regularization or centring.
Impact encoding is also referred to as a target or James-Stein encoding.

## GLMM Encoding

Smoothed target encoding [@micci_barreca_2001] can be interpreted as a simple (generalized) linear mixed model (glmm) in which the target is predicted by a random intercept for each feature level in addition to a fixed global intercept.
This connection is described in [@kuhn_2019].
To achieve regularized impact encoding, we constructed our own glmm encoders for regression, binary, and multiclass classification which are described in Algorithm \ref{alg:glmm-regr} (regression) and in the Supplementary Material (classification).
The encoded value for each level is based on the spherical conditional mode estimates.
In regression, the conditional modes are similar to the mean target value for each level, weighted by the relative observed frequency of that level in the training set [@gelman_2006]. For technical details, see @bates_2018.
Additionally, the estimate of the fixed intercept can be used during the prediction phase to encode new feature levels that were not observed in the training set.
In multiclass classification, we fit $C$ one vs. rest glmms resulting in one encoded feature per class.
An important advantage of using a glmm over impact encoding with a smoothing parameter is that a reasonable amount of regularization is determined automatically and tuning the complete ML pipeline is not necessary.

A further strategy to avoid overfitting in target encoding is to combine it with CV to train the encoder on independent observations without limiting the data to train the ML model.
To our knowledge, we provide the first implementation which combines target encoding based on glmms with CV.
During the training phase, we partition the data using CV into $n.folds$ and fit a glmm on each resulting training set.
For each observation, there is exactly one glmm that did not use that observation for model fitting and can be safely used for encoding.
Note that the $n.folds$ CV models (for $n.folds$ > 1) are only used during the training phase.
In the prediction phase, feature values are always encoded by a single glmm fitted to the complete training set.
We study this method in three different settings: without CV (noCV), with $5-$ (5CV) and with $10-$ fold CV (10CV).
In our study, we use the lmer (regression) and glmer (classification) functions from the lme4 package in R [@bates_2015] as an efficient way to fit glmms.
\begin{algorithm}
  \caption{GLMM Encoding Regression}
  \label{alg:glmm-regr}
  \begin{algorithmic}
    \State \underline{Training:} require $n.folds \in \mathbb{N}$
     \State fit simple random intercept model: $y^{train}_i = \beta_{0l} + \epsilon_{i} = \gamma_{oo} + u_l + \epsilon_{i}$ on $\mathcal{D}^{train}$
     \State with $u_l \overset{iid}{\sim} N(0,\tau^2)$, $\epsilon_i \overset{iid}{\sim} N(0,\sigma^2)$ and $x^{train}_i = l$, $l \in \mathcal{L}^{train}$
    \If{$n.folds = 1$}
      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
        \State $\hat{x}^{train}_i = \hat{\beta}_{0l}^{\mathcal{D}^{train}}$ with $x^{train}_i = l$
      \EndFor
    \Else\ {use $n.folds$ cross-validation scheme to make training sets $\mathcal{D}^{train}_1, \dots, \mathcal{D}^{train}_{n.folds}$}
      \State and fit simple random intercept model on each $\mathcal{D}^{train}_{m}$
      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
        \State $\hat{x}^{train}_i = \hat{\beta}_{0l}^{\mathcal{D}^{train}}$ with $x^{train}_i = l$ based on the model m
        \State with $(x^{train}_i, y^{train}_i) \notin \mathcal{D}^{train}_m$
      \EndFor
    \EndIf
    \State \underline{Prediction:}
    \For{$x^{new}$}
      \If{$x^{new} \in \mathcal{L}^{train}$}
        \State $\hat{x}^{new} = \hat{\beta}_{0l}^{\mathcal{D}^{train}}$ with $x^{new} = l$ based on full model fitted on $\mathcal{D}^{train}$
      \Else\ {$\hat{x}^{new} = \hat{\gamma}_{00}$ based on full model fitted on $\mathcal{D}^{train}$}
      \EndIf
    \EndFor
  \end{algorithmic}
\end{algorithm}

## Control Conditions

We include three control conditions to better understand the effectiveness of the investigated encoders:
The performance of a featureless learner (**FL** condition) was estimated as a conservative baseline for each dataset.
In regression problems, FL predicts the mean of the target variable in the training set for each observation in the test set.
In classification problems, the most frequent class of the target within the training set is predicted.
For each dataset, we also consider a random forest model without encoding (**none** condition), to compare the use of encoding methods in a random forest with a natural categorical splitting approach.
We use the random forest in ranger [@wright_2017] which provides efficient categorical feature support by ordering levels once before starting the tree growing algorithm [@wright_2019].
In the **remove** high cardinality features control condition, we remove features with a high number of levels above some threshold and use one-hot encoding (without collapsing rare levels) for the remaining features.
This condition reflects on whether including high cardinality features does indeed improve predictive performance.
Otherwise, the best encoding might just provide the least impairment compared to not including any high cardinality features.
We include an overview over available implementations in widely used machine learning frameworks for R and python in the supplementary material.

Benchmark Setup
===============

## Datasets

```{r, results='hide', message=FALSE}
library(data.table)
library(stringi)
library(kableExtra)
descr_dat = as.data.table(readRDS("../analysis/high_cardinality_benchmark/descr_dat.rds"))
# remove unfinished datasets
descr_dat = descr_dat[!(Name %in% c("KDD98", "Traffic_violations", "sf-police-incidents")),]
```

```{r, eval=FALSE, echo=FALSE}
# Collect info (do not run, just here for reproducibility)
if (FALSE) {
  # Collect cardinality information
  library(OpenML)
  setOMLConfig(arff.reader = "farff")
  out = list()
  for (id in descr_dat$OmlId) {
    d = getOMLDataSet(id)
    fcts = (sapply(d$data, class) == "factor") & (names(d$data) != d$desc$default.target.attribute)
    out[[as.character(id)]] = lapply(d$data[fcts], table)
    saveRDS(out, "../analysis/high_cardinality_benchmark/dataset_factor_levels.rds")
    gc()
  }

  # Restart
  setOMLConfig(arff.reader = "RWeka") # farff seems to break for one dataset
  out = readRDS("../analysis/high_cardinality_benchmark/dataset_factor_levels.rds")
  for (id in sample(setdiff(descr_dat$OmlId, as.numeric(names(out))))) {
    d = getOMLDataSet(id)
    fcts = (sapply(d$data, class) == "factor") & (names(d$data) != d$desc$default.target.attribute)
    out[[as.character(id)]] = lapply(d$data[fcts], table)
    saveRDS(out, "../analysis/high_cardinality_benchmark/dataset_factor_levels.rds")
    gc()
  }
}
```

```{r}
# Normalized Shannon entropy
shannonEntropy <- function(counts.table, log.base = getOption("GeneFamilies.entropy.log.base",
    base::exp(1))) {
    if (length(counts.table) <= 1)
        return(0)
    c.t.s <- sum(counts.table)
    -sum(sapply(counts.table, function(x) x/c.t.s * log(x/c.t.s, base = log.base)))/log(length(counts.table),
        base = log.base)
}
out = readRDS("../analysis/high_cardinality_benchmark/dataset_factor_levels.rds")
ents = lapply(out, function(x) lapply(x, shannonEntropy))
ent2 = lapply(out, function(x) lapply(x, function(x) min(x)/max(x)))
ents = data.table(OmlId = as.numeric(names(ents)), entropies = ents)
descr_dat = merge(descr_dat, ents)
```

```{r}
descr_dat2 = descr_dat
descr_dat2[Classes == 0, task_type := rep("regr", .N)]
descr_dat2[Classes == 2, task_type := rep("bin_class", .N)]
descr_dat2[Classes > 2, task_type  := rep("multi_class", .N)]
descr_dat2[, task_type := factor(task_type, levels = c("regr", "bin_class", "multi_class"))]
setorder(descr_dat2, task_type, Obs)
descr_dat2$HighCardLevels = stri_replace_all_fixed(descr_dat2$HighCardLevels, " ", "")
descr_dat2[Name == "KDDCup09_upselling", HighCardLevels := "14,...,5073,5713,13990,15415,15415"]
#hcl_list = lapply(descr_dat$HighCardLevels, function(x) {log(as.numeric(unlist(strsplit(x, #split = ","))))})
#descr_dat2$logNHCL = ""
```

Table \ref{tab:datasets} summarizes the datasets used in our benchmark study.
We specifically investigate datasets that contain categorical variables with a large number of levels, including many well-known datasets
from previous studies [@cerda_2018; @prokhorenkova_2018; @coors_2018; @kuhn_2019].
All datasets can be downloaded from the `OpenML` platform [@vanschoren_2013] based on the name or the displayed OmlId.
The datasets include `r nrow(descr_dat[Classes == 0])` regression, `r nrow(descr_dat[Classes == 2])` binary classification, and `r nrow(descr_dat[Classes > 2])` multiclass classification problems (between `r descr_dat[Classes > 2][, min(Classes)]` and `r descr_dat[Classes > 2][, max(Classes)]` classes).
To assess the imbalance in categorical variables we computed the normalized entropy for each categorical variable. The maximum normalized entropy is $1$, which corresponds to a uniform distribution, while a lower number indicates a larger imbalance.
Sample sizes range between `r descr_dat[, min(Obs)]` and `r descr_dat[, max(Obs)]` observations.
The total number of features ranges between `r descr_dat[, min(NumFeats + BinFeats + CatFeats)]` and `r descr_dat[, max(NumFeats + BinFeats + CatFeats)]`.
Datasets contain between `r descr_dat[, min(sapply(strsplit(HighCardLevels, " "), length))]` and `r descr_dat[, max(sapply(strsplit(HighCardLevels, " "), length))]` categorical features with more than $10$ levels, with the maximum number of levels for one feature in a dataset ranging between `r descr_dat[, min(MaxCatLvls)]` and `r descr_dat[, max(MaxCatLvls)]`.
Missing values are present in about half of the datasets.

```{r datasets, warning=FALSE, results="asis"}
library(kableExtra)
landscape(
  knitr::kable(
    descr_dat2[, .(OmlId, Name, Cl = Classes, N = Obs, `NA%` = NAs*100/(Obs*(NumFeats + BinFeats + CatFeats)),
                Num = as.integer(NumFeats), Bin = as.integer(BinFeats), Cat = as.integer(CatFeats), HighCardLevels, Entropy = "")],
    digits=2L,
    caption="Benchmark Datasets and Dataset Characteristics",
    label="datasets"
  ) %>%
  column_spec(10, image = spec_boxplot(lapply(descr_dat2$entropies, unlist))) %>%
  footnote(general = "OmlId = Id on OpenML, Name = name on OpenML, Cl = classes (0: regression), N = observations, NA% = percentage of missing values, Num = numeric features, Bin = binary features, Cat = categorical features, HighCardLevels = number of levels for each categorical feature with at least 10 levels (some levels are not displayed for KDDCup09_upselling), Entropy = box-plot of normalized Shannon-entropy across levels (smaller = larger imbalance).", threeparttable = TRUE) %>%
  kable_styling(latex_options="scale_down")
)
```

## High Cardinality Threshold
\label{sec:HCT}

<!--
A relevant question we aim to answer is what the minimum number of levels is, for which more involved encoding methods should be used. WE CANNOT REALLY ANSWER THIS WITH ONLY 3 DIFFERENT HCT VALUES...
-->
It is widely assumed that more involved encoding methods are only advantageous for variables with a high number of levels, while simple indicator encoding is more appropriate for a small number of levels.
To make our benchmark results more realistic, a high cardinality threshold (HCT) parameter with values of $10$, $25$, and $125$ was introduced which determined varying configurations for the different encoders.
For indicator encoding, the $HCT - 1$ most frequent levels are encoded together with a single collapsed category for the remaining levels.
For integer, frequency, hash, leaf, impact, and glmm encoders, only features with more than HCT levels in the training set were encoded with the respective strategy, while the remaining categorical features were one-hot encoded.
HCT is used as the hash size in hash encoding.
In the remove control condition, features with more than HCT levels in the training set are removed from the feature set.
Based on the number of categorical features and levels per feature, some HCT settings were removed from the benchmark for some combinations of dataset $\times$ encoder.
This made sure that encoders always affect at least one feature and that the remove condition always removes at least one feature.
If several HCT settings of an encoder would lead to identical encoding strategies for all features of a dataset, we only kept the condition with the smallest HCT value.

## Machine Learning Pipeline and Algorithms
\label{sec:pipeline}

Because we investigate strategies that can be used together with different models, we separately report results for different ML algorithms.
We kept tuning the hyperparameters of the subsequent ML methods to a minimum because we were interested in the effect of the encoding techniques instead of a comparison of the ML models.
In total, we investigate $5$ ML methods:
Regularized linear models (LASSO) were fitted with glmnet [@friedman_2010], internally tuning the regularization using 5-fold CV.
Random forests (RF) were trained using ranger [@wright_2017].
Because the random forest algorithm can be expected to give reasonable results without tuning [@probst_2019], we only fixed the number of fitted trees to $500$.
Gradient boosting models (GB) were trained using xgboost [@chen_2018].
We set the learning rate to $0.01$ and determine the number of iterations using early stopping on a 20% holdout set.
K-nearest neighbours (KNN) was taken from package knn [@schliep_2016] standardizing all features and using a constant $k=15$ for the number of nearest neighbours, together with an information gain filter [@brown2012] to limit the number of features to $25$.
Support vector machines (SVM) with radial basis function kernel were trained with liquidSVM [@steinwart_2017].
The bandwidth and regularization parameters were internally tuned using 5-fold CV, and we used a one-vs-all approach for multiclass-settings.

The ML pipeline outlined below was used for all experimental conditions.
It is not trivial and was carefully designed to ensure consistent results for extreme conditions (e.g. some levels only existing in the training data).

- **Imputation I**: Create a new factor level for missing values in categorical features with more than two categories. Impute missing values in binary features using the mode and missing values in numerical features using the mean feature value in the training data.
- **Encoding**: Transform the complete categorical features by the respective encoder.
  In the \textit{no encoding} condition, the encoder simply passes on its input.
  The leaf and remove conditions still return categorical features, while the remaining encoders return only numerical features. Encoders only affect categorical variables above the specified HCT value.
- **Imputation II**: To handle new levels observed during prediction, impute missing values obtained during encoding.
- **Drop constants**: Drop features that are constant during training. As none of the original datasets includes constant columns, this step only removes constant features that are produced by the encoders or the CV splitting procedure.
- **Final one-hot encoding**: Transform all remaining categorical features via one-hot encoding (skipped for no encoding condition).
- **Learner**: Use the transformed data from each training set to fit the respective ML algorithm.
  In the prediction phase, transformed feature values (based on the trained encoder) for new observations in each test set are fed into the trained model to compute predictions.

## Performance Evaluation

Throughout our analysis, we use implementations of the various methods for the open-source statistical software R [@r_2018].
The pre-processing, as well as final ML algorithm described in \ref{sec:pipeline}, was trained and resampled using the `mlr` framework [@bischl_2016] together with the `batchtools` package [@lang_2017] to scale the benchmark analysis to HPC compute infrastructure.
All materials for this study, including reproducible code for this manuscript and the necessary result objects can be downloaded from our repository (see above).

Throughout our experiments, we use 5-fold CV to obtain estimates of the predictive performance.
Depending on the target variable, we report root mean squared error (RMSE) for regression, area under the curve (AUC) for binary classification or its extension AUNU [@ferri_2009] for multiclass problems:
\begin{align*}
RMSE = \sqrt{\frac{1}{N}\sum_{i=1}^N(y_i-\hat{y_i})^2},\
AUC = \frac{U}{N_{pos} \cdot N_{neg}},\
AUNU = \frac{\sum_{c = 1}^C AUC_c}{C}
\end{align*}
$N$: number of observations, $N_{pos}$ and $N_{neg}$: number of true labels in both classes, $U$: test statistic of the Mann-Whitney-U test, and $AUC_c$: one vs. rest $AUC$ of class $c$.

Throughout the benchmark study, each encoding method listed in Section \ref{sec:methods} is applied together with all ML algorithms (c.f. Section \ref{sec:pipeline}) across
high cardinality thresholds (HCT) $10$, $25$, $125$.
While we only report results for the best HCT setting for each ML algorithm $\times$ dataset combination in Section \ref{sec:results}, we study the effect of the HCT parameter in more detail in section \ref{sec:hct_study}.

Benchmark Results
=================
\label{sec:results}

```{r}
res = readRDS("../analysis/high_cardinality_benchmark/results.rds")

# remove cluster and random forest encoders
res = res[algorithm != "cluster",]
res = res[algorithm != "ranger",]

# keep old algorithm column for later code
res[, algorithm2 := algorithm]

# separate dummy and one-hot encoding into two algorithms
res[algorithm == "dummy" & dummy.enc == FALSE, algorithm := "one-hot"]
# separate glmm conditions into three algorithms
res[algorithm == "lmer" & n.folds == 1, algorithm := "glmm-noCV"]
res[algorithm == "lmer" & n.folds == 5, algorithm := "glmm-5CV"]
res[algorithm == "lmer" & n.folds == 10, algorithm := "glmm-10CV"]

# create factor variables with intended names, in the intended order
res[, algorithm := factor(algorithm,
  levels = c("integer", "frequency", "dummy", "one-hot", "hash", "leaf", "impact", "glmm-noCV", "glmm-5CV", "glmm-10CV", "none", "remove"))]
res[, lrn.id := factor(lrn.id,
  levels = c("cvglmnet", "ranger", "xgboost.earlystop.wrap", "kknn", "liquidsvm", "featureless"),
  labels = c("LASSO", "RF", "GB", "KNN", "SVM", "FL"))]

# plotting function
library(ggplot2)
library(gridExtra)
library(grid)
library(ggpubr)
plotResultsGrid = function(dat, mes, legend = FALSE, xlabels = FALSE, ytext = TRUE, ylab = " ",
  colors = box_colors) {
  p = ggplot(dat, aes_string(x = "lrn.id", y = paste0(mes, ".test.mean"),
    color = "algorithm", shape = "algorithm")) +
    geom_pointrange(aes_string(ymin = paste0(mes, ".test.min"), ymax = paste0(mes, ".test.max")),
      position = position_dodge(width = 0.8)) +
    ggtitle(label = unique(dat$Name)) +
    scale_shape_manual(values = 0:11) +
    ylab(ylab) + theme_bw() +
    guides(color = guide_legend(nrow = 1, title = NULL, label.position = "top"),
      shape = guide_legend(nrow = 1, title = NULL, label.position = "top")) +
    theme(axis.title.x = element_blank(), legend.spacing.x = unit(0.5, "cm"))
  if (!xlabels) {
    p = p + theme(axis.text.x=element_blank())
  }
  if (!ytext) {
    p = p + theme(axis.title.y=element_blank())
  }
  if (!legend) {
    p = p + theme(legend.position="none")
  }
  if (mes %in% c("rmse")) {
    p = p + scale_y_reverse()
  }
  if (!is.null(colors)) {
    p = p + scale_colour_manual(values = colors)
  }
  p
}
```

```{r}
# keep only the best high.card.thresh condition of each encoding on each dataset for each learner
res_max = rbind(
  res[Classes == 2][
    , .SD[which.max(auc.test.mean)], by = c("lrn.id", "algorithm", "problem")],
  res[Classes == 0][
    , .SD[which.min(rmse.test.mean)], by = c("lrn.id", "algorithm", "problem")],
  res[Classes > 2][
    , .SD[which.max(multiclass.aunu.test.mean)], by = c("lrn.id", "algorithm", "problem")]
)

# conditions in which no primary performance measure could be computed (only SVM for some datasets)
svm_errors = res[is.na(auc.test.mean) & is.na(rmse.test.mean) & is.na(multiclass.aunu.test.mean), .(job.id, lrn.id, algorithm, high.card.thresh, problem, Name)]
# flag datasets for SVM where not all algorithms could be computed (remove them to avoid biasing ranks)
inds_rm_SVM = res_max$problem %in% svm_errors$problem & res_max$lrn.id == "SVM"

# compute ranks (low rank = good performance) of encoders within each combination of learner and dataset
res_max[!(inds_rm_SVM | lrn.id == "FL"), `:=` (rank_auc = rank(-auc.test.mean, na.last = "keep"),
  rank_rmse = rank(rmse.test.mean, na.last = "keep"),
  rank_aunu = rank(-multiclass.aunu.test.mean, na.last = "keep")),
  by = c("lrn.id", "problem")]

# create rank variable containing ranks based on the primary measure
res_max[Classes == 2, rank_mes1 := rank_auc]
res_max[Classes == 0, rank_mes1 := rank_rmse]
res_max[Classes > 2, rank_mes1 := rank_aunu]
```

The overarching question we aim to answer in our study is which encoding methods generally work well across various datasets.
We report results for regression and classification datasets separately, as the associated metrics differ in scale.

For  `r length(svm_errors[, unique(Name)])` datasets, some conditions with the SVM lead to unexpected crashes due to memory problems or numerical errors. We completely removed those datasets for the SVM when computing ranks or other statistics to compare encodings across datasets.

## Encoder Performance on Each Dataset

(ref:regr-res) Performance estimates from 5-CV for regression (mean, min, max). For each combination, only the best HCT condition is displayed. Note the reversed y-axis to ease visual interpretation.

```{r regr-res, fig.cap='(ref:regr-res)', out.width="\\textwidth", fig.pos="!h", fig.width=9, fig.height=8.1}
regr = res_max[Classes == 0 & lrn.id != "FL"]

lay1 = rbind(c(1, 2), c(3, 4), c(5, 6))
lay2 = rbind(c(1, 2))

tmp <- ggplot_gtable(ggplot_build(plotResultsGrid(regr[problem == "41211"], "rmse", xlabels = TRUE, legend = TRUE)))
leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
legend <- tmp$grobs[[leg]]

grob1 = arrangeGrob(
  plotResultsGrid(regr[problem == "41211"], "rmse", ylab = "RMSE"), plotResultsGrid(regr[problem == "41445"], "rmse", ytext = FALSE),
  plotResultsGrid(regr[problem == "41210"], "rmse", ylab = "RMSE"), plotResultsGrid(regr[problem == "41437"], "rmse", ytext = FALSE),
  plotResultsGrid(regr[problem == "41444"], "rmse", ylab = "RMSE"), plotResultsGrid(regr[problem == "41267"], "rmse", ytext = FALSE),
  layout_matrix = lay1
)
grob2 = arrangeGrob(
  plotResultsGrid(regr[problem == "41251"], "rmse", xlabels = TRUE, ylab = "RMSE"), plotResultsGrid(regr[problem == "41255"], "rmse", xlabels = TRUE, ytext = FALSE),
  layout_matrix = lay2
)
grid.arrange(legend, grob1, grob2, nrow = 3, heights = c(0.1, 1, 0.38))
```

(ref:bincl-res) Performance estimates from 5-CV for binary classification (mean, min and max). For each combination, only the best HCT condition is displayed.

```{r bincl-res, fig.cap='(ref:bincl-res)', out.width="\\textwidth", fig.pos="!h", fig.width=9, fig.height=10}
bin_cl = res_max[Classes == 2 & lrn.id != "FL"]

lay1 = rbind(c(1, 2), c(3, 4), c(5, 6), c(7, 8))
lay2 = rbind(c(1, 2))

tmp <- ggplot_gtable(ggplot_build(plotResultsGrid(bin_cl[problem == "41283"], "auc", xlabels = TRUE, legend = TRUE)))
leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
legend <- tmp$grobs[[leg]]

grob1 = arrangeGrob(
  plotResultsGrid(bin_cl[problem == "41283"], "auc", ylab = "AUC"), plotResultsGrid(bin_cl[problem == "981"], "auc", ytext = FALSE),
  plotResultsGrid(bin_cl[problem == "4135"], "auc", ylab = "AUC"), plotResultsGrid(bin_cl[problem == "41434"], "auc", ytext = FALSE),
  plotResultsGrid(bin_cl[problem == "1590"], "auc", ylab = "AUC"), plotResultsGrid(bin_cl[problem == "1114"], "auc", ytext = FALSE),
  plotResultsGrid(bin_cl[problem == "41162"], "auc", ylab = "AUC"), plotResultsGrid(bin_cl[problem == "41442"], "auc", ytext = FALSE),
  layout_matrix = lay1
)
grob2 = arrangeGrob(
  plotResultsGrid(bin_cl[problem == "41447"], "auc", xlabels = TRUE, ylab = "AUC"), plotResultsGrid(bin_cl[problem == "41224"], "auc", xlabels = TRUE, ytext = FALSE),
  layout_matrix = lay2
)
grid.arrange(legend, grob1, grob2, nrow = 3, heights = c(0.1, 1, 0.28))
```

(ref:multcl-res) Performance estimates from 5-CV for multiclass classification (mean, min and max). For each combination, only the best HCT condition is displayed.

```{r multcl-res, fig.cap='(ref:multcl-res)', out.width="\\textwidth", fig.pos="!h", fig.width=9, fig.height=6}
mult_cl = res_max[Classes > 2 & lrn.id != "FL"]

lay1 = rbind(c(1, 2), c(3, 4))
lay2 = rbind(c(1, 2))

tmp <- ggplot_gtable(ggplot_build(plotResultsGrid(mult_cl[problem == "188"], "multiclass.aunu", xlabels = TRUE, legend = TRUE)))
leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
legend <- tmp$grobs[[leg]]

grob1 = arrangeGrob(
  plotResultsGrid(mult_cl[problem == "188"], "multiclass.aunu", ylab = "AUNU"), plotResultsGrid(mult_cl[problem == "41446"], "multiclass.aunu", ytext = FALSE),
  plotResultsGrid(mult_cl[problem == "41212"], "multiclass.aunu", ylab = "AUNU"), plotResultsGrid(mult_cl[problem == "41216"], "multiclass.aunu", ytext = FALSE),
  layout_matrix = lay1
)
grob2 = arrangeGrob(
  plotResultsGrid(mult_cl[problem == "41440"], "multiclass.aunu", xlabels = TRUE, ylab = "AUNU"), plotResultsGrid(mult_cl[problem == "4541"], "multiclass.aunu", xlabels = TRUE, ytext = FALSE),
  layout_matrix = lay2
)
grid.arrange(legend, grob1, grob2, nrow = 3, heights = c(0.15, 1, 0.55))
```

Mean performance estimates along with minimum and maximum performance from the 5-CV folds are reported for all datasets in Figures \@ref(fig:regr-res) (regression), \@ref(fig:bincl-res) (binary classification), and \@ref(fig:multcl-res) (multiclass classification).
To reduce the complexity induced by the hyperparameter HCT we only display the parameter condition with the best performance for each combination of dataset $\times$ encoding $\times$ ML algorithm.
The y-axis differs for all datasets and is reversed for the RMSE for better visual comparison.
For some datasets, the remove condition performed very similar to the other encodings (e.g. *ames-housing*, *porto-seguro*), suggesting that categorical features were not informative.
Performance in some folds was below the FL learner for *flight-delay-usa-dec-2017* ($RMSE_{FL} =$ `r res[lrn.id == "FL" & Name == "flight-delay-usa-dec-2017", rmse.test.mean]`) and *nyc-taxi-green-dec-2016* ($RMSE_{FL} =$ `r res[lrn.id == "FL" & Name == "nyc-taxi-green-dec-2016", rmse.test.mean]`).
<!-- den satz können wir nicht rauslassen, sonst gibt es zum FL keine Information im ganzen Paper -->

On datasets where substantive performance differences could be observed, target encoding with the glmm encoder was generally *most* effective.
The *worst* encoder depended on the ML algorithm and respective dataset.
For datasets *Click_prediction_small*, *KDDCup09_upselling*, *kick*, and *okcupid-stem* some combinations of encoder $\times$ ML algorithm performed worse than simply removing high cardinality features.

## Meta Rankings and Dataset Clustering

```{r, include=FALSE}
library(batchtools)
library(BBmisc)
library(relations)
library(ggdendro)

# load registry in read-only mode
reg = loadRegistry(file.dir = "../analysis/high_cardinality_benchmark/registry", writeable = FALSE)

# save all test set evaluations as data.table: list column result, each entry is a data.frame
res0 = reduceResultsDataTable(reg = reg, fun = function(x) x$measures.test)

# new unwrapped data.table with each row referring one test set evaluation
dat = data.table()
for(i in seq_row(res0)) {
  dat = rbind(dat, res0[i, data.table(job.id, rbindlist(result, fill = TRUE))], fill = TRUE)
}

# merge job parameters, descriptives, measures
descr_dat$OmlId = as.character(descr_dat$OmlId)
dat = unwrap(getJobTable(reg = reg))[descr_dat, on = c("problem==OmlId")][
  dat, on = "job.id"]

# remove cluster and random forest encoders
dat = dat[algorithm != "cluster",]
dat = dat[algorithm != "ranger",]

dat_max = dat[job.id %in% res_max$job.id]
# keep old algorithm column for later code
dat_max[, algorithm2 := algorithm]
# separate dummy and one-hot encoding into two algorithms
dat_max[algorithm == "dummy" & dummy.enc == FALSE, algorithm := "one-hot"]
# separate glmm conditions into three algorithms
dat_max[algorithm == "lmer" & n.folds == 1, algorithm := "glmm-noCV"]
dat_max[algorithm == "lmer" & n.folds == 5, algorithm := "glmm-5CV"]
dat_max[algorithm == "lmer" & n.folds == 10, algorithm := "glmm-10CV"]

# create factor variables with intended names, in the intended order
dat_max[, algorithm := factor(algorithm,
  levels = c("integer", "frequency", "dummy", "one-hot", "hash", "leaf", "impact", "glmm-noCV", "glmm-5CV", "glmm-10CV", "none", "remove"))]
dat_max[, lrn.id := factor(lrn.id,
  levels = c("cvglmnet", "ranger", "xgboost.earlystop.wrap", "kknn", "liquidsvm", "featureless"),
  labels = c("LASSO", "RF", "GB", "KNN", "SVM", "FL"))]

# define primary performance measures for all task types
dat_max[Classes == 2, mes1 := auc]
dat_max[Classes == 0, mes1 := - rmse]
dat_max[Classes > 2, mes1 := multiclass.aunu]

# remove problematic datasets with SVM and FL learner
dat_max = dat_max[!(lrn.id == "SVM" & problem %in% unique(svm_errors$problem))]
dat_max = dat_max[lrn.id != "FL"]

## t-test for cross-validation (nadeau & bengio, 2000)
t.test.cor.paired = function(x1, x2, N, alpha = 0.05, lower.tail = FALSE) {
  n = n.folds = length(x1)
  N_train = (n.folds - 1) / n.folds * N
  N_test = 1 / n.folds * N
  d = x1 - x2
  m = mean(d)
  v = var(d)

  t = m / sqrt(v * (1/n + N_test/N_train))
  df = n - 1

  p = pt(t, df, lower.tail = lower.tail)
  as.numeric(p < alpha) # x1 dominates x2
}

# compute relations between encodings for all learners and datasets
rel_list = sapply(unique(dat_max$lrn.id), function(lname) {
  sapply(unique(dat_max[lrn.id == lname, Name]), function(dname) {
    dt = dat_max[lrn.id == lname & Name == dname]
    grid = CJ(x1 = levels(droplevels(dt$algorithm)), x2 = levels(droplevels(dt$algorithm)))
    grid$x1_dom_x2 = apply(grid, 1, function(x) t.test.cor.paired(x1 = dt[algorithm %in% x[1], mes1], x2 = dt[algorithm %in% x[2], mes1], N = dt[,unique(Obs)]))
    grid[x1 == x2, x1_dom_x2 := 0]
    m = matrix(grid$x1_dom_x2, byrow = TRUE, ncol = length(unique(grid$x1)), nrow = length(unique(grid$x1)))
    dimnames(m) = list(unique(grid$x1), unique(grid$x1))
    as.relation(m)
  }, simplify = FALSE, USE.NAMES = TRUE)
}, simplify = FALSE)
names(rel_list) = unique(dat_max$lrn.id)

## compute consensus relations
set.seed(1)
cons_list = lapply(rel_list, function(lrn) {
  ens = relation_ensemble(list = lrn)
  relation_consensus(ens, "symdiff/W", control = list(n = "all"))
  })

## compute meta cluster analysis across all learners

# remove none encoding to ensure same relation domain for all learners
dat_max = dat_max[algorithm != "none"]

# compute relations between encodings for all datasets and learners
rel_list2 = sapply(unique(dat_max$Name), function(dname) {
  sapply(unique(dat_max[Name == dname, lrn.id]), function(lname) {
    dt = dat_max[lrn.id == lname & Name == dname]
    grid = CJ(x1 = levels(droplevels(dt$algorithm)), x2 = levels(droplevels(dt$algorithm)))
    grid$x1_dom_x2 = apply(grid, 1, function(x) t.test.cor.paired(x1 = dt[algorithm %in% x[1], mes1], x2 = dt[algorithm %in% x[2], mes1], N = dt[,unique(Obs)]))
    grid[x1 == x2, x1_dom_x2 := 0]
    m = matrix(grid$x1_dom_x2, byrow = TRUE, ncol = length(unique(grid$x1)), nrow = length(unique(grid$x1)))
    dimnames(m) = list(unique(grid$x1), unique(grid$x1))
    as.relation(m)
  }, simplify = FALSE, USE.NAMES = TRUE)
}, simplify = FALSE, USE.NAMES = TRUE)

# compute consensus relation for each dataset across learners
set.seed(1)
lrn_cons_list = lapply(rel_list2, function(dt) {
  ens = relation_ensemble(list = dt)
  relation_consensus(ens, "symdiff/O")
})

# compute cluster analysis of datasets
meta_ens = relation_ensemble(list = lrn_cons_list)
meta_d = relation_dissimilarity(meta_ens)
meta_clust = hclust(meta_d)
```

To further summarize the results, we computed meta rankings for each ML algorithm:
First, we defined an encoder relation within each dataset, based on corrected resample t-tests [@nadeau_2003].
An encoding was defined to beat another encoding if the one-sided p-value of the t-test was $< .05$.
This allowed us to compute a weak-order consensus ranking $R$ defined by the optimization problem:

$$ \underset{R \in \mathcal{C}}{arg\ min} \ \sum_{b=1}^B d(R_b, R)$$

where $d$ is the symmetric difference distance and $R_b$ is the relation for dataset $b$ [@hornik_2007; @meyer_2018].
The symmetric difference between two relations is the number of cases one encoding beats another encoding in one relation but not in the other one.
This procedure resulted in the meta rankings based on the consensus ranking for each ML algorithm detailed in Figure \ref{fig:rank-res}.

(ref:rank-res) Consensus rankings across all datasets for each algorithm.

```{r, rank-res, out.height=".55\\textwidth", fig.cap='(ref:rank-res)', out.width="\\textwidth"}
plotdt = rbindlist(lapply(names(cons_list), function(x) {
  r = relation_class_ids(cons_list[[x]][[1]])
  data.frame(method = names(r), rank = r, algorithm = x)
}))
set(plotdt, which(plotdt$method == "remove" & plotdt$algorithm == "RF"), "rank", 11)
plotdt = plotdt[-which(plotdt$method == "none"),]
p = ggplot(plotdt, aes(x = rank, y = algorithm, fill = method)) +
 geom_tile(size = .5, width =.93, height =.93, color = "black")
p = p +
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    legend.spacing.x = unit(0.1, "cm"),
    legend.key.size = unit(.9, "cm")
  ) +
  theme(axis.title.y=element_blank()) +
  theme(legend.background = element_blank()) +
  theme(legend.key = element_blank()) +
  theme(legend.title = element_blank()) +
  scale_fill_manual(values = box_colors) +
  theme(panel.grid.minor.y = element_blank()) +
  theme(panel.grid.major.y = element_blank()) +
  theme(panel.grid.minor.x = element_blank()) +
  theme(panel.grid.major.x = element_blank()) +
  scale_x_continuous(breaks = c(1:11), expand = c(0, 0))
p
```
<!--

- **LASSO:** `r paste0(names(sort(relation_class_ids(cons_list$LASSO[[1]]))), collapse = " > ")`

- **RF:** `r paste0(names(sort(relation_class_ids(cons_list$RF[[1]]))), collapse = " > ")`

- **GB:** `r paste0(names(sort(relation_class_ids(cons_list$GB[[1]]))), collapse = " > ")`

- **KNN:** `r paste0(names(sort(relation_class_ids(cons_list$KNN[[1]]))), collapse = " > ")`

- **SVM:** `r paste0(names(sort(relation_class_ids(cons_list$SVM[[1]]))), collapse = " > ")`
-->

Although the presented solutions of the optimization problem are not unique, rankings were highly stable for the high and low ranks. Meta rankings seem to be highly consistent with the individual patterns of encoder performances reflected in Figures \@ref(fig:regr-res) to \@ref(fig:multcl-res).
Looking at meta-rankings, approaches based on GLMM's in combinaion with cross-validation outperform all other approaches across all algorithms. A further interesting detail ommited in Figure \ref{fig:rank-res} for clarity is that the \textit{none} encoding strategy for the random forest was beaten by all strategies except for the \textit{remove} condition. This implies, that even if the algorithm includes a mechanic for treating categorical variables, it might often be optimal to use a different strategy instead.

In an additional exploratory analysis, we tried to find clusters of datasets, which show systematic patterns of encoder performance (independent of the employed ML algorithm).
We computed a partial-order consensus relation across ML algorithms for each dataset and hierarchically clustered the dataset consensus relations using the symmetric difference distance in combination with the complete linkage agglomeration method.
The resulting dendrogram is displayed in Figure \@ref(fig:dendrogram).
Although the cluster structure is somewhat ambiguous, roughly three clusters can be described:
The first 9 datasets from the top of the dendrogram are characterized by a medium to a high number of levels, low performance of the remove condition (indicating the importance of high cardinality features) and clear performance advantage of target encoders.
For the next 13 datasets which contain the smallest number of levels, traditional encodings can compete with target encoding.
This biggest cluster also includes 9 datasets with zero distances, in which no significant performance differences could be observed between any encoding conditions (nor with the remove condition, indicating that high cardinality features are not very informative).
The last two datasets with the highest number of levels formed a separate cluster, in which target encoding without strong regularization (impact, glmm-noCV) showed severe overfitting.
Note that clusters were not determined by problem type, again suggesting that encoder rankings are somewhat similar for regression and classification settings.

(ref:dendro) Hierarchical cluster analysis of benchmark datasets. The symmetric difference distance between two datasets reflect differences in performance patterns between encodings (independent of the employed ML algorithm).

```{r dendrogram, fig.cap='(ref:dendro)', out.width="\\textwidth", fig.pos="!h", fig.width=9, fig.height=6}
# plot dendrogram
ggdendrogram(meta_clust, rotate = TRUE) + theme(text = element_text(size=20))
```

## Summary of Encoder Performance

Our results indicate that regularized target encoding was superior or at least competitive on all datasets.
Especially effective was glmm encoding with 5-fold-CV, which ranked first place for all ML algorithms except SVM.
While the impact encoding used no regularization, we compared different amounts of regularization for the glmm encoder.
When performance differences between target encoders were observed, more regularization seemed beneficial.
In those datasets, impact encoding not only performed worse than glmm with CV but was sometimes also inferior compared to other encoders.
We could not observe a setting in which regularized target encoding was convincingly beaten by target agnostic methods.
When looking for a simple default method, indicator encoding in combination with collapsing small levels seems a robust alternative, although the glmm strategies should be preferred.

We observed several interesting patterns, which we aim to summarize:
Integer encoding did not perform well together with XGBoost in our benchmark in contrast to what is stated in the documentation of other boosting libraries\footnote{\url{https://lightgbm.readthedocs.io/en/latest/Advanced-Topics.html\#categorical-feature-support}}.
In contrast, target-based encoders (especially the glmm encoder) seem to be a compelling strategy.
For LASSO, previous studies have suggested that indicator encoding works well, even with a very high number of levels [@cerda_2018].
Although the glmm encoders ranked first in our benchmark for the LASSO, it was the only algorithm where indicator encoding ranked second.
Note that for computational reasons we limited the maximum amount of indicator variables to 125 in our experimental design.
The HCT = 125 setting performed best for LASSO with indicator encoding in a large number of datasets, indicating that performance might have further improved with higher values.
Both KNN and SVM rely on numerical distances in feature space and tend to perform poorly in the presence of high dimensionality.
Thus, we expected that target encoding should work well here as it transforms categories into a single, smooth numerical feature. This was backed by our benchmark results. For some datasets (*medical_charges*, *road-safety-drivers-sex*, and *Midwest_survey*) KNN (without tuning of the optimal number of nearest neighbours) could compete with the more sophisticated ML algorithms when combined with target encoding, but performed poorly with other encoders.
Although consistent with the big picture, SVM results have to be considered with care, as some experimental conditions resulted in unexpected computational errors.
In RF, a widely used strategy to deal with categorical features is to order levels by average target statistics for a given level.
For a small number of levels, this approach has been reported superior to indicator and integer encoding [@wright_2019], while we observed poor performance for datasets with a larger number of levels.

In line with earlier research [@micci_barreca_2001; @prokhorenkova_2018], target encoding with regularization (glmm) performed better or equally well in comparison with the unregularized impact encoder.
The glmm method was the overall winner based on the encoder rankings across datasets and outperformed impact encoding on some datasets.
Our results show that combining target encoding based on glmms with 5-fold CV led to an improved predictive performance on many datasets.
Performance often did not improve further when using 10 folds, suggesting that 5 folds might be a good default in practice.
The glmm encoder has a clear advantage over target encoding with a smoothing hyperparameter [@micci_barreca_2001], as costly tuning the whole ML pipeline with different smoothing values is not required.

On a side note, we found that one-hot encoding usually gave a slightly better performance than dummy encoding.
This phenomenon has also been observed by @chiquet_2016.
@tutz_2016 commented that in generalized linear models "the naive combination of usual dummy coding with reference category and standard penalties is often a bad idea" (p.254) unless the reference category is of special interest.
In our benchmark, the reference category was chosen arbitrarily as is usually the case in ML.
Our results suggest that one-hot encoding is the better standard compared to dummy encoding when applying nonlinear regularized models like RF, GB or SVM with (high cardinal) categorical features.
We furthermore provide a more in-detail anaysis of the performances of one-hot and dummy encoding in the supplementary material.

```{r}
# # Cluster Table
# cluster_ids = cutree(meta_clust, k = 3)
# clust_descr = merge(descr_dat, data.table(Name = names(cluster_ids), MetaClust = unname(cluster_ids)), by = "Name", all.x = TRUE)
# # Names = paste0(unique(OmlId), collapse = ", ")
# clust_descr = clust_descr[!is.na(MetaClust), .(NDat = .N, `Type%` = paste(round(sum(Classes == 0)/.N*100), round(sum(Classes == 2)/.N*100),
#   round(sum(Classes > 2)/.N*100), sep = ","), N = as.numeric(median(Obs)), `NA%` = as.numeric(median(NAs*100/(Obs*(NumFeats + BinFeats + CatFeats)))),
#     `N+BFts` = as.numeric(median(NumFeats + BinFeats)), CFts = as.numeric(median(CatFeats)), MdLvls = as.numeric(median(MdHighCardLvls)),
#       MxLvls = as.numeric(median(MaxCatLvls))), keyby = MetaClust]
# knitr::kable(clust_descr)
```

## Runtime Analysis

```{r time-ranks, results="asis"}
run_time = merge(res_max[!(inds_rm_SVM | lrn.id == "FL")],
  res_max[!(inds_rm_SVM | lrn.id == "FL" | algorithm != "one-hot"), .(lrn.id, problem,
    time.running.onehot = time.running)], by = c("lrn.id", "problem"), all.x = TRUE)
run_time[, time.running.stand := as.numeric(time.running) / as.numeric(time.running.onehot)]
run_time = run_time[, .(M = median(time.running.stand), mn = min(time.running.stand), mx = max(time.running.stand)), by = c("algorithm", "lrn.id")]
run_time[, M := paste0("$", round(M,2), "_{", round(mn,1), "}^{", round(mx,1), "}$")][, mn := NULL][, mx := NULL]
run_time = dcast(run_time, ... ~ lrn.id, value.var = "M")
colnames(run_time)[1] = "Encoding"
kbl(run_time, digits=2L, caption = "Proportional Increase in Runtime Compared to One-Hot Encoding",
    escape = FALSE) %>%
  footnote("Median $_{min}^{max}$ across datasets of the proportional increase in runtime from 5-CV, when comparing the respective encoder with one-hot encoding. Only the best HCT conditions are reported.", threeparttable = TRUE, escape = FALSE)
```

To determine whether traditional encodings might be worth it when facing limited computational resources, we further analyzed the runtimes of the whole analysis pipeline for different encoders and ML algorithms.
Aggregated results are shown in Table \@ref(tab:time-ranks).
To enable a meaningful comparison, we report runtime as the fraction of a full pipeline's strategy compared to running simple one-hot encoding for the same pipeline and then further aggregate across datasets using the median.
We again only report the **best** HCT setting.
Absolute runtimes are hard to interpret and aggregate because runtime distributions across datasets are heavily skewed as proportionally larger runtimes are observed for big datasets.
Furthermore, while a pipeline's runtime can be dominated by the encoder for small datasets, the training of the consecutive ML algorithm is dominating for large datasets, which might render differences during encoding irrelevant.
Therefore we aim to report what is important in practice, the time differences for training the **full pipeline**.
The results clearly show that regularized target encoding does not consistently yield slower runtimes compared to simple strategies like indicator encoding. Reasons for this might be, that more efficient representations yielded by target encoders lead to faster runtimes of subsequent ML algorithms.
This suggests that a possible runtime vs. predictive performance trade-off might also be in favour of target encoding, especially for large datasets where a high number of indicator variables increases computational load.
We did observe a substantial increase in runtimes when using regularized glmm with GB, where runtimes are comparably short in comparison to the time required to fit categorical encoders. In other settings (LASSO, RF), the glmm encoder can also be faster, as it allows for a more efficient data representation for the ML algorithm.

## Analysing High Cardinality Thresholds

\label{sec:hct_study}

```{r hct-freq, results="asis"}
#thresh = res_max[!inds_rm_SVM][algorithm != "none", .(N10 = sum(high.card.thresh == 10),
#  N25 = sum(high.card.thresh == 25), N125 = sum(high.card.thresh == 125)), by = c("algorithm", "lrn.id")]
#setorder(thresh, algorithm, lrn.id)

#num_thresh = res[!(problem %in% unique(svm_errors$problem) & lrn.id == "SVM")][
#  lrn.id != "FL"][algorithm != "none"][,
#  .N, by = c("algorithm", "lrn.id", "high.card.thresh", "dummy.enc", "num.trees", "n.folds")][
#    n.folds == 1 | is.na(n.folds)][num.trees == 10 | is.na(num.trees)][dummy.enc == TRUE | is.na(dummy.enc)]
#num_thresh = dcast(num_thresh, ... ~ high.card.thresh, value.var = "N")[, !c("dummy.enc", "num.trees", "n.folds"), with = FALSE]
#setorder(num_thresh, algorithm, lrn.id)

#thresh[, c("N10", "N25", "N125")] = thresh[, c("N10", "N25", "N125")]/num_thresh[, c("10", "25", "125")]
#thresh[, c("N10", "N25", "N125")] = thresh[, lapply(.SD, round, digits = 2), .SDcols = c("N10", "N25", "N125")]
#thresh[, c("N10", "N25", "N125")] = thresh[, lapply(.SD, function(x)x*100), .SDcols = c("N10", "N25", "N125")]

#thresh[, HCT := paste(N10, N25, N125, sep = ",")]
#thresh = setnames(dcast(thresh, algorithm ~ lrn.id, value.var = "HCT"), "algorithm", "Encoder")

#knitr::kable(thresh)
```

Until now, we ignored the HCT parameter by reporting only the condition with the best performance.
An interesting question was whether target encoding is only useful for features with a very high number of levels or if it might also have advantages for fewer levels, where most practitioners would routinely use indicator encoding.
We tested this using HCT thresholds of $10$, $25$ and $125$.
In general, the optimal threshold seemed to strongly depend on the dataset, but we also observed some weak patterns:
LASSO improved for larger HCT, indicating that its internal regularization can efficiently deal with the sparseness induced by indicator encoding.
In comparison, other methods generally yielded better performance if features above the very low HCT of $10$ were encoded using one of the target encoding strategies.
Furthermore, different regularization strategies for the target based encoders seemed not to prefer strongly different HCT values.

Discussion
==========

In our benchmark, we compared encoding strategies for high cardinality features on a variety of regression, binary and multiclass classification datasets with different ML algorithms.
Regularized target encoding (glmm) was superior across most datasets and ML algorithms.
Although the performance of other encoding strategies was comparable in some conditions, target encoding was never outperformed.
In general, our results suggest that regularized target encoding works well for all kinds of algorithms and should be considered when working with high cardinality features.
In terms of runtime, using target encoding together with 5-fold CV sometimes leads to slightly longer runtimes (in comparison to indicator encoding), but especially for larger datasets this is often offset by the more efficient representation yielded from target encoding.

What constitutes a "high" cardinality problem is a difficult question that might not only depend on the number of levels but also other characteristics of the dataset and its features. Unfortunately, the number of datasets in our study was too small to discover consistent patterns between encoder performance and dataset characteristics.
Target encoding features with as little as 10 levels seemed to be effective in a substantive number of conditions, but for other datasets, higher HCT values performed better.
Thus, some form of hyperparameter tuning seems necessary to decide the level threshold for target encoding at this point, as no suitable defaults seem to be available.
Note that we compared different HCT values but then used the same encoding strategy for all affected features alongside one-hot encoding for the remaining ones.
Although this strategy is reasonable, it might be beneficial to decide whether to use target encoding on a feature by feature basis.
ML pipelines could introduce a categorical hyperparameter for each feature that represents which encoding is used.
To make this complicated meta optimization problem feasible, only a small number of encoders can be included.
Our study can help to decide which encoders could be safely omitted from consideration.

## Limitations

In our study, several decisions were necessary to make answering our research questions technically and computationally feasible.
First, we only used minimal tuning for our ML algorithms, as we were not interested in comparing their performance against each other.
This probably led to suboptimal performance for the GB, KNN, and SVM learners.
When interpreting our results, we assume that the encoder rankings are comparable when more extensive tuning is used.
This seems plausible, considering the high stability of our meta rankings between algorithms.
We only use 5-fold-CV without repetitions to estimate predictive performance, but due to a small variance between folds, we could confidently detect performance differences of encoders for many datasets.
As we are interested in model-agnostic methods that can be combined with any supervised ML algorithm, we do not investigate several strategies proposed in the recent literature [@guo_2016; @prokhorenkova_2018].
Additionally, we only investigate univariate encodings, i.e. we always encode each variable separately.
Levels with comparable main effects can therefore not be distinguished based on the transformed feature, which prevents the consecutive ML algorithm from learning interactions for specific levels.
Methods that can jointly encode several variables and therefore leverage correlation structures between the different variables are an interesting avenue of research in the future. Another interesting avenue is using target-based encoding strategies to learn a vector-valued representation of a level, similar to entity embeddings [@guo_2016].
We furthermore only investigate traditional categorical variables and do not extend our analysis to multi-categorical variables or text-strings [@cerda_2018; @cerda_2020], where other encoding techniques can harvest additional information.

## Conclusion

This benchmark study compared the predictive performance of a variety of strategies to encode categorical features with a high number of unordered levels for different supervised machine learning algorithms.
Regularized versions of target encoding, which uses predictions of the target variable as numeric feature values, performed better than traditional strategies like integer or indicator encoding.
Most effective was a target encoder which combines simple generalized linear mixed models with cross-validation and does not require tuning hyperparameters.

Acknowledgements {-}
==========
This work has been funded by the German Federal Ministry of Education and Research (BMBF) under Grant No. 01IS18036A and by the Bavarian Ministry of Economic Affairs, Regional Development and Energy through the Center for Analytics – Data – Applications (ADA-Center) within the framework of „BAYERN DIGITAL II“ (20-3410-2-9-8).
The authors of this work take full responsibilities for its content.

Declarations of interest: none

References {#references .unnumbered}
==========
\raggedright

### 2021-04-01 publication/supplementary_material.Rmd
inital commit

---
title: |
  | Supplementary Material\:
  | Regularized target encoding outperforms traditional methods in supervised machine learning with high cardinality features
numbersections: true
csl: elsevier-harvard.csl
output: pdf_document
header-includes:
   - \usepackage{algorithm}
   - \usepackage{array}
   - \usepackage[noend]{algpseudocode}
   - \usepackage{amsmath}
   - \usepackage{amssymb}
   - \usepackage{amsfonts}
   - \usepackage{mathtools}
   - \usepackage{bm}
   - \usepackage{tikz}
   - \usepackage{threeparttable}
   - \usepackage{booktabs}
   - \usetikzlibrary{trees,arrows, chains, fit, positioning, calc, shapes, shadows}
   - \setlength\parindent{0pt}
   - \usepackage{pdflscape}
   - \usepackage{float}
---

# Implementations

Feature encoding techniques are implemented in several general preprocessing packages.
We provide an overview of existing implementations for R in Table \ref{tab:impls}.
For python, an extension to scikit-learn [@pedregosa_2011], *category encoders* [@will_mcginnis_2018_1157110] is available.
To enable a fair and reliable comparison in our study, we implemented all methods outlined above on top of the `mlrCPO` package.
All code can be found in our **online repository** \footnote{\url{https://github.com/compstat-lmu/paper_2021_categorical_feature_encodings}}.

\begin{table}
\centering
\begin{tabular}{|ccp{0.43\linewidth}|}
\hline
encoding method  & regularization & package \\
\hline
indicator        & --                &  \tt{stats} \newline \tt{embed}  \newline \tt{mlrCPO} \newline \tt{mlr3pipelines} \\
\hline
hash             & --                &  \tt{FeatureHashing} \newline \tt{embed} \\
\hline
impact           & smoothing         & \tt{embed} \newline \tt{mlrCPO} \newline \tt{mlr3pipelines} \\
\hline
impact           & cross-validation  & \tt{vtreat} \newline \texttt{mlr3pipelines} (via vtreat) \\
\hline
regularized impact (glmm) & --            & \tt{embed} \newline \tt{mlr3pipelines} \\
\hline
\end{tabular}
\caption{
  Overview over existing encoding implementations in R. Implementations can deviate from the algorithms described in section \ref{sec:methods} by minor implementation details.
  Indicator encoding encloses a variety of methods (one-hot, dummy, helmert encoding etc.)}
\label{tab:impls}
\end{table}

# Pseudocode for all encoding strategies \label{algos}

Algorithm $1$ to $10$ contain the pseudocode for all encoding strategies studied in our manuscript in order to improve reproducibility and to provide further hints towards subtle differences in encoders and implementations.

\begin{algorithm}[H]
  \caption{Integer Encoding}
  \begin{algorithmic}
    \State \underline{Training:}
    \State compute random permutation $\boldsymbol{int} = (int_1, \dots, int_k, \dots, int_L)^T$ of $(1, \dots, L)^T$
      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}\ {$\hat{x}^{train}_i = int_k$ with $x^{train}_i = l_k$, $k = 1, \dots, L$}
      \EndFor
    \State \underline{Prediction:}
      \For{$x^{new}$}
        \State \algorithmicif\ {$x^{new} \in \mathcal{L}^{train}$}\ \algorithmicthen\ {$\hat{x}^{new} = int_k$ with $x^{new} = l_k$, $k = 1, \dots, L$}\ \algorithmicelse\ {$\hat{x}^{new} = NA$}
      \EndFor
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
  \caption{Frequency Encoding}
  \begin{algorithmic}
    \State \underline{Training:}
      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}\ {$\hat{x}^{train}_i = \frac{N_l}{N}$ with $x^{train}_i = l$}
      \EndFor
    \State \underline{Prediction:}
      \For{$x^{new}$}
        \State \algorithmicif\ {$x^{new} \in \mathcal{L}^{train}$}\ \algorithmicthen\ {$\hat{x}^{new} = \frac{N_l}{N}$ with $x^{new} = l$}\ \algorithmicelse\ {$\hat{x}^{new} = 1$}
      \EndFor
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
  \caption{One-Hot Encoding}
  \begin{algorithmic}
    \State \underline{Training:}
      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
        \ForAll{$l \in \mathcal{L}^{train}$}\ $\hat{x}^{train}_{il} = I(x^{train}_i = l)$
        \EndFor
      \EndFor
    \State \underline{Prediction:}
    \For{$x^{new}$}
      \ForAll{$l \in \mathcal{L}^{train}$}
        \State \algorithmicif\ {$x^{new} \in \mathcal{L}^{train}$}\ \algorithmicthen\ {$\hat{x}^{new}_l = I(x^{new} = l)$}\ \algorithmicelse\ {$\hat{x}^{new}_l = 0$}
      \EndFor
    \EndFor
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
  \caption{Dummy Encoding}
  \begin{algorithmic}
    \State \underline{Training:}
      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
        \ForAll{$l \in \mathcal{L}^{train} \setminus l_{ref}$}\ {$\hat{x}^{train}_{il} = I(x^{train}_i = l)$}
        \EndFor
      \EndFor
    \State \underline{Prediction:}
    \For{$x^{new}$}
      \ForAll{$l \in \mathcal{L}^{train} \setminus l_{ref}$}
        \State \algorithmicif\ {$x^{new} \in \mathcal{L}^{train}$}\ \algorithmicthen\ {$\hat{x}^{new}_l = I(x^{new} = l)$}\ \algorithmicelse\ {$\hat{x}^{new}_l = NA$}
      \EndFor
    \EndFor
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
  \caption{Hash Encoding}
  \begin{algorithmic}
    \State \underline{Training:} require $hash.size \in \mathbb{N}$
      \ForAll{$l \in \mathcal{L}^{train}$}\ {$ind_l = (hash(l) \mod hash.size) + 1$, $ind_l \in \mathbb{N}$, $hash(l) \in \mathbb{N}$}
      \EndFor
    \State 1. define matrix $\boldsymbol{D}^{N \times hash.size}$ with $d_{ih} = 1$ if $ind_l = h$ and $d_{ih} = 0$ if $ind_l \neq h$, $x^{train}_i = l$
    \State 2. $\boldsymbol{D} \rightarrow \tilde{\boldsymbol{D}}^{N \times V}$ with $V \leq hash.size$, by dropping constant columns in $\boldsymbol{D}$
      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}\ {$\hat{x}^{train}_{iv} = \tilde{d}_{iv}$}
      \EndFor
    \State \underline{Prediction:}
    \For{$x^{new}$}
      \State $ind^{new} = (hash(x^{new}) \mod hash.size) + 1$
      \State $\boldsymbol{d}^{new}$ of length $hash.size$ with $d^{new}_h = 1$ if $ind^{new} = h$ and $d^{new}_h = 1$ if $ind^{new} \neq h$
      \State $\boldsymbol{d}^{new} \rightarrow \tilde{\boldsymbol{d}}^{new}$ of length $V$, by dropping columns which were constant in $\boldsymbol{D}$
        \ForAll{$V$ columns in $\tilde{\boldsymbol{D}}$}\ {$\hat{x}^{new}_v = \tilde{d}^{new}_v$}
        \EndFor
    \EndFor
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
  \caption{Leaf Encoding}
  \label{alg:leaf}
  \begin{algorithmic}
    \State \underline{Training:} require number of cross-validation folds $K \in \mathbb{N}$
    \State fit CART tree on $\mathcal{D}^{train}$ with complexity pruning based on $K$-fold cross-validation
      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
        $\tilde{x}_i = t$ with $x^{train}_i$ in terminal node $t$
      \EndFor
    \State \underline{Prediction:}
    \For{$x^{new}$}
        \If{$x^{new} \in \mathcal{L}^{train}$}\ {$\tilde{x}^{new} = t$ with $x^{new}$ in terminal node $t$}
        \Else\ {$\tilde{x}^{new} = b$ where $b$ indicates the biggest terminal node}
        \EndIf
    \EndFor
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
  \caption{Impact Encoding Regression}
  \begin{algorithmic}
    \State \underline{Training:} require smoothing parameter $\epsilon \in \mathbb{R}$
    \ForAll{$l \in \mathcal{L}^{train}$}\ {$\delta_l = \frac{\sum_{i:x^{train}_i = l} y^{train}_i + \epsilon \cdot \bar{y}^{train}}{N_l + \epsilon} -\bar{y}^{train}$ with $\bar{y}^{train} = \frac{\sum^{N}_{i=1} y^{train}_i}{N}$}
    \EndFor
    \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}\ {$\hat{x}^{train}_{i} = \delta_{l}$ with $x^{train}_i = l$}
    \EndFor
    \State \underline{Prediction:}
      \For{$x^{new}$}
        \State \algorithmicif\ {$x^{new} \in \mathcal{L}^{train}$}\ \algorithmicthen\ {$\hat{x}^{new} = \delta_{l}$ with $x^{new} = l$}\ \algorithmicelse\ {$\hat{x}^{new} = 0$}
      \EndFor
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
  \caption{Impact Encoding Classification}
  \begin{algorithmic}
    \State \underline{Training:} require smoothing parameter $\epsilon \in \mathbb{R}$
      \ForAll{$c \in \mathcal{C}$}
        \State $p_c = \frac{N_c}{N}$, $p^{new}_c = \frac{N_c + \epsilon}{N + 2\epsilon}$, $logit_c = \log(\frac{p_c}{1-p_c})$, $logit^{new}_c = \log(\frac{p^{new}_c}{1 - p^{new}_c})$
        \State $\delta^{new}_c = logit^{new}_c - logit_c$
        \ForAll{$l \in \mathcal{L}^{train}$}
          \State $p_{lc} = \frac{\sum_{i: x^{train}_i = l} I(y^{train}_i = c) + \epsilon}{N_l + 2 \epsilon}$, $logit_{lc} = \log(\frac{p_{lc}}{1 - p_{lc}})$
          \State $\delta_{lc} = logit_{lc} - logit_c$
        \EndFor
      \EndFor
      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}\ {$\hat{x}^{train}_{ic} = \delta_{lc}$ with $x^{train}_i = l$}
      \EndFor
    \State \underline{Prediction:}
      \For{$x^{new}$}
        \ForAll{$c$ in $\mathcal{C}$}
          \State \algorithmicif\ {$x^{new} \in \mathcal{L}^{train}$}\ \algorithmicthen\ {$\hat{x}^{new}_{c} = \delta_{lc}$ with $x^{new} = l$}\ \algorithmicelse\ {$\hat{x}^{new}_c = \delta^{new}_c$}
        \EndFor
      \EndFor
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
  \caption{GLMM Encoding Regression}
  \label{alg:glmm-regr}
  \begin{algorithmic}
    \State \underline{Training:} require $n.folds \in \mathbb{N}$
     \State fit simple random intercept model: $y^{train}_i = \beta_{0l} + \epsilon_{i} = \gamma_{oo} + u_l + \epsilon_{i}$ on $\mathcal{D}^{train}$
     \State with $u_l \overset{iid}{\sim} N(0,\tau^2)$, $\epsilon_i \overset{iid}{\sim} N(0,\sigma^2)$ and $x^{train}_i = l$, $l \in \mathcal{L}^{train}$
    \If{$n.folds = 1$}
      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
        \State $\hat{x}^{train}_i = \hat{\beta}_{0l}^{\mathcal{D}^{train}}$ with $x^{train}_i = l$
      \EndFor
    \Else\ {use $n.folds$ cross-validation scheme to make training sets $\mathcal{D}^{train}_1, \dots, \mathcal{D}^{train}_{n.folds}$}
      \State and fit simple random intercept model on each $\mathcal{D}^{train}_{m}$
      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
        \State $\hat{x}^{train}_i = \hat{\beta}_{0l}^{\mathcal{D}^{train}}$ with $x^{train}_i = l$ based on the model m
        \State with $(x^{train}_i, y^{train}_i) \notin \mathcal{D}^{train}_m$
      \EndFor
    \EndIf
    \State \underline{Prediction:}
    \For{$x^{new}$}
      \If{$x^{new} \in \mathcal{L}^{train}$}
        \State $\hat{x}^{new} = \hat{\beta}_{0l}^{\mathcal{D}^{train}}$ with $x^{new} = l$ based on full model fitted on $\mathcal{D}^{train}$
      \Else\ {$\hat{x}^{new} = \hat{\gamma}_{00}$ based on full model fitted on $\mathcal{D}^{train}$}
      \EndIf
    \EndFor
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
  \caption{GLMM Encoding Binary Classification}
  \label{alg:glmm-bincl}
  \begin{algorithmic}
    \State \underline{Training:} require $n.folds \in \mathbb{N}$
    \State fit simple glmm: $E(y^{train}_i) = h(\eta_i) = \frac{\exp(\eta_i)}{1 + \exp(\eta_i)}$, $\eta_i = \beta_{0l} = \gamma_{00} + u_l$ on $\mathcal{D}^{train}$
    \State with $u_l \overset{iid}{\sim} N(0,\sigma^2)$, $y^{train}_i \overset{ind}{\sim} Be(h(\eta_i))$ and $x^{train}_i = l$, $l \in \mathcal{L}^{train}$
    \If{$n.folds = 1$}
      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
        \State $\hat{x}^{train}_i = \hat{\beta}_{0l}^{\mathcal{D}^{train}}$ with $x^{train}_i = l$
      \EndFor
    \Else\ {use $n.folds$ cross-validation scheme to make training sets $\mathcal{D}^{train}_1, \dots, \mathcal{D}^{train}_{n.folds}$}
      \State and fit simple glmm on each $\mathcal{D}^{train}_{m}$
      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
        \State $\hat{x}^{train}_i = \hat{\beta}_{0l}^{\mathcal{D}^{train}}$ with $x^{train}_i = l$ based on the model $m$
        \State with $(x^{train}_i, y^{train}_i) \notin \mathcal{D}^{train}_m$
      \EndFor
    \EndIf
    \State \underline{Prediction:}
    \For{$x^{new}$}
      \If{$x^{new} \in \mathcal{L}^{train}$}
        \State $\hat{x}^{new} = \hat{\beta}_{0l}^{\mathcal{D}^{train}}$ with $x^{new} = l$ based on full model fitted on $\mathcal{D}^{train}$
      \Else\ {$\hat{x}^{new} = \hat{\gamma}_{00}$ based on full model fitted on $\mathcal{D}^{train}$}
      \EndIf
    \EndFor
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
  \caption{GLMM Encoding Multiclass Classification}
  \label{alg:glmm-multcl}
  \begin{algorithmic}
    \State \underline{Training:} require $n.folds \in \mathbb{N}$
    \State define response matrix $\boldsymbol{Y}^{N \times C}$ with $y_{ic} = 1$ if $y^{train}_i = c$ and $y_{ic} = 0$ if $y^{train}_i \neq c$:
    \State $\tilde{\mathcal{D}}^{train} = \{(x^{train}_i, y^{train}_{i1}, \dots, y^{train}_{iC}), \dots, (x^{train}_N, y^{train}_{N1}, \dots, y^{train}_{NC})\}$
    \ForAll{$C$ classes}
      \State fit simple glmm: $E(y^{train}_{ic}) = h(\eta_i) = \frac{\exp(\eta_i)}{1 + \exp(\eta_i)}$, $\eta_i = \beta_{0l} = \gamma_{00} + u_l$ on   $\boldsymbol{y}^{train}_c$ from $\tilde{\mathcal{D}}^{train}$
      \State with $u_l \overset{iid}{\sim} N(0,\sigma^2)$, $y^{train}_{ic} \overset{ind}{\sim} Be(h(\eta_i))$ and $x^{train}_i = l$, $l \in \mathcal{L}^{train}$
    \EndFor
    \If{$n.folds = 1$}
      \ForAll{$C$ models}
        \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
          \State $\hat{x}^{train}_{ic} = \hat{\beta}_{0l}^{\tilde{\mathcal{D}}^{train}}$ with $x^{train}_i = l$
        \EndFor
      \EndFor
    \Else\ {use $n.folds$ cross-validation scheme to make training sets $\tilde{\mathcal{D}}^{train}_1, \dots, \tilde{\mathcal{D}}^{train}_{n.folds}$}
    \ForAll{$C$ classes}
      \State fit simple glmm on $\boldsymbol{y}^{train}_c$ from each $\tilde{\mathcal{D}}^{train}_{m}$
      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
        \State $\hat{x}^{train}_{ic} = \hat{\beta}_{0l}^{\tilde{\mathcal{D}}^{train}}$ with $x^{train}_i = l$ based on the model $m$ for class $c$
        \State with $(x^{train}_i, y^{train}_i) \notin \tilde{\mathcal{D}}^{train}_m$
      \EndFor
    \EndFor
    \EndIf
    \State \underline{Prediction:}
    \For{$x^{new}$}
      \ForAll{$C$ class models}
        \If{$x^{new} \in \mathcal{L}^{train}$}
          \State $\hat{x}^{new}_c = \hat{\beta}_{0l}^{\tilde{\mathcal{D}}^{train}}$ with $x^{new} = l$ based on full model for class $c$
        \Else\ {$\hat{x}^{new}_c = \hat{\gamma}_{00}$ based on full model for class $c$}
        \EndIf
      \EndFor
    \EndFor
  \end{algorithmic}
\end{algorithm}

*Note on GLMM Encoders:* To speed up the computation of the GLMM encoders, we followed the performance tips from the lme4 vignette (<https://cran.r-project.org/web/packages/lme4/vignettes/lmerperf.html>): we did not compute derivations (`calc.derivs = FALSE`) and used the `NLOPT_LN_BOBYQA` optimizer from the nloptr package with liberal stopping criteria (`maxeval = 1000`, `xtol_abs = 1e-6`, `ftol_abs = 1e-6`).

# Comparison of One-Hot and Dummy Encoding

A frequently asked question is, whether one-hot encoding or dummy encoding is to be preferred. While dummy encoding clearly is to be preferred for non-regularized linear models in order to avoid multi-colinearity, the answer is less clear for general machine learning algorithms studied in our setting.

```{r, include = FALSE}
library(data.table)
library(ggplot2)
library(gridExtra)
library(grid)
library(ggpubr)
library(papaja)
res = readRDS("../analysis/high_cardinality_benchmark/results.rds")
res[, algorithm := factor(algorithm,
  levels = c("integer", "frequency", "dummy", "hash", "leaf", "cluster", "impact", "lmer", "ranger", "none", "remove"),
  labels = c("integer", "frequency", "indicator", "hash", "leaf", "cluster", "impact", "glmm", "rf", "none", "remove"))]
res[, lrn.id := factor(lrn.id,
  levels = c("cvglmnet", "ranger", "xgboost.earlystop.wrap", "kknn", "liquidsvm", "featureless"),
  labels = c("LASSO", "RF", "GB", "KNN", "SVM", "FL"))]
indicator = res[algorithm == "indicator", .(Classes, auc.test.mean, multiclass.aunu.test.mean, rmse.test.mean),
  keyby = c("lrn.id", "problem", "high.card.thresh", "dummy.enc")]
indicator = dcast(indicator, ... ~ dummy.enc, value.var = c("auc.test.mean", "multiclass.aunu.test.mean", "rmse.test.mean"))
indicator[Classes == 2 & !is.na(auc.test.mean_FALSE > auc.test.mean_TRUE),
  oneh_beats_dummy := auc.test.mean_FALSE > auc.test.mean_TRUE, by = "lrn.id"]
indicator[Classes > 2 & !is.na(multiclass.aunu.test.mean_FALSE > multiclass.aunu.test.mean_TRUE),
  oneh_beats_dummy := multiclass.aunu.test.mean_FALSE > multiclass.aunu.test.mean_TRUE, by = "lrn.id"]
indicator[Classes == 0 & !is.na(rmse.test.mean_FALSE < rmse.test.mean_TRUE),
  oneh_beats_dummy := rmse.test.mean_FALSE < rmse.test.mean_TRUE, by = "lrn.id"]
```

```{r oneh-win, echo = FALSE, results="asis"}
oneh_win = indicator[Classes == 2 & !is.na(auc.test.mean_FALSE > auc.test.mean_TRUE),
  .(BinCl = sum(auc.test.mean_FALSE > auc.test.mean_TRUE)/.N*100), by = "lrn.id"][
indicator[Classes > 2 & !is.na(multiclass.aunu.test.mean_FALSE > multiclass.aunu.test.mean_TRUE),
  .(MultCl = sum(multiclass.aunu.test.mean_FALSE > multiclass.aunu.test.mean_TRUE)/.N*100), by = "lrn.id"],
    on = "lrn.id"][
indicator[Classes == 0 & !is.na(rmse.test.mean_FALSE < rmse.test.mean_TRUE),
  .(Regr = sum(rmse.test.mean_FALSE < rmse.test.mean_TRUE)/.N*100), by = "lrn.id"],
      on = "lrn.id"]
setnames(oneh_win, "lrn.id", "Learner")
setnames(oneh_win, "BinCl", "Binary Class")
setnames(oneh_win, "MultCl", "Multi Class")
setnames(oneh_win, "Regr", "Regression")
apa_table(oneh_win, digits = 0,
  caption = "Win Percentages of One-hot over Dummy Encoding",
  note = "Percentage of encoding conditions in which one-hot performs better than dummy per task setting.")
```

From the results shown in\ref{fig:oneh-win} we can observe, that for most algorithms and datasets, one-hot encoding is to be preferred over dummy-encoding.

# Additional encoders not mentioned in the manuscript

In addition to the encoders mentioned in the manuscript, the benchmark results available in our online repository (<https://github.com/compstat-lmu/paper_2021_categorical_feature_encodings>) contain two additional experimental encoders that were newly developed. We do not mention them in our manuscript because they did not perform well, and in hindsight, we are not satisfied with their design. We briefly mention them here for transparency.

\begin{algorithm}[H]
  \caption{Cluster Encoding}
  \label{alg:cluster}
  \begin{algorithmic}
    \State \underline{Training:} require number of desired levels $V \in \mathbb{N}$
    \If{task is regression}
      \ForAll{$l \in \mathcal{L}^{train}$}\ {$\bar{y}^{train}_l = \frac{\sum_{i:x^{train}_i = l} y^{train}_i}{N_l}$}
      \State define $\boldsymbol{v}_l = (\bar{y}^{train}_l, \delta_l)^T$ with $\delta_l = \left| N_l - \frac{\sum^{L}_{k=1} N_k}{L} \right|$
      \EndFor
    \EndIf
    \If{task is classification}
      \ForAll{$l \in \mathcal{L}^{train}$}
        \ForAll{$c \in \mathcal{C}$}\ {$s_{lc} = \sum_{i: x^{train}_i = l} I(y^{train}_i = c)$}
        \EndFor
        \State define $\boldsymbol{v}_l = (s_{l1}, \dots, s_{lC}, \delta_l)^T$ with $\delta_l = \left| N_l - \frac{\sum^{L}_{k=1} N_k}{L} \right|$
      \EndFor
    \EndIf
    \State 1. compute distance matrix $\boldsymbol{D}^{L \times L}$ with $d_{jk} = \left|\left|\boldsymbol{v}_j - \boldsymbol{v}_k\right|\right|$
    \State 2. fit hierarchical cluster analysis on $\boldsymbol{D}$
    \State 3. prune dendrogram to obtain $V$ combined levels
    \item[]
    \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}\ {$\tilde{x}^{train}_i = v$ with $x^{train}_i$ in leaf $v$}
    \EndFor
    \State \underline{Prediction:}
    \For{$x^{new}$}
        \State \algorithmicif\ {$x^{new} \in \mathcal{L}^{train}$}\ \algorithmicthen\ {$\tilde{x}^{new} = v$ with $x^{new}$ in leaf $v$}\ \algorithmicelse\ {$\tilde{x}^{new} = NA$}
    \EndFor
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
  \caption{RF Encoding Regression and Binary Classification}
  \label{alg:rf-regrbincl}
  \begin{algorithmic}
    \State \underline{Training:} require number of trees $B \in \mathbb{N}$
    \State fit random forest with trees $T_1, ..., T_B$
    \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
      \If{$x^{train}_i$ inbag for all $B$ trees}\ {$\hat{x}^{train}_i = \frac{1}{B}\sum^B_{b=1}T_b(x^{train}_i)$}
      \EndIf
      \State \algorithmicelse\ {$\hat{x}^{train}_i = \frac{1}{B^{OOB}_i}\sum_{b:\ x^{train}_i\ OOB\ T_b}T_b(x^{train}_i)$}
    \EndFor
    \State \underline{Prediction:}
    \For{$x^{new}$}
      \If{$x^{new} \in \mathcal{L}^{train}$}\ {$\hat{x}^{new} = \frac{1}{B}\sum^B_{b=1}T_b(x^{new})$}
      \EndIf
      \State \algorithmicelse\ $\hat{x}^{new} =  \frac{1}{B} \sum^B_{b=1} \frac{1}{L} \sum_{l \in \mathcal{L}^{train}} T_b(l)$
    \EndFor
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
  \caption{RF Encoding Multiclass Classification}
  \label{alg:rf-multcl}
  \begin{algorithmic}
    \State \underline{Training:} require number of trees $B \in \mathbb{N}$
    \State fit random forest with trees $T_1, ..., T_B$
    \ForAll{$C$ classes}
    \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
      \If{$x^{train}_i$ inbag for all $B$ trees}\ {$\hat{x}^{train}_{ic} = \frac{1}{B}\sum^B_{b=1}T^c_b(x^{train}_i)$}
      \EndIf
      \State \algorithmicelse\ {$\hat{x}^{train}_{ic} = \frac{1}{B^{OOB}_i}\sum_{b:\ x^{train}_i\ OOB\ T_b}T^c_b(x^{train}_i)$}
    \EndFor
    \EndFor
    \State \underline{Prediction:}
    \For{$x^{new}$}
    \ForAll{$C$ classes}
      \If{$x^{new} \in \mathcal{L}^{train}$}\ {$\hat{x}^{new}_c = \frac{1}{B}\sum^B_{b=1}T^c_b(x^{new})$}
      \EndIf
      \State \algorithmicelse\ $\hat{x}^{new}_c =  \frac{1}{B} \sum^B_{b=1} \frac{1}{L} \sum_{l \in \mathcal{L}^{train}} T^c_b(l)$
    \EndFor
    \EndFor
  \end{algorithmic}
\end{algorithm}

\clearpage

## Additional datasets not mentioned in the manuscript

In addition to the datasets mentioned in the manuscript, the benchmark results available in our online repository (<https://github.com/compstat-lmu/paper_2021_categorical_feature_encodings>) contain three additional datasets: *KDD98* (OmlId: 41435), *sf-police-incidents* (OmlId: 41436), *Traffic_violations* (OmlId: 41443).
A substantive amount of conditions failed for those rather big datasets due to memory problems, which is why we remove them from the results discussed in our manuscript. We briefly mention them here to provide full transparency.

References {#references .unnumbered}
==========
\raggedright

### 2021-04-01 analysis/high_cardinality_benchmark/algorithms.R

@@ -1,0 +1,350 @@
+# algorithm function for each encoding
+
+resampleWithDummyEncoder = function(job, data, instance, lrn.id, ...) {
+
+  encoder = list(
+    cpo = cpoCollapseFact2() %>>%  
+      cpoDummyEncode(),
+    ps = makeParamSet(makeLogicalParam("dummyencode.reference.cat"),
+      makeDiscreteParam("collapse.fact.2.n.levels", values = HIGH_CARD_THRESH))
+  )
+  
+  pipeline = constructPipeline(lrn.id, instance, encoder)
+  resamplePipeline(instance, pipeline$cpo, pipeline$ps)
+}
+
+resampleWithIntegerEncoder = function(job, data, instance, lrn.id, ...) {
+
+  caseAsNumeric2 = makeCPOCase(pSS(high.card.thresh: integer),
+    properties.needed = "numerics",
+    cpo.build = function (data, target, high.card.thresh) {
+      high_card_feats = getHighCardFeats(instance, high.card.thresh)
+      cpoAsNumeric2(affect.names = high_card_feats)})
+  
+  encoder = list(
+    cpo = caseAsNumeric2(),
+    ps = makeParamSet(makeDiscreteParam("case.high.card.thresh", 
+      values = HIGH_CARD_THRESH))
+  )
+
+  pipeline = constructPipeline(lrn.id, instance, encoder)
+  resamplePipeline(instance, pipeline$cpo, pipeline$ps)
+}
+
+resampleWithFrequencyEncoder = function(job, data, instance, lrn.id, ...) {
+
+  caseFreqEncode = makeCPOCase(pSS(high.card.thresh: integer), 
+    properties.needed = "numerics",
+    cpo.build = function (data, target, high.card.thresh) {
+      high_card_feats = getHighCardFeats(instance, high.card.thresh)
+      cpoFreqEncode(rank = RANK, affect.names = high_card_feats)})
+  
+  encoder = list(
+    cpo = caseFreqEncode(),
+    ps = makeParamSet(makeDiscreteParam("case.high.card.thresh", 
+      values = HIGH_CARD_THRESH))
+  )
+  
+  pipeline = constructPipeline(lrn.id, instance, encoder)
+  resamplePipeline(instance, pipeline$cpo, pipeline$ps)
+}
+
+resampleWithHashEncoder = function(job, data, instance, lrn.id, ...) {
+
+  caseHashEncode = makeCPOCase(pSS(high.card.thresh: integer), 
+    properties.needed = "numerics",
[…]

### 2021-04-01 analysis/high_cardinality_benchmark/batchtools.conf.R

@@ -1,0 +1,6 @@
+cluster.functions = batchtools::makeClusterFunctionsSlurm("slurm-lrz")
+default.resources = list(walltime = 10000L, # in seconds
+  memory = 1700L, nodes = 1L, ntasks = 1L, ncpus = 1L, 
+  clusters = "serial", partition = "serial_mpp2")
+max.concurrent.jobs = 1000L
+

### 2021-04-01 analysis/high_cardinality_benchmark/collect_results.R

@@ -1,0 +1,74 @@
+library(batchtools)
+
+# load registry in read-only mode
+reg = loadRegistry(file.dir = "analysis/high_cardinality_benchmark/registry", writeable = FALSE)
+
+# save all test set evaluations as data.table: list column result, each entry is a data.frame
+res = reduceResultsDataTable(reg = reg, fun = function(x) x$measures.test)
+  
+# new unwrapped data.table with each row refering one test set evaluation
+dat = data.table()
+for(i in seq_row(res)) {
+  dat = rbind(dat, res[i, data.table(job.id, rbindlist(result, fill = TRUE))], fill = TRUE)
+}
+
+# compute test mean, sd, min, max for each job.id
+# aggregate dataset on job.id level
+mes_names = colnames(dat[, multiclass.aunp:brier])
+dat = dat[, c(lapply(.SD, mean), lapply(.SD, sd), lapply(.SD, min), lapply(.SD, max)), 
+  by = job.id, .SDcols = mes_names]
+setnames(dat, old = 2:ncol(dat), paste(mes_names, rep(c("test.mean", "test.sd", "test.min", "test.max"), 
+  each = length(mes_names)), sep = "."))
+
+# load dataset descriptives
+descr_dat = readRDS("analysis/high_cardinality_benchmark/descr_dat.rds")
+# merge job parameters, descriptives, measures
+descr_dat$OmlId = as.character(descr_dat$OmlId)
+dat = unwrap(getJobTable(reg = reg))[descr_dat, on = c("problem==OmlId")][
+  dat, on = "job.id"]
+
+# remove problematic high.card.thresh conditions (in doubt remove highest one)
+# make sure that:
+# no condition is effectively computed multiple times
+# algorithm == "remove" always removes something
+# algorithm %in% thresh_encs alway encode anything
+rmJobs= function(by, dt) {
+  df = dt[, tstrsplit(HighCardLevels, ", ")]
+  df = df[, lapply(.SD, as.numeric)]
+  df = df[, lapply(.SD, function(x) x > dt$high.card.thresh)]
+    
+  ind_1 = rowSums(df) == 0
+  ind_2 = duplicated(df)
+    
+  if (by == "ReInFrLeImLmRa") {
+    return(dt[!(ind_1 | ind_2)])    
+  } else if (by == "Du") {
+    return(dt[!(ind_1 & ind_2)])
[…]

### 2021-04-01 analysis/high_cardinality_benchmark/config.R

@@ -1,0 +1,11 @@
+options(java.parameters = "-Xmx8000m")
+
+library(OpenML)
+library(mlrCPO)
+library(BBmisc)
+library(checkmate)
+
+configureMlr(on.learner.error = "warn", on.learner.warning = "warn",
+  on.par.without.desc = "warn", on.par.out.of.bounds = "warn",
+  on.measure.not.applicable = "stop", show.learner.output = TRUE,
+  on.error.dump = FALSE, show.info = TRUE)

### 2021-04-01 analysis/high_cardinality_benchmark/global_params.R

@@ -1,0 +1,59 @@
+# load OpenML dataset ids defined by doc/high_card_final_datasets.Rmd
+OML_DATA_IDS = readRDS("analysis/high_cardinality_benchmark/oml_ids.rds")
+
+MLR_LRN_IDS = c("featureless", "cvglmnet", "kknn", "ranger", 
+  "liquidSVM", "xgboost.earlystop.wrap", "keras")
+
+CLASSIF_RDESC = makeResampleDesc("CV", iters = 5L, stratify = TRUE)
+REGR_RDESC = makeResampleDesc("CV", iters = 5L)
+
+INNER_CLASSIF_RDESC = makeResampleDesc("CV", iters = 3L, stratify = TRUE)
+INNER_REGR_RDESC = makeResampleDesc("CV", iters = 3L)
+
+TWO_CLASSIF_MES = list(
+  auc, brier, lsr, ssr, ber, kappa,
+  mmce, acc, tp, tn, fp, fn, 
+  timetrain, timepredict)
+MULTI_CLASSIF_MES = list(
+  multiclass.aunu, multiclass.aunp, multiclass.au1p, multiclass.au1u, 
+  multiclass.brier, lsr, ssr, ber, kappa, mmce, acc,
+  timetrain, timepredict
+)
+REGR_MES = list(rmse, mse, medse, mae, medae, rsq,
+  kendalltau, spearmanrho, 
+  timetrain, timepredict)
+
+INNER_TWO_CLASSIF_MES = auc
+INNER_MULTI_CLASSIF_MES = multiclass.aunu
+INNER_REGR_MES = rmse
+
+TUNE_THRESH = FALSE
+CONTROL = makeTuneControlGrid(tune.threshold = TUNE_THRESH)
+
+KEEP_MODELS = FALSE
+KEEP_PREDS = FALSE
+
+SEED = 250319
+REPLS = 1
+
+TUNE_PIPE = TRUE
+
+## encoder hyperpars
+HIGH_CARD_THRESH = c(10L, 25L, 125L)
+
+# dummy
+DUMMY_ENC = c(TRUE, FALSE)
+# frequency
+RANK = c(FALSE)
+# leaf
+PRUNE = c(TRUE)
+# impact
+SMOOTHING = c(1e-04)
+# lmer
+N_FOLDS = c(1L, 5L, 10L)
+# ranger
+NUM_TREES = c(10L, 25L, 50L)
[…]

### 2021-04-01 analysis/high_cardinality_benchmark/learners/RLearner_classif_embed_kerasff_clean.R

@@ -1,0 +1,291 @@
+#' @export
+makeRLearner.classif.embed_kerasff = function() {
+  makeRLearnerClassif(
+    cl = "classif.embed_kerasff",
+    package = "keras",
+    par.set = makeParamSet(
+      makeIntegerLearnerParam(id = "epochs", lower = 0L, default = 30L),
+      makeIntegerLearnerParam(id = "early_stopping_patience", lower = 0L, default = 1L),
+      makeDiscreteLearnerParam(id = "optimizer",  default = "adam",
+        values = c("sgd", "rmsprop", "adagrad", "adadelta", "adam", "nadam")),
+      makeNumericLearnerParam(id = "lr", lower = 0, upper = 1, default = 0.001),
+      makeNumericLearnerParam(id = "decay", lower = 0, upper = 1, default = 0),
+      makeNumericLearnerParam(id = "momentum", lower = 0, upper = 1, default = 0,
+        requires = quote(optimizer == "sgd")),
+      makeNumericLearnerParam(id = "rho", lower = 0, upper = 1, default = 0.001,
+        requires = quote(optimizer == "rmsprop")),
+      makeNumericLearnerParam(id = "beta_1", lower = 0, upper = 1, default = 0.9,
+        requires = quote(optimizer %in% c("adam", "nadam"))),
+      makeNumericLearnerParam(id = "beta_2", lower = 0, upper = 1, default = 0.999,
+        requires = quote(optimizer %in% c("adam", "nadam"))),
+      makeIntegerLearnerParam(id = "batch_size", lower = 1L, upper = Inf, default = 1L),
+      makeIntegerLearnerParam(id = "n_layers", lower = 1L, upper = 4L, default = 1L),
+      makeNumericLearnerParam(id = "dropout_rate", default = 0, lower = 0, upper = 1),
+      makeIntegerLearnerParam(id = "embed_size", lower = 1, upper = Inf, default = NULL, special.vals = list(NULL)),
+      makeNumericLearnerParam(id = "embed_dropout_rate", default = 0.05, lower = 0, upper = 1),
+      # Neurons / Layers
+      makeIntegerLearnerParam(id = "units_layer1", lower = 1L, default = 1L),
+      makeIntegerLearnerParam(id = "units_layer2", lower = 1L, default = 1L,
+        requires = quote(layers >= 2)),
+      makeIntegerLearnerParam(id = "units_layer3", lower = 1L, default = 1L,
+        requires = quote(layers >= 3)),
[…]

### 2021-04-01 analysis/high_cardinality_benchmark/learners/RLearner_regr_embed_kerasff_clean.R

@@ -1,0 +1,280 @@
+#' @export
+makeRLearner.regr.embed_kerasff = function() {
+  makeRLearnerRegr(
+    cl = "regr.embed_kerasff",
+    package = "keras",
+    par.set = makeParamSet(
+      makeDiscreteLearnerParam(id = "loss", default = "mean_squared_error",
+        values = c("mean_squared_error", "mean_squared_logarithmic_error", "mean_absolute_error")),
+      makeIntegerLearnerParam(id = "epochs", lower = 0L, default = 30L),
+      makeIntegerLearnerParam(id = "early_stopping_patience", lower = 0L, default = 1L),
+      makeDiscreteLearnerParam(id = "optimizer",  default = "adam",
+        values = c("sgd", "rmsprop", "adagrad", "adadelta", "adam", "nadam")),
+      makeNumericLearnerParam(id = "lr", lower = 0, upper = 1, default = 0.001),
+      makeNumericLearnerParam(id = "decay", lower = 0, upper = 1, default = 0),
+      makeNumericLearnerParam(id = "momentum", lower = 0, upper = 1, default = 0,
+        requires = quote(optimizer == "sgd")),
+      makeNumericLearnerParam(id = "rho", lower = 0, upper = 1, default = 0.001,
+        requires = quote(optimizer == "rmsprop")),
+      makeNumericLearnerParam(id = "beta_1", lower = 0, upper = 1, default = 0.9,
+        requires = quote(optimizer %in% c("adam", "nadam"))),
+      makeNumericLearnerParam(id = "beta_2", lower = 0, upper = 1, default = 0.999,
+        requires = quote(optimizer %in% c("adam", "nadam"))),
+      makeIntegerLearnerParam(id = "batch_size", lower = 1L, upper = Inf, default = 1L),
+      makeIntegerLearnerParam(id = "n_layers", lower = 1L, upper = 4L, default = 1L),
+      makeNumericLearnerParam(id = "dropout_rate", default = 0, lower = 0, upper = 1),
+      makeIntegerLearnerParam(id = "embed_size", lower = 1, upper = Inf, default = NULL, special.vals = list(NULL)),
+      makeNumericLearnerParam(id = "embed_dropout_rate", default = 0.05, lower = 0, upper = 1),
+      # Neurons / Layers
+      makeIntegerLearnerParam(id = "units_layer1", lower = 1L, default = 1L),
+      makeIntegerLearnerParam(id = "units_layer2", lower = 1L, default = 1L,
+        requires = quote(layers >= 2)),
[…]

### 2021-04-01 analysis/high_cardinality_benchmark/learners/RLearner_regr_kerasff.R

@@ -1,0 +1,134 @@
+#' @export
+makeRLearner.regr.kerasff = function() {
+  makeRLearnerRegr(
+    cl = "regr.kerasff",
+    package = "keras",
+    par.set = makeParamSet(
+      makeIntegerLearnerParam(id = "epochs", lower = 1L, default = 30L),
+      makeIntegerLearnerParam(id = "early_stopping_patience", lower = 0L, default = 2L),
+      makeDiscreteLearnerParam(id = "optimizer",  default = "sgd",
+        values = c("sgd", "rmsprop", "adagrad", "adadelta", "adam", "nadam")),
+      makeNumericLearnerParam(id = "lr", lower = 0, upper = 1, default = 0.001),
+      makeNumericLearnerParam(id = "decay", lower = 0, upper = 1, default = 0),
+      makeNumericLearnerParam(id = "momentum", lower = 0, upper = 1, default = 0,
+        requires = quote(optimizer == "sgd")),
+      makeNumericLearnerParam(id = "rho", lower = 0, upper = 1, default = 0.001,
+        requires = quote(optimizer == "rmsprop")),
+      makeNumericLearnerParam(id = "beta_1", lower = 0, upper = 1, default = 0.9,
+        requires = quote(optimizer %in% c("adam", "nadam"))),
+      makeNumericLearnerParam(id = "beta_2", lower = 0, upper = 1, default = 0.999,
+        requires = quote(optimizer %in% c("adam", "nadam"))),
+      makeDiscreteLearnerParam(id = "loss",
+        values = c("categorical_crossentropy", "sparse_categorical_crossentropy"),
+        default = "categorical_crossentropy"),
+      makeIntegerLearnerParam(id = "batch_size", lower = 1L, upper = Inf, default = 1L),
+      makeIntegerLearnerParam(id = "layers", lower = 1L, upper = 4L, default = 1L),
+      makeDiscreteLearnerParam(id = "batchnorm_dropout",
+        values = c("batchnorm", "dropout", "none"), default = "none"),
+      makeNumericLearnerParam(id = "input_dropout_rate", default = 0, lower = 0, upper = 1, requires = quote(batchnorm_dropout == "dropout")),
+      makeNumericLearnerParam(id = "dropout_rate", default = 0, lower = 0, upper = 1, requires = quote(batchnorm_dropout == "dropout")),
+      # Neurons / Layers
[…]

### 2021-04-01 analysis/high_cardinality_benchmark/learners/liquidsvm.R

@@ -1,0 +1,126 @@
+# copied from https://github.com/liquidSVM/liquidSVM/blob/03df5c8841dac0466cb7a7f487e44e82858832f6/bindings/R/liquidSVM/R/mlr.R
+# change the names to liquidsvm to avoid strange clashes with the mlr-integrated liquidSVM learners
+
+commonParamSet <- function() {
+  if(!requireNamespace('mlr', quietly=TRUE)) stop("this function needs mlr to be installed")
+  if(!requireNamespace('ParamHelpers', quietly=TRUE)) stop("this function needs ParamHelpers to be installed")
+  ParamHelpers::makeParamSet(
+    ParamHelpers::makeLogicalLearnerParam(id = "scale", default = TRUE),
+    ParamHelpers::makeDiscreteLearnerParam(id = "kernel", default = "gauss_rbf",
+                                           values = c("gauss_rbf","poisson")),
+    ParamHelpers::makeIntegerLearnerParam(id = "partition_choice", default = 0, lower = 0, upper = 6),
+    ParamHelpers::makeNumericLearnerParam(id = "partition_param", default = -1,
+                                          requires = quote(partition_choice >= 1L)),
+    ParamHelpers::makeIntegerLearnerParam(id = "grid_choice", default = 0, lower = -2, upper = 2),
+    ParamHelpers::makeIntegerLearnerParam(id = "folds", default = 5, lower = 1),
+    ParamHelpers::makeNumericLearnerParam(id = "min_gamma", lower=0),
+    ParamHelpers::makeNumericLearnerParam(id = "max_gamma", lower=0,requires = quote(min_gamma <= max_gamma)),
+    ParamHelpers::makeIntegerLearnerParam(id = "gamma_steps", lower=0),
+    ParamHelpers::makeNumericLearnerParam(id = "min_lambda", lower=0),
+    ParamHelpers::makeNumericLearnerParam(id = "max_lambda", lower=0,requires = quote(min_lambda <= max_lambda)),
+    ParamHelpers::makeIntegerLearnerParam(id = "lambda_steps", lower=0),
+    ParamHelpers::makeDiscreteLearnerParam(id = "retrain_method", default = "select_on_each_fold",
+                                           values = c("select_on_entire_train_Set","select_on_each_fold")),
+    ParamHelpers::makeLogicalLearnerParam(id = "store_solutions_internally", default = TRUE),
[…]

### 2021-04-01 analysis/high_cardinality_benchmark/learners/xgboost.earlystop.wrap.R

@@ -1,0 +1,237 @@
+# adapted from https://github.com/ja-thomas/autoxgboost/tree/master/R
+# replace param early_stopping_data with early_stopping_split to automatically select
+# the early_stopping data based on the current training set
+# NOTE: the learners cannot handle weights any more
+
+makeRLearner.classif.xgboost.earlystop.wrap = function() {
+  makeRLearnerClassif(
+    cl = "classif.xgboost.earlystop.wrap",
+    package = "xgboost",
+    par.set = makeParamSet(
+      makeNumericLearnerParam(id = "eta", default = 0.3, lower = 0, upper = 1),
+      makeNumericLearnerParam(id = "gamma", default = 0, lower = 0),
+      makeIntegerLearnerParam(id = "max_depth", default = 6L, lower = 1L),
+      makeNumericLearnerParam(id = "min_child_weight", default = 1, lower = 0),
+      makeNumericLearnerParam(id = "subsample", default = 1, lower = 0, upper = 1),
+      makeNumericLearnerParam(id = "colsample_bytree", default = 1, lower = 0, upper = 1),
+      makeNumericLearnerParam(id = "colsample_bylevel", default = 1, lower = 0, upper = 1),
+      makeIntegerLearnerParam(id = "num_parallel_tree", default = 1L, lower = 1L),
+      makeNumericLearnerParam(id = "lambda", default = 0, lower = 0),
+      makeNumericLearnerParam(id = "lambda_bias", default = 0, lower = 0),
+      makeNumericLearnerParam(id = "alpha", default = 0, lower = 0),
+      makeUntypedLearnerParam(id = "objective", default = NULL, tunable = FALSE),
+      makeUntypedLearnerParam(id = "eval_metric", default = NULL, tunable = FALSE),
+      makeNumericLearnerParam(id = "base_score", default = 0.5, tunable = FALSE),
+      makeNumericLearnerParam(id = "early_stopping_split", default = 0.8, lower = 0, upper = 1),
+      makeIntegerLearnerParam(id = "early_stopping_rounds", default = 1, lower = 1L, tunable = FALSE),
+      makeIntegerLearnerParam(id = "max.nrounds", default = 10^6L, lower = 1L, upper = 10^7L),
+      makeLogicalLearnerParam(id = "maximize", default = NULL, special.vals = list(NULL), tunable = FALSE),
+      makeNumericLearnerParam(id = "scale_pos_weight", lower = 0),
[…]

### 2021-04-01 analysis/high_cardinality_benchmark/main.R

@@ -1,0 +1,74 @@
+library(batchtools)
+library(OpenML)
+library(mlr)
+
+library(data.table)
+setDTthreads(threads = 1)
+
+# source globals params
+source("analysis/high_cardinality_benchmark/global_params.R")
+
+# configure OpenML
+saveOMLConfig(cachedir = "~/.openml/cache", arff.reader = "RWeka", overwrite = TRUE)
+
+# download OML datasets and store them locally
+populateOMLCache(data.ids = OML_DATA_IDS, overwrite = FALSE)
+
+# make batchtools registry
+reg_dir = "analysis/high_cardinality_benchmark/registry"
+reg = makeExperimentRegistry(reg_dir, 
+  source = c("analysis/high_cardinality_benchmark/config.R",
+    "analysis/high_cardinality_benchmark/global_params.R",
+    "analysis/high_cardinality_benchmark/problems.R", 
+    "analysis/high_cardinality_benchmark/algorithms.R",
+    "analysis/high_cardinality_benchmark/pipeline.R",
+    "analysis/high_cardinality_benchmark/mlrCPOs.R",
+    "analysis/high_cardinality_benchmark/learners/xgboost.earlystop.wrap.R",
+    "analysis/high_cardinality_benchmark/learners/RLearner_classif_kerasff.R",
+    "analysis/high_cardinality_benchmark/learners/RLearner_regr_kerasff.R",
+    "analysis/high_cardinality_benchmark/learners/RLearner_classif_embed_kerasff_clean.R",
+    "analysis/high_cardinality_benchmark/learners/RLearner_regr_embed_kerasff_clean.R"),
+  seed = SEED)
+setDefaultRegistry(loadRegistry(reg_dir, writeable = TRUE))
+
+# add problems
+for (did in OML_DATA_IDS) {
+  addProblem(name = as.character(did), data = did, fun = provideTaskWithPreProc,
+    seed = SEED)
+}
+
+# add algorithms
+addAlgorithm(name = "dummy", fun = resampleWithDummyEncoder)
+addAlgorithm(name = "integer", fun = resampleWithIntegerEncoder)
+addAlgorithm(name = "frequency", fun = resampleWithFrequencyEncoder)
[…]

### 2021-04-01 analysis/high_cardinality_benchmark/mlrCPOs.R

@@ -1,0 +1,1508 @@
+#' @title Convert All Features to Numerics
+#'
+#' @template cpo_doc_intro
+#'
+#' @description
+#' Converts all \code{feature} columns to randomized (integer) \code{numeric} 
+#' columns. Missing values are ignored by the CPO. New factor levels are converted
+#' to NAs during prediction.
+#'
+#'#' @section CPOTrained State:
+#' The state's \code{$control} slot is a named list, containing for each factorial
+#' feature a named integer vector indicating the numeric value for each observed
+#' factor level.
+#'
+#' @template cpo_doc_outro
+#' @export
+cpoAsNumeric2 = makeCPO("as.numeric.2", properties.adding = c("factors", "ordered"), 
+  properties.needed = "numerics",
+  dataformat = "factor",
+  fix.factors = TRUE,
+  cpo.train = function(data, target){
+    lapply(data, function(col) {
+      setNames(1:nlevels(col), nm = sample(levels(col)))
+    })
+  }, 
+  cpo.retrafo = function(data, control) {
+    data[] = lapply(colnames(data), function(cname) {
+      as.numeric(control[[cname]][as.character(data[[cname]])])
+      })
+    data
+  })
+
+#' @title Standardize Feature Names
+#'
+#' @template cpo_doc_intro
+#'
+#' @description
+#' Create standardized feature names (Feat1, Feat2, ...) to avoid rare bugs
+#' if some learners cannot handle special symbols in column names (e.g. liquidSVM). 
+#' Useful after CPO encoders that name new features based on factor levels.
+#'
+#' @section CPOTrained State:
+#' NULL
+#' 
+#' @template cpo_doc_outro
+#' @export
+cpoStandardizeFeatNames = makeCPO("standardize.feat.names",
+  dataformat = "df.features",
+  cpo.train = NULL,
+  cpo.retrafo = function(data) {
+    if (ncol(data) > 0) {
+      names(data) = paste0("..Feat", 1:ncol(data), "..")
+    }
+    data
+  })
+
+#' @title New Factor Level for Missing Values
+#'
+#' @template cpo_doc_intro
+#'
+#' @description
+#' This CPO creates a new factor level labeled ..NA.. for factorial columns
+#' containing missing values.
+#'
+#' @section CPOTrained State:
+#' NULL
+#' 
+#' @template cpo_doc_outro
+#' @export
+cpoNaToFactLvl = makeCPO("na.to.fact.lvl",
[…]

### 2021-04-01 analysis/high_cardinality_benchmark/pipeline.R

@@ -1,0 +1,139 @@
+# functions to build the pipeline, the learners, and the resampling
+
+constructPipeline = function(lrn.id, task, encoder) {
+  lrn = constructLearnerCPO(lrn.id, task)
+  pipeline = 
+    # impute all missing values and new levels
+    cpoImputeAll(classes = list(numeric = imputeMean(), factor = imputeConstant("..NA..")), 
+      impute.new.levels = TRUE, recode.factor.levels = TRUE) %>>%
+    # add encoding
+    encoder$cpo %>>% 
+    # remove constant features
+    cpoDropConstants() %>>%
+    # one-hot encode the remaining factors with nlevels < high.card.thresh
+    cpoDummyEncode(reference.cat = FALSE) %>>%
+    # avoid strange bugs with some learners (liquidsvm)
+    cpoStandardizeFeatNames() %>>%
+    # add learner
+    lrn$cpo 
+  pipeline_ps = c(lrn$ps, encoder$ps)
+  return(list(cpo = pipeline, ps = pipeline_ps))
+}
+
+
+constructLearnerCPO = function(lrn.id, task) {
+  task_type = getTaskType(task)
+  pred_type = ifelse(task_type == "regr", "response", "prob")
+  lrn_id = paste(task_type, lrn.id, sep = ".")
+  switch(lrn.id,
+    featureless = {
+      par_vals = list()
+      lrn = makeLearner(lrn_id, predict.type = pred_type, par.vals = par_vals)
+      ps = makeParamSet()
+    },
+    cvglmnet = {
+      par_vals = list(alpha = 1, nfolds = 5, nlambda = 100, s = "lambda.min", 
+        intercept = TRUE, standardize = TRUE)
+      lrn = makeLearner(lrn_id, predict.type = pred_type, par.vals = par_vals)
+      ps = makeParamSet()
+    },
+    kknn = {
+      par_vals = list(k = 15, distance = 2, kernel = "rectangular", scale = TRUE)
+      lrn = makeLearner(lrn_id, predict.type = pred_type, par.vals = par_vals)
+      lrn = cpoFilterFeatures(method = "FSelectorRcpp_information.gain", abs = 25) %>>% lrn
+      ps = makeParamSet()
+    },
+    ranger = {
+      par_vals = list(num.trees = 500, respect.unordered.factors = "order", 
[…]

### 2021-04-01 analysis/high_cardinality_benchmark/problems.R

@@ -1,0 +1,13 @@
+# problem function to prepare the datasets
+
+provideTaskWithPreProc = function(job, data) {
+  # read dataset from cache
+  oml_dat = getOMLDataSet(data.id = data, cache.only = TRUE)
+  # convert dataset to task
+  task = convertOMLDataSetToMlr(oml_dat)
+  task = task %>>% 
+    # avoid strange bugs with cpoDummyEncode
+    cpoStandardizeFeatNames()
+  
+  return(task)
+}

### 2021-04-01 analysis/install.R

@@ -1,0 +1,5 @@
+install.packages(c("devtools", "RWeka",
+  "glmnet", "xgboost", "kknn", "liquidSVM", 
+  "FSelectorRcpp", "abind", "FeatureHashing", "lme4", "fastcluster"))
+devtools::install_github(c("mlr-org/mlr", "mlr-org/mlrCPO", "openml/openml-r", "mllg/batchtools", "imbs-hl/ranger"))
+install.packages(c("ggplot2", "knitr", "kableExtra"))

### 2021-04-01 analysis/upload_datasets/Midwest_survey.R

@@ -1,0 +1,58 @@
+options(java.parameters = "-Xmx8000m")
+
+library(OpenML)
+library(data.table)
+
+# load OML config file to access server
+loadOMLConfig(path = "~/.openml/config", assign = TRUE)
+
+# Midwest_survey
+oml_dat = getOMLDataSet(41234)
+desc = oml_dat$desc
+dat = as.data.table(oml_dat$data)
+
+# remove variables
+dat[, c("In.what.ZIP.code.is.your.home.located...enter.5.digit.ZIP.code..for.example..00544.or.94305.") := NULL]
+
+# recode all features as factors
+nominal_vars = colnames(dat[, dat[, sapply(.SD, is.character)], with = FALSE])
+dat[, c(nominal_vars) := lapply(.SD, factor), .SDcols = nominal_vars]
+
+# create new target category for missing values (as was probably done in Cerda et al., 2018)
+dat$Location..Census.Region. = addNA(dat$Location..Census.Region, ifany = TRUE)
+# rename NA level as it might not work with some learners (e.g. ranger)
+levels(dat$Location..Census.Region.)[is.na(levels(dat$Location..Census.Region.))] = "Not Available"
+
+# check dataset for inconsistencies
+str(dat)
+
+# OML dataset
+new_desc = makeOMLDataSetDescription(
+  name = desc$name,
+  creator = desc$creator,
+  collection.date = desc$collection.date,
+  language = desc$language,
+  description = paste(desc$description, 
+                      "For this version, some features were removed and all remaining character features were recoded as nominal factor variables.
+                      The variable 'Location..Census.Region.' is used as target by default. For missing values in the target variable, a new target level was created labeled 'Not Available'."),
+  original.data.url = desc$url,
+  default.target.attribute = "Location..Census.Region.",
+  licence = desc$licence,
+  visibility = desc$visibility
+  )
+
+new_oml_dat = makeOMLDataSet(
+  desc = new_desc,
+  data = dat,
+  colnames.old = colnames(dat),
+  colnames.new = colnames(dat),
+  target.features = "Location..Census.Region.")
+
+# upload dataset to the OML server
+# !DO NOT ACCIDENTALLY RUN THIS!
+
+# uploadOMLDataSet(new_oml_dat)
+
+# Uploading data set to server.
+# Uploading to 'http://www.openml.org/api/v1/data'.
+# Data set successfully uploaded. Data set ID: 41446

### 2021-04-01 analysis/upload_datasets/Traffic_violations.R

@@ -1,0 +1,148 @@
+options(java.parameters = "-Xmx8000m")
+
+library(OpenML)
+library(data.table)
+
+# load OML config file to access server
+loadOMLConfig(path = "~/.openml/config", assign = TRUE)
+
+# Traffic_violations
+oml_dat = getOMLDataSet(41246)
+desc = oml_dat$desc
+dat = as.data.table(oml_dat$data)
+
+# remove variables
+dat[, c("Date.Of.Stop", "Time.Of.Stop", "Agency", "SubAgency", "Location", "Latitude", "Longitude", "Accident", "Article", "Geolocation") := NULL]
+
+# collapse SERO and ESERO to SERO
+levels(dat$Violation.Type) = c("Citation", "SERO", "SERO", "Warning")
+
+# fix factors
+nominal_vars = colnames(dat[, dat[, sapply(.SD, is.character)], with = FALSE])
+
+# remove punctuation
+dat[, c(nominal_vars) := 
+      lapply(.SD, stringr::str_replace_all, pattern = "[[:punct:]]", replacement = ""), 
+    .SDcols = nominal_vars]
+
+dat[, c(nominal_vars) := 
+      lapply(.SD, stringr::str_replace_all, pattern = "`", replacement = ""), 
+    .SDcols = c(nominal_vars)]
+
+dat[, c(nominal_vars) := 
+      lapply(.SD, stringr::str_replace_all, pattern = "-", replacement = ""), 
+    .SDcols = c(nominal_vars)]
+
+dat[, c(nominal_vars) := 
+      lapply(.SD, stringr::str_squish), 
+    .SDcols = nominal_vars]
+
+dat[dat == ""] = NA
+
+# recode all features as factors
+dat[, c(nominal_vars) := lapply(.SD, factor), .SDcols = nominal_vars]
+
+# check dataset for inconsistencies
+str(dat)
+
+# OML dataset
+new_desc = makeOMLDataSetDescription(
+  name = desc$name,
+  creator = desc$creator,
+  collection.date = desc$collection.date,
+  language = desc$language,
+  description = paste(desc$description, 
+                      "For this version, some features were removed and all remaining character features were recoded as nominal factor variables. All punctuation characters were removed from factor levels.
+                      The variable 'Violation.Type' is used as target by default. The smaller target categories 'SERO' and 'ESERO' were collapsed into one category labeled 'SERO'."),
[…]

### 2021-04-01 analysis/upload_datasets/airlines.R

@@ -1,0 +1,71 @@
+# strange memory option for RWeka
+options(java.parameters = "-Xmx8000m")
+
+library(mlr)
+library(mlrCPO)
+library(OpenML)
+
+# source("analysis/high_cardinality_benchmark/mlrCPOs.R")
+
+# This should work for the final datasets!
+
+# configure OpenML
+# always use RWeka as farff does not work with too many factor levels!!
+#saveOMLConfig(cachedir = "~/.openml/cache", arff.reader = "RWeka")
+
+oml_id = 1169
+highcard_vars = c("AirportFrom", "AirportTo", "Flight")
+
+oml_dat = getOMLDataSet(data.id = oml_id)
+# prepare data
+dat = oml_dat$data
+dat$Flight = as.factor(dat$Flight)
+
+oml_dat$data = dat
+
+# keep the extreme Flight feature in the dataset to detect encoding strategies that tend to overfit:
+# lmer encoder overfits with n.folds = 1 but is fine with n.folds = 5 or 10
+# oml_dat$desc$ignore.attribute = "Flight"
+
+task = convertOMLDataSetToMlr(oml_dat)
+set.seed(42)
+task = downsample(task, perc = 0.05)
+# task_no_highcard = subsetTask(task, features = setdiff(getTaskFeatureNames(task), highcard_vars))
+# 
+# # for binary classif, change learner to "classif.ranger" and cpoLmerEncodeRegr() to
+# # cpoLmerEncodeMultiClassif( or cpoLmerEncodeMultiClassif() 
+# 
+# lrn = cpoImputeConstant("NA", affect.type = "factor") %>>% 
+#   cpoImputeMean(affect.type = "numeric") %>>%
+#   cpoDummyEncode(affect.names = setdiff(getTaskFeatureNames(task), highcard_vars)) %>>% 
+#   cpoLmerEncodeTwoClassif(n.folds = 5) %>>% 
+#   makeLearner("classif.ranger", num.threads = 10, predict.type = "prob")
+# 
+# # check whether performance drops if high cardinality features are removed!
+# # (if it takes too long to compute, holdout should be enough)
+# res = crossval(lrn, task, iters = 5, measures = list(auc, mmce)) 
+# res_no_highcard = crossval(lrn, task_no_highcard, iters = 5, measures = list(auc, mmce))
+# # compare performance between conditions
+# (res_no_highcard$aggr - res$aggr) > 0
+
+desc = oml_dat$desc
+new_desc = makeOMLDataSetDescription(
+  name = desc$name,
[…]

### 2021-04-01 analysis/upload_datasets/ames_housing.R

@@ -1,0 +1,39 @@
+library("OpenML")
+library(data.table)
+
+# load OML config file to access server
+loadOMLConfig(path = "~/.openml/config", assign = TRUE)
+
+# load dataset
+library(AmesHousing)
+dat = as.data.table(make_ames())
+
+# check dataset for inconsistencies
+summary(dat)
+
+# OML description
+desc = makeOMLDataSetDescription(
+  name = "ames-housing",
+  description = "A processed version of the 'Ames Iowa Housing' dataset as provided by the make_ames() function in the R-package 'AmesHousing' [Max Kuhn (2017). AmesHousing: The Ames Iowa Housing Data. R package version 0.0.3. https://CRAN.R-project.org/package=AmesHousing]. The original data was published in [De Cock, D. (2011). 'Ames, Iowa: Alternative to the Boston Housing Data as an End of Semester Regression Project', Journal of Statistics Education, Volume 19, Number 3]. For a description of the dataset checkout either the documentation of the R-package or the original publication. The variable 'Sale_Price' was chosen as target variable. Note that all factors are unordered in this version of the dataset as provided by the make_ames() function, in contrast to the version provided by the make_ordinal_ames() function.",
+  licence = NA,
+  paper.url = "https://www.tandfonline.com/doi/abs/10.1080/10691898.2011.11889627",
+  default.target.attribute = "Sale_Price",
+  visibility = "Everyone"
+)
+
+# create OML dataset
+oml_dat = makeOMLDataSet(
+  desc = desc,
+  data = dat,
+  colnames.old = colnames(dat),
+  colnames.new = colnames(dat),
+  target.features = "Sale_Price")
+
+# upload dataset to the OML server
+# !DO NOT ACCIDENTALLY RUN THIS!
+
+# uploadOMLDataSet(oml_dat)
+
+# Uploading data set to server.
+# Uploading to 'http://www.openml.org/api/v1/data'.
+# Data set successfully uploaded. Data set ID: 41211

### 2021-04-01 analysis/upload_datasets/avocado-sales.R

@@ -1,0 +1,46 @@
+library("OpenML")
+library(data.table)
+
+# load OML config file to access server
+loadOMLConfig(path = "~/.openml/config", assign = TRUE)
+
+# load dataset
+dat = fread(file = "data/avocado_sales/avocado.csv")
+
+# basic preprocessing
+
+dat[, "V1" := NULL]
+
+nominal_vars = c("type", "year", "region")
+dat[, c(nominal_vars) := lapply(.SD, factor), .SDcols = nominal_vars]
+
+# check dataset for inconsistencies
+summary(dat)
+
+# OML description
+desc = makeOMLDataSetDescription(
+  name = "avocado-sales",
+  description = "Historical data on avocado prices and sales volume in multiple US markets. Downloaded from Kaggle [https://www.kaggle.com/neuromusic/avocado-prices/home] on 29.10.2018. The original data stems from the HASS AVOCADO BOARD [http://www.hassavocadoboard.com/retail/volume-and-price-data]. The Kaggle dataset was licensed under the Open Database License (ODbL) [https://opendatacommons.org/licenses/odbl/1.0/]. The variable 'AveragePrice' was selected as target variable. For a description of all variables checkout the Kaggle dataset repo or the original dataset description by the HASS AVOCADO BOARD. 'Year' is coded as a categorical features as the dataset covers only the years 2015-2018. The dataset also includes a 'Date' variable (ignored by default) which can be used to construct additional month or day features. The ID variable from the Kaggle version was removed from the dataset.",
+  licence = "Open Database License (ODbL)",
+  original.data.url = "http://www.hassavocadoboard.com/retail/volume-and-price-data",
+  default.target.attribute = "AveragePrice",
+  ignore.attribute = "Date",
+  visibility = "Everyone"
+)
+
+# create OML dataset
+oml_dat = makeOMLDataSet(
+  desc = desc,
+  data = dat,
+  colnames.old = colnames(dat),
+  colnames.new = colnames(dat),
+  target.features = "AveragePrice")
+
+# upload dataset to the OML server
+# !DO NOT ACCIDENTALLY RUN THIS!
+
+# uploadOMLDataSet(oml_dat)
+
+# Uploading data set to server.
+# Uploading to 'http://www.openml.org/api/v1/data'.
+# Data set successfully uploaded. Data set ID: 41210

### 2021-04-01 analysis/upload_datasets/beer.R

@@ -1,0 +1,77 @@
+# strange memory option for RWeka
+options(java.parameters = "-Xmx8000m")
+
+library(mlr)
+library(mlrCPO)
+library(OpenML)
+
+# source("analysis/high_cardinality_benchmark/mlrCPOs.R")
+
+# This should work for the final datasets!
+
+# configure OpenML
+# always use RWeka as farff does not work with too many factor levels!!
+#saveOMLConfig(cachedir = "~/.openml/cache", arff.reader = "RWeka")
+
+oml_id = 42078
+highcard_vars = c("brewery_id", "review_profilename", "beer_beerid", "beer_style")
+
+oml_dat = getOMLDataSet(data.id = oml_id)
+# prepare data
+dat = oml_dat$data
+dat$brewery_id = as.factor(dat$brewery_id)
+dat$review_profilename = as.factor(dat$review_profilename)
+dat$beer_beerid = as.factor(dat$beer_beerid)
+dat$review_time = dat$review_time / (60 * 60 * 24 * 365) + 1970
+dat$beer_style = as.factor(dat$beer_style)
+
+oml_dat$data = dat
+oml_dat$desc$ignore.attribute = c("brewery_name", "review_aroma", 
+                                  "review_apparance", "review_palate", 
+                                  "review_taste", "beer_abv", "beer_name")
+oml_dat$target.features = "review_overall"
+oml_dat$desc$default.target.attribute = "review_overall"
+
+task = convertOMLDataSetToMlr(oml_dat)
+# downsample to 2% because performance and tendency for overfitting is high enough
+set.seed(42)
+task = downsample(task, perc = 0.02)
+# task_no_highcard = subsetTask(task, features = setdiff(getTaskFeatureNames(task), highcard_vars))
+# 
+# # for binary classif, change learner to "classif.ranger" and cpoLmerEncodeRegr() to
+# # cpoLmerEncodeMultiClassif( or cpoLmerEncodeMultiClassif() 
+# 
+# lrn = cpoImputeConstant("NA", affect.type = "factor") %>>% 
+#   cpoImputeMean(affect.type = "numeric") %>>%
+#   cpoDummyEncode(affect.names = setdiff(getTaskFeatureNames(task), highcard_vars)) %>>% 
+#   cpoLmerEncodeRegr(n.folds = 10) %>>% 
+#   makeLearner("regr.ranger", num.threads = 10)
+# 
+# # check whether performance drops if high cardinality features are removed!
+# # (if it takes too long to compute, holdout should be enough)
[…]

### 2021-04-01 analysis/upload_datasets/churn.R

@@ -1,0 +1,36 @@
+library(OpenML)
+
+# churn
+oml_dat = getOMLDataSet(40701)
+desc = oml_dat$desc
+dat = oml_dat$data
+
+# fix dataset
+dat$state = factor(dat$state)
+dat$phone_number = NULL
+
+new_desc = makeOMLDataSetDescription(
+  name = desc$name,
+  description = paste(desc$description, 
+    "For this version, the feature 'phone_number' was removed and 'state' was recoded as a nominal factor variable."),
+  original.data.url = desc$url,
+  default.target.attribute = desc$default.target.attribute,
+  licence = desc$licence,
+  visibility = "Everyone"
+)
+
+new_oml_dat = makeOMLDataSet(
+  desc = new_desc,
+  data = dat,
+  colnames.old = colnames(dat),
+  colnames.new = colnames(dat),
+  target.features = "class")
+
+# upload dataset to the OML server
+# !DO NOT ACCIDENTALLY RUN THIS!
+
+# uploadOMLDataSet(new_oml_dat)
+
+# Uploading data set to server.
+# Uploading to 'http://www.openml.org/api/v1/data'.
+# Data set successfully uploaded. Data set ID: 41283

### 2021-04-01 analysis/upload_datasets/click_prediction_small.R

@@ -1,0 +1,46 @@
+library(OpenML)
+library(data.table)
+
+# load OML config file to access server
+loadOMLConfig(path = "~/.openml/config", assign = TRUE)
+
+# Click_prediction_small
+oml_dat = getOMLDataSet(1220)
+desc = oml_dat$desc
+dat = as.data.table(oml_dat$data)
+
+# code id features as factors
+more_nominal_vars = c("ad_id", "advertiser_id", "keyword_id", "title_id", 
+  "description_id", "user_id")
+dat[, c(more_nominal_vars) := lapply(.SD, factor), .SDcols = more_nominal_vars]
+
+# check dataset for inconsistencies
+summary(dat)
+
+# OML description
+new_desc = makeOMLDataSetDescription(
+  name = desc$name,
+  description = "This is the same data as version 5 (OpenML ID = 1220) with '_id' features coded as nominal factor variables.",
+  original.data.url = desc$url,
+  default.target.attribute = desc$default.target.attribute,
+  ignore.attribute = c("url_hash", "query_id"),
+  licence = desc$licence,
+  visibility = "Everyone"
+)
+
+# OML dataset
+new_oml_dat = makeOMLDataSet(
+  desc = new_desc,
+  data = dat,
+  colnames.old = colnames(dat),
+  colnames.new = colnames(dat),
+  target.features = "click")
+
+# upload dataset to the OML server
+# !DO NOT ACCIDENTALLY RUN THIS!
+
+# uploadOMLDataSet(new_oml_dat)
+
+# Uploading data set to server.
+# Uploading to 'http://www.openml.org/api/v1/data'.
+# Data set successfully uploaded. Data set ID: 41434

### 2021-04-01 analysis/upload_datasets/delays_zurich.R

@@ -1,0 +1,73 @@
+# strange memory option for RWeka
+options(java.parameters = "-Xmx8000m")
+
+library(mlr)
+library(mlrCPO)
+library(OpenML)
+
+# source("analysis/high_cardinality_benchmark/mlrCPOs.R")
+
+# This should work for the final datasets!
+
+# configure OpenML
+# always use RWeka as farff does not work with too many factor levels!!
+#saveOMLConfig(cachedir = "~/.openml/cache", arff.reader = "RWeka")
+
+oml_id = 40753
+highcard_vars = c("stop_id", "line_number")
+
+oml_dat = getOMLDataSet(data.id = oml_id)
+# prepare data
+dat = oml_dat$data
+
+anything_to_factor = c("vehicle_type", "line_number", "direction", "stop_id",
+                       "weekday")
+dat[, anything_to_factor] = lapply(dat[, anything_to_factor], factor)
+dat$hours = as.numeric(substr(dat$time, 12, 13))
+dat$minutes = as.numeric(substr(dat$time, 15, 16))
+dat$seconds = as.numeric(substr(dat$time, 18, 19))
+dat$time = dat$hours * 60 + dat$minutes + dat$seconds / 60
+
+oml_dat$data = dat
+
+task = convertOMLDataSetToMlr(oml_dat)
+set.seed(47)
+task = downsample(task, perc = 0.005)
+# task_no_highcard = subsetTask(task, features = setdiff(getTaskFeatureNames(task), highcard_vars))
+# 
+# # for binary classif, change learner to "classif.ranger" and cpoLmerEncodeRegr() to
+# # cpoLmerEncodeTwoClassif() or cpoLmerEncodeMultiClassif() 
+# 
+# lrn = cpoImputeConstant("NA", affect.type = "factor") %>>% 
+#   cpoImputeMean(affect.type = "numeric") %>>%
+#   cpoDummyEncode(affect.names = setdiff(getTaskFeatureNames(task), highcard_vars)) %>>% 
+#   cpoLmerEncodeRegr(n.folds = 5) %>>% 
+#   makeLearner("regr.ranger", num.threads = 10)
+# 
+# # check whether performance drops if high cardinality features are removed!
+# # (if it takes too long to compute, holdout should be enough)
+# res = crossval(lrn, task, iters = 5, measures = list(rsq, rmse)) 
+# res_no_highcard = crossval(lrn, task_no_highcard, iters = 5, measures = list(rsq, rmse))
+# # compare performance between conditions
[…]

### 2021-04-01 analysis/upload_datasets/employee_salaries.R

@@ -1,0 +1,54 @@
+options(java.parameters = "-Xmx8000m")
+
+library(OpenML)
+library(data.table)
+
+# load OML config file to access server
+loadOMLConfig(path = "~/.openml/config", assign = TRUE)
+
+# employee_salaries
+oml_dat = getOMLDataSet(41204)
+desc = oml_dat$desc
+dat = as.data.table(oml_dat$data)
+
+# remove variables
+dat[, c("Full_Name", "X2016_Gross_Pay_Received", "X2016_Overtime_Pay", "Department", "Underfilled_Job_Title", "Date_First_Hired") := NULL]
+
+# recode all features as factors
+nominal_vars = colnames(dat[, dat[, sapply(.SD, is.character)], with = FALSE])
+dat[, c(nominal_vars) := lapply(.SD, factor), .SDcols = nominal_vars]
+
+# check dataset for inconsistencies
+str(dat)
+
+# OML dataset
+new_desc = makeOMLDataSetDescription(
+  name = desc$name,
+  creator = desc$creator,
+  collection.date = desc$collection.date,
+  language = desc$language,
+  description = paste(desc$description, 
+                      "For this version, some features were removed and all remaining character features were recoded as nominal factor variables.
+                      The variable 'Current_Annual_Salary' is used as target by default."),
+  original.data.url = desc$url,
+  default.target.attribute = "Current_Annual_Salary",
+  licence = desc$licence,
+  visibility = desc$visibility
+  )
+
+new_oml_dat = makeOMLDataSet(
+  desc = new_desc,
+  data = dat,
+  colnames.old = colnames(dat),
+  colnames.new = colnames(dat),
+  target.features = "Current_Annual_Salary")
+
+# upload dataset to the OML server
+# !DO NOT ACCIDENTALLY RUN THIS!
+
+# uploadOMLDataSet(new_oml_dat)
+
+# Uploading data set to server.
+# Uploading to 'http://www.openml.org/api/v1/data'.
+# Data set successfully uploaded. Data set ID: 41445
+

### 2021-04-01 analysis/upload_datasets/flight_delays.R

@@ -1,0 +1,49 @@
+library("OpenML")
+library(data.table)
+
+# load OML config file to access server
+loadOMLConfig(path = "~/.openml/config", assign = TRUE)
+
+# load dataset
+dat = fread(file = "data/flight_delays/660669749_T_ONTIME_REPORTING_2017_12.csv")
+
+# basic preprocessing
+
+dat[, c("V15", "CANCELLED", "DIVERTED") := NULL]
+dat = dat[!is.na(ARR_DELAY),]
+
+nominal_vars = c("DAY_OF_MONTH", "DAY_OF_WEEK", "OP_UNIQUE_CARRIER", "ORIGIN", 
+  "ORIGIN_STATE_NM", "DEST", "DEST_STATE_NM")
+dat[, c(nominal_vars) := lapply(.SD, factor), .SDcols = nominal_vars]
+
+# check dataset for inconsistencies
+summary(dat)
+
+# OML description
+desc = makeOMLDataSetDescription(
+  name = "flight-delay-usa-dec-2017",
+  description = "Predicting US flight delay in December 2017 based on airline on-time performance data provided by the Bureau of Transportation Statistics (BTS) [https://www.transtats.bts.gov/Tables.asp?DB_ID=120]. Data was downloaded on 03.11.2018. For a description of all variables included in this dataset (many more are available from the BTS website) checkout [https://www.transtats.bts.gov/Fields.asp]. All flights which got canceled or diverted were removed. The dataset includes the variables 'FL_DATE' and 'CRS_DEP_TIME' (ignored by default) which can be used to construct additional time features.",
+  original.data.url = "https://www.transtats.bts.gov/Tables.asp?DB_ID=120",
+  default.target.attribute = "ARR_DELAY",
+  ignore.attribute = c("FL_DATE", "CRS_DEP_TIME"),
+  visibility = "Everyone"
+)
+
+# create OML dataset
+oml_dat = makeOMLDataSet(
+  desc = desc,
+  data = dat,
+  colnames.old = colnames(dat),
+  colnames.new = colnames(dat),
+  target.features = "ARR_DELAY")
+
+# upload dataset to the OML server
+# !DO NOT ACCIDENTALLY RUN THIS!
+
+# uploadOMLDataSet(oml_dat)
+
+# Uploading data set to server.
+# Uploading to 'http://www.openml.org/api/v1/data'.
+# Data set successfully uploaded. Data set ID: 41251
+
+

### 2021-04-01 analysis/upload_datasets/hpc-job-scheduling.R

@@ -1,0 +1,39 @@
+library("OpenML")
+library(data.table)
+
+# load OML config file to access server
+loadOMLConfig(path = "~/.openml/config", assign = TRUE)
+
+# load dataset
+library(AppliedPredictiveModeling)
+data(schedulingData)
+dat = as.data.table(schedulingData)
+
+# check dataset for inconsistencies
+summary(dat)
+
+# OML description
+desc = makeOMLDataSetDescription(
+  name = "hpc-job-scheduling",
+  description = "HPC Job Scheduling Data as included in the R-package 'AppliedPredictiveModeling' [Max Kuhn and Kjell Johnson (2018). AppliedPredictiveModeling: Functions and Data Sets for 'Applied Predictive Modeling'. R package version 1.1-7. https://CRAN.R-project.org/package=AppliedPredictiveModeling]. To obtain the dataset directly in R, call data(schedulingData, package = 'AppliedPredictiveModeling'). For a description of the dataset checkout the documentation of the R-package. The variable 'Class' was chosen as target variable.",
+  licence = NA,
+  default.target.attribute = "Class",
+  visibility = "Everyone"
+)
+
+# create OML dataset
+oml_dat = makeOMLDataSet(
+  desc = desc,
+  data = dat,
+  colnames.old = colnames(dat),
+  colnames.new = colnames(dat),
+  target.features = "Class")
+
+# upload dataset to the OML server
+# !DO NOT ACCIDENTALLY RUN THIS!
+
+# uploadOMLDataSet(oml_dat)
+
+# Uploading data set to server.
+# Uploading to 'http://www.openml.org/api/v1/data'.
+# Data set successfully uploaded. Data set ID: 41212

### 2021-04-01 analysis/upload_datasets/kdd98.R

@@ -1,0 +1,107 @@
+library(OpenML)
+library(data.table)
+
+# load OML config file to access server
+loadOMLConfig(path = "~/.openml/config", assign = TRUE)
+
+# KDD98
+oml_dat = getOMLDataSet(23513)
+desc = oml_dat$desc
+dat = as.data.table(oml_dat$data)
+
+# fix dataset
+dat$TARGET_B = factor(dat$TARGET_B)
+
+more_nominal_vars = c("HPHONE_D", "MHUC2", "INCOME", "WEALTH1", "WEALTH2")
+dat[, c(more_nominal_vars) := lapply(.SD, factor), .SDcols = more_nominal_vars]
+
+# remove constant variable
+dat[, RFA_2R := NULL]
+
+new_desc = makeOMLDataSetDescription(
+  name = desc$name,
+  description = paste(desc$description, 
+    "For this version, the target was correctly encoded as a binary factor. The features 'HPHONE_D', 'MHUC2', 'INCOME', 'WEALTH1', 'WEALTH2' were recoded as nominal factor variables and the constant feature 'RFA_2R' was removed from the dataset."),
+  original.data.url = desc$url,
+  default.target.attribute = desc$default.target.attribute,
+  licence = desc$licence,
+  visibility = "Everyone"
+)
+
+new_oml_dat = makeOMLDataSet(
+  desc = new_desc,
+  data = dat,
+  colnames.old = colnames(dat),
+  colnames.new = colnames(dat),
+  target.features = "TARGET_B")
+
+# upload dataset to the OML server
+# !DO NOT ACCIDENTALLY RUN THIS!
+
+# uploadOMLDataSet(new_oml_dat)
+
+# Uploading data set to server.
+# Uploading to 'http://www.openml.org/api/v1/data'.
+# Data set successfully uploaded. Data set ID: 41435
+
+
+# prepare smaller version (undersample dataset)
+
+options(java.parameters = "-Xmx8000m")
+
+library(OpenML)
+library(mlrCPO)
+
+loadOMLConfig(path = "~/.openml/config", assign = TRUE)
+
+oml_dat = getOMLDataSet(41435)
+desc = oml_dat$desc
+
+task = convertOMLDataSetToMlr(oml_dat)
+# undersample the larger class
+set.seed(42)
+task = undersample(task, rate = 0.4, cl = "0")
+# drop unused levels
+task = task %>>% cpoFixFactors(drop.unused.levels = TRUE, fix.factors.prediction = FALSE)
+
+# # test performance of new dataset
[…]

### 2021-04-01 analysis/upload_datasets/medical_charges.R

@@ -1,0 +1,54 @@
+options(java.parameters = "-Xmx8000m")
+
+library(OpenML)
+library(data.table)
+
+# load OML config file to access server
+loadOMLConfig(path = "~/.openml/config", assign = TRUE)
+
+# medical_charges
+oml_dat = getOMLDataSet(41239)
+desc = oml_dat$desc
+dat = as.data.table(oml_dat$data)
+
+# remove variables
+dat[, c("Provider.Id", "Provider.Street.Address", "Provider.Zip.Code", "Average.Total.Payments", "Average.Covered.Charges") := NULL]
+
+# recode all features as factors
+nominal_vars = colnames(dat[, dat[, sapply(.SD, is.character)], with = FALSE])
+dat[, c(nominal_vars) := lapply(.SD, factor), .SDcols = nominal_vars]
+
+# check dataset for inconsistencies
+str(dat)
+
+# OML dataset
+new_desc = makeOMLDataSetDescription(
+  name = desc$name,
+  creator = desc$creator,
+  collection.date = desc$collection.date,
+  language = desc$language,
+  description = paste(desc$description, 
+                      "For this version, some features were removed and all remaining character features were recoded as nominal factor variables.
+                      The variable 'Average.Medicare.Payments' is used as target by default."),
+  original.data.url = desc$url,
+  default.target.attribute = "Average.Medicare.Payments",
+  licence = desc$licence,
+  visibility = desc$visibility
+  )
+
+new_oml_dat = makeOMLDataSet(
+  desc = new_desc,
+  data = dat,
+  colnames.old = colnames(dat),
+  colnames.new = colnames(dat),
+  target.features = "Average.Medicare.Payments")
+
+# upload dataset to the OML server
+# !DO NOT ACCIDENTALLY RUN THIS!
+
+# uploadOMLDataSet(new_oml_dat)
+
+# Uploading data set to server.
+# Uploading to 'http://www.openml.org/api/v1/data'.
+# Data set successfully uploaded. Data set ID: 41444
+

### 2021-04-01 analysis/upload_datasets/nyc_taxi.R

@@ -1,0 +1,60 @@
+library("OpenML")
+library(data.table)
+
+# load OML config file to access server
+loadOMLConfig(path = "~/.openml/config", assign = TRUE)
+
+# load dataset
+dat = fread(file = "data/nyc_taxi/green_tripdata_2016-12.csv")
+dat = dat[, 1:19]
+colnames(dat) = as.character(
+  read.csv(file = "data/nyc_taxi/green_tripdata_2016-12.csv", 
+    header = F, nrows = 1, as.is = T)
+)
+
+# basic preprocessing
+
+dat[, "ehail_fee" := NULL]
+
+# select only payment type 1 (credit card) 
+# tips are not recorded most remaining types
+dat = dat[payment_type == 1,]
+
+nominal_vars = c("VendorID", "store_and_fwd_flag", "RatecodeID", "PULocationID", 
+  "DOLocationID", "extra", "mta_tax", "improvement_surcharge", 
+  "payment_type", "trip_type")
+dat[, c(nominal_vars) := lapply(.SD, factor), .SDcols = nominal_vars]
+
+# remove variables to increase the importance of categorical features
+rm_vars = c("trip_distance", "payment_type", "fare_amount")
+dat[, (rm_vars) := NULL]
+
+# check dataset for inconsistencies
+summary(dat)
+
+# OML description
+desc = makeOMLDataSetDescription(
+  name = "nyc-taxi-green-dec-2016",
[…]

### 2021-04-01 analysis/upload_datasets/okcupid.R

@@ -1,0 +1,66 @@
+library("OpenML")
+library(data.table)
+
+# load OML config file to access server
+loadOMLConfig(path = "~/.openml/config", assign = TRUE)
+
+# load dataset
+dat = fread(file = "data/jse_okcupid/profiles.csv")
+
+# basic preprocessing
+
+dat[dat == ""] = NA
+dat[dat == -1] = NA
+
+dat = dat[!is.na(job) & !(job %in% c("unemployed", "retired", "rather not say"))]
+dat = dat[, -grep("essay", names(dat), value = TRUE), with = FALSE]
+
+nominal_vars = c("body_type", "diet", "drinks", "drugs", "education", "ethnicity",
+  "income", "job", "location", "offspring", "orientation", "pets", "religion", 
+  "sex", "sign", "smokes", "speaks", "status")
+dat[, c(nominal_vars) := lapply(.SD, factor), .SDcols = nominal_vars]
+
+stem_jobs = c("computer / hardware / software", "science / tech / engineering")
+levels(dat$job) = list(
+  stem = stem_jobs, 
+  non_stem = setdiff(levels(dat$job), c(stem_jobs, "student")), 
+  student = c("student")
+)
+
+# check dataset for inconsistencies
+summary(dat)
+
+# OML description
+desc = makeOMLDataSetDescription(
+  name = "okcupid-stem",
+  description = "User profile data for San Francisco OkCupid users published in [Kim, A. Y., & Escobedo-Land, A. (2015). OKCupid data for introductory statistics and data science courses. Journal of Statistics Education, 23(2).]. The curated dataset was downloaded from [https://github.com/rudeboybert/JSE_OkCupid]. The original dataset was created with the use of a python script that pulled the data from public profiles on www.okcupid.com on 06/30/2012. It includes people (n = 59946) within a 25 mile radius of San Francisco, who were online in the last year (06/30/2011), with at least one profile picture. Permission to use this data was obtained by the author of the original paper from OkCupid president and co-founder Christian Rudder under the condition that the dataset remains public. As target, the variable 'job' was collapsed into three categories: 'stem', 'non_stem', and 'student'. STEM jobs were defined as 'job' %in% c('computer / hardware / software', 'science / tech / engineering'). Observations with 'job' %in% c('unemployed', 'retired', 'rather not say') or missing values in 'job' were removed. The original dataset also included ten open text variables 'essay
[…]

### 2021-04-01 analysis/upload_datasets/open_payments.R

@@ -1,0 +1,42 @@
+library(OpenML)
+library(data.table)
+
+# load OML config file to access server
+loadOMLConfig(path = "~/.openml/config", assign = TRUE)
+
+# open_payments
+oml_dat = getOMLDataSet(41243)
+desc = oml_dat$desc
+dat = as.data.table(oml_dat$data)
+
+# recode all features as factors
+dat[, c(colnames(dat)) := lapply(.SD, factor), .SDcols = colnames(dat)]
+
+new_desc = makeOMLDataSetDescription(
+  name = desc$name,
+  creator = desc$creator,
+  collection.date = desc$collection.date,
+  language = desc$language,
+  description = paste(desc$description, 
+                      "For this version, all features were recoded as nominal factor variables."),
+  original.data.url = desc$url,
+  default.target.attribute = "status",
+  licence = desc$licence,
+  visibility = "public"
+)
+
+new_oml_dat = makeOMLDataSet(
+  desc = new_desc,
+  data = dat,
+  colnames.old = colnames(dat),
+  colnames.new = colnames(dat),
+  target.features = "status")
+
+# upload dataset to the OML server
+# !DO NOT ACCIDENTALLY RUN THIS!
+
+# uploadOMLDataSet(new_oml_dat)
+
+# Uploading data set to server.
+# Uploading to 'http://www.openml.org/api/v1/data'.
+# Data set successfully uploaded. Data set ID: 41442

### 2021-04-01 analysis/upload_datasets/particulate-matter-ukair.R

@@ -1,0 +1,81 @@
+library(OpenML)
+library(data.table)
+library(fasttime)
+
+# load OML config file to access server
+loadOMLConfig(path = "~/.openml/config", assign = TRUE)
+
+# library(rdefra)
+# library(parallelMap)
+# 
+# stations_raw = as.data.table(ukair_catalogue())
+# stations = as.data.table(ukair_get_coordinates(stations_raw))
+# stations$SiteID = ukair_get_site_id(stations$UK.AIR.ID)
+# 
+# parallelStart(mode = "multicore", cpus = 4)
+# 
+# fun = function(x){
+#   tryCatch(as.data.table(ukair_get_hourly_data(x, years = 2017)),
+#     error = function(e)return(data.table()))
+# }
+# dat = parallelLapply(names(table(stations$SiteID, useNA = "no")), fun)
+# 
+# parallelStop()
+# 
+# dat = rbindlist(dat, fill = TRUE)
+# dat = merge(dat, stations)
+
+# !DO NOT ACCIDENTALLY OVERWRITE THIS!
+# fwrite(dat, file = "data/ukair_2017/ukair_2017.csv")
+
+# load dataset
+dat = fread(file = "data/ukair_2017/ukair_2017.csv")
+
+# preprocessing
+
+dat[, datetime := fastPOSIXct(datetime, tz = "GMT")]
+dat[, c("Month", "DayofWeek", "Hour") := .(month(datetime), wday(datetime), hour(datetime))]
+dat[, datetime := as.character(datetime)]
+
+dat = dat[, .(datetime, Hour, Month, DayofWeek, 
+  Site.Name, Environment.Type, Zone, Latitude, Longitude, Altitude..m.,  
+  PM.sub.10..sub..particulate.matter..Hourly.measured., PM.sub.2.5..sub..particulate.matter..Hourly.measured.)]
+dat = dat[complete.cases(dat)]
+
+nominal_vars = c("Month", "DayofWeek", "Site.Name", "Environment.Type", "Zone")
+dat[, c(nominal_vars) := lapply(.SD, factor), .SDcols = nominal_vars]
+
+# remove Longitude and Latitude to increase the importance of Zone and Site.Name
+geo_vars = c("Longitude", "Latitude")
+dat[, (geo_vars) := NULL]
+
+# check dataset for inconsistencies
+summary(dat)
+
+# OML description
+desc = makeOMLDataSetDescription(
+  name = "particulate-matter-ukair-2017",
[…]

### 2021-04-01 analysis/upload_datasets/porto_seguro.R

@@ -1,0 +1,50 @@
+library("OpenML")
+library(data.table)
+
+# load OML config file to access server
+loadOMLConfig(path = "~/.openml/config", assign = TRUE)
+
+# load dataset
+dat = fread(file = "data/porto_seguro/train.csv")
+
+# basic preprocessing
+
+dat[dat == -1] = NA
+  
+nominal_vars = grep("bin|cat", colnames(dat), value = TRUE)
+nominal_vars = c(nominal_vars, "target")
+dat[, c(nominal_vars) := lapply(.SD, factor), .SDcols = nominal_vars]
+
+# remove useless calc variables, see https://www.kaggle.com/c/porto-seguro-safe-driver-prediction/kernels
+calc_vars = grep("calc", colnames(dat), value = TRUE)
+dat[, (calc_vars) := NULL]
+
+# check dataset for inconsistencies
+summary(dat)
+
+# OML description
+desc = makeOMLDataSetDescription(
+  name = "porto-seguro",
+  description = "Training dataset of the 'Porto Seguros Safe Driver Prediction' Kaggle challenge [https://www.kaggle.com/c/porto-seguro-safe-driver-prediction]. The goal was to predict whether a driver will file an insurance claim next year. The official rules of the challenge explicitely state that the data may be used for 'academic research and education, and other non-commercial purposes' [https://www.kaggle.com/c/porto-seguro-safe-driver-prediction/rules]. For a description of all variables checkout the Kaggle dataset repository [https://www.kaggle.com/c/porto-seguro-safe-driver-prediction/data]. It states that numeric features with integer values that do not contain 'bin' or 'cat' in their variable names are in fact ordinal features which could be treated as ordinal factors in R. For further information on effective preprocessing and feature engineering checkout the 'Kernels' section of the Kaggle challenge website [https://www.kaggle.com/c/porto-seguro-safe-driver-prediction/kernels]. For this version we removed all 'calc' variables, as the Kaggle forum indicates that they do not carry much information.",
+  original.data.url = "https://www.kaggle.com/c/porto-seguro-safe-driver-prediction/data",
+  default.target.attribute = "target",
+  row.id.attribute = "id", 
+  visibility = "Everyone"
+)
+
+# create OML dataset
+oml_dat = makeOMLDataSet(
+  desc = desc,
+  data = dat,
+  colnames.old = colnames(dat),
+  colnames.new = colnames(dat),
[…]

### 2021-04-01 analysis/upload_datasets/road-safety-drivers-sex.R

@@ -1,0 +1,62 @@
+library("OpenML")
+library(data.table)
+
+# load OML config file to access server
+loadOMLConfig(path = "~/.openml/config", assign = TRUE)
+
+# load dataset
+dat = fread(file = "data/road-safety-drivers-sex/MakeModel2016.csv")
+
+# basic preprocessing
+dat[dat == -1] = NA
+
+dat = dat[, .(Was_Vehicle_Left_Hand_Drive, `Engine_Capacity_(CC)`, Propulsion_Code, 
+  Age_of_Vehicle, make, model, Sex_of_Driver)]
+
+colnames(dat)[colnames(dat) == "Engine_Capacity_(CC)"] = "Engine_Capacity"
+
+nominal_vars = c("Was_Vehicle_Left_Hand_Drive", "Propulsion_Code", "make", 
+  "model", "Sex_of_Driver")
+
+dat[, c(nominal_vars) := 
+  lapply(.SD, stringr::str_replace_all, pattern = "'", replacement = ""), 
+  .SDcols = c(nominal_vars)]
+
+# remove rows with NAs in Sex_of_Driver
+dat[Sex_of_Driver == 3, Sex_of_Driver := NA]
+dat = dat[!is.na(Sex_of_Driver)]
+dat[, Sex_of_Driver := factor(Sex_of_Driver, labels = c("male", "female"))]
+
+# recode variables as factors
+dat[, c(nominal_vars) := lapply(.SD, factor), 
+  .SDcols = nominal_vars]
+
+# check dataset for inconsistencies
+str(dat)
+
+# OML description
+desc = makeOMLDataSetDescription(
+  name = "road-safety-drivers-sex",
+  description = "Road Safety - Vehicles by Make and Model 2016. Predict the sex of drivers involved in personal injury road accidents in Great Britain in 2016, based on characteristics of their vehicles. The data was downloaded on 27.01.2019 from [http://data.dft.gov.uk/road-accidents-safety-data/MakeModel2016.zip]. The original data was published under the Open Government Licence (OGL) [http://www.nationalarchives.gov.uk/doc/open-government-licence/version/3/]. The variable 'Sex_of_Driver' was selected as target variable. For a description of all variables, checkout the original dataset description at [https://data.gov.uk/dataset/cb7ae6f0-4be6-4935-9277-47e5ce24a11f/road-safety-data].",
+  licence = "Open Government Licence (OGL)",
[…]

### 2021-04-01 analysis/upload_datasets/seattlecrime6.R

@@ -1,0 +1,121 @@
+# strange memory option for RWeka
+options(java.parameters = "-Xmx8000m")
+
+library(mlr)
+library(mlrCPO)
+library(OpenML)
+
+# source("analysis/high_cardinality_benchmark/mlrCPOs.R")
+
+# This should work for the final datasets!
+
+# configure OpenML
+# always use RWeka as farff does not work with too many factor levels!!
+#saveOMLConfig(cachedir = "~/.openml/cache", arff.reader = "RWeka")
+
+oml_id = 41960
+highcard_vars = c("Primary_Offense_Description")
+
+oml_dat = getOMLDataSet(data.id = oml_id)
+# prepare data
+dat = oml_dat$data
+# date columns are non meaningful: perform extraction
+dat$Occurred_Time = as.character(dat$Occurred_Time)
+dat$Occurred_hour = 0
+dat$Occurred_hour = ifelse(nchar(dat$Occurred_Time) == 4, 
+                           substr(dat$Occurred_Time, 1, 2), dat$Occurred_hour)
+dat$Occurred_hour = ifelse(nchar(dat$Occurred_Time) == 3, 
+                           substr(dat$Occurred_Time, 2, 2), dat$Occurred_hour)
+dat$Occurred_min = 0
+dat$Occurred_min = ifelse(nchar(dat$Occurred_Time) == 4, 
+                          substr(dat$Occurred_Time, 3, 4), dat$Occurred_min)
+dat$Occurred_min = ifelse(nchar(dat$Occurred_Time) == 3, 
+                          substr(dat$Occurred_Time, 2, 3), dat$Occurred_min)
+dat$Occurred_min = ifelse(nchar(dat$Occurred_Time) == 2, 
+                          substr(dat$Occurred_Time, 1, 2), dat$Occurred_min)
+dat$Occurred_min = ifelse(nchar(dat$Occurred_Time) == 1, 
+                          substr(dat$Occurred_Time, 1, 1), dat$Occurred_hour)
+
+dat$Reported_Time = as.character(dat$Reported_Time)
+dat$Reported_hour = 0
+dat$Reported_hour = ifelse(nchar(dat$Reported_Time) == 4, 
+                           substr(dat$Reported_Time, 1, 2), dat$Reported_hour)
+dat$Reported_hour = ifelse(nchar(dat$Reported_Time) == 3, 
[…]

### 2021-04-01 analysis/upload_datasets/sf-police-incidents.R

@@ -1,0 +1,138 @@
+library("OpenML")
+library(data.table)
+
+# load OML config file to access server
+loadOMLConfig(path = "~/.openml/config", assign = TRUE)
+
+# load dataset
+dat = fread(file = "data/police_incidents/Police_Department_Incident_Reports__Historical_2003_to_May_2018.csv")
+
+# preprocessing
+
+dat[, ViolentCrime := Category %in% c("ASSAULT", "ROBBERY", "SEX OFFENSES, FORCIBLE", "KIDNAPPING") |
+  Descript %in% c("GRAND THEFT PURSESNATCH", "ATTEMPTED GRAND THEFT PURSESNATCH")]
+
+dat[, ViolentCrime := factor(ViolentCrime, levels = c("TRUE", "FALSE"), labels = c("Yes", "No"))]
+
+dat[, Hour := hour(as.ITime(Time, format = "%H:%M"))]
+  
+dat[, Date := as.IDate(Date, format = "%m/%d/%Y")]
+
+dat[, DayOfWeek := wday(Date)]
+dat[, Month := month(Date)]
+dat[, Year := year(Date)]
+
+# dat[, Address := stri_replace_all(dat$Address, replacement = "_", regex = "/")]
+# dat[, Address := abbreviate(Address, minlength = 15)]
+
+# dat[, Address := match(Address, unique(na.omit(Address)))]
+
+nominal_vars = c("DayOfWeek", "Month", "Year", "PdDistrict", "Address")
+dat[, c(nominal_vars) := lapply(.SD, factor), .SDcols = nominal_vars]
+
+# dat = dat[sample(nrow(dat), size = 50000, replace = FALSE), c("IncidntNum", nominal_vars, "X", "Y", "ViolentCrime"), with = FALSE]
+dat = dat[, c("IncidntNum", "Hour", nominal_vars, "X", "Y", "ViolentCrime"), with = FALSE]
+
+dat = dat[PdDistrict != "",]
+
+# check dataset for inconsistencies
+summary(dat)
+
+# OML description
+desc = makeOMLDataSetDescription(
+  name = "sf-police-incidents",
[…]

### 2021-04-01 analysis/upload_datasets/upload_dataset.R

@@ -1,0 +1,38 @@
+library("OpenML")
+library(data.table)
+
+# load OML config file to access server
+loadOMLConfig(path = "~/.openml/config", assign = TRUE)
+
+# load dataset
+dat = fread(file = "")
+
+# basic preprocessing
+
+# check dataset for inconsistencies
+summary(dat)
+
+# OML description
+desc = makeOMLDataSetDescription(
+  name = "",
+  description = "",
+  creator = "",
+  licence = "",
+  url = "",
+  default.target.attribute = "",
+  citation = ""
+)
+
+# create OML dataset
+oml_dat = makeOMLDataSet(
+  desc = desc,
+  data = dat,
+  colnames.old = colnames(dat),
+  colnames.new = colnames(dat),
+  target.features = "")
+
+# upload dataset to the OML server
+# !DO NOT ACCIDENTALLY RUN THIS!
+
+# uploadOMLDataSet(oml_dat)
+

### 2021-04-01 analysis/upload_datasets/video-game-sales.R

@@ -1,0 +1,48 @@
+library("OpenML")
+library(data.table)
+
+# load OML config file to access server
+loadOMLConfig(path = "~/.openml/config", assign = TRUE)
+
+# load dataset
+dat = fread(file = "data/videogames_sales/vgsales.csv")
+
+# basic preprocessing
+
+dat[dat == "N/A"] = NA
+
+dat[, Year := as.numeric(Year)]
+
+nominal_vars = c("Platform", "Genre", "Publisher")
+dat[, c(nominal_vars) := lapply(.SD, factor), .SDcols = nominal_vars]
+
+# check dataset for inconsistencies
+summary(dat)
+
+# OML description
+desc = makeOMLDataSetDescription(
+  name = "video-game-sales",
+  description = "Kaggle dataset containing a list of video games with sales greater than 100,000 copies [https://www.kaggle.com/gregorut/videogamesales#vgsales.csv], which was generated by a scrape of vgchartz.com [https://github.com/GregorUT/vgchartzScrape]. For a description of the dataset, checkout the home of the kaggle dataset repo [https://www.kaggle.com/gregorut/videogamesales/home]. In contrast to the intention of the original author, the categorical variable 'Genre' was selected as the target here. Note that the variable 'Name' (ignored by default) is not unique for each row, as many games are published on more than one 'Platform'. The original dataset was sorted deacreasingly by the variable 'Global_Sales' (which was not changed here), which is why the ID variable 'Rank' directly reflects the rank order of 'Global_Sales'.",
+  original.data.url = "https://www.kaggle.com/gregorut/videogamesales#vgsales.csv",
+  default.target.attribute = "Genre",
+  row.id.attribute = "Rank", 
+  ignore.attribute = "Name",
+  visibility = "Everyone"
+)
+
+# create OML dataset
+oml_dat = makeOMLDataSet(
+  desc = desc,
+  data = dat,
+  colnames.old = colnames(dat),
+  colnames.new = colnames(dat),
+  target.features = "Genre")
+
+# upload dataset to the OML server
+# !DO NOT ACCIDENTALLY RUN THIS!
+
+# uploadOMLDataSet(oml_dat)
+
+# Uploading data set to server.
+# Uploading to 'http://www.openml.org/api/v1/data'.
+# Data set successfully uploaded. Data set ID: 41216

### 2021-04-01 analysis/upload_datasets/wine-reviews.R

@@ -1,0 +1,80 @@
+library(OpenML)
+library(data.table)
+
+# load OML config file to access server
+loadOMLConfig(path = "~/.openml/config", assign = TRUE)
+
+# load dataset
+dat = fread(file = "data/wine_reviews/winemag-data-130k-v2.csv")
+
+# basic preprocessing
+
+dat[, V1 := NULL]
+
+dat[dat == ""] = NA
+
+nominal_vars = c("country", "province", "region_1", 
+  "taster_name", "variety", "winery")
+
+text_vars = c("designation", "region_2", "description", "taster_twitter_handle", "title")
+
+dat[, c(text_vars) := lapply(.SD, stringi::stri_trans_general, id = "latin-ascii"), 
+   .SDcols = text_vars]
+
+dat[, c(text_vars) := 
+  lapply(.SD, stringr::str_replace_all, pattern = "[^[:alnum:]^[:blank:]^[:punct:]]", replacement = ""), 
+  .SDcols = text_vars]
+
+# "'" are not allowed in nominal factors by OpenML
+dat[, c(text_vars, nominal_vars) := 
+  lapply(.SD, stringr::str_replace_all, pattern = "'", replacement = " "), 
+  .SDcols = c(text_vars, nominal_vars)]
+
+dat[, c(text_vars) := 
+  lapply(.SD, stringr::str_squish), 
+  .SDcols = text_vars]
+
+# dat[, c(nominal_vars) := lapply(.SD, function(x) factor(match(x, unique(na.omit(x))))), 
+#   .SDcols = nominal_vars]
+
+dat[, c(nominal_vars) := lapply(.SD, factor), 
+  .SDcols = nominal_vars]
+
+
+# check dataset for inconsistencies
+summary(dat)
+
+# OML description
+desc = makeOMLDataSetDescription(
+  name = "wine-reviews",
[…]

### 2021-04-01 publication/manuscript.Rmd
more typos

@@
 Meta rankings seem to be highly consistent with the individual patterns of encoder performances reflected in Figures \@ref(fig:regr-res) to \@ref(fig:multcl-res).
-Looking at meta-rankings, approaches based on GLMM's in combinaion with cross-validation outperform all other approaches across all algorithms.
-A further interesting detail ommited in Figure \ref{fig:rank-res} for clarity is that the \textit{none} encoding strategy for the random forest was beaten by all strategies except for the \textit{remove} condition.
+Looking at meta-rankings, approaches based on GLMM's in combination with cross-validation outperform all other approaches across all algorithms.
+A further interesting detail omited in Figure \ref{fig:rank-res} for clarity is that the \textit{none} encoding strategy for the random forest was beaten by all strategies except for the \textit{remove} condition.
 This implies, that even if the algorithm includes a mechanic for treating categorical variables, it might often be optimal to use a different strategy instead.
@@
 Our results suggest that one-hot encoding is the better standard compared to dummy encoding when applying nonlinear regularized models like RF, GB or SVM with (high cardinal) categorical features.
-We furthermore provide a more in-detail anaysis of the performances of one-hot and dummy encoding in the supplementary material.
+We furthermore provide a more in-detail analysis of the performances of one-hot and dummy encoding in the supplementary material.
 

### 2021-04-01 publication/manuscript.Rmd
iterate on supplement and manuscript

@@
 Otherwise, the best encoding might just provide the least impairment compared to not including any high cardinality features.
-We include an overview over available implementations in widely used machine learning frameworks for R and python in the supplementary material.
+We include an overview over available implementations in widely used machine learning frameworks for R and python in the Supplementary Material.
 
@@
 Throughout our analysis, we use implementations of the various methods for the open-source statistical software R [@r_2018].
+To enable a fair and reliable comparison in our study, we implemented all encoding methods on top of the `mlrCPO` package [@binder_2018].
 The pre-processing, as well as final ML algorithm described in \ref{sec:pipeline}, was trained and resampled using the `mlr` framework [@bischl_2016] together with the `batchtools` package [@lang_2017] to scale the benchmark analysis to HPC compute infrastructure.
-All materials for this study, including reproducible code for this manuscript and the necessary result objects can be downloaded from our repository (see above).
+All materials for this study, including reproducible code for this manuscript and the necessary result objects can be downloaded from our **online repository** \footnote{\url{https://github.com/compstat-lmu/paper_2021_categorical_feature_encodings}}.
 
@@
 Our results suggest that one-hot encoding is the better standard compared to dummy encoding when applying nonlinear regularized models like RF, GB or SVM with (high cardinal) categorical features.
-We furthermore provide a more in-detail analysis of the performances of one-hot and dummy encoding in the supplementary material.
+We furthermore provide a more in-detail analysis of the performances of one-hot and dummy encoding in the Supplementary Material.
 
@@
 
-Declarations of interest: none
+Declarations of interest {-}
+==========
+none
 

### 2021-04-01 publication/supplementary_material.Rmd
iterate on supplement and manuscript

@@
 csl: elsevier-harvard.csl
+bibliography: literature.bib
 output: pdf_document
@@
 To enable a fair and reliable comparison in our study, we implemented all methods outlined above on top of the `mlrCPO` package.
-All code can be found in our **online repository** \footnote{\url{https://github.com/compstat-lmu/paper_2021_categorical_feature_encodings}}.
+All code can be found in our **online repository** (<https://github.com/compstat-lmu/paper_2021_categorical_feature_encodings>).
 
-\begin{table}
+\begin{table}[H]
 \centering
@@
 \caption{
-  Overview over existing encoding implementations in R. Implementations can deviate from the algorithms described in section \ref{sec:methods} by minor implementation details.
+  Overview over existing encoding implementations in R. Implementations can deviate from the algorithms described in our manuscript by minor implementation details.
   Indicator encoding encloses a variety of methods (one-hot, dummy, helmert encoding etc.)}
@@
 
+\newpage
+\clearpage
+
 # Comparison of One-Hot and Dummy Encoding
@@
 A frequently asked question is, whether one-hot encoding or dummy encoding is to be preferred.
-While dummy encoding clearly is to be preferred for non-regularized linear models in order to avoid multi-colinearity, the answer is less clear for general machine learning algorithms studied in our setting.
+While dummy encoding clearly is commonly preferred for non-regularized linear models, the answer is less clear for general machine learning algorithms studied in our setting.
 
@@
 
-```{r oneh-win, echo = FALSE, results="asis"}
+```{r oneh-win, echo = FALSE, results="asis", fig.pos="!H"}
 oneh_win = indicator[Classes == 2 & !is.na(auc.test.mean_FALSE > auc.test.mean_TRUE),
@@
 
-From the results shown in\ref{fig:oneh-win} we can observe, that for most algorithms and datasets, one-hot encoding is to be preferred over dummy-encoding.
+From the results shown in Table 2 we can observe, that for most algorithms and datasets, one-hot encoding is to be preferred over dummy-encoding.
 

### 2021-04-01 publication/manuscript.Rmd
fix typos

@@
 using target predictions based on the feature levels in the training set as a new numerical feature) consistently provided the best results.
-Traditional encodings often make unreasonable assumptions to map levels to integers (e.g.
-integer encoding) or reduce the number of levels (possibly based on target information, e.g.
+Traditional encodings that make unreasonable assumptions to map levels to integers (e.g.
+integer encoding) or to reduce the number of levels (possibly based on target information, e.g.
 leaf encoding) before creating binary indicator variables (one-hot or dummy encoding) were not as effective.
@@
 Although domain knowledge can sometimes be used to reduce the number of theoretically relevant levels, finding strategies that work well on a large variety of problems is highly important for many applications as well as in automated ML [@thomas_2018; @thornton_2013; @feurer_2015].
-Optimally, strategies should be model-agnostic, as benchmarking encoding methods together with ML algorithms from different classes is often necessary for applications.
-While a variety of strategies exists, there are very few benchmarks that can be used to decide which technique is expected to yield good predictive performance.
+Optimally, strategies should be model-agnostic because benchmarking encoding methods together with ML algorithms from different classes is often necessary for applications.
+While a variety of strategies exist, there are very few benchmarks that can be used to decide which technique is expected to yield good predictive performance.
 Furthermore, there has recently been increasing attention on scientific benchmark studies that compare different methods to provide a clearer picture in light of a large number of methods available to practitioners [@delgado; @bommert], as they can provide at least partial answers to such questions.
-The goal of this study is to provide an overview of existing approaches for encoding categorical variables and to study their effect on a model's predictive performance.
+The goal of this study is to provide an overview of existing approaches for encoding categorical predictor variables and to study their effect on a model's predictive performance.
 
@@
 
-We consider the classical setting of supervised learning from an $i.i.d$ tabular dataset $\mathcal{D}$ of size $N$ sampled from a joint distribution $\mathcal{P}(\bm{x}, y)$ of a set of features $\bm{x}$ and an associated target variable $y$.
+We consider the classical setting of supervised learning from an $i.i.d.$ tabular dataset $\mathcal{D}$ of size $N$ sampled from a joint distribution $\mathcal{P}(\bm{x}, y)$ of a set of features $\bm{x}$ and an associated target variable $y$.
 Here, $\bm{x}$ consists of a mix of numeric (real-valued or integer-valued) features and categorical features, the latter of which we seek to transform feature-wise to numeric features using a categorical encoding technique.
@@
 It holds either $y \in \mathbb{R}$ (regression), $y \in \mathcal{C}$ from a finite class space $\mathcal{C} = \{c_1, ..., c_C\}$ with $C = 2$ (binary classification) or $C > 2$ (multiclass classification).
-We always assume to observe all $C$ classes in our training sample, however we might only observe a subset $\mathcal{L}^{train} \subseteq \mathcal{X}$ of a features' available $L$ levels, $\mathcal{L}^{train} = \{l_1, ..., l_L\}$ for categorical features.
+We always assume to observe all $C$ classes in our training sample, however we might only observe a subset $\mathcal{L}^{train} \subseteq \mathcal{X}$ of a feature's available $L$ levels, $\mathcal{L}^{train} = \{l_1, ..., l_L\}$ for categorical features.
 We denote the observed frequency of class $c$ in the training set with $N_c$ and the observed frequency of a level $l$ in the training set with $N_l$.
 As most ML algorithms require numerical features, we use categorical encoding techniques to transform each nominal feature $x^{train}$ into numerical features  $\hat{x}^{train}$ which are then used for training.
-If clear from the context, we use $\hat{x}_l$ as the encoded value for observation with level $l$.
-Although datasets might contain multiple high cardinal features, we encode each feature separately but using the same strategy.
+If clear from the context, we use $\hat{x}_l$ as the encoded value for an observation with level $l$.
+Although datasets might contain multiple high cardinality features, we encode each feature separately but with the same strategy.
 
@@
 
-We broadly categorize feature encoding techniques into \textit{target-agnostic} methods and methods that incorporate target statistics [@micci_barreca_2001].
+We broadly categorize feature encoding techniques into *target-agnostic* methods and *target-based* methods [@micci_barreca_2001].
 Figure \ref{fig:overview} contains a taxonomy of the encoding methods we consider in this work.
-\textit{Target-agnostic} methods do not rely on any information about the target variable and can therefore also be used in unsupervised settings.
+*Target-agnostic* methods do not rely on any information about the target variable and can therefore also be used in unsupervised settings.
 Simple strategies from this domain e.g.
-one-hot or dummy encoding are widely used to embed variables for classical ML algorithms as well as (deep) neural networks in the scientific literature [@kuhn_2019; @hancock] but also on  Kaggle\footnote{\url{https://www.kaggle.com/}}.
+one-hot or dummy encoding are widely used -- in the scientific literature [@kuhn_2019; @hancock] but also on  Kaggle\footnote{\url{https://www.kaggle.com}} -- to embed variables for classical ML algorithms as well as (deep) neural networks.
 Such indicator methods map each level of a categorical variable to a set of dichotomous features encoding the presence or absence of a particular level.
@@
 
-Target-based methods try to incorporate information about the target values associated with a given level.
-Early strategies aimed to reduce the number of levels by methods like hierarchical clustering or decision trees based on statistics of the target variable, although this has been rarely described in the scientific literature [for a short mention, see @micci_barreca_2001].
-The basic idea of more advanced methods called target, impact, mean, or likelihood encoding is to use the training set to make a simple prediction of the target for each level of the categorical feature, and to use the prediction as to the numerical feature value $\hat{x}_l$ for the respective level.
+*Target-based* methods try to incorporate information about the target values associated with a given level.
+Early strategies aimed to reduce the number of levels by methods like hierarchical clustering or decision trees based on statistics of the target variable, although this has been rarely described in the scientific literature [for a brief mention, see @micci_barreca_2001].
+The basic idea of more advanced methods called target, impact, mean, or likelihood encoding is to use the training set to make a simple prediction of the target for each level of the categorical feature, and to use the prediction as the numerical feature value $\hat{x}_l$ for the respective level.
 One of the earliest formal descriptions of this strategy is @micci_barreca_2001.
-A very similar approach is called *weights of evidence* in the credit scoring literature, see @hand_1997.
-In basic target encoding for regression problems, the mean target value in the training set from all observations with a certain feature level is used as the numeric value to encode that level for all observations: $\hat{x}_l = \frac{\sum_{i:x^{train}_i = l}y^{train}_i}{N_l}$.
-Simple target encoding often does not perform well with rare levels, where it tends to overfit to the training data and fail to generalize well for new observations.
+A very similar approach is called *weights of evidence* in the credit scoring literature @hand_1997.
+In simple target encoding for regression problems, the mean target value in the training set from all observations with a certain feature level is used as the numeric value to encode that level for all observations: $\hat{x}_l = \frac{\sum_{i:x^{train}_i = l}y^{train}_i}{N_l}$.
+Simple target encoding often does not perform well with rare levels, where it tends to overfit to the training data and fails to generalize well for new observations.
 In the extreme case of a categorical feature with unique values (e.g.
@@
 Different versions of target encoding are compared with indicator, integer, and frequency encoding.
-They recommend combining the smoothed version of target encoding with 4 or 5-fold CV, never using simple target encoding and use indicator encoding only for small datasets.
+They recommend combining the smoothed version of target encoding with 4- or 5-fold CV, never using simple target encoding and use indicator encoding only for small datasets.
 Interestingly, frequency encoding did perform well in many cases.
@@
 Another line of work studies variable encodings employed within specific ML models.
-@wright_2019 study treatments for categorical variables in random forests together with dummy and integer encoding (18 datasets, cardinality between $3$ and $38$), concluding that indicator and integer encoding perform subpar in comparison to methods that re-order levels according to the target variables.
+@wright_2019 study treatments for categorical variables in random forests together with dummy and integer encoding (18 datasets, cardinality between $3$ and $38$), concluding that indicator and integer encoding perform subpar in comparison to methods that re-order levels according to the target variable.
 @prokhorenkova_2018 compared their new CatBoost variant of target encoding to smoothed target encoding with no CV, hold-out, and leave-one-out CV on $8$ datasets.
 While the CatBoost method performed best, hold-out came second and target encoding without CV performed worst.
-An overview of encoding techniques tailored to be used together with Neural Networks is provided in [@hancock].
+An overview of encoding techniques tailored to be used together with Neural Networks is provided in @hancock.
 Amongst others, they survey indicator-based methods as well as embedding-based methods but do not conduct a benchmark study.
-One widely adopted approach, in particular, has been proposed by [@guo_2016] under the term \textit{entity embeddings}.
+One widely adopted approach, in particular, has been proposed by @guo_2016 under the term *entity embeddings*.
 In contrast, our work focuses on high cardinality variables and techniques that are agnostic to the subsequent ML method, which has not been done before.
@@
 
-The main goal of our study is to assess the impact different categorical encoding techniques have on subsequent models predictive performance.
-As the optimal encoding might differ depending on the ML algorithm, we consider various state-of-the-art algorithms, \textit{regularized linear models}, \textit{random forests}, \textit{gradient tree boosting}, \textit{k-nearest neighbors}, and \textit{support vector machines}.
-To find default settings for high cardinality features, we analyse a variety of datasets with different characteristics including regression, binary classification and multiclass classification problems.
+The main goal of our study is to assess the impact different categorical encoding techniques have on subsequent models' predictive performance.
+As the optimal encoding might differ depending on the ML algorithm, we consider various state-of-the-art algorithms, *regularized linear models*, *random forests*, *gradient tree boosting*, *k-nearest neighbors*, and *support vector machines*.
+To find default settings for high cardinality features, we analyze a variety of datasets with different characteristics including regression, binary classification and multiclass classification problems.
 As our methods vary in runtime and complexity, we aim to learn whether more complex methods, in general, are to be preferred, or if simpler approaches suffice.
@@
 
-We use indicator encoding as an umbrella term for the two most common strategies to encode categorical features with a small to moderate the number of levels: one-hot and dummy encoding.
+We use indicator encoding as an umbrella term for the two most common strategies to encode categorical features with a small to moderate number of levels: one-hot and dummy encoding.
 One-hot encoding transforms the original feature into $L$ binary indicator columns, each representing one original level.
@@
 In datasets with a high total number of feature levels, constructing all indicator variables can be detrimental to the predictive performance or at least increase the computational load.
-As indicator encoding is practically infeasible for high cardinality variables (a feature with $10^5$ levels would add $10^5$ new columns), we limit the number of indicator coded features by collapsing rare levels beyond a varying threshold to a single \textit{other} category before encoding.
+As indicator encoding is practically infeasible for high cardinality variables (a feature with $10^5$ levels would add $10^5$ new columns), we limit the number of indicator coded features by collapsing rare levels beyond a varying threshold to a single *other* category before encoding.
 Closely related to dummy encoding in classical analysis of variance and regression are Helmert and Polynomial contrasts [@chambers_1992].
@@
 Hash Encoding can be used to compute indicator variables based on a hash function [@weinberger_2009].
-The basic idea is to transform each feature level $l$ to an integer $hash(l) \in \mathbb{N}$, based on its label.
-This integer is then transformed to an indicator representation, with $1$ in indicator column number $(hash(l) \mod hash.size) + 1$ and $0$ in all remaining columns [@kuhn_2019].
+The basic idea is to transform each feature level $l$ into an integer $hash(l) \in \mathbb{N}$, based on its label.
+This integer is then transformed into an indicator representation, with $1$ in indicator column number $(hash(l) \mod hash.size) + 1$ and $0$ in all remaining columns [@kuhn_2019].
 Some levels will be hashed to the same indicator representation.
@@
 Thus, our leaf encoder automatically uses an "optimal" number of new levels.
-To speed up the computation for multiclass classification, our implementation uses the ordering approach presented in [@wright_2019].
+To speed up the computation for multiclass classification, our implementation uses the ordering approach presented in @wright_2019.
 New levels are encoded with the arbitrary number of the terminal node with most observations during training.
 Encoded values are treated as a new categorical feature and encoded by one-hot encoding.
-Our leaf encoder can be thought of as a simplification of the approach suggested by [@grabczewski_2003].
+Our leaf encoder can be thought of as a simplification of the approach suggested by @grabczewski_2003.
 
@@
 Similar techniques have been proposed in the literature:
-Weights of evidence encoding from the credit scoring classification literature [@hand_1997] are almost identical to impact encoding, but does not
-consider any regularization or centring.
+Weights of evidence encoding from the credit scoring classification literature [@hand_1997] is almost identical to impact encoding, but does not
+consider any regularization or centering.
 Impact encoding is also referred to as a target or James-Stein encoding.
@@
 Smoothed target encoding [@micci_barreca_2001] can be interpreted as a simple (generalized) linear mixed model (glmm) in which the target is predicted by a random intercept for each feature level in addition to a fixed global intercept.
-This connection is described in [@kuhn_2019].
+This connection is described in @kuhn_2019.
 To achieve regularized impact encoding, we constructed our own glmm encoders for regression, binary, and multiclass classification which are described in Algorithm \ref{alg:glmm-regr} (regression) and in the Supplementary Material (classification).
@@
 Otherwise, the best encoding might just provide the least impairment compared to not including any high cardinality features.
-We include an overview over available implementations in widely used machine learning frameworks for R and python in the Supplementary Material.
+We include an overview of available implementations in widely used machine learning frameworks for R and python in the Supplementary Material.
 
@@
 In total, we investigate $5$ ML methods:
-Regularized linear models (LASSO) were fitted with glmnet [@friedman_2010], internally tuning the regularization using 5-fold CV.
-Random forests (RF) were trained using ranger [@wright_2017].
+Regularized linear models (LASSO) were fitted with `glmnet` [@friedman_2010], internally tuning the regularization using 5-fold CV.
+Random forests (RF) were trained using `ranger` [@wright_2017].
 Because the random forest algorithm can be expected to give reasonable results without tuning [@probst_2019], we only fixed the number of fitted trees to $500$.
-Gradient boosting models (GB) were trained using xgboost [@chen_2018].
+Gradient boosting models (GB) were trained using `xgboost` [@chen_2018].
 We set the learning rate to $0.01$ and determine the number of iterations using early stopping on a 20% holdout set.
-K-nearest neighbours (KNN) was taken from package knn [@schliep_2016] standardizing all features and using a constant $k=15$ for the number of nearest neighbours, together with an information gain filter [@brown2012] to limit the number of features to $25$.
-Support vector machines (SVM) with radial basis function kernel were trained with liquidSVM [@steinwart_2017].
+K-nearest neighbours (KNN) was taken from package `kknn` [@schliep_2016] standardizing all features and using a constant $k=15$ for the number of nearest neighbours, together with an information gain filter [@brown2012] to limit the number of features to $25$.
+Support vector machines (SVM) with radial basis function kernel were trained with `liquidSVM` [@steinwart_2017].
 The bandwidth and regularization parameters were internally tuned using 5-fold CV, and we used a one-vs-all approach for multiclass-settings.
@@
 To enable a fair and reliable comparison in our study, we implemented all encoding methods on top of the `mlrCPO` package [@binder_2018].
-The pre-processing, as well as final ML algorithm described in \ref{sec:pipeline}, was trained and resampled using the `mlr` framework [@bischl_2016] together with the `batchtools` package [@lang_2017] to scale the benchmark analysis to HPC compute infrastructure.
+The pre-processing, as well as the final ML algorithm described in \ref{sec:pipeline}, were trained and resampled using the `mlr` framework [@bischl_2016] together with the `batchtools` package [@lang_2017] to scale the benchmark analysis to HPC compute infrastructure.
 All materials for this study, including reproducible code for this manuscript and the necessary result objects can be downloaded from our **online repository** \footnote{\url{https://github.com/compstat-lmu/paper_2021_categorical_feature_encodings}}.
@@
 
-For  `r length(svm_errors[, unique(Name)])` datasets, some conditions with the SVM lead to unexpected crashes due to memory problems or numerical errors.
+For  `r length(svm_errors[, unique(Name)])` datasets, some conditions with the SVM led to unexpected crashes due to memory problems or numerical errors.
 We completely removed those datasets for the SVM when computing ranks or other statistics to compare encodings across datasets.
@@
 
-On datasets where substantive performance differences could be observed, target encoding with the glmm encoder was generally *most* effective.
-The *worst* encoder depended on the ML algorithm and respective dataset.
+On datasets where substantive performance differences could be observed, target encoding with the glmm encoder was generally **most** effective.
+The **worst** encoder depended on the ML algorithm and respective dataset.
 For datasets *Click_prediction_small*, *KDDCup09_upselling*, *kick*, and *okcupid-stem* some combinations of encoder $\times$ ML algorithm performed worse than simply removing high cardinality features.
@@
 (ref:rank-res) Consensus rankings across all datasets for each algorithm.
+Lower ranks indicate better performance.
+The rank of the *none* control condition of the random forest (rank 11) was omitted from the figure.
 
@@
 (ref:dendro) Hierarchical cluster analysis of benchmark datasets.
-The symmetric difference distance between two datasets reflect differences in performance patterns between encodings (independent of the employed ML algorithm).
+The symmetric difference distance between two datasets reflects differences in performance patterns between encodings (independent of the employed ML algorithm).
 
@@
 For LASSO, previous studies have suggested that indicator encoding works well, even with a very high number of levels [@cerda_2018].
-Although the glmm encoders ranked first in our benchmark for the LASSO, it was the only algorithm where indicator encoding ranked second.
+Although the glmm encoders ranked first in our benchmark for the LASSO, it was the only algorithm where the indicator encoders achieved the next best ranking.
 Note that for computational reasons we limited the maximum amount of indicator variables to 125 in our experimental design.
@@
 Our results suggest that one-hot encoding is the better standard compared to dummy encoding when applying nonlinear regularized models like RF, GB or SVM with (high cardinal) categorical features.
-We furthermore provide a more in-detail analysis of the performances of one-hot and dummy encoding in the Supplementary Material.
+We provide a more in-detail analysis of the performances of one-hot and dummy encoding in the Supplementary Material.
 
@@
 predictive performance trade-off might also be in favour of target encoding, especially for large datasets where a high number of indicator variables increases computational load.
-We did observe a substantial increase in runtimes when using regularized glmm with GB, where runtimes are comparably short in comparison to the time required to fit categorical encoders.
-In other settings (LASSO, RF), the glmm encoder can also be faster, as it allows for a more efficient data representation for the ML algorithm.
+We did observe a substantial increase in runtimes when using regularized glmm with GB (with little tuning), where runtimes are relatively short in comparison to the time required to fit categorical encoders.
+In other settings (LASSO, RF), the glmm encoders have been observed to be even faster than indicator encoding.
 
@@
 An interesting question was whether target encoding is only useful for features with a very high number of levels or if it might also have advantages for fewer levels, where most practitioners would routinely use indicator encoding.
-We tested this using HCT thresholds of $10$, $25$ and $125$.
+We tested this using HCT thresholds of $10$, $25$ and $125$, but the results are not easy to interpret.
 In general, the optimal threshold seemed to strongly depend on the dataset, but we also observed some weak patterns:
@@
 Although the performance of other encoding strategies was comparable in some conditions, target encoding was never outperformed.
-In general, our results suggest that regularized target encoding works well for all kinds of algorithms and should be considered when working with high cardinality features.
+In general, our results suggest that regularized target encoding works well for all kinds of algorithms and could be considered by default when working with high cardinality features.
 In terms of runtime, using target encoding together with 5-fold CV sometimes leads to slightly longer runtimes (in comparison to indicator encoding), but especially for larger datasets this is often offset by the more efficient representation yielded from target encoding.
 
-What constitutes a "high" cardinality problem is a difficult question that might not only depend on the number of levels but also other characteristics of the dataset and its features.
+What constitutes a "high" cardinality problem is a difficult question that might not only depend on the number of levels but also on other characteristics of the dataset and its features.
 Unfortunately, the number of datasets in our study was too small to discover consistent patterns between encoder performance and dataset characteristics.
@@
 Regularized versions of target encoding, which uses predictions of the target variable as numeric feature values, performed better than traditional strategies like integer or indicator encoding.
-Most effective was a target encoder which combines simple generalized linear mixed models with cross-validation and does not require tuning hyperparameters.
+Most effective, with a consistent superior performance across Ml algorithms and datasets, was a target encoder which combines simple generalized linear mixed models with cross-validation and does not require tuning hyperparameters.
 

### 2021-04-28 publication/manuscript.Rmd
remove acknowledgements; add declarations

@@
 
-Acknowledgements {-}
-==========
+## Declarations
+
+### Funding
 This work has been funded by the German Federal Ministry of Education and Research (BMBF) under Grant No.
 01IS18036A and by the Bavarian Ministry of Economic Affairs, Regional Development and Energy through the Center for Analytics – Data – Applications (ADA-Center) within the framework of „BAYERN DIGITAL II“ (20-3410-2-9-8).
-The authors of this work take full responsibilities for its content.
-
-Declarations of interest {-}
-==========
-none
+
+### Conflicts of Interest
+None.
+
+### Availability of Data and Material
+\url{https://github.com/compstat-lmu/paper_2021_categorical_feature_encodings}
+
+### Code Availability
+\url{https://github.com/compstat-lmu/paper_2021_categorical_feature_encodings}
 

### 2021-04-28 publication/manuscript.Rmd
switch to springer template

@@
 title: Regularized target encoding outperforms traditional methods in supervised machine learning with high cardinality features
-author:
-  - name: Florian Pargent
+titlerunning: Encoding high cardinality features in supervised ML
+authors:
+  - name: Florian Pargent \orcidlink{0000-0002-2388-553X}
     email: florian.pargent@psy.lmu.de
-    affiliation: LMU Munich 1
+    address: Department of Psychology, Psychological Methods and Assessment, LMU Munich, Leopoldstraße 13, 80802 Munich, Germany
     footnote: 1
   - name: Florian Pfisterer
-    #email: florian.pfisterer@stat.uni-muenchen.de
-    affiliation: LMU Munich 2
-  - name: Janek Thomas
-    #email: janek.thomas@iis.fraunhofer.de
-    affiliation: Fraunhofer
-  - name: Bernd Bischl
-    #email: bernd.bischl@stat.uni-muenchen.de
-    affiliation: LMU Munich 2
-address:
-  - code: LMU Munich 1
-    address: Department of Psychology, Psychological Methods and Assessment, LMU Munich, Leopoldstraße 13, 80802 Munich, Germany
-  - code: LMU Munich 2
     address: Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstraße 33, 80539 Munich, Germany
-  - code: Fraunhofer
+  - name: Janek Thomas \orcidlink{0000-0003-4511-6245}
     address: Fraunhofer Institute for Integrated Circuits, Hansastraße 32, 8068 Munich, Germany
-footnote:
-  - code: 1
-    text: "Corresponding Author"
+  - name: Bernd Bischl \orcidlink{0000-0001-6002-6980}
+    address: Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstraße 33, 80539 Munich, Germany
+
 abstract: |
@@
 leaf encoding) before creating binary indicator variables (one-hot or dummy encoding) were not as effective.
-keywords: supervised machine learning, benchmark, high-cardinality categorical features, target encoding, impact encoding, dummy encoding
-
-journal: "Computational Statistics and Data Analysis"
-date: "`r Sys.Date()`"
+keywords:
+- supervised machine learning
+- benchmark
+- high-cardinality categorical features
+- target encoding
+- impact encoding
+- dummy encoding
+
+#date: "`r Sys.Date()`"
 bibliography: literature.bib
@@
 numbersections: true
-csl: elsevier-harvard.csl
 output:
@@
   bookdown::pdf_book:
-    base_format: rticles::elsevier_article
-#output: rticles::elsevier_article
+    base_format: rticles::springer_article
+bibstyle: spphys
 
@@
    - \usepackage{pdflscape}
+   - \usepackage[T1]{fontenc}
+   - \usepackage{orcidlink}
 ---

### 2021-04-28 publication/manuscript.Rmd
add orcid id

@@
     footnote: 1
-  - name: Florian Pfisterer
+  - name: Florian Pfisterer \orcidlink{0000-0001-8867-762X}
     address: Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstraße 33, 80539 Munich, Germany

### 2021-04-28 publication/manuscript.Rmd
update availability of data section
update code availability section

@@
 
-### Availability of Data and Material
-\url{https://github.com/compstat-lmu/paper_2021_categorical_feature_encodings}
-
 ### Code Availability
-\url{https://github.com/compstat-lmu/paper_2021_categorical_feature_encodings}
+All analysis code is available from \url{https://github.com/compstat-lmu/paper_2021_categorical_feature_encodings}
+
+### Availability of Data
+Our analysis code downloads all benchmark datasets from \url{https://www.openml.org/}.
 

### 2021-04-28 publication/manuscript.Rmd
switch citation style

@@
     base_format: rticles::springer_article
-bibstyle: spphys
+#bibstyle: spphys
+csl: elsevier-harvard.csl
 

### 2021-04-28 publication/manuscript.Rmd
add text

@@
 plotdt = plotdt[-which(plotdt$method == "none"),]
+
+swit = Vectorize(function(x) {
+  switch(x,
+  "frequency" = "frq",
+  "one-hot" = "oh",
+  "remove" = "rm",
+  "integer" = "int",
+  "dummy" = "dumm",
+  "leaf" = "leaf",
+  "impact" = "imp",
+  "glmm-10CV" = "CV10",
+  "glmm-5CV" = "CV5",
+  "glmm-noCV" = "noCV",
+  "hash" = "hash"
+)})
+plotdt$lab = unlist(swit(plotdt$method))
+
 p = ggplot(plotdt, aes(x = rank, y = algorithm, fill = method)) +
- geom_tile(size = .5, width =.93, height =.93, color = "black")
+ geom_tile(size = .5, width =.93, height =.93, color = "black") +
+ geom_text(aes(x=rank, y = algorithm, label=lab), size = 3)
+
 p = p +

### 2021-04-29 publication/manuscript.Rmd
manuscript

@@
 plotdt = plotdt[-which(plotdt$method == "none"),]
-
+plotdt$method = factor(plotdt$method,
+ levels = c("integer", "frequency", "dummy", "one-hot", "hash", "leaf", "impact",
+    "glmm-noCV", "glmm-5CV", "glmm-10CV", "remove")
+)
+plotdt[, lab := method]
 swit = Vectorize(function(x) {
@@
 )})
-plotdt$lab = unlist(swit(plotdt$method))
+levels(plotdt$lab) = unlist(swit(levels(plotdt$method)))
 
@@
  geom_tile(size = .5, width =.93, height =.93, color = "black") +
- geom_text(aes(x=rank, y = algorithm, label=lab), size = 3)
+ geom_text(aes(x=rank, y = algorithm, label=lab), size = 3) +
+ scale_fill_manual(values = box_colors)
 

### 2021-04-30 publication/manuscript.Rmd
fix figure 5

@@
 
-```{r, rank-res, out.height=".55\\textwidth", fig.cap='(ref:rank-res)', out.width="\\textwidth"}
+```{r, rank-res, out.height=".55\\textwidth", fig.cap='(ref:rank-res)', out.width="\\textwidth", message = FALSE}
 plotdt = rbindlist(lapply(names(cons_list), function(x) {

### 2021-04-30 publication/manuscript.Rmd
update keywords

@@
 - target encoding
-- impact encoding
 - dummy encoding
+- generalized linear mixed models
 

### 2021-04-30 publication/manuscript.Rmd
fix declarations and dois

@@
 ### Code Availability
-All analysis code is available from \url{https://github.com/compstat-lmu/paper_2021_categorical_feature_encodings}
+All analysis code is publicly available at \url{https://github.com/compstat-lmu/paper_2021_categorical_feature_encodings}.
 
 ### Availability of Data
-Our analysis code downloads all benchmark datasets from \url{https://www.openml.org/}.
+All benchmark datasets are publicly available at \url{https://www.openml.org/}.
 

### 2021-09-07 publication/manuscript.Rmd
first reviewer changes; add boulesteix neutral benchmarking references to bibfile

@@
 An overview of encoding techniques tailored to be used together with Neural Networks is provided in @hancock.
-Amongst others, they survey indicator-based methods as well as embedding-based methods but do not conduct a benchmark study.
+They present a survey of indicator-based and embedding-based methods but do not conduct a benchmark study.
 One widely adopted approach, in particular, has been proposed by @guo_2016 under the term *entity embeddings*.
@@
 
-Throughout our analysis, we use implementations of the various methods for the open-source statistical software R [@r_2018].
+Throughout our analysis, we use implementations of the various methods for the open-source statistical software R [@r_2021].
 To enable a fair and reliable comparison in our study, we implemented all encoding methods on top of the `mlrCPO` package [@binder_2018].
@@
 dat_max = dat_max[lrn.id != "FL"]
+
+## Reviewer Question: repeat consensus relation analyses for each task type?
+# dat_max = dat_max[Classes == 0]
+## unfortunately, consensus relations do not converge then...
 
@@
 Looking at meta-rankings, approaches based on GLMM's in combination with cross-validation outperform all other approaches across all algorithms.
-A further interesting detail omited in Figure \ref{fig:rank-res} for clarity is that the \textit{none} encoding strategy for the random forest was beaten by all strategies except for the \textit{remove} condition.
+A further interesting detail omitted in Figure \ref{fig:rank-res} for clarity is that the \textit{none} encoding strategy for the random forest was beaten by all strategies except for the \textit{remove} condition.
 This implies, that even if the algorithm includes a mechanic for treating categorical variables, it might often be optimal to use a different strategy instead.

### 2021-09-08 publication/manuscript.Rmd
add optimal scaling references to manuscript and bib

@@
 Levels with comparable main effects can therefore not be distinguished based on the transformed feature, which prevents the consecutive ML algorithm from learning interactions for specific levels.
-Methods that can jointly encode several variables and therefore leverage correlation structures between the different variables are an interesting avenue of research in the future.
-Another interesting avenue is using target-based encoding strategies to learn a vector-valued representation of a level, similar to entity embeddings [@guo_2016].
+Methods that can jointly encode several variables and therefore leverage correlation structures between the different variables are an interesting avenue of research in the future:
+Recent approaches use target-based encoding strategies to learn a vector-valued representation of a level, similar to entity embeddings [@guo_2016].
+Those should be compared to classical approaches from the optimal scaling literature [e.g.
+@deleeuw_1976] leading to the *aspect* framework [@mair_2010], which is unfortunately unfamiliar to many ML researchers.
 We furthermore only investigate traditional categorical variables and do not extend our analysis to multi-categorical variables or text-strings [@cerda_2018; @cerda_2020], where other encoding techniques can harvest additional information.

### 2021-09-08 publication/manuscript.Rmd
add footnote on convergence issues for problem type specific meta rankings

@@
 The last two datasets with the highest number of levels formed a separate cluster, in which target encoding without strong regularization (impact, glmm-noCV) showed severe overfitting.
-Note that clusters were not determined by problem type, again suggesting that encoder rankings are somewhat similar for regression and classification settings.
+Note that clusters were not determined by problem type, again suggesting that encoder rankings are somewhat similar for regression and classification settings.\footnote{We tried to corroborate this observation by comparing meta-rankings between problem types.
+Unfortunately, problem type specific consensus relations did not converge (probably due to the reduced number of datasets) and were uninterpretable.}
 

### 2021-09-08 publication/manuscript.Rmd
add limitation section on ordinal encoding methods

@@
 As indicator encoding is practically infeasible for high cardinality variables (a feature with $10^5$ levels would add $10^5$ new columns), we limit the number of indicator coded features by collapsing rare levels beyond a varying threshold to a single *other* category before encoding.
-Closely related to dummy encoding in classical analysis of variance and regression are Helmert and Polynomial contrasts [@chambers_1992].
-However, these encodings focus on ordinal categorical features, which we did not consider in our study.
 
@@
 As we are interested in model-agnostic methods that can be combined with any supervised ML algorithm, we do not investigate several strategies proposed in the recent literature [@guo_2016; @prokhorenkova_2018].
-Additionally, we only investigate univariate encodings, i.e. we always encode each variable separately.
+We do not differentiate between nominal and ordinal categorical features:
+Most publicly available datasets simply do not contain any high cardinal ordinal features.
+For many applications, ordinal feature information is not available as metadata which makes it less relevant for some applications like automatic ML.
+Thus, our benchmark does not include specific ordinal encoding methods like Helmert and Polynomial contrasts [@chambers_1992] which are closely related to dummy encoding or modern ordinal penalization methods [@tutz_2016].
+We only investigate univariate encodings, i.e. we always encode each variable separately.
 Levels with comparable main effects can therefore not be distinguished based on the transformed feature, which prevents the consecutive ML algorithm from learning interactions for specific levels.
-Methods that can jointly encode several variables and therefore leverage correlation structures between the different variables are an interesting avenue of research in the future:
+Methods that can jointly encode several variables and therefore leverage correlation structures between the different variables are an interesting avenue for future research:
 Recent approaches use target-based encoding strategies to learn a vector-valued representation of a level, similar to entity embeddings [@guo_2016].
@@
 @deleeuw_1976] leading to the *aspect* framework [@mair_2010], which is unfortunately unfamiliar to many ML researchers.
-We furthermore only investigate traditional categorical variables and do not extend our analysis to multi-categorical variables or text-strings [@cerda_2018; @cerda_2020], where other encoding techniques can harvest additional information.
+We only investigate traditional categorical variables and do not extend our analysis to multi-categorical variables or text-strings [@cerda_2018; @cerda_2020], where other encoding techniques can harvest additional information.
 

### 2021-09-09 publication/manuscript.Rmd
shorten text

@@
 We denote the observed frequency of class $c$ in the training set with $N_c$ and the observed frequency of a level $l$ in the training set with $N_l$.
-As most ML algorithms require numerical features, we use categorical encoding techniques to transform each nominal feature $x^{train}$ into numerical features  $\hat{x}^{train}$ which are then used for training.
+We investigate categorical encoding techniques to transform each nominal feature $x^{train}$ into numerical features  $\hat{x}^{train}$ which are then used for training.
 If clear from the context, we use $\hat{x}_l$ as the encoded value for an observation with level $l$.
@@
 We broadly categorize feature encoding techniques into *target-agnostic* methods and *target-based* methods [@micci_barreca_2001].
-Figure \ref{fig:overview} contains a taxonomy of the encoding methods we consider in this work.
+Figure \ref{fig:overview} contains a taxonomy of our considered encoding methods.
 *Target-agnostic* methods do not rely on any information about the target variable and can therefore also be used in unsupervised settings.
 Simple strategies from this domain e.g.
-one-hot or dummy encoding are widely used -- in the scientific literature [@kuhn_2019; @hancock] but also on  Kaggle\footnote{\url{https://www.kaggle.com}} -- to embed variables for classical ML algorithms as well as (deep) neural networks.
+one-hot or dummy encoding are widely used -- in the scientific literature [@kuhn_2019; @hancock] but also on  Kaggle\footnote{\url{https://www.kaggle.com}} to embed variables for classical ML algorithms as well as (deep) neural networks.
 Such indicator methods map each level of a categorical variable to a set of dichotomous features encoding the presence or absence of a particular level.
 An obvious drawback of indicator encoding is that it adds one additional feature per level of a categorical variable.
-When the number of levels increases to a point where indicator encoding leads to an unreasonable number of features, levels are often simply mapped to integer values with random order (integer encoding).
+When indicator encoding leads to an unreasonable number of features, levels are often mapped to integer values with random order (integer encoding).
 Alternatively, the "hashing trick" [@weinberger_2009] can be used to randomly collapse feature levels into a smaller number of indicator variables [@kuhn_2019], or levels can be encoded by using the observed frequency of a given level in the dataset (frequency encoding).
@@
 The basic idea of more advanced methods called target, impact, mean, or likelihood encoding is to use the training set to make a simple prediction of the target for each level of the categorical feature, and to use the prediction as the numerical feature value $\hat{x}_l$ for the respective level.
-One of the earliest formal descriptions of this strategy is @micci_barreca_2001.
-A very similar approach is called *weights of evidence* in the credit scoring literature @hand_1997.
+An early formal description of this strategy is @micci_barreca_2001.
+Similar approaches are called *weights of evidence* in the credit scoring literature @hand_1997.
 In simple target encoding for regression problems, the mean target value in the training set from all observations with a certain feature level is used as the numeric value to encode that level for all observations: $\hat{x}_l = \frac{\sum_{i:x^{train}_i = l}y^{train}_i}{N_l}$.
@@
 
-Several small-scale studies that investigate aspects of our research questions have been previously conducted.
-However, those works do not yield conclusive results due to narrower or different scopes or not considering high cardinality variables.
+Several small-scale studies have been previously conducted.
+Those works did not yield conclusive results due to narrower scopes or not considering high cardinality variables.
 One benchmark ($6$ datasets) on encoding high cardinality features (cardinality between $103$ and $9095$) in combination with gradient boosting has been published on the Kaggle forums [@prokopev_2018].
 Different versions of target encoding are compared with indicator, integer, and frequency encoding.
-They recommend combining the smoothed version of target encoding with 4- or 5-fold CV, never using simple target encoding and use indicator encoding only for small datasets.
+They recommend combining the smoothed version of target encoding with 4- or 5-fold CV, never using simple target encoding and using indicator encoding only for small datasets.
 Interestingly, frequency encoding did perform well in many cases.

### 2021-09-09 publication/manuscript.Rmd
shorten text

@@
 We focus on the impact of those techniques on a subsequent algorithm's predictive performance, and -- if possible -- derive best practices on when to use which technique.
-We conducted a large-scale benchmark experiment, where we compared different encoding strategies together with five ML algorithms (lasso, random forest, gradient boosting, k-nearest neighbours, support vector machine) using datasets from regression, binary- and multiclass- classification settings.
+We conducted a large-scale benchmark experiment, where we compared different encoding strategies together with five ML algorithms (lasso, random forest, gradient boosting, k-nearest neighbors, support vector machine) using datasets from regression, binary- and multiclass- classification settings.
 Throughout our study, regularized versions of target encoding (i.e.
@@
 An early formal description of this strategy is @micci_barreca_2001.
-Similar approaches are called *weights of evidence* in the credit scoring literature @hand_1997.
 In simple target encoding for regression problems, the mean target value in the training set from all observations with a certain feature level is used as the numeric value to encode that level for all observations: $\hat{x}_l = \frac{\sum_{i:x^{train}_i = l}y^{train}_i}{N_l}$.
@@
 ML models would place a high priority on such an encoded feature during training but would perform badly on test data.
-To avoid this, practitioners often use regularized target encoding by adding a smoothing parameter $\epsilon$ which shrinks those effects towards the global mean [@micci_barreca_2001].
+To avoid this, practitioners often use regularized target encoding with a smoothing parameter that shrinks those effects towards the global mean [@micci_barreca_2001].
 An alternative strategy is to combine target encoding with cross-validation (CV) techniques [@prokhorenkova_2018].
@@
 @wright_2019 study treatments for categorical variables in random forests together with dummy and integer encoding (18 datasets, cardinality between $3$ and $38$), concluding that indicator and integer encoding perform subpar in comparison to methods that re-order levels according to the target variable.
-@prokhorenkova_2018 compared their new CatBoost variant of target encoding to smoothed target encoding with no CV, hold-out, and leave-one-out CV on $8$ datasets.
+@prokhorenkova_2018 compared their new CatBoost variant of target encoding to smoothed target encoding without CV, hold-out, and leave-one-out CV on $8$ datasets.
 While the CatBoost method performed best, hold-out came second and target encoding without CV performed worst.
-An overview of encoding techniques tailored to be used together with Neural Networks is provided in @hancock.
-They present a survey of indicator-based and embedding-based methods but do not conduct a benchmark study.
-One widely adopted approach, in particular, has been proposed by @guo_2016 under the term *entity embeddings*.
-In contrast, our work focuses on high cardinality variables and techniques that are agnostic to the subsequent ML method, which has not been done before.
-We furthermore study this problem on a sufficiently large variety of datasets and settings to allow for more compelling conclusions.
+An overview of encoding techniques tailored towards neural networks (e.g. the widely adopted *entity embeddings* by @guo_2016) is provided in @hancock.
+They present a survey of indicator- and embedding-based methods but no benchmark study.
+In contrast, our work is the first to focus on high cardinality variables and techniques that are agnostic to the subsequent ML method.
+We study this problem on a larger variety of datasets and settings.
 \newline
@@
 The main goal of our study is to assess the impact different categorical encoding techniques have on subsequent models' predictive performance.
-As the optimal encoding might differ depending on the ML algorithm, we consider various state-of-the-art algorithms, *regularized linear models*, *random forests*, *gradient tree boosting*, *k-nearest neighbors*, and *support vector machines*.
-To find default settings for high cardinality features, we analyze a variety of datasets with different characteristics including regression, binary classification and multiclass classification problems.
-As our methods vary in runtime and complexity, we aim to learn whether more complex methods, in general, are to be preferred, or if simpler approaches suffice.
-Furthermore, the relationship between a feature's cardinality and the choice of encoding technique has not been investigated, which we study by varying the minimum number of levels above which features are transformed.
+As the optimal encoding might differ depending on the ML algorithm, we consider various state-of-the-art algorithms, *regularized linear models* (LASSO), *random forests* (RF), *gradient tree boosting* (GB), *k-nearest neighbors* (KNN), and *support vector machines* (SVM).
+To find default settings for high cardinality features, we analyze a variety of datasets with different characteristics including regression, binary classification, and multiclass classification problems.
+Because our methods vary in runtime and complexity, we are interested whether more complex methods are to be preferred, or if simpler approaches suffice.
+We also study the relationship between a feature's cardinality and the choice of encoding technique by varying the minimum number of levels above which features are transformed.
 
@@
 
-We first survey various categorical encoding techniques and conduct a comprehensive benchmark study of a broad set of categorical encoding techniques with a focus on high cardinality features.
+We survey a broad set of categorical encoding techniques and conduct a comprehensive benchmark study with a focus on high cardinality features.
 We study $7$ different encoding techniques in conjunction with $5$ commonly used ML algorithms across $24$ datasets, both from a classification and a regression regime. 
@@
 After a discussion of results concerning predictive performance, we provide further analyses seeking to inform practitioners which methods to apply.
-This includes an important discussion on runtimes.
+This includes an important discussion of runtimes.
 
@@
 Pseudocode is presented in the main text (for non-standard encodings), or the Supplementary Material (for standard methods).
-An important detail is how encoding techniques treat new levels during the prediction phase, as this is often not obvious.
+An important detail is how encoding techniques treat new levels during the prediction phase.
 
@@
 
-The simplest strategy for categorical features is integer encoding, which is also often referred to as ordinal encoding.
-To make sure that the order of the observed levels does not affect our results, we randomly map the observed levels from the training set to the integers $1$ to $L$.
-Although new levels could be mapped to $L+1$ or $0$, model predictions would be completely arbitrary as the order of the integers does not contain any information.
+The simplest strategy for categorical features is integer encoding (also called ordinal encoding).
+We randomly map the observed levels from the training set to the integers $1$ to $L$.
+Although new levels could be mapped to $L+1$ or $0$, model predictions would be arbitrary as the integer order does not carry information.
 Thus, we encode new levels as missing values and use mode imputation to obtain the integer which matches the most frequent level in the training set.
-Integer encoding should be an acceptable strategy for tree-based models, which can separate all original levels with repeated splits, but less so for other algorithms.
+Integer encoding should only be an acceptable strategy for tree-based models, which can separate all original levels with repeated splits.
 
@@
 
-Frequency encoding simply maps each level to its observed frequency in the training set ($\hat{x}_l = N_l$).
-On the one hand, this assumes a functional relationship between the frequency of a level and the target.
-On the other hand, it implicitly reduces the number of levels, as the subsequent model can best differentiate between levels with dissimilar frequencies.
-This approach is typically used e.g.
-in natural language processing to encode token or n-gram counts.
-We encode new levels with a frequency of $1$, although choosing the minimal observed frequency for any level during training might be another reasonable strategy.
+Frequency encoding maps each level to its observed frequency in the training set ($\hat{x}_l = N_l$).
+This assumes a functional relationship between the frequency of a level and the target.
+It implicitly reduces the number of levels, and the subsequent model can best differentiate between levels with dissimilar frequencies.
+This approach is heavily used in natural language processing to encode token or n-gram counts.
+We encode new levels with a frequency of $1$.
 
@@
 
-We use indicator encoding as an umbrella term for the two most common strategies to encode categorical features with a small to moderate number of levels: one-hot and dummy encoding.
+We use indicator encoding as an umbrella term for two common strategies to encode categorical features with a small to moderate number of levels: one-hot and dummy encoding.
 One-hot encoding transforms the original feature into $L$ binary indicator columns, each representing one original level.
@@
 For one-hot encoding, the zero vector can be used to encode new levels which were not observed during training.
-For dummy encoding, this is not useful as it collapses new levels to the (often arbitrary) reference category.
-In our study, the first level in alphabetical order is used as the reference category for dummy encoding.
+For dummy encoding, it is not useful to collapse new levels to the often arbitrary reference category; in our case the first level in alphabetical order.
 We replace new levels in the prediction phase with the most frequent level in the training set.
@@
 This integer is then transformed into an indicator representation, with $1$ in indicator column number $(hash(l) \mod hash.size) + 1$ and $0$ in all remaining columns [@kuhn_2019].
-Some levels will be hashed to the same indicator representation.
+Some levels are hashed to the same indicator representation.
 The smaller the $hash.size$, the higher the number of collapsed levels.
 The number of indicators is often effectively lower than $hash.size$, as some indicators can be constant in the training set (we remove those columns for both training and prediction).
-Although hashing would allow us to jointly hash multiple features, we hash each feature separately to improve comparability with the other encoders.
+Although we could jointly hash multiple features, we hash each feature separately to improve comparability with the other encoders.
 
@@
 
-In leaf encoding, a decision tree is fitted on the training set to predict the target based on the categorical feature.
+Leaf encoding fits a decision tree on the training set to predict the target based on the categorical feature.
 Each level is encoded by the number of the terminal node, in which an observation with the respective level ends up.
@@
 
-The first formal description of a so-called impact or target encoder was provided by @micci_barreca_2001.
+An early formal description of a so-called impact, target or James-Stein encoder was provided by @micci_barreca_2001.
 The basic idea is to encode each feature level with the conditional target mean (regression) or the conditional relative frequency of one or more target classes (classification).
@@
 A smoothing parameter $\epsilon$ is introduced to avoid division by zero.
-This parameter could be further used to regularize towards the unconditional mean.
-To allow comparing regularized (introduced below) with simple target encoding, we always choose a small $\epsilon = 0.0001$ for the impact encoder.
-Similar techniques have been proposed in the literature:
-Weights of evidence encoding from the credit scoring classification literature [@hand_1997] is almost identical to impact encoding, but does not
-consider any regularization or centering.
-Impact encoding is also referred to as a target or James-Stein encoding.
+This parameter could be used to further regularize towards the unconditional mean.
+We always choose a small $\epsilon = 0.0001$ as we want to compare simple target encoding with the regularized encoder introduced next.
+Weights of evidence encoding from the credit scoring classification literature [@hand_1997] is almost identical to impact encoding, but without regularization or centering.
 
@@
 This connection is described in @kuhn_2019.
-To achieve regularized impact encoding, we constructed our own glmm encoders for regression, binary, and multiclass classification which are described in Algorithm \ref{alg:glmm-regr} (regression) and in the Supplementary Material (classification).
+To achieve regularized impact encoding, we implemented glmm encoders for regression, binary, and multiclass classification which are described in Algorithm \ref{alg:glmm-regr} (regression) and in the Supplementary Material (classification).
 The encoded value for each level is based on the spherical conditional mode estimates.
@@
 A further strategy to avoid overfitting in target encoding is to combine it with CV to train the encoder on independent observations without limiting the data to train the ML model.
-To our knowledge, we provide the first implementation which combines target encoding based on glmms with CV.
+We provide an implementation which combines target encoding based on glmms with CV.
 During the training phase, we partition the data using CV into $n.folds$ and fit a glmm on each resulting training set.
@@
 In classification problems, the most frequent class of the target within the training set is predicted.
-For each dataset, we also consider a random forest model without encoding (**none** condition), to compare the use of encoding methods in a random forest with a natural categorical splitting approach.
-We use the random forest in ranger [@wright_2017] which provides efficient categorical feature support by ordering levels once before starting the tree growing algorithm [@wright_2019].
-In the **remove** high cardinality features control condition, we remove features with a high number of levels above some threshold and use one-hot encoding (without collapsing rare levels) for the remaining features.
+For each dataset, we also consider a RF without encoding (**none** condition), to compare the use of encoding methods with a natural categorical splitting approach.
+The ranger [@wright_2017] implementation provides efficient categorical feature support by ordering levels once before starting the tree growing algorithm [@wright_2019].
+In the **remove** high cardinality features control condition, we omit features with a high number of levels above some threshold and use one-hot encoding (without collapsing rare levels) for the remaining features.
 This condition reflects on whether including high cardinality features does indeed improve predictive performance.
 Otherwise, the best encoding might just provide the least impairment compared to not including any high cardinality features.
-We include an overview of available implementations in widely used machine learning frameworks for R and python in the Supplementary Material.
+We include an overview of available implementations in widely used ML frameworks for R and python in the Supplementary Material.
 
@@
 
-Table \ref{tab:datasets} summarizes the datasets used in our benchmark study.
+Table \ref{tab:datasets} summarizes our benchmark datasets.
 We specifically investigate datasets that contain categorical variables with a large number of levels, including many well-known datasets
@@
 -->
-It is widely assumed that more involved encoding methods are only advantageous for variables with a high number of levels, while simple indicator encoding is more appropriate for a small number of levels.
+It is often assumed that advanced encoding methods are only advantageous for variables with a high number of levels, while simple indicator encoding is more appropriate for a small number of levels.
 To make our benchmark results more realistic, a high cardinality threshold (HCT) parameter with values of $10$, $25$, and $125$ was introduced which determined varying configurations for the different encoders.
@@
 HCT is used as the hash size in hash encoding.
-In the remove control condition, features with more than HCT levels in the training set are removed from the feature set.
-Based on the number of categorical features and levels per feature, some HCT settings were removed from the benchmark for some combinations of dataset $\times$ encoder.
+In the remove control condition, features with more than HCT levels in the training set are removed from the feature set.\footnote{Based on the number of categorical features and levels per feature, some HCT settings were removed from the benchmark for some combinations of dataset $\times$ encoder.
 This made sure that encoders always affect at least one feature and that the remove condition always removes at least one feature.
-If several HCT settings of an encoder would lead to identical encoding strategies for all features of a dataset, we only kept the condition with the smallest HCT value.
+If several HCT settings of an encoder would lead to identical encoding strategies for all features of a dataset, we only kept the condition with the smallest HCT value.}
 
@@
 
-Because we investigate strategies that can be used together with different models, we separately report results for different ML algorithms.
-We kept tuning the hyperparameters of the subsequent ML methods to a minimum because we were interested in the effect of the encoding techniques instead of a comparison of the ML models.
-In total, we investigate $5$ ML methods:
-Regularized linear models (LASSO) were fitted with `glmnet` [@friedman_2010], internally tuning the regularization using 5-fold CV.
-Random forests (RF) were trained using `ranger` [@wright_2017].
-Because the random forest algorithm can be expected to give reasonable results without tuning [@probst_2019], we only fixed the number of fitted trees to $500$.
-Gradient boosting models (GB) were trained using `xgboost` [@chen_2018].
-We set the learning rate to $0.01$ and determine the number of iterations using early stopping on a 20% holdout set.
-K-nearest neighbours (KNN) was taken from package `kknn` [@schliep_2016] standardizing all features and using a constant $k=15$ for the number of nearest neighbours, together with an information gain filter [@brown2012] to limit the number of features to $25$.
-Support vector machines (SVM) with radial basis function kernel were trained with `liquidSVM` [@steinwart_2017].
-The bandwidth and regularization parameters were internally tuned using 5-fold CV, and we used a one-vs-all approach for multiclass-settings.
+In total, we investigate $5$ ML algorithms.
+We kept tuning their hyperparameters to a minimum because we were interested in the effect of the encoding techniques instead of a comparison of ML algorithms.
+LASSOs were fitted with `glmnet` [@friedman_2010], internally tuning the regularization using 5-fold CV.
+RFs with $500$ trees were trained using `ranger` [@wright_2017] without tuning, because RFs can be expected to give reasonable results with default settings [@probst_2019].
+GB models were trained using `xgboost` [@chen_2018].
+We set the learning rate to $0.01$ and determined the number of iterations using early stopping on a 20% holdout set.
+KNN was taken from package `kknn` [@schliep_2016].
+We standardized all features and used a constant $k=15$ for the number of nearest neighbors, together with an information gain filter [@brown2012] to limit the number of features to $25$.
+SVMs with radial basis function kernel were trained with `liquidSVM` [@steinwart_2017].
+The bandwidth and regularization parameters were internally tuned using 5-fold CV. We used a one-vs-all approach for multiclass-settings.
 
 The ML pipeline outlined below was used for all experimental conditions.
-It is not trivial and was carefully designed to ensure consistent results for extreme conditions (e.g. some levels only existing in the training data).
+It was carefully designed to ensure consistent results for extreme conditions (e.g. some levels only existing in the training data).
 
@@
 
-Throughout our analysis, we use implementations of the various methods for the open-source statistical software R [@r_2021].
+We perform all analyses in the open-source statistical software R [@r_2021].
 To enable a fair and reliable comparison in our study, we implemented all encoding methods on top of the `mlrCPO` package [@binder_2018].
 The pre-processing, as well as the final ML algorithm described in \ref{sec:pipeline}, were trained and resampled using the `mlr` framework [@bischl_2016] together with the `batchtools` package [@lang_2017] to scale the benchmark analysis to HPC compute infrastructure.
-All materials for this study, including reproducible code for this manuscript and the necessary result objects can be downloaded from our **online repository** \footnote{\url{https://github.com/compstat-lmu/paper_2021_categorical_feature_encodings}}.
-
-Throughout our experiments, we use 5-fold CV to obtain estimates of the predictive performance.
-Depending on the target variable, we report root mean squared error (RMSE) for regression, area under the curve (AUC) for binary classification or its extension AUNU [@ferri_2009] for multiclass problems:
+All materials for this study, including reproducible code for this manuscript and result objects can be downloaded from our **online repository** \footnote{\url{https://github.com/compstat-lmu/paper_2021_categorical_feature_encodings}}.
+
+Throughout our experiments, we use 5-fold CV to obtain estimates of predictive performance.
+Depending on the target variable, we report root mean squared error (RMSE) for regression, area under the curve (AUC) for binary classification, and its extension AUNU [@ferri_2009] for multiclass problems:
 \begin{align*}
@@
 
-Throughout the benchmark study, each encoding method listed in Section \ref{sec:methods} is applied together with all ML algorithms (c.f. Section \ref{sec:pipeline}) across
+In our benchmark study, each encoding method listed in Section \ref{sec:methods} is combined with all ML algorithms (c.f. Section \ref{sec:pipeline}) across
 high cardinality thresholds (HCT) $10$, $25$, $125$.
@@
 
-The overarching question we aim to answer in our study is which encoding methods generally work well across various datasets.
+Our main question is which encoding methods generally work well across various datasets.
 We report results for regression and classification datasets separately, as the associated metrics differ in scale.
@@
 For `r length(svm_errors[, unique(Name)])` datasets, some conditions with the SVM led to unexpected crashes due to memory problems or numerical errors.
-We completely removed those datasets for the SVM when computing ranks or other statistics to compare encodings across datasets.
+We completely removed those datasets for the SVM when computing ranks or other statistics that compare encodings across datasets.
 
@@
 
-Mean performance estimates along with minimum and maximum performance from the 5-CV folds are reported for all datasets in Figures \@ref(fig:regr-res) (regression), \@ref(fig:bincl-res) (binary classification), and \@ref(fig:multcl-res) (multiclass classification).
+Mean performance estimates along with minimum and maximum performance are reported for all datasets in Figures \@ref(fig:regr-res) (regression), \@ref(fig:bincl-res) (binary classification), and \@ref(fig:multcl-res) (multiclass classification).
 To reduce the complexity induced by the hyperparameter HCT we only display the parameter condition with the best performance for each combination of dataset $\times$ encoding $\times$ ML algorithm.
@@
 For some datasets, the remove condition performed very similar to the other encodings (e.g. *ames-housing*, *porto-seguro*), suggesting that categorical features were not informative.
-Performance in some folds was below the FL learner for *flight-delay-usa-dec-2017* ($RMSE_{FL} =$ `r res[lrn.id == "FL" & Name == "flight-delay-usa-dec-2017", rmse.test.mean]`) and *nyc-taxi-green-dec-2016* ($RMSE_{FL} =$ `r res[lrn.id == "FL" & Name == "nyc-taxi-green-dec-2016", rmse.test.mean]`).
+Performance in some CV folds was below the FL learner for *flight-delay-usa-dec-2017* ($RMSE_{FL} =$ `r res[lrn.id == "FL" & Name == "flight-delay-usa-dec-2017", rmse.test.mean]`) and *nyc-taxi-green-dec-2016* ($RMSE_{FL} =$ `r res[lrn.id == "FL" & Name == "nyc-taxi-green-dec-2016", rmse.test.mean]`).
 <!-- den satz können wir nicht rauslassen, sonst gibt es zum FL keine Information im ganzen Paper -->
 
-On datasets where substantive performance differences could be observed, target encoding with the glmm encoder was generally **most** effective.
-The **worst** encoder depended on the ML algorithm and respective dataset.
-For datasets *Click_prediction_small*, *KDDCup09_upselling*, *kick*, and *okcupid-stem* some combinations of encoder $\times$ ML algorithm performed worse than simply removing high cardinality features.
+On datasets with substantive performance differences, target encoding with the glmm encoder was generally **most** effective.
+The **worst** encoder differed by ML algorithm and dataset.
+For datasets *Click_prediction_small*, *KDDCup09_upselling*, *kick*, and *okcupid-stem* some encoder $\times$ ML algorithm conditions performed worse than simply removing high cardinality features.
 
@@
 
-To further summarize the results, we computed meta rankings for each ML algorithm:
+To further summarize, present meta rankings for each ML algorithm in Figure \ref{fig:rank-res}.
 First, we defined an encoder relation within each dataset, based on corrected resample t-tests [@nadeau_2003].
@@
 The symmetric difference between two relations is the number of cases one encoding beats another encoding in one relation but not in the other one.
-This procedure resulted in the meta rankings based on the consensus ranking for each ML algorithm detailed in Figure \ref{fig:rank-res}.
 
@@
 Lower ranks indicate better performance.
-The rank of the *none* control condition of the random forest (rank 11) was omitted from the figure.
+The rank of the *none* control condition of the RF (rank 11) was omitted from the figure.
 
@@
 Looking at meta-rankings, approaches based on GLMM's in combination with cross-validation outperform all other approaches across all algorithms.
-A further interesting detail omitted in Figure \ref{fig:rank-res} for clarity is that the \textit{none} encoding strategy for the random forest was beaten by all strategies except for the \textit{remove} condition.
-This implies, that even if the algorithm includes a mechanic for treating categorical variables, it might often be optimal to use a different strategy instead.
-
-In an additional exploratory analysis, we tried to find clusters of datasets, which show systematic patterns of encoder performance (independent of the employed ML algorithm).
+A further interesting detail omitted in Figure \ref{fig:rank-res} for clarity is that the \textit{none} encoding strategy for the RF was beaten by all strategies except for the \textit{remove} condition.
+This implies, that even if the algorithm provides a mechanic for treating categorical variables, it might often be optimal to use a different strategy instead.
+
+In an exploratory analysis, we tried to find clusters of datasets based on systematic patterns of encoder performance (independent of the employed ML algorithm).
 We computed a partial-order consensus relation across ML algorithms for each dataset and hierarchically clustered the dataset consensus relations using the symmetric difference distance in combination with the complete linkage agglomeration method.
@@
 Although the cluster structure is somewhat ambiguous, roughly three clusters can be described:
-The first 9 datasets from the top of the dendrogram are characterized by a medium to a high number of levels, low performance of the remove condition (indicating the importance of high cardinality features) and clear performance advantage of target encoders.
+The first 9 datasets from the top of the dendrogram are characterized by a medium to high number of levels, low performance of the remove condition (indicating the importance of high cardinality features) and clear performance advantage of target encoders.
 For the next 13 datasets which contain the smallest number of levels, traditional encodings can compete with target encoding.
-This biggest cluster also includes 9 datasets with zero distances, in which no significant performance differences could be observed between any encoding conditions (nor with the remove condition, indicating that high cardinality features are not very informative).
+This biggest cluster also includes 9 datasets with zero distances, in which no significant performance differences could be observed between any encoding conditions (nor with the remove condition, indicating that high cardinality features are less informative).
 The last two datasets with the highest number of levels formed a separate cluster, in which target encoding without strong regularization (impact, glmm-noCV) showed severe overfitting.
@@
 While the impact encoding used no regularization, we compared different amounts of regularization for the glmm encoder.
-When performance differences between target encoders were observed, more regularization seemed beneficial.
+When performance differences between target encoders were observed, regularization seemed beneficial.
 In those datasets, impact encoding not only performed worse than glmm with CV but was sometimes also inferior compared to other encoders.
@@
 
-We observed several interesting patterns, which we aim to summarize:
-Integer encoding did not perform well together with XGBoost in our benchmark in contrast to what is stated in the documentation of other boosting libraries\footnote{\url{https://lightgbm.readthedocs.io/en/latest/Advanced-Topics.html\#categorical-feature-support}}.
+Furthermore, we observed several interesting patterns:
+Integer encoding did not perform well with GB in our benchmark in contrast to what is stated in the documentation of some boosting libraries\footnote{\url{https://lightgbm.readthedocs.io/en/latest/Advanced-Topics.html\#categorical-feature-support}}.
 In contrast, target-based encoders (especially the glmm encoder) seem to be a compelling strategy.
@@
 Although the glmm encoders ranked first in our benchmark for the LASSO, it was the only algorithm where the indicator encoders achieved the next best ranking.
-Note that for computational reasons we limited the maximum amount of indicator variables to 125 in our experimental design.
+Note that for computational reasons we limited the maximum amount of indicator variables per feature to 125 in our experimental design.
 The HCT = 125 setting performed best for LASSO with indicator encoding in a large number of datasets, indicating that performance might have further improved with higher values.
@@
 This was backed by our benchmark results.
-For some datasets (*medical_charges*, *road-safety-drivers-sex*, and *Midwest_survey*) KNN (without tuning of the optimal number of nearest neighbours) could compete with the more sophisticated ML algorithms when combined with target encoding, but performed poorly with other encoders.
+For some datasets (*medical_charges*, *road-safety-drivers-sex*, and *Midwest_survey*) KNN (without tuning of the optimal number of nearest neighbors) could compete with the more sophisticated ML algorithms when combined with target encoding, but performed poorly with other encoders.
 Although consistent with the big picture, SVM results have to be considered with care, as some experimental conditions resulted in unexpected computational errors.
@@
 
+<!-- heavily shorten this section or combine with others? -->
 In line with earlier research [@micci_barreca_2001; @prokhorenkova_2018], target encoding with regularization (glmm) performed better or equally well in comparison with the unregularized impact encoder.
@@
 
-On a side note, we found that one-hot encoding usually gave a slightly better performance than dummy encoding.
-This phenomenon has also been observed by @chiquet_2016.
-@tutz_2016 commented that in generalized linear models "the naive combination of usual dummy coding with reference category and standard penalties is often a bad idea" (p.254) unless the reference category is of special interest.
-In our benchmark, the reference category was chosen arbitrarily as is usually the case in ML.
+We found that one-hot encoding usually gave a slightly better performance than dummy encoding, which has also been observed by @chiquet_2016 and commented on in
+@tutz_2016 (p.254).
 Our results suggest that one-hot encoding is the better standard compared to dummy encoding when applying nonlinear regularized models like RF, GB or SVM with (high cardinal) categorical features.
-We provide a more in-detail analysis of the performances of one-hot and dummy encoding in the Supplementary Material.
+We provide a more detailed comparison for indicator encoding in the Supplementary Material.
 
@@
 
-To determine whether traditional encodings might be worth it when facing limited computational resources, we further analyzed the runtimes of the whole analysis pipeline for different encoders and ML algorithms.
+To determine whether traditional encodings are preferred when facing limited computational resources, we further analyzed runtimes of the whole analysis pipeline for different encoders and ML algorithms.
 Aggregated results are shown in Table \@ref(tab:time-ranks).
-To enable a meaningful comparison, we report runtime as the fraction of a full pipeline's strategy compared to running simple one-hot encoding for the same pipeline and then further aggregate across datasets using the median.
-We again only report the **best** HCT setting.
+To enable a meaningful comparison, we report runtime as the fraction of a full pipeline's strategy compared to the one-hot encoding condition and then further aggregate across datasets using the median.
+Again, we only report the **best** HCT setting.
 Absolute runtimes are hard to interpret and aggregate because runtime distributions across datasets are heavily skewed as proportionally larger runtimes are observed for big datasets.
-Furthermore, while a pipeline's runtime can be dominated by the encoder for small datasets, the training of the consecutive ML algorithm is dominating for large datasets, which might render differences during encoding irrelevant.
+While a pipeline's runtime can be dominated by the encoder for small datasets, the training of the consecutive ML algorithm is dominating for large datasets, which might render differences during encoding irrelevant.
 Therefore we aim to report what is important in practice, the time differences for training the **full pipeline**.
 The results clearly show that regularized target encoding does not consistently yield slower runtimes compared to simple strategies like indicator encoding.
-Reasons for this might be, that more efficient representations yielded by target encoders lead to faster runtimes of subsequent ML algorithms.
+Supposedly, the more efficient representations produced by target encoding lead to faster runtimes of subsequent ML algorithms.
 This suggests that a possible runtime vs.
@@
 Until now, we ignored the HCT parameter by reporting only the condition with the best performance.
-An interesting question was whether target encoding is only useful for features with a very high number of levels or if it might also have advantages for fewer levels, where most practitioners would routinely use indicator encoding.
+An interesting question is whether target encoding is only useful for features with a very high number of levels or also for fewer levels, where most practitioners would routinely use indicator encoding.
 We tested this using HCT thresholds of $10$, $25$ and $125$, but the results are not easy to interpret.
@@
 In comparison, other methods generally yielded better performance if features above the very low HCT of $10$ were encoded using one of the target encoding strategies.
-Furthermore, different regularization strategies for the target based encoders seemed not to prefer strongly different HCT values.
+Furthermore, different regularization for the target based encoders seemed not to prefer strongly different HCT values.
 
@@
 
-This benchmark study compared the predictive performance of a variety of strategies to encode categorical features with a high number of unordered levels for different supervised machine learning algorithms.
+This benchmark study compared the predictive performance of a variety of strategies to encode categorical features with a high number of unordered levels for different supervised ML algorithms.
 Regularized versions of target encoding, which uses predictions of the target variable as numeric feature values, performed better than traditional strategies like integer or indicator encoding.

### 2021-09-09 publication/manuscript.Rmd
shorten text

@@
 In comparison, other methods generally yielded better performance if features above the very low HCT of $10$ were encoded using one of the target encoding strategies.
-Furthermore, different regularization for the target based encoders seemed not to prefer strongly different HCT values.
+Differences in regularization for the target based encoders seemed not to prefer different HCT values.
 
@@
 
-In our benchmark, we compared encoding strategies for high cardinality features on a variety of regression, binary and multiclass classification datasets with different ML algorithms.
-Regularized target encoding (glmm) was superior across most datasets and ML algorithms.
+In our benchmark, we compared encoding strategies for high cardinality categorical features on a variety of regression, binary, and multiclass classification datasets with different ML algorithms.
+Regularized target encoding was superior across most datasets and ML algorithms.
 Although the performance of other encoding strategies was comparable in some conditions, target encoding was never outperformed.
-In general, our results suggest that regularized target encoding works well for all kinds of algorithms and could be considered by default when working with high cardinality features.
-In terms of runtime, using target encoding together with 5-fold CV sometimes leads to slightly longer runtimes (in comparison to indicator encoding), but especially for larger datasets this is often offset by the more efficient representation yielded from target encoding.
+In general, our results suggest that regularized target encoding based on glmms with 5-fold CV (glmm-5CV) works well for all kinds of algorithms and should be a reasonable default strategy.
+It sometimes leads to slightly longer runtimes (in comparison to indicator encoding), but especially for larger datasets this is often offset by the more efficient representation produced by target encoding.
 
 What constitutes a "high" cardinality problem is a difficult question that might not only depend on the number of levels but also on other characteristics of the dataset and its features.
-Unfortunately, the number of datasets in our study was too small to discover consistent patterns between encoder performance and dataset characteristics.
-Target encoding features with as little as 10 levels seemed to be effective in a substantive number of conditions, but for other datasets, higher HCT values performed better.
+Supposedly, the number of datasets in our study was too small to discover consistent patterns between encoder performance and dataset characteristics.
+Target encoding features with only 10 levels seemed to be effective in a substantive number of conditions, but for other datasets, higher HCT values performed better.
 Thus, some form of hyperparameter tuning seems necessary to decide the level threshold for target encoding at this point, as no suitable defaults seem to be available.
 Note that we compared different HCT values but then used the same encoding strategy for all affected features alongside one-hot encoding for the remaining ones.
-Although this strategy is reasonable, it might be beneficial to decide whether to use target encoding on a feature by feature basis.
+A further improvement could be to decide whether to use target encoding on a feature by feature basis.
 ML pipelines could introduce a categorical hyperparameter for each feature that represents which encoding is used.
@@
 
-In our study, several decisions were necessary to make answering our research questions technically and computationally feasible.
-First, we only used minimal tuning for our ML algorithms, as we were not interested in comparing their performance against each other.
+Several design decisions were necessary to make answering our research questions computationally feasible:
+We used minimal tuning for our ML algorithms, as we were not interested in comparing their performance against each other.
 This probably led to suboptimal performance for the GB, KNN, and SVM learners.
 When interpreting our results, we assume that the encoder rankings are comparable when more extensive tuning is used.
-This seems plausible, considering the high stability of our meta rankings between algorithms.
-We only use 5-fold-CV without repetitions to estimate predictive performance, but due to a small variance between folds, we could confidently detect performance differences of encoders for many datasets.
-As we are interested in model-agnostic methods that can be combined with any supervised ML algorithm, we do not investigate several strategies proposed in the recent literature [@guo_2016; @prokhorenkova_2018].
+This is plausible because the meta rankings between algorithms were highly stable.
+We only used 5-fold-CV without repetitions to estimate predictive performance, but due to a small variance between folds, we could confidently detect performance differences of encoders for many datasets.
+Because we are interested in model-agnostic methods that can be combined with any supervised ML algorithm, we did not investigate recent model-specific strategies [@guo_2016; @prokhorenkova_2018].
 We do not differentiate between nominal and ordinal categorical features:
@@
 For many applications, ordinal feature information is not available as metadata which makes it less relevant for some applications like automatic ML.
-Thus, our benchmark does not include specific ordinal encoding methods like Helmert and Polynomial contrasts [@chambers_1992] which are closely related to dummy encoding or modern ordinal penalization methods [@tutz_2016].
+Thus, our benchmark does not include specific ordinal encoding methods like Helmert and Polynomial contrasts [@chambers_1992] or ordinal penalization methods [@tutz_2016].
+We only investigate traditional categorical variables and do not extend our analysis to multi-categorical variables or text-strings [@cerda_2018; @cerda_2020], where other encoding techniques can harvest additional information.
 We only investigate univariate encodings, i.e. we always encode each variable separately.
-Levels with comparable main effects can therefore not be distinguished based on the transformed feature, which prevents the consecutive ML algorithm from learning interactions for specific levels.
-Methods that can jointly encode several variables and therefore leverage correlation structures between the different variables are an interesting avenue for future research:
+Levels with comparable main effects can not be distinguished based on the transformed feature, which prevents the consecutive ML algorithm from learning interactions for specific levels.
+Methods that jointly encode features and therefore leverage correlational structures between features are an interesting avenue for future research:
 Recent approaches use target-based encoding strategies to learn a vector-valued representation of a level, similar to entity embeddings [@guo_2016].
 Those should be compared to classical approaches from the optimal scaling literature [e.g.
-@deleeuw_1976] leading to the *aspect* framework [@mair_2010], which is unfortunately unfamiliar to many ML researchers.
-We only investigate traditional categorical variables and do not extend our analysis to multi-categorical variables or text-strings [@cerda_2018; @cerda_2020], where other encoding techniques can harvest additional information.
+@deleeuw_1976] or the subsequent *aspect* framework [@mair_2010], which are unfortunately unfamiliar to many ML researchers.
 
@@
 Regularized versions of target encoding, which uses predictions of the target variable as numeric feature values, performed better than traditional strategies like integer or indicator encoding.
-Most effective, with a consistent superior performance across Ml algorithms and datasets, was a target encoder which combines simple generalized linear mixed models with cross-validation and does not require tuning hyperparameters.
+Most effective, with a consistently superior performance across Ml algorithms and datasets, was a target encoder which combines simple generalized linear mixed models with cross-validation and does not require hyperparameter tuning.
 

### 2021-09-09 publication/manuscript.Rmd
remove seed warning
restructure and shorten summary of results

@@
 
-```{r}
+```{r, warning=FALSE}
 res = readRDS("../analysis/high_cardinality_benchmark/results.rds")
@@
 
-Our results indicate that regularized target encoding was superior or at least competitive on all datasets.
+Regularized target encoding was superior or at least competitive on all datasets.
+We could not observe a setting in which regularized target encoding was convincingly beaten by target agnostic methods.
 Especially effective was glmm encoding with 5-fold-CV, which ranked first place for all ML algorithms except SVM.
-While the impact encoding used no regularization, we compared different amounts of regularization for the glmm encoder.
-When performance differences between target encoders were observed, regularization seemed beneficial.
-In those datasets, impact encoding not only performed worse than glmm with CV but was sometimes also inferior compared to other encoders.
-We could not observe a setting in which regularized target encoding was convincingly beaten by target agnostic methods.
-When looking for a simple default method, indicator encoding in combination with collapsing small levels seems a robust alternative, although the glmm strategies should be preferred.
-
-Furthermore, we observed several interesting patterns:
-Integer encoding did not perform well with GB in our benchmark in contrast to what is stated in the documentation of some boosting libraries\footnote{\url{https://lightgbm.readthedocs.io/en/latest/Advanced-Topics.html\#categorical-feature-support}}.
-In contrast, target-based encoders (especially the glmm encoder) seem to be a compelling strategy.
+Performance often did not improve with glmm-10CV, suggesting that 5 folds might be a good regularization default in practice.
+In line with earlier research [@micci_barreca_2001; @prokhorenkova_2018], target encoding with regularization (glmm) performed better or equally well in comparison with the unregularized impact encoder.
+When glmm performed better, impact encoding not only performed worse than glmm with CV but was sometimes also inferior compared to other encoders.
+
+The following observations were also interesting:
+Surprisingly, integer encoding did not perform well with GB and target-based encoders (especially the glmm encoder) seem to be preferable.
 For LASSO, previous studies have suggested that indicator encoding works well, even with a very high number of levels [@cerda_2018].
@@
 For a small number of levels, this approach has been reported superior to indicator and integer encoding [@wright_2019], while we observed poor performance for datasets with a larger number of levels.
-
-<!-- heavily shorten this section or combine with others? -->
-In line with earlier research [@micci_barreca_2001; @prokhorenkova_2018], target encoding with regularization (glmm) performed better or equally well in comparison with the unregularized impact encoder.
-The glmm method was the overall winner based on the encoder rankings across datasets and outperformed impact encoding on some datasets.
-Our results show that combining target encoding based on glmms with 5-fold CV led to an improved predictive performance on many datasets.
-Performance often did not improve further when using 10 folds, suggesting that 5 folds might be a good default in practice.
-The glmm encoder has a clear advantage over target encoding with a smoothing hyperparameter [@micci_barreca_2001], as costly tuning the whole ML pipeline with different smoothing values is not required.
-
+When looking for a simple default encoding, indicator encoding in combination with collapsing small levels seems a robust alternative, although the glmm encoders performed better.
 We found that one-hot encoding usually gave a slightly better performance than dummy encoding, which has also been observed by @chiquet_2016 and commented on in
@@
 It sometimes leads to slightly longer runtimes (in comparison to indicator encoding), but especially for larger datasets this is often offset by the more efficient representation produced by target encoding.
+The glmm encoder has a clear advantage over target encoding with a smoothing hyperparameter [@micci_barreca_2001], as costly tuning the whole ML pipeline with different smoothing values is not required.
 

### 2021-09-13 publication/manuscript.Rmd
first pass

@@
 abstract: |
-  Because most machine learning (ML) algorithms are designed for numerical inputs, efficiently encoding categorical variables is a crucial aspect during data analysis.
-An often encountered problem are high cardinality features, i.e.
+  Since most machine learning (ML) algorithms are designed for numerical inputs, efficiently encoding categorical variables is a crucial aspect in data analysis.
+A common problem are high cardinality features, i.e.
 unordered categorical predictor variables with a high number of levels.
 We study techniques that yield numeric representations of categorical variables which can then be used in subsequent ML applications.
-We focus on the impact of those techniques on a subsequent algorithm's predictive performance, and -- if possible -- derive best practices on when to use which technique.
+We focus on the impact of these techniques on a subsequent algorithm's predictive performance, and -- if possible -- derive best practices on when to use which technique.
 We conducted a large-scale benchmark experiment, where we compared different encoding strategies together with five ML algorithms (lasso, random forest, gradient boosting, k-nearest neighbors, support vector machine) using datasets from regression, binary- and multiclass- classification settings.
-Throughout our study, regularized versions of target encoding (i.e.
+In our study, regularized versions of target encoding (i.e.
 using target predictions based on the feature levels in the training set as a new numerical feature) consistently provided the best results.
-Traditional encodings that make unreasonable assumptions to map levels to integers (e.g.
+Traditionally widely used encodings that make unreasonable assumptions to map levels to integers (e.g.
 integer encoding) or to reduce the number of levels (possibly based on target information, e.g.
-leaf encoding) before creating binary indicator variables (one-hot or dummy encoding) were not as effective.
+leaf encoding) before creating binary indicator variables (one-hot or dummy encoding) were not as effective in comparison.
 keywords:
@@
    - \usepackage{bm}
+   - \usepackage{float}
    - \usepackage{tikz}
@@
 
-<!--
-Possible titles:
-- Target encoding outperforms traditional methods in supervised machine learning with high cardinality features
-- Experimental evaluation of categorical encoding techniques
-- Encoding Categorical Variables -- An empirical study
-- A Benchmark Experiment on How to Encode High Cardinality Features in Supervised Machine Learning
-- Encoding Categorical Variables -- An empirical study
-- Your algorithm can not handle categorical variables? 10 solutions you would have never thought of.
-- You will not believe how these categorical variable encoding techniques perform in practice!
--->
-
 ```{r setup, include=FALSE}
@@
 The goal of this study is to provide an overview of existing approaches for encoding categorical predictor variables and to study their effect on a model's predictive performance.
+In order to achieve this, we carefully design a benchmark setting allowing us to discern the effect of encoding strategies (and their hyperparamters) along with their interaction with subsequently applied ML methods from an unbiased perspective.
+Since we do not introduce a new competitor method, we do not risk cherry picking \cite{dehghani2021benchmark} or reporting over-optimistic results \cite{niesl2021overoptimism}.
 
@@
 An early formal description of this strategy is @micci_barreca_2001.
-In simple target encoding for regression problems, the mean target value in the training set from all observations with a certain feature level is used as the numeric value to encode that level for all observations: $\hat{x}_l = \frac{\sum_{i:x^{train}_i = l}y^{train}_i}{N_l}$.
+In simple target encoding for regression problems, the mean target value in the training set from all observations with a certain feature level is used to encode that level for all observations: $\hat{x}_l = \frac{\sum_{i:x^{train}_i = l}y^{train}_i}{N_l}$.
 Simple target encoding often does not perform well with rare levels, where it tends to overfit to the training data and fails to generalize well for new observations.
@@
 
-## Previous Encoding Benchmarks
+## Categorical Encoding Benchmarks
 
@@
 To find default settings for high cardinality features, we analyze a variety of datasets with different characteristics including regression, binary classification, and multiclass classification problems.
-Because our methods vary in runtime and complexity, we are interested whether more complex methods are to be preferred, or if simpler approaches suffice.
+Because our methods vary in runtime and complexity, we are interested in whether more complex methods are to be preferred, or if simpler approaches suffice.
 We also study the relationship between a feature's cardinality and the choice of encoding technique by varying the minimum number of levels above which features are transformed.
@@
 We survey a broad set of categorical encoding techniques and conduct a comprehensive benchmark study with a focus on high cardinality features.
-We study $7$ different encoding techniques in conjunction with $5$ commonly used ML algorithms across $24$ datasets, both from a classification and a regression regime. 
+We carefully design a benchmark scenario as well as a preprocessing scheme allowing us to study $7$ different encoding techniques in conjunction with $5$ commonly used ML algorithms across $24$ diverse datasets, both from a classification and a regression regime.
 We give a detailed description of our study design to highlight important considerations for studying high cardinality features.
@@
 
-<!--
-how many different encoders do we count?
-7? 10? 11?
--->
-
 Encodings
@@
 The simplest strategy for categorical features is integer encoding (also called ordinal encoding).
-We randomly map the observed levels from the training set to the integers $1$ to $L$.
+Observed levels from the training set are mapped to the integers $1$ to $L$.
 Although new levels could be mapped to $L+1$ or $0$, model predictions would be arbitrary as the integer order does not carry information.
@@
 We replace new levels in the prediction phase with the most frequent level in the training set.
-In datasets with a high total number of feature levels, constructing all indicator variables can be detrimental to the predictive performance or at least increase the computational load.
-As indicator encoding is practically infeasible for high cardinality variables (a feature with $10^5$ levels would add $10^5$ new columns), we limit the number of indicator coded features by collapsing rare levels beyond a varying threshold to a single *other* category before encoding.
+As constructing all indicator variables is practically infeasible for high cardinality variables, we limit their number by collapsing rare levels beyond a varying threshold to a single *other* category before encoding.
 
@@
 This parameter could be used to further regularize towards the unconditional mean.
-We always choose a small $\epsilon = 0.0001$ as we want to compare simple target encoding with the regularized encoder introduced next.
+We choose a small $\epsilon = 0.0001$ as we want to compare simple target encoding with the regularized encoder introduced next.
 Weights of evidence encoding from the credit scoring classification literature [@hand_1997] is almost identical to impact encoding, but without regularization or centering.
@@
 The encoded value for each level is based on the spherical conditional mode estimates.
-In regression, the conditional modes are similar to the mean target value for each level, weighted by the relative observed frequency of that level in the training set [@gelman_2006].
-For technical details, see @bates_2018.
-Additionally, the estimate of the fixed intercept can be used during the prediction phase to encode new feature levels that were not observed in the training set.
+In regression, the conditional modes are similar to the mean target value for each level, weighted by the relative observed frequency of that level in the training set [@gelman_2006, @bates_2018].
+The estimate of the fixed intercept can be used during the prediction phase to encode new feature levels not observed in the training set.
 In multiclass classification, we fit $C$ one vs. rest glmms resulting in one encoded feature per class.
@@
 
-A further strategy to avoid overfitting in target encoding is to combine it with CV to train the encoder on independent observations without limiting the data to train the ML model.
+Overfitting can be further avoided through combination with cross-validation (CV) to train the encoder on independent observations without limiting the data to train the ML model.
 We provide an implementation which combines target encoding based on glmms with CV.
@@
 
-Table \ref{tab:datasets} summarizes our benchmark datasets.
+Table \ref{tab:datasets} summarizes datasets used in our benchmark.
 We specifically investigate datasets that contain categorical variables with a large number of levels, including many well-known datasets
@@
 The total number of features ranges between `r descr_dat[, min(NumFeats + BinFeats + CatFeats)]` and `r descr_dat[, max(NumFeats + BinFeats + CatFeats)]`.
-Datasets contain between `r descr_dat[, min(sapply(strsplit(HighCardLevels, " "), length))]` and `r descr_dat[, max(sapply(strsplit(HighCardLevels, " "), length))]` categorical features with more than $10$ levels, with the maximum number of levels for one feature in a dataset ranging between `r descr_dat[, min(MaxCatLvls)]` and `r descr_dat[, max(MaxCatLvls)]`.
+Datasets contain between `r descr_dat[, min(sapply(strsplit(HighCardLevels, " "), length))]` and `r descr_dat[, max(sapply(strsplit(HighCardLevels, " "), length))]` categorical features with more than $10$ levels, with the maximum number of levels for a feature ranging between `r descr_dat[, min(MaxCatLvls)]` and `r descr_dat[, max(MaxCatLvls)]`.
 Missing values are present in about half of the datasets.
 
-```{r datasets, warning=FALSE, results="asis"}
+## High Cardinality Threshold
+\label{sec:HCT}
+
+It is often assumed that advanced encoding methods are only advantageous for variables with a high number of levels, while simple indicator encoding is more appropriate for a small number of levels.
+To reflect this in our benchmark, a high cardinality threshold (HCT) parameter with values of $10$, $25$, and $125$ was introduced determining varying configurations for the different encoders.
+For indicator encoding, the $HCT - 1$ most frequent levels are encoded together with a single collapsed category for the remaining levels.
+For integer, frequency, hash, leaf, impact, and glmm encoders, only features with more than HCT levels in the training set were encoded with the respective strategy, while the remaining categorical features were one-hot encoded.
+HCT is used as the hash size in hash encoding.
+In the remove control condition, features with more than HCT levels in the training set are removed from the feature set.\footnote{Based on the number of categorical features and levels per feature, some HCT settings were removed from the benchmark for some combinations of dataset $\times$ encoder.
+This ensured that encoders always affect at least one feature and the remove condition always removes at least one feature.
+If settings where aan encoder would lead to identical encoding strategies for all features of a dataset, we only kept the condition with the smallest HCT value.}
+
+## Machine Learning Pipeline and Algorithms
+\label{sec:pipeline}
+
+In total, we investigate $5$ ML algorithms.
+We kept tuning their hyperparameters to a minimum because we were interested in the effect of the encoding techniques instead of a comparison of ML algorithms.
+LASSOs were fitted with `glmnet` [@friedman_2010], internally tuning the regularization using 5-fold CV.
+RFs with $500$ trees were trained using `ranger` [@wright_2017] without tuning, because RFs can be expected to give reasonable results with default settings [@probst_2019].
+GB models were trained using `xgboost` [@chen_2018], setting the learning rate to $0.01$ and determining the number of iterations using early stopping on a 20% holdout set.
+KNN was taken from package `kknn` [@schliep_2016] standardizing features and using a constant $k=15$ for the number of nearest neighbors, together with an information gain filter [@brown2012] to limit the number of features to $25$.
+SVMs with radial basis function kernel were trained with `liquidSVM` [@steinwart_2017].
+The bandwidth and regularization parameters were internally tuned using 5-fold CV. We used a one-vs-all approach for multiclass-settings.
+
+The ML pipeline outlined below was used for all experimental conditions.
+It was carefully designed to ensure consistent results for extreme conditions (e.g. some levels only existing in the training data).
+
+- **Imputation I**: Create a new factor level for missing values in categorical features with more than two categories.
+Impute missing values in binary features using the mode and missing values in numerical features using the mean feature value in the training data.
+- **Encoding**: Transform the complete categorical features by the respective encoder.
+  In the \textit{no encoding} condition, the encoder simply passes on its input.
+  The leaf and remove conditions still return categorical features, while the remaining encoders return only numerical features.
+Encoders only affect categorical variables above the specified HCT value.
+- **Imputation II**: To handle new levels observed during prediction, impute missing values obtained during encoding.
+- **Drop constants**: Drop features that are constant during training.
+As none of the original datasets includes constant columns, this step only removes constant features that are produced by the encoders or the CV splitting procedure.
+- **Final one-hot encoding**: Transform all remaining categorical features via one-hot encoding (skipped for no encoding condition).
+- **Learner**: Use the transformed data from each training set to fit the respective ML algorithm.
+  In the prediction phase, transformed feature values (based on the trained encoder) for new observations in each test set are fed into the trained model to compute predictions.
+
+```{r datasets, warning=FALSE, results="asis", fig.pos="h"}
 library(kableExtra)
@@
 ```
-
-## High Cardinality Threshold
-\label{sec:HCT}
-
-<!--
-A relevant question we aim to answer is what the minimum number of levels is, for which more involved encoding methods should be used. WE CANNOT REALLY ANSWER THIS WITH ONLY 3 DIFFERENT HCT VALUES...
--->
-It is often assumed that advanced encoding methods are only advantageous for variables with a high number of levels, while simple indicator encoding is more appropriate for a small number of levels.
-To make our benchmark results more realistic, a high cardinality threshold (HCT) parameter with values of $10$, $25$, and $125$ was introduced which determined varying configurations for the different encoders.
-For indicator encoding, the $HCT - 1$ most frequent levels are encoded together with a single collapsed category for the remaining levels.
-For integer, frequency, hash, leaf, impact, and glmm encoders, only features with more than HCT levels in the training set were encoded with the respective strategy, while the remaining categorical features were one-hot encoded.
-HCT is used as the hash size in hash encoding.
-In the remove control condition, features with more than HCT levels in the training set are removed from the feature set.\footnote{Based on the number of categorical features and levels per feature, some HCT settings were removed from the benchmark for some combinations of dataset $\times$ encoder.
-This made sure that encoders always affect at least one feature and that the remove condition always removes at least one feature.
-If several HCT settings of an encoder would lead to identical encoding strategies for all features of a dataset, we only kept the condition with the smallest HCT value.}
-
-## Machine Learning Pipeline and Algorithms
-\label{sec:pipeline}
-
-In total, we investigate $5$ ML algorithms.
-We kept tuning their hyperparameters to a minimum because we were interested in the effect of the encoding techniques instead of a comparison of ML algorithms.
-LASSOs were fitted with `glmnet` [@friedman_2010], internally tuning the regularization using 5-fold CV.
-RFs with $500$ trees were trained using `ranger` [@wright_2017] without tuning, because RFs can be expected to give reasonable results with default settings [@probst_2019].
-GB models were trained using `xgboost` [@chen_2018].
-We set the learning rate to $0.01$ and determined the number of iterations using early stopping on a 20% holdout set.
-KNN was taken from package `kknn` [@schliep_2016].
-We standardized all features and used a constant $k=15$ for the number of nearest neighbors, together with an information gain filter [@brown2012] to limit the number of features to $25$.
-SVMs with radial basis function kernel were trained with `liquidSVM` [@steinwart_2017].
-The bandwidth and regularization parameters were internally tuned using 5-fold CV. We used a one-vs-all approach for multiclass-settings.
-
-The ML pipeline outlined below was used for all experimental conditions.
-It was carefully designed to ensure consistent results for extreme conditions (e.g. some levels only existing in the training data).
-
-- **Imputation I**: Create a new factor level for missing values in categorical features with more than two categories.
-Impute missing values in binary features using the mode and missing values in numerical features using the mean feature value in the training data.
-- **Encoding**: Transform the complete categorical features by the respective encoder.
-  In the \textit{no encoding} condition, the encoder simply passes on its input.
-  The leaf and remove conditions still return categorical features, while the remaining encoders return only numerical features.
-Encoders only affect categorical variables above the specified HCT value.
-- **Imputation II**: To handle new levels observed during prediction, impute missing values obtained during encoding.
-- **Drop constants**: Drop features that are constant during training.
-As none of the original datasets includes constant columns, this step only removes constant features that are produced by the encoders or the CV splitting procedure.
-- **Final one-hot encoding**: Transform all remaining categorical features via one-hot encoding (skipped for no encoding condition).
-- **Learner**: Use the transformed data from each training set to fit the respective ML algorithm.
-  In the prediction phase, transformed feature values (based on the trained encoder) for new observations in each test set are fed into the trained model to compute predictions.
 
@@
 
-## Encoder Performance on Each Dataset
+## Encoder Performance
 
@@
 When looking for a simple default encoding, indicator encoding in combination with collapsing small levels seems a robust alternative, although the glmm encoders performed better.
-We found that one-hot encoding usually gave a slightly better performance than dummy encoding, which has also been observed by @chiquet_2016 and commented on in
-@tutz_2016 (p.254).
+We found that one-hot encoding usually gave a slightly better performance than dummy encoding, which has also been observed by @chiquet_2016 and @tutz_2016 (p.254).
 Our results suggest that one-hot encoding is the better standard compared to dummy encoding when applying nonlinear regularized models like RF, GB or SVM with (high cardinal) categorical features.
@@
 ### Conflicts of Interest
-None.
+The authors declare no conflict of interest.
 

### 2021-09-14 publication/manuscript.Rmd
add sentences

@@
   - name: Janek Thomas \orcidlink{0000-0003-4511-6245}
-    address: Fraunhofer Institute for Integrated Circuits, Hansastraße 32, 8068 Munich, Germany
+    address: Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstraße 33, 80539 Munich, Germany
   - name: Bernd Bischl \orcidlink{0000-0001-6002-6980}
@@
 While categorical variables with only a small number of possible levels can often be efficiently dealt with using standard techniques such as one-hot encoding, this approach becomes inefficient as the number of levels increases.
+Despite this inefficiency, simpler strategies are often favored in practice because other methods are either not known, implementations are missing or because of a lack of trust due to missing validation studies.
 Although domain knowledge can sometimes be used to reduce the number of theoretically relevant levels, finding strategies that work well on a large variety of problems is highly important for many applications as well as in automated ML [@thomas_2018; @thornton_2013; @feurer_2015].
@@
 
-- **Imputation I**: Create a new factor level for missing values in categorical features with more than two categories.
+**Imputation I**: Create a new factor level for missing values in categorical features with more than two categories.
 Impute missing values in binary features using the mode and missing values in numerical features using the mean feature value in the training data.
-- **Encoding**: Transform the complete categorical features by the respective encoder.
+
+**Encoding**: Transform the complete categorical features by the respective encoder.
   In the \textit{no encoding} condition, the encoder simply passes on its input.
@@
 Encoders only affect categorical variables above the specified HCT value.
-- **Imputation II**: To handle new levels observed during prediction, impute missing values obtained during encoding.
-- **Drop constants**: Drop features that are constant during training.
+
+**Imputation II**: To handle new levels observed during prediction, impute missing values obtained during encoding.
+
+**Drop constants**: Drop features that are constant during training.
 As none of the original datasets includes constant columns, this step only removes constant features that are produced by the encoders or the CV splitting procedure.
-- **Final one-hot encoding**: Transform all remaining categorical features via one-hot encoding (skipped for no encoding condition).
-- **Learner**: Use the transformed data from each training set to fit the respective ML algorithm.
+
+**Final one-hot encoding**: Transform all remaining
+categorical features via one-hot encoding (skipped for no encoding condition).
+
+**Learner**: Use the transformed data from each training set to fit the respective ML algorithm.
   In the prediction phase, transformed feature values (based on the trained encoder) for new observations in each test set are fed into the trained model to compute predictions.

### 2021-09-14 publication/manuscript.Rmd
adding new regression encoding benchmark; iterate on neutral benchmark study section and fix references

@@
 The goal of this study is to provide an overview of existing approaches for encoding categorical predictor variables and to study their effect on a model's predictive performance.
-In order to achieve this, we carefully design a benchmark setting allowing us to discern the effect of encoding strategies (and their hyperparamters) along with their interaction with subsequently applied ML methods from an unbiased perspective.
-Since we do not introduce a new competitor method, we do not risk cherry picking \cite{dehghani2021benchmark} or reporting over-optimistic results \cite{niesl2021overoptimism}.
+Following calls for neutral benchmark studies [@boulesteix_2017], which do not introduce a new method, thus reducing the risk of cherry picking methods [@dehghani2021benchmark] and reporting over-optimistic performance [@niesl2021overoptimism], we present a carefully designed experimental setting to discern the effect of encoding strategies and their interaction with different ML algorithms.
 
@@
 As those studies only consider gradient boosting and a limited amount of datasets, it is unclear whether results generalize to other datasets and ML algorithms.
+A recent study ($15$ datasets) found good performance of target encoding, but they only investigated categorical features in regression settings [@seca_2021]. 
 Several other publications studied encoding text data based on similarity [@cerda_2018; @cerda_2020] and employed indicator or target encoding as baselines.

### 2021-09-21 publication/manuscript.Rmd
remove performance measure formulas

@@
 Throughout our experiments, we use 5-fold CV to obtain estimates of predictive performance.
-Depending on the target variable, we report root mean squared error (RMSE) for regression, area under the curve (AUC) for binary classification, and its extension AUNU [@ferri_2009] for multiclass problems:
-\begin{align*}
-RMSE = \sqrt{\frac{1}{N}\sum_{i=1}^N(y_i-\hat{y_i})^2},\
-AUC = \frac{U}{N_{pos} \cdot N_{neg}},\
-AUNU = \frac{\sum_{c = 1}^C AUC_c}{C}
-\end{align*}
-$N$: number of observations, $N_{pos}$ and $N_{neg}$: number of true labels in both classes, $U$: test statistic of the Mann-Whitney-U test, and $AUC_c$: one vs. rest $AUC$ of class $c$.
+Depending on the target variable, we report root mean squared error (RMSE) for regression, area under the curve (AUC) for binary classification, and its extension AUNU [@ferri_2009] for multiclass problems.
 

### 2021-09-22 publication/manuscript.Rmd
move encoder algos and dataset table to the supplementary material

@@
 
-Pseudocode is presented in the main text (for non-standard encodings), or the Supplementary Material (for standard methods).
+Pseudocode for all encoders is presented in the Supplementary Material.
 An important detail is how encoding techniques treat new levels during the prediction phase.
@@
 
-\begin{algorithm}
-  \caption{Leaf Encoding}
-  \label{alg:leaf}
-  \begin{algorithmic}
-    \State \underline{Training:} require number of cross-validation folds $K \in \mathbb{N}$
-    \State fit CART tree on $\mathcal{D}^{train}$ with complexity pruning based on $K$-fold cross-validation
-      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
-        $\tilde{x}_i = t$ with $x^{train}_i$ in terminal node $t$
-      \EndFor
-    \State \underline{Prediction:}
-    \For{$x^{new}$}
-        \If{$x^{new} \in \mathcal{L}^{train}$}\ {$\tilde{x}^{new} = t$ with $x^{new}$ in terminal node $t$}
-        \Else\ {$\tilde{x}^{new} = b$ where $b$ indicates the biggest terminal node}
-        \EndIf
-    \EndFor
-  \end{algorithmic}
-\end{algorithm}
-
 ## Impact Encoding
@@
 This connection is described in @kuhn_2019.
-To achieve regularized impact encoding, we implemented glmm encoders for regression, binary, and multiclass classification which are described in Algorithm \ref{alg:glmm-regr} (regression) and in the Supplementary Material (classification).
+To achieve regularized impact encoding, we implemented glmm encoders for regression, binary, and multiclass classification.
 The encoded value for each level is based on the spherical conditional mode estimates.
@@
 In our study, we use the lmer (regression) and glmer (classification) functions from the lme4 package in R [@bates_2015] as an efficient way to fit glmms.
-\begin{algorithm}
-  \caption{GLMM Encoding Regression}
-  \label{alg:glmm-regr}
-  \begin{algorithmic}
-    \State \underline{Training:} require $n.folds \in \mathbb{N}$
-     \State fit simple random intercept model: $y^{train}_i = \beta_{0l} + \epsilon_{i} = \gamma_{oo} + u_l + \epsilon_{i}$ on $\mathcal{D}^{train}$
-     \State with $u_l \overset{iid}{\sim} N(0,\tau^2)$, $\epsilon_i \overset{iid}{\sim} N(0,\sigma^2)$ and $x^{train}_i = l$, $l \in \mathcal{L}^{train}$
-    \If{$n.folds = 1$}
-      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
-        \State $\hat{x}^{train}_i = \hat{\beta}_{0l}^{\mathcal{D}^{train}}$ with $x^{train}_i = l$
-      \EndFor
-    \Else\ {use $n.folds$ cross-validation scheme to make training sets $\mathcal{D}^{train}_1, \dots, \mathcal{D}^{train}_{n.folds}$}
-      \State and fit simple random intercept model on each $\mathcal{D}^{train}_{m}$
-      \ForAll{$x^{train}_i \in \boldsymbol{x}^{train}$}
-        \State $\hat{x}^{train}_i = \hat{\beta}_{0l}^{\mathcal{D}^{train}}$ with $x^{train}_i = l$ based on the model m
-        \State with $(x^{train}_i, y^{train}_i) \notin \mathcal{D}^{train}_m$
-      \EndFor
-    \EndIf
-    \State \underline{Prediction:}
-    \For{$x^{new}$}
-      \If{$x^{new} \in \mathcal{L}^{train}$}
-        \State $\hat{x}^{new} = \hat{\beta}_{0l}^{\mathcal{D}^{train}}$ with $x^{new} = l$ based on full model fitted on $\mathcal{D}^{train}$
-      \Else\ {$\hat{x}^{new} = \hat{\gamma}_{00}$ based on full model fitted on $\mathcal{D}^{train}$}
-      \EndIf
-    \EndFor
-  \end{algorithmic}
-\end{algorithm}
 
@@
 
-Table \ref{tab:datasets} summarizes datasets used in our benchmark.
+A table showing a detailed summary of all benchmark datasets can be found in the Supplementary Material.
 We specifically investigate datasets that contain categorical variables with a large number of levels, including many well-known datasets
@@
 
-```{r datasets, warning=FALSE, results="asis", fig.pos="h"}
+```{r datasets, warning=FALSE, results="hide", fig.pos="h"}
 library(kableExtra)
@@
 Lower ranks indicate better performance.
-The rank of the *none* control condition of the RF (rank 11) was omitted from the figure.
+The rank of the *none* control condition of the RF (rank 11 of 12) was omitted from the figure.
 

### 2021-09-22 publication/supplementary_material.Rmd
move encoder algos and dataset table to the supplementary material

@@
 
+# Benchmark Datasets
+
+```{r, echo=FALSE, results='hide', message=FALSE}
+library(data.table)
+library(stringi)
+library(kableExtra)
+descr_dat = as.data.table(readRDS("../analysis/high_cardinality_benchmark/descr_dat.rds"))
+# remove unfinished datasets
+descr_dat = descr_dat[!(Name %in% c("KDD98", "Traffic_violations", "sf-police-incidents")),]
+```
+
+```{r, eval=FALSE, echo=FALSE}
+# Collect info (do not run, just here for reproducibility)
+if (FALSE) {
+  # Collect cardinality information
+  library(OpenML)
+  setOMLConfig(arff.reader = "farff")
+  out = list()
+  for (id in descr_dat$OmlId) {
+    d = getOMLDataSet(id)
+    fcts = (sapply(d$data, class) == "factor") & (names(d$data) != d$desc$default.target.attribute)
+    out[[as.character(id)]] = lapply(d$data[fcts], table)
+    saveRDS(out, "../analysis/high_cardinality_benchmark/dataset_factor_levels.rds")
+    gc()
+  }
+
+  # Restart
+  setOMLConfig(arff.reader = "RWeka") # farff seems to break for one dataset
+  out = readRDS("../analysis/high_cardinality_benchmark/dataset_factor_levels.rds")
+  for (id in sample(setdiff(descr_dat$OmlId, as.numeric(names(out))))) {
+    d = getOMLDataSet(id)
+    fcts = (sapply(d$data, class) == "factor") & (names(d$data) != d$desc$default.target.attribute)
+    out[[as.character(id)]] = lapply(d$data[fcts], table)
+    saveRDS(out, "../analysis/high_cardinality_benchmark/dataset_factor_levels.rds")
+    gc()
+  }
+}
+```
+
+```{r, echo=FALSE}
+# Normalized Shannon entropy
+shannonEntropy <- function(counts.table, log.base = getOption("GeneFamilies.entropy.log.base",
+    base::exp(1))) {
+    if (length(counts.table) <= 1)
+        return(0)
+    c.t.s <- sum(counts.table)
+    -sum(sapply(counts.table, function(x) x/c.t.s * log(x/c.t.s, base = log.base)))/log(length(counts.table),
+        base = log.base)
+}
+out = readRDS("../analysis/high_cardinality_benchmark/dataset_factor_levels.rds")
+ents = lapply(out, function(x) lapply(x, shannonEntropy))
+ent2 = lapply(out, function(x) lapply(x, function(x) min(x)/max(x)))
+ents = data.table(OmlId = as.numeric(names(ents)), entropies = ents)
+descr_dat = merge(descr_dat, ents)
+```
+
+```{r, echo=FALSE}
+descr_dat2 = descr_dat
+descr_dat2[Classes == 0, task_type := rep("regr", .N)]
+descr_dat2[Classes == 2, task_type := rep("bin_class", .N)]
+descr_dat2[Classes > 2, task_type  := rep("multi_class", .N)]
+descr_dat2[, task_type := factor(task_type, levels = c("regr", "bin_class", "multi_class"))]
+setorder(descr_dat2, task_type, Obs)
+descr_dat2$HighCardLevels = stri_replace_all_fixed(descr_dat2$HighCardLevels, " ", "")
+descr_dat2[Name == "KDDCup09_upselling", HighCardLevels := "14,...,5073,5713,13990,15415,15415"]
+#hcl_list = lapply(descr_dat$HighCardLevels, function(x) {log(as.numeric(unlist(strsplit(x, #split = ","))))})
+#descr_dat2$logNHCL = ""
+```
+
+```{r datasets, echo=FALSE, warning=FALSE, results="asis", fig.pos="h"}
+library(kableExtra)
+landscape(
+  knitr::kable(
+    descr_dat2[, .(OmlId, Name, Cl = Classes, N = Obs, `NA%` = NAs*100/(Obs*(NumFeats + BinFeats + CatFeats)),
+                Num = as.integer(NumFeats), Bin = as.integer(BinFeats), Cat = as.integer(CatFeats), HighCardLevels, Entropy = "")],
+    digits=2L,
+    caption="Benchmark Datasets and Dataset Characteristics",
+    label="datasets"
+  ) %>%
+  column_spec(10, image = spec_boxplot(lapply(descr_dat2$entropies, unlist))) %>%
+  footnote(general = "OmlId = Id on OpenML, Name = name on OpenML, Cl = classes (0: regression), N = observations, NA% = percentage of missing values, Num = numeric features, Bin = binary features, Cat = categorical features, HighCardLevels = number of levels for each categorical feature with at least 10 levels (some levels are not displayed for KDDCup09_upselling), Entropy = box-plot of normalized Shannon-entropy across levels (smaller = larger imbalance).", threeparttable = TRUE) %>%
+  kable_styling(latex_options="scale_down")
+)
+```
+
 \newpage
@@
 
-From the results shown in Table 2 we can observe, that for most algorithms and datasets, one-hot encoding is to be preferred over dummy-encoding.
+From the results shown in Table \ref{tab:oneh-win} we can observe, that for most algorithms and datasets, one-hot encoding is to be preferred over dummy-encoding.
 

### 2022-02-02 publication/manuscript.Rmd
add references for second revision

@@
 Methods that jointly encode features and therefore leverage correlational structures between features are an interesting avenue for future research:
-Recent approaches use target-based encoding strategies to learn a vector-valued representation of a level, similar to entity embeddings [@guo_2016].
+Recent approaches use target-based encoding strategies to learn a vector-valued representation of a level, similar to neural network embeddings [@guo_2016;@rodriguez_2018].
 Those should be compared to classical approaches from the optimal scaling literature [e.g.
-@deleeuw_1976] or the subsequent *aspect* framework [@mair_2010], which are unfortunately unfamiliar to many ML researchers.
+@deleeuw_1976;@young_1976] or the subsequent *aspect* framework [@mair_2010], which are unfortunately unfamiliar to many ML researchers.
 

### 2022-02-04 publication/manuscript.Rmd
make additions for second revision

@@
 The goal of this study is to provide an overview of existing approaches for encoding categorical predictor variables and to study their effect on a model's predictive performance.
-Following calls for neutral benchmark studies [@boulesteix_2017], which do not introduce a new method, thus reducing the risk of cherry picking methods [@dehghani2021benchmark] and reporting over-optimistic performance [@niesl2021overoptimism], we present a carefully designed experimental setting to discern the effect of encoding strategies and their interaction with different ML algorithms.
+Following calls in the computational statistics community for neutral benchmark studies [@boulesteix_2017], which do not introduce a new method, thus reducing the risk of cherry picking methods [@dehghani2021benchmark] and reporting over-optimistic performance [@niesl2021overoptimism], we present a carefully designed experimental setting to discern the effect of encoding strategies and their interaction with different ML algorithms.
 
@@
 While the CatBoost method performed best, hold-out came second and target encoding without CV performed worst.
-An overview of encoding techniques tailored towards neural networks (e.g. the widely adopted *entity embeddings* by @guo_2016) is provided in @hancock.
+An overview of encoding techniques tailored towards neural networks [e.g. the widely adopted *entity embeddings* by @guo_2016] is provided in @hancock.
 They present a survey of indicator- and embedding-based methods but no benchmark study.
@@
 
-To further summarize, present meta rankings for each ML algorithm in Figure \ref{fig:rank-res}.
-First, we defined an encoder relation within each dataset, based on corrected resample t-tests [@nadeau_2003].
+To further analyze our results, we used statistical inference methods inspired by the benchmark community in computational statistics [e.g., @hothorn_2005]. 
+First, we present meta rankings for each ML algorithm in Figure \ref{fig:rank-res}:
+We defined an encoder relation within each dataset, based on corrected resample t-tests [@nadeau_2003].
 An encoding was defined to beat another encoding if the one-sided p-value of the t-test was $< .05$.
@@
 
-In an exploratory analysis, we tried to find clusters of datasets based on systematic patterns of encoder performance (independent of the employed ML algorithm).
+In a second exploratory analysis, we tried to find clusters of datasets based on systematic patterns of encoder performance (independent of the employed ML algorithm).
 We computed a partial-order consensus relation across ML algorithms for each dataset and hierarchically clustered the dataset consensus relations using the symmetric difference distance in combination with the complete linkage agglomeration method.
@@
 Because we are interested in model-agnostic methods that can be combined with any supervised ML algorithm, we did not investigate recent model-specific strategies [@guo_2016; @prokhorenkova_2018].
+We did not include deep neural networks in our benchmark because they are still rarely applied in the statistical learning community and more dominant in computer science. 
 We do not differentiate between nominal and ordinal categorical features:
@@
 Most effective, with a consistently superior performance across Ml algorithms and datasets, was a target encoder which combines simple generalized linear mixed models with cross-validation and does not require hyperparameter tuning.
+GLMMs are a major workhorse in applied statistics but not well understood and often neglected by the ML community.
+Refining target encoders that use statistical models for more efficient regularization and studying their theoretical properties could be a valuable research topic for the computational statistics community.
 

### 2022-02-04 publication/supplementary_material.Rmd
add author and journal information to supplementary materials document

@@
   | Regularized target encoding outperforms traditional methods in supervised machine learning with high cardinality features
+author:
+  - Florian Pargent
+  - Florian Pfisterer
+  - Janek Thomas
+  - Bernd Bischl
+subtitle: Published in Computational Statistics
 numbersections: true
@@
 We provide an overview of existing implementations for R in Table \ref{tab:impls}.
-For python, an extension to scikit-learn [@pedregosa_2011], *category encoders* [@will_mcginnis_2018_1157110] is available.
+For python, an extension to `scikit-learn` [@pedregosa_2011], *category encoders* [@will_mcginnis_2018_1157110] is available.
 To enable a fair and reliable comparison in our study, we implemented all methods outlined above on top of the `mlrCPO` package.
@@
 
-*Note on GLMM Encoders:* To speed up the computation of the GLMM encoders, we followed the performance tips from the lme4 vignette (<https://cran.r-project.org/web/packages/lme4/vignettes/lmerperf.html>): we did not compute derivations (`calc.derivs = FALSE`) and used the `NLOPT_LN_BOBYQA` optimizer from the nloptr package with liberal stopping criteria (`maxeval = 1000`, `xtol_abs = 1e-6`, `ftol_abs = 1e-6`).
+*Note on GLMM Encoders:* To speed up the computation of the GLMM encoders, we followed the performance tips from the `lme4` vignette (<https://cran.r-project.org/web/packages/lme4/vignettes/lmerperf.html>): we did not compute derivations (`calc.derivs = FALSE`) and used the `NLOPT_LN_BOBYQA` optimizer from the `nloptr` package with liberal stopping criteria (`maxeval = 1000`, `xtol_abs = 1e-6`, `ftol_abs = 1e-6`).
 
@@
 
-A frequently asked question is, whether one-hot encoding or dummy encoding is to be preferred.
+A frequently asked question is whether one-hot encoding or dummy encoding is to be preferred.
 While dummy encoding clearly is commonly preferred for non-regularized linear models, the answer is less clear for general machine learning algorithms studied in our setting.
@@
 
-From the results shown in Table \ref{tab:oneh-win} we can observe, that for most algorithms and datasets, one-hot encoding is to be preferred over dummy-encoding.
+From the results shown in Table \ref{tab:oneh-win} we can observe that for most algorithms and datasets, one-hot encoding is to be preferred over dummy-encoding.
 

### 2022-02-04 publication/supplementary_material.Rmd
update link to github repository

@@
 
-In addition to the datasets mentioned in the manuscript, the benchmark results available in our online repository (<https://github.com/compstat-lmu/paper_2021_categorical_feature_encodings>) contain three additional datasets: *KDD98* (OmlId: 41435), *sf-police-incidents* (OmlId: 41436), *Traffic_violations* (OmlId: 41443).
+In addition to the datasets mentioned in the manuscript, the benchmark results available in our online repository (<https://github.com/slds-lmu/paper_2021_categorical_feature_encodings>) contain three additional datasets: *KDD98* (OmlId: 41435), *sf-police-incidents* (OmlId: 41436), *Traffic_violations* (OmlId: 41443).
 A substantive amount of conditions failed for those rather big datasets due to memory problems, which is why we remove them from the results discussed in our manuscript.

### 2022-02-04 publication/manuscript.Rmd
fix typo and update github link

@@
 The pre-processing, as well as the final ML algorithm described in \ref{sec:pipeline}, were trained and resampled using the `mlr` framework [@bischl_2016] together with the `batchtools` package [@lang_2017] to scale the benchmark analysis to HPC compute infrastructure.
-All materials for this study, including reproducible code for this manuscript and result objects can be downloaded from our **online repository** \footnote{\url{https://github.com/compstat-lmu/paper_2021_categorical_feature_encodings}}.
+All materials for this study, including reproducible code for this manuscript and result objects can be downloaded from our **online repository** \footnote{\url{https://github.com/slds-lmu/paper_2021_categorical_feature_encodings}}.
 
@@
 Regularized versions of target encoding, which uses predictions of the target variable as numeric feature values, performed better than traditional strategies like integer or indicator encoding.
-Most effective, with a consistently superior performance across Ml algorithms and datasets, was a target encoder which combines simple generalized linear mixed models with cross-validation and does not require hyperparameter tuning.
+Most effective, with a consistently superior performance across ML algorithms and datasets, was a target encoder which combines simple generalized linear mixed models with cross-validation and does not require hyperparameter tuning.
 GLMMs are a major workhorse in applied statistics but not well understood and often neglected by the ML community.
@@
 ### Code Availability
-All analysis code is publicly available at \url{https://github.com/compstat-lmu/paper_2021_categorical_feature_encodings}.
+All analysis code is publicly available at \url{https://github.com/slds-lmu/paper_2021_categorical_feature_encodings}.
 

### 2022-02-07 publication/manuscript.Rmd
prepare all files for LATEX upload of second revision to Computational Statistics

@@
     base_format: rticles::springer_article
+  keep_md: true
 #bibstyle: spphys
@@
 set.seed(42)
-knitr::opts_chunk$set(cache.extra = knitr::rand_seed, echo = FALSE)
+knitr::opts_chunk$set(cache.extra = knitr::rand_seed, echo = FALSE, fig.path = "", dev=c('pdf', 'tiff'))
 options(knitr.kable.NA = " ")
 # options(papaja.na_string = "")
+options(tinytex.clean = FALSE)
 ```

### 2022-02-07 publication/manuscript.tex
prepare all files for LATEX upload of second revision to Computational Statistics

% !TeX program = pdfLaTeX
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}

\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}

%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%

%% load any required packages here

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{algorithm}
\usepackage{array}
\usepackage[noend]{algpseudocode}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{float}
\usepackage{tikz}
\usepackage{threeparttable}
\usetikzlibrary{trees,arrows, chains, fit, positioning, calc, shapes, shadows}
\setlength\parindent{0pt}
\usepackage{pdflscape}
\usepackage[T1]{fontenc}
\usepackage{orcidlink}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\begin{document}

\title{Regularized target encoding outperforms traditional methods in supervised machine learning with high cardinality features }

    \titlerunning{Encoding high cardinality features in supervised ML}

\author{  Florian Pargent \orcidlink{0000-0002-2388-553X} \and  Florian Pfisterer \orcidlink{0000-0001-8867-762X} \and  Janek Thomas \orcidlink{0000-0003-4511-6245} \and  Bernd Bischl \orcidlink{0000-0001-6002-6980} \and  }

\institute{
        Florian Pargent \orcidlink{0000-0002-2388-553X} \at
     Department of Psychology, Psychological Methods and Assessment, LMU Munich, Leopoldstra{\ss}e 13, 80802 Munich, Germany \\
     \email{\href{mailto:florian.pargent@psy.lmu.de}{\nolinkurl{florian.pargent@psy.lmu.de}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Florian Pfisterer \orcidlink{0000-0001-8867-762X} \at
     Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstra{\ss}e 33, 80539 Munich, Germany \\
     %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Janek Thomas \orcidlink{0000-0003-4511-6245} \at
     Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstra{\ss}e 33, 80539 Munich, Germany \\
     %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Bernd Bischl \orcidlink{0000-0001-6002-6980} \at
     Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstra{\ss}e 33, 80539 Munich, Germany \\
     %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}
Since most machine learning (ML) algorithms are designed for numerical inputs, efficiently encoding categorical variables is a crucial aspect in data analysis. A common problem are high cardinality features, i.e.~unordered categorical predictor variables with a high number of levels. We study techniques that yield numeric representations of categorical variables which can then be used in subsequent ML applications. We focus on the impact of these techniques on a subsequent algorithm's predictive performance, and -- if possible -- derive best practices on when to use which technique. We conducted a large-scale benchmark experiment, where we compared different encoding strategies together with five ML algorithms (lasso, random forest, gradient boosting, k-nearest neighbors, support vector machine) using datasets from regression, binary- and multiclass- classification settings. In our study, regularized versions of target encoding (i.e.~using target predictions based on the feature levels in the training set as a new numerical feature) consistently provided the best results. Traditionally widely used encodings that make unreasonable assumptions to map levels to integers (e.g.~integer encoding) or to reduce the number of levels (possibly based on target information, e.g.~leaf encoding) before creating binary indicator variables (one-hot or dummy encoding) were not as effective in comparison.
\\
\keywords{
        supervised machine learning \and
        benchmark \and
        high-cardinality categorical features \and
        target encoding \and
        dummy encoding \and
        generalized linear mixed models \and
    }

\end{abstract}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

While increasing sample size is usually considered the most important step to improve the predictive performance of a machine learning (ML) model, using effective feature engineering comes as a close second. One remaining challenge is how to handle high cardinality features -- categorical predictor variables with a high number of different levels but without any natural ordering.
While categorical variables with only a small number of possible levels can often be efficiently dealt with using standard techniques such as one-hot encoding, this approach becomes inefficient as the number of levels increases.
Despite this inefficiency, simpler strategies are often favored in practice because other methods are either not known, implementations are missing or because of a lack of trust due to missing validation studies.
Although domain knowledge can sometimes be used to reduce the number of theoretically relevant levels, finding strategies that work well on a large variety of problems is highly important for many applications as well as in automated ML (Feurer et al., 2015; Thomas et al., 2018; Thornton et al., 2013). Optimally, strategies should be model-agnostic because benchmarking encoding methods together with ML algorithms from different classes is often necessary for applications.
While a variety of strategies exist, there are very few benchmarks that can be used to decide which technique is expected to yield good predictive performance.
Furthermore, there has recently been increasing attention on scientific benchmark studies that compare different methods to provide a clearer picture in light of a large number of methods available to practitioners (Bommert et al., 2020; Fernández-Delgado et al., 2014), as they can provide at least partial answers to such questions.
The goal of this study is to provide an overview of existing approaches for encoding categorical predictor variables and to study their effect on a model's predictive performance.
Following calls in the computational statistics community for neutral benchmark studies (Boulesteix et al., 2017), which do not introduce a new method, thus reducing the risk of cherry picking methods (Dehghani et al., 2021) and reporting over-optimistic performance (Nie{\ss}l et al., 2021), we present a carefully designed experimental setting to discern the effect of encoding strategies and their interaction with different ML algorithms.

\hypertarget{notation}{%
\subsection{Notation}\label{notation}}

We consider the classical setting of supervised learning from an \(i.i.d.\) tabular dataset \(\mathcal{D}\) of size \(N\) sampled from a joint distribution \(\mathcal{P}(\bm{x}, y)\) of a set of features \(\bm{x}\) and an associated target variable \(y\).
Here, \(\bm{x}\) consists of a mix of numeric (real-valued or integer-valued) features and categorical features, the latter of which we seek to transform feature-wise to numeric features using a categorical encoding technique.
Let \(x\) be a single unordered categorical feature from a feature space \(\mathcal{X}\) with cardinality \(card(\mathcal{X}) \leq card(\mathbb{N})\).
It holds either \(y \in \mathbb{R}\) (regression), \(y \in \mathcal{C}\) from a finite class space \(\mathcal{C} = \{c_1, ..., c_C\}\) with \(C = 2\) (binary classification) or \(C > 2\) (multiclass classification). We always assume to observe all \(C\) classes in our training sample, however we might only observe a subset \(\mathcal{L}^{train} \subseteq \mathcal{X}\) of a feature's available \(L\) levels, \(\mathcal{L}^{train} = \{l_1, ..., l_L\}\) for categorical features. We denote the observed frequency of class \(c\) in the training set with \(N_c\) and the observed frequency of a level \(l\) in the training set with \(N_l\).
We investigate categorical encoding techniques to transform each nominal feature \(x^{train}\) into numerical features \(\hat{x}^{train}\) which are then used for training. If clear from the context, we use \(\hat{x}_l\) as the encoded value for an observation with level \(l\). Although datasets might contain multiple high cardinality features, we encode each feature separately but with the same strategy.

\hypertarget{related-work}{%
\subsection{Related Work}\label{related-work}}

We broadly categorize feature encoding techniques into \emph{target-agnostic} methods and \emph{target-based} methods (Micci-Barreca, 2001).
Figure \ref{fig:overview} contains a taxonomy of our considered encoding methods. \emph{Target-agnostic} methods do not rely on any information about the target variable and can therefore also be used in unsupervised settings.
Simple strategies from this domain e.g.~one-hot or dummy encoding are widely used -- in the scientific literature (Hancock and Khoshgoftaar, 2020; Kuhn and Johnson, 2019) but also on Kaggle\footnote{\url{https://www.kaggle.com}} to embed variables for classical ML algorithms as well as (deep) neural networks.
Such indicator methods map each level of a categorical variable to a set of dichotomous features encoding the presence or absence of a particular level.
An obvious drawback of indicator encoding is that it adds one additional feature per level of a categorical variable.
When indicator encoding leads to an unreasonable number of features, levels are often mapped to integer values with random order (integer encoding). Alternatively, the ``hashing trick'' (Weinberger et al., 2009) can be used to randomly collapse feature levels into a smaller number of indicator variables (Kuhn and Johnson, 2019), or levels can be encoded by using the observed frequency of a given level in the dataset (frequency encoding).

\tikzset{
  basic/.style  = {draw, text width=4cm, drop shadow, font=\sffamily \scriptsize, rectangle},
  root/.style   = {basic, rounded corners=2pt, thin, align=center, fill=white},
  level-2/.style = {basic, rounded corners=5pt, thin,align=center, fill=white, text width=2.3cm},
  level-3/.style = {basic, thin, align=center, fill=white, text width=1.8cm},
  level-3-focus/.style = {basic, thick, align=center, fill=white, text width=1.8cm}
}
\begin{figure}[ht]
  \centering
  \begin{tikzpicture}[
    level 1/.style={sibling distance=8em, level distance=3em},
    edge from parent/.style={->,solid,black,thick,draw},
    edge from parent path={(\tikzparentnode.south) -- (\tikzchildnode.north)},
    >=latex, node distance=0.8cm, edge from parent fork down]
    \node[root] {\textbf{Taxonomy of categorical encoding techniques}}
      child {node[level-2, xshift=-10pt] (c1) {\textbf{Target-agnostic}}}
      child {node[level-2, xshift=10pt] (c2) {\textbf{Target-based}}};

    \begin{scope}[every node/.style={level-3}]
      \node [below of = c1, xshift=5pt] (c11) {indicator: dummy, one-hot};
      \node [below of = c11] (c12) {integer};
      \node [below of = c12] (c13) {frequency};
      \node [below of = c13] (c14) {hash};

      \node [below of = c2, xshift=5pt](c21) {leaf};
      \node [below of = c21] (c22) {impact};
      \node [below of = c22] (c23) {regularized impact};
    \end{scope}

    \foreach \value in {1,2,3,4}
      \draw[->] (c1.192) |- (c1\value.west);

    \foreach \value in {1,...,3}
      \draw[->] (c2.192) |- (c2\value.west);
  \end{tikzpicture}
  \caption{Taxonomy of common categorical variable encoding techniques.}
  \label{fig:overview}
\end{figure}

\emph{Target-based} methods try to incorporate information about the target values associated with a given level.
Early strategies aimed to reduce the number of levels by methods like hierarchical clustering or decision trees based on statistics of the target variable, although this has been rarely described in the scientific literature (for a brief mention, see Micci-Barreca, 2001).
The basic idea of more advanced methods called target, impact, mean, or likelihood encoding is to use the training set to make a simple prediction of the target for each level of the categorical feature, and to use the prediction as the numerical feature value \(\hat{x}_l\) for the respective level.
An early formal description of this strategy is Micci-Barreca (2001).
In simple target encoding for regression problems, the mean target value in the training set from all observations with a certain feature level is used to encode that level for all observations: \(\hat{x}_l = \frac{\sum_{i:x^{train}_i = l}y^{train}_i}{N_l}\).
Simple target encoding often does not perform well with rare levels, where it tends to overfit to the training data and fails to generalize well for new observations.
In the extreme case of a categorical feature with unique values (e.g.~some hashed ID variable) studied in Prokhorenkova et al. (2018), the mean target for each level of this feature in simple target encoding is similar to the true target value of a single observation.
Based on the encoded feature, all observations can be predicted perfectly in the training set, even if the original variable did not contain any useful information.
ML models would place a high priority on such an encoded feature during training but would perform badly on test data.
To avoid this, practitioners often use regularized target encoding with a smoothing parameter that shrinks those effects towards the global mean (Micci-Barreca, 2001).
An alternative strategy is to combine target encoding with cross-validation (CV) techniques (Prokhorenkova et al., 2018).

\hypertarget{categorical-encoding-benchmarks}{%
\subsection{Categorical Encoding Benchmarks}\label{categorical-encoding-benchmarks}}

Several small-scale studies have been previously conducted.
Those works did not yield conclusive results due to narrower scopes or not considering high cardinality variables.
One benchmark (\(6\) datasets) on encoding high cardinality features (cardinality between \(103\) and \(9095\)) in combination with gradient boosting has been published on the Kaggle forums (Prokopev, 2018).
Different versions of target encoding are compared with indicator, integer, and frequency encoding.
They recommend combining the smoothed version of target encoding with 4- or 5-fold CV, never using simple target encoding and using indicator encoding only for small datasets.
Interestingly, frequency encoding did perform well in many cases.
Coors (2018) performed a benchmark (\(12\) datasets) on encoding high cardinality features (maximum number of levels per dataset between 10 and 25847) when developing the automatic gradient boosting (autoxgboost) library (Thomas et al., 2018). They compared different variants of target encoding with integer and indicator encoding. In their benchmarks, target encoding only improved over target-agnostic methods on \(2\) datasets, while it led to worse results on 4 datasets. For a smaller number of levels, indicator and integer encoding yielded similar results.
As those studies only consider gradient boosting and a limited amount of datasets, it is unclear whether results generalize to other datasets and ML algorithms.
A recent study (\(15\) datasets) found good performance of target encoding, but they only investigated categorical features in regression settings (Seca and Mendes-Moreira, 2021).
Several other publications studied encoding text data based on similarity (Cerda et al., 2018; Cerda and Varoquaux, 2020) and employed indicator or target encoding as baselines.
Another line of work studies variable encodings employed within specific ML models.
Wright and König (2019) study treatments for categorical variables in random forests together with dummy and integer encoding (18 datasets, cardinality between \(3\) and \(38\)), concluding that indicator and integer encoding perform subpar in comparison to methods that re-order levels according to the target variable.
Prokhorenkova et al. (2018) compared their new CatBoost variant of target encoding to smoothed target encoding without CV, hold-out, and leave-one-out CV on \(8\) datasets.
While the CatBoost method performed best, hold-out came second and target encoding without CV performed worst.
An overview of encoding techniques tailored towards neural networks (e.g.~the widely adopted \emph{entity embeddings} by Guo and Berkhahn, 2016) is provided in Hancock and Khoshgoftaar (2020).
They present a survey of indicator- and embedding-based methods but no benchmark study.
In contrast, our work is the first to focus on high cardinality variables and techniques that are agnostic to the subsequent ML method.
We study this problem on a larger variety of datasets and settings.
\newline

The main goal of our study is to assess the impact different categorical encoding techniques have on subsequent models' predictive performance.
As the optimal encoding might differ depending on the ML algorithm, we consider various state-of-the-art algorithms, \emph{regularized linear models} (LASSO), \emph{random forests} (RF), \emph{gradient tree boosting} (GB), \emph{k-nearest neighbors} (KNN), and \emph{support vector machines} (SVM).
To find default settings for high cardinality features, we analyze a variety of datasets with different characteristics including regression, binary classification, and multiclass classification problems.
Because our methods vary in runtime and complexity, we are interested in whether more complex methods are to be preferred, or if simpler approaches suffice.
We also study the relationship between a feature's cardinality and the choice of encoding technique by varying the minimum number of levels above which features are transformed.

\hypertarget{contributions}{%
\subsection{Contributions}\label{contributions}}

We survey a broad set of categorical encoding techniques and conduct a comprehensive benchmark study with a focus on high cardinality features.
We carefully design a benchmark scenario as well as a preprocessing scheme allowing us to study \(7\) different encoding techniques in conjunction with \(5\) commonly used ML algorithms across \(24\) diverse datasets, both from a classification and a regression regime.
We give a detailed description of our study design to highlight important considerations for studying high cardinality features.
Our results provide an overview of the performance of various approaches heavily used in the literature.
After a discussion of results concerning predictive performance, we provide further analyses seeking to inform practitioners which methods to apply.
This includes an important discussion of runtimes.

\hypertarget{encodings}{%
\section{Encodings}\label{encodings}}

\label{sec:methods}

Pseudocode for all encoders is presented in the Supplementary Material.
An important detail is how encoding techniques treat new levels during the prediction phase.

\hypertarget{integer-encoding}{%
\subsection{Integer Encoding}\label{integer-encoding}}

The simplest strategy for categorical features is integer encoding (also called ordinal encoding).
Observed levels from the training set are mapped to the integers \(1\) to \(L\).
Although new levels could be mapped to \(L+1\) or \(0\), model predictions would be arbitrary as the integer order does not carry information.
Thus, we encode new levels as missing values and use mode imputation to obtain the integer which matches the most frequent level in the training set.
Integer encoding should only be an acceptable strategy for tree-based models, which can separate all original levels with repeated splits.

\hypertarget{frequency-encoding}{%
\subsection{Frequency Encoding}\label{frequency-encoding}}

Frequency encoding maps each level to its observed frequency in the training set (\(\hat{x}_l = N_l\)).
This assumes a functional relationship between the frequency of a level and the target.
It implicitly reduces the number of levels, and the subsequent model can best differentiate between levels with dissimilar frequencies.
This approach is heavily used in natural language processing to encode token or n-gram counts.
We encode new levels with a frequency of \(1\).

\hypertarget{indicator-encoding}{%
\subsection{Indicator Encoding}\label{indicator-encoding}}

We use indicator encoding as an umbrella term for two common strategies to encode categorical features with a small to moderate number of levels: one-hot and dummy encoding.
One-hot encoding transforms the original feature into \(L\) binary indicator columns, each representing one original level.
An observation is coded with \(1\) for the indicator column representing its level (\(x^{train}_i = l\)) and \(0\) for all other indicators.
Dummy encoding results in only \(L-1\) indicator columns.
A reference feature level is chosen that is encoded with \(0\) in all indicator columns.
For one-hot encoding, the zero vector can be used to encode new levels which were not observed during training.
For dummy encoding, it is not useful to collapse new levels to the often arbitrary reference category; in our case the first level in alphabetical order.
We replace new levels in the prediction phase with the most frequent level in the training set.
As constructing all indicator variables is practically infeasible for high cardinality variables, we limit their number by collapsing rare levels beyond a varying threshold to a single \emph{other} category before encoding.

\hypertarget{hash-encoding}{%
\subsection{Hash Encoding}\label{hash-encoding}}

Hash Encoding can be used to compute indicator variables based on a hash function (Weinberger et al., 2009).
The basic idea is to transform each feature level \(l\) into an integer \(hash(l) \in \mathbb{N}\), based on its label.
This integer is then transformed into an indicator representation, with \(1\) in indicator column number \((hash(l) \mod hash.size) + 1\) and \(0\) in all remaining columns (Kuhn and Johnson, 2019).
Some levels are hashed to the same indicator representation.
The smaller the \(hash.size\), the higher the number of collapsed levels.
The number of indicators is often effectively lower than \(hash.size\), as some indicators can be constant in the training set (we remove those columns for both training and prediction).
Although we could jointly hash multiple features, we hash each feature separately to improve comparability with the other encoders.

\hypertarget{leaf-encoding}{%
\subsection{Leaf Encoding}\label{leaf-encoding}}

Leaf encoding fits a decision tree on the training set to predict the target based on the categorical feature.
Each level is encoded by the number of the terminal node, in which an observation with the respective level ends up.
In that way, leaf encoding combines feature levels with similar target values.
We use the rpart package in R (Therneau and Atkinson, 2018) which grows CARTs with categorical feature support that can be pruned based on internal performance estimates from 10-fold CV.
Thus, our leaf encoder automatically uses an ``optimal'' number of new levels.
To speed up the computation for multiclass classification, our implementation uses the ordering approach presented in Wright and König (2019).
New levels are encoded with the arbitrary number of the terminal node with most observations during training.
Encoded values are treated as a new categorical feature and encoded by one-hot encoding.
Our leaf encoder can be thought of as a simplification of the approach suggested by Grąbczewski and Jankowski (2003).

\hypertarget{impact-encoding}{%
\subsection{Impact Encoding}\label{impact-encoding}}

An early formal description of a so-called impact, target or James-Stein encoder was provided by Micci-Barreca (2001).
The basic idea is to encode each feature level with the conditional target mean (regression) or the conditional relative frequency of one or more target classes (classification).
The impact encoder for classification uses a logit link and transforms the original feature into \(C\) numeric features, each representing one target class.
A smoothing parameter \(\epsilon\) is introduced to avoid division by zero.
This parameter could be used to further regularize towards the unconditional mean.
We choose a small \(\epsilon = 0.0001\) as we want to compare simple target encoding with the regularized encoder introduced next.
Weights of evidence encoding from the credit scoring classification literature (Hand and Henley, 1997) is almost identical to impact encoding, but without regularization or centering.

\hypertarget{glmm-encoding}{%
\subsection{GLMM Encoding}\label{glmm-encoding}}

Smoothed target encoding (Micci-Barreca, 2001) can be interpreted as a simple (generalized) linear mixed model (glmm) in which the target is predicted by a random intercept for each feature level in addition to a fixed global intercept.
This connection is described in Kuhn and Johnson (2019).
To achieve regularized impact encoding, we implemented glmm encoders for regression, binary, and multiclass classification.
The encoded value for each level is based on the spherical conditional mode estimates.
In regression, the conditional modes are similar to the mean target value for each level, weighted by the relative observed frequency of that level in the training set Bates (2020).
The estimate of the fixed intercept can be used during the prediction phase to encode new feature levels not observed in the training set.
In multiclass classification, we fit \(C\) one vs.~rest glmms resulting in one encoded feature per class.
An important advantage of using a glmm over impact encoding with a smoothing parameter is that a reasonable amount of regularization is determined automatically and tuning the complete ML pipeline is not necessary.

Overfitting can be further avoided through combination with cross-validation (CV) to train the encoder on independent observations without limiting the data to train the ML model.
We provide an implementation which combines target encoding based on glmms with CV.
During the training phase, we partition the data using CV into \(n.folds\) and fit a glmm on each resulting training set.
For each observation, there is exactly one glmm that did not use that observation for model fitting and can be safely used for encoding.
Note that the \(n.folds\) CV models (for \(n.folds\) \textgreater{} 1) are only used during the training phase.
In the prediction phase, feature values are always encoded by a single glmm fitted to the complete training set.
We study this method in three different settings: without CV (noCV), with \(5-\) (5CV) and with \(10-\) fold CV (10CV).
In our study, we use the lmer (regression) and glmer (classification) functions from the lme4 package in R (Bates et al., 2015) as an efficient way to fit glmms.

\hypertarget{control-conditions}{%
\subsection{Control Conditions}\label{control-conditions}}

We include three control conditions to better understand the effectiveness of the investigated encoders:
The performance of a featureless learner (\textbf{FL} condition) was estimated as a conservative baseline for each dataset.
In regression problems, FL predicts the mean of the target variable in the training set for each observation in the test set.
In classification problems, the most frequent class of the target within the training set is predicted.
For each dataset, we also consider a RF without encoding (\textbf{none} condition), to compare the use of encoding methods with a natural categorical splitting approach.
The ranger (Wright and Ziegler, 2017) implementation provides efficient categorical feature support by ordering levels once before starting the tree growing algorithm (Wright and König, 2019).
In the \textbf{remove} high cardinality features control condition, we omit features with a high number of levels above some threshold and use one-hot encoding (without collapsing rare levels) for the remaining features.
This condition reflects on whether including high cardinality features does indeed improve predictive performance.
Otherwise, the best encoding might just provide the least impairment compared to not including any high cardinality features.
We include an overview of available implementations in widely used ML frameworks for R and python in the Supplementary Material.

\hypertarget{benchmark-setup}{%
\section{Benchmark Setup}\label{benchmark-setup}}

\hypertarget{datasets}{%
\subsection{Datasets}\label{datasets}}

A table showing a detailed summary of all benchmark datasets can be found in the Supplementary Material.
We specifically investigate datasets that contain categorical variables with a large number of levels, including many well-known datasets
from previous studies (Cerda et al., 2018; Coors, 2018; Kuhn and Johnson, 2019; Prokhorenkova et al., 2018).
All datasets can be downloaded from the \texttt{OpenML} platform (Vanschoren et al., 2013) based on the name or the displayed OmlId.
The datasets include 8 regression, 10 binary classification, and 6 multiclass classification problems (between 3 and 12 classes).
To assess the imbalance in categorical variables we computed the normalized entropy for each categorical variable.
The maximum normalized entropy is \(1\), which corresponds to a uniform distribution, while a lower number indicates a larger imbalance.
Sample sizes range between 736 and 1224158 observations.
The total number of features ranges between 5 and 208.
Datasets contain between 1 and 20 categorical features with more than \(10\) levels, with the maximum number of levels for a feature ranging between 14 and 30114. Missing values are present in about half of the datasets.

\hypertarget{high-cardinality-threshold}{%
\subsection{High Cardinality Threshold}\label{high-cardinality-threshold}}

\label{sec:HCT}

It is often assumed that advanced encoding methods are only advantageous for variables with a high number of levels, while simple indicator encoding is more appropriate for a small number of levels.
To reflect this in our benchmark, a high cardinality threshold (HCT) parameter with values of \(10\), \(25\), and \(125\) was introduced determining varying configurations for the different encoders.
For indicator encoding, the \(HCT - 1\) most frequent levels are encoded together with a single collapsed category for the remaining levels.
For integer, frequency, hash, leaf, impact, and glmm encoders, only features with more than HCT levels in the training set were encoded with the respective strategy, while the remaining categorical features were one-hot encoded. HCT is used as the hash size in hash encoding.
In the remove control condition, features with more than HCT levels in the training set are removed from the feature set.\footnote{Based on the number of categorical features and levels per feature, some HCT settings were removed from the benchmark for some combinations of dataset $\times$ encoder.
This ensured that encoders always affect at least one feature and the remove condition always removes at least one feature.
If settings where aan encoder would lead to identical encoding strategies for all features of a dataset, we only kept the condition with the smallest HCT value.}

\hypertarget{machine-learning-pipeline-and-algorithms}{%
\subsection{Machine Learning Pipeline and Algorithms}\label{machine-learning-pipeline-and-algorithms}}

\label{sec:pipeline}

In total, we investigate \(5\) ML algorithms. We kept tuning their hyperparameters to a minimum because we were interested in the effect of the encoding techniques instead of a comparison of ML algorithms.
LASSOs were fitted with \texttt{glmnet} (Friedman et al., 2010), internally tuning the regularization using 5-fold CV.
RFs with \(500\) trees were trained using \texttt{ranger} (Wright and Ziegler, 2017) without tuning, because RFs can be expected to give reasonable results with default settings (Probst et al., 2019).
GB models were trained using \texttt{xgboost} (Chen et al., 2018), setting the learning rate to \(0.01\) and determining the number of iterations using early stopping on a 20\% holdout set.
KNN was taken from package \texttt{kknn} (Schliep and Hechenbichler, 2016) standardizing features and using a constant \(k=15\) for the number of nearest neighbors, together with an information gain filter (Brown et al., 2012) to limit the number of features to \(25\).
SVMs with radial basis function kernel were trained with \texttt{liquidSVM} (Steinwart and Thomann, 2017).
The bandwidth and regularization parameters were internally tuned using 5-fold CV. We used a one-vs-all approach for multiclass-settings.

The ML pipeline outlined below was used for all experimental conditions.
It was carefully designed to ensure consistent results for extreme conditions (e.g.~some levels only existing in the training data).

\textbf{Imputation I}: Create a new factor level for missing values in categorical features with more than two categories. Impute missing values in binary features using the mode and missing values in numerical features using the mean feature value in the training data.

\textbf{Encoding}: Transform the complete categorical features by the respective encoder.
In the \textit{no encoding} condition, the encoder simply passes on its input.
The leaf and remove conditions still return categorical features, while the remaining encoders return only numerical features. Encoders only affect categorical variables above the specified HCT value.

\textbf{Imputation II}: To handle new levels observed during prediction, impute missing values obtained during encoding.

\textbf{Drop constants}: Drop features that are constant during training. As none of the original datasets includes constant columns, this step only removes constant features that are produced by the encoders or the CV splitting procedure.

\textbf{Final one-hot encoding}: Transform all remaining
categorical features via one-hot encoding (skipped for no encoding condition).

\textbf{Learner}: Use the transformed data from each training set to fit the respective ML algorithm.
In the prediction phase, transformed feature values (based on the trained encoder) for new observations in each test set are fed into the trained model to compute predictions.

\hypertarget{performance-evaluation}{%
\subsection{Performance Evaluation}\label{performance-evaluation}}

We perform all analyses in the open-source statistical software R (R Core Team, 2021).
To enable a fair and reliable comparison in our study, we implemented all encoding methods on top of the \texttt{mlrCPO} package (Binder, 2018).
The pre-processing, as well as the final ML algorithm described in \ref{sec:pipeline}, were trained and resampled using the \texttt{mlr} framework (Bischl et al., 2016) together with the \texttt{batchtools} package (Lang et al., 2017) to scale the benchmark analysis to HPC compute infrastructure.
All materials for this study, including reproducible code for this manuscript and result objects can be downloaded from our \textbf{online repository} \footnote{\url{https://github.com/slds-lmu/paper_2021_categorical_feature_encodings}}.

Throughout our experiments, we use 5-fold CV to obtain estimates of predictive performance.
Depending on the target variable, we report root mean squared error (RMSE) for regression, area under the curve (AUC) for binary classification, and its extension AUNU (Ferri et al., 2009) for multiclass problems.

In our benchmark study, each encoding method listed in Section \ref{sec:methods} is combined with all ML algorithms (c.f. Section \ref{sec:pipeline}) across
high cardinality thresholds (HCT) \(10\), \(25\), \(125\).
While we only report results for the best HCT setting for each ML algorithm \(\times\) dataset combination in Section \ref{sec:results}, we study the effect of the HCT parameter in more detail in section \ref{sec:hct_study}.

\hypertarget{benchmark-results}{%
\section{Benchmark Results}\label{benchmark-results}}

\label{sec:results}

Our main question is which encoding methods generally work well across various datasets.
We report results for regression and classification datasets separately, as the associated metrics differ in scale.

For 6 datasets, some conditions with the SVM led to unexpected crashes due to memory problems or numerical errors.
We completely removed those datasets for the SVM when computing ranks or other statistics that compare encodings across datasets.

\hypertarget{encoder-performance}{%
\subsection{Encoder Performance}\label{encoder-performance}}

\begin{figure}[!h]
\includegraphics[width=\textwidth]{regr-res-1} \caption{Performance estimates from 5-CV for regression (mean, min, max). For each combination, only the best HCT condition is displayed. Note the reversed y-axis to ease visual interpretation.}\label{fig:regr-res}
\end{figure}

\begin{figure}[!h]
\includegraphics[width=\textwidth]{bincl-res-1} \caption{Performance estimates from 5-CV for binary classification (mean, min and max). For each combination, only the best HCT condition is displayed.}\label{fig:bincl-res}
\end{figure}

\begin{figure}[!h]
\includegraphics[width=\textwidth]{multcl-res-1} \caption{Performance estimates from 5-CV for multiclass classification (mean, min and max). For each combination, only the best HCT condition is displayed.}\label{fig:multcl-res}
\end{figure}

Mean performance estimates along with minimum and maximum performance are reported for all datasets in Figures \ref{fig:regr-res} (regression), \ref{fig:bincl-res} (binary classification), and \ref{fig:multcl-res} (multiclass classification).
To reduce the complexity induced by the hyperparameter HCT we only display the parameter condition with the best performance for each combination of dataset \(\times\) encoding \(\times\) ML algorithm.
The y-axis differs for all datasets and is reversed for the RMSE for better visual comparison.
For some datasets, the remove condition performed very similar to the other encodings (e.g.~\emph{ames-housing}, \emph{porto-seguro}), suggesting that categorical features were not informative.
Performance in some CV folds was below the FL learner for \emph{flight-delay-usa-dec-2017} (\(RMSE_{FL} =\) 48.8052624) and \emph{nyc-taxi-green-dec-2016} (\(RMSE_{FL} =\) 2.2152442).

On datasets with substantive performance differences, target encoding with the glmm encoder was generally \textbf{most} effective.
The \textbf{worst} encoder differed by ML algorithm and dataset.
For datasets \emph{Click\_prediction\_small}, \emph{KDDCup09\_upselling}, \emph{kick}, and \emph{okcupid-stem} some encoder \(\times\) ML algorithm conditions performed worse than simply removing high cardinality features.

\hypertarget{meta-rankings-and-dataset-clustering}{%
\subsection{Meta Rankings and Dataset Clustering}\label{meta-rankings-and-dataset-clustering}}

To further analyze our results, we used statistical inference methods inspired by the benchmark community in computational statistics (e.g., Hothorn et al., 2005).
First, we present meta rankings for each ML algorithm in Figure \ref{fig:rank-res}:
We defined an encoder relation within each dataset, based on corrected resample t-tests (Nadeau and Bengio, 2003).
An encoding was defined to beat another encoding if the one-sided p-value of the t-test was \(< .05\).
This allowed us to compute a weak-order consensus ranking \(R\) defined by the optimization problem:

\[ \underset{R \in \mathcal{C}}{arg\ min} \ \sum_{b=1}^B d(R_b, R)\]

where \(d\) is the symmetric difference distance and \(R_b\) is the relation for dataset \(b\) (Hornik and Meyer, 2007; Meyer and Hornik, 2018).
The symmetric difference between two relations is the number of cases one encoding beats another encoding in one relation but not in the other one.

\begin{figure}
\includegraphics[width=\textwidth,height=.55\textwidth]{rank-res-1} \caption{Consensus rankings across all datasets for each algorithm. Lower ranks indicate better performance. The rank of the \emph{none} control condition of the RF (rank 11 of 12) was omitted from the figure.}\label{fig:rank-res}
\end{figure}

Although the presented solutions of the optimization problem are not unique, rankings were highly stable for the high and low ranks.
Meta rankings seem to be highly consistent with the individual patterns of encoder performances reflected in Figures \ref{fig:regr-res} to \ref{fig:multcl-res}.
Looking at meta-rankings, approaches based on GLMM's in combination with cross-validation outperform all other approaches across all algorithms.
A further interesting detail omitted in Figure \ref{fig:rank-res} for clarity is that the \textit{none} encoding strategy for the RF was beaten by all strategies except for the \textit{remove} condition.
This implies, that even if the algorithm provides a mechanic for treating categorical variables, it might often be optimal to use a different strategy instead.

In a second exploratory analysis, we tried to find clusters of datasets based on systematic patterns of encoder performance (independent of the employed ML algorithm).
We computed a partial-order consensus relation across ML algorithms for each dataset and hierarchically clustered the dataset consensus relations using the symmetric difference distance in combination with the complete linkage agglomeration method.
The resulting dendrogram is displayed in Figure \ref{fig:dendrogram}.
Although the cluster structure is somewhat ambiguous, roughly three clusters can be described:
The first 9 datasets from the top of the dendrogram are characterized by a medium to high number of levels, low performance of the remove condition (indicating the importance of high cardinality features) and clear performance advantage of target encoders.
For the next 13 datasets which contain the smallest number of levels, traditional encodings can compete with target encoding.
This biggest cluster also includes 9 datasets with zero distances, in which no significant performance differences could be observed between any encoding conditions (nor with the remove condition, indicating that high cardinality features are less informative).
The last two datasets with the highest number of levels formed a separate cluster, in which target encoding without strong regularization (impact, glmm-noCV) showed severe overfitting.
Note that clusters were not determined by problem type, again suggesting that encoder rankings are somewhat similar for regression and classification settings.\footnote{We tried to corroborate this observation by comparing meta-rankings between problem types. Unfortunately, problem type specific consensus relations did not converge (probably due to the reduced number of datasets) and were uninterpretable.}

\begin{figure}[!h]
\includegraphics[width=\textwidth]{dendrogram-1} \caption{Hierarchical cluster analysis of benchmark datasets. The symmetric difference distance between two datasets reflects differences in performance patterns between encodings (independent of the employed ML algorithm).}\label{fig:dendrogram}
\end{figure}

\hypertarget{summary-of-encoder-performance}{%
\subsection{Summary of Encoder Performance}\label{summary-of-encoder-performance}}

Regularized target encoding was superior or at least competitive on all datasets.
We could not observe a setting in which regularized target encoding was convincingly beaten by target agnostic methods.
Especially effective was glmm encoding with 5-fold-CV, which ranked first place for all ML algorithms except SVM.
Performance often did not improve with glmm-10CV, suggesting that 5 folds might be a good regularization default in practice.
In line with earlier research (Micci-Barreca, 2001; Prokhorenkova et al., 2018), target encoding with regularization (glmm) performed better or equally well in comparison with the unregularized impact encoder.
When glmm performed better, impact encoding not only performed worse than glmm with CV but was sometimes also inferior compared to other encoders.

The following observations were also interesting:
Surprisingly, integer encoding did not perform well with GB and target-based encoders (especially the glmm encoder) seem to be preferable.
For LASSO, previous studies have suggested that indicator encoding works well, even with a very high number of levels (Cerda et al., 2018).
Although the glmm encoders ranked first in our benchmark for the LASSO, it was the only algorithm where the indicator encoders achieved the next best ranking.
Note that for computational reasons we limited the maximum amount of indicator variables per feature to 125 in our experimental design.
The HCT = 125 setting performed best for LASSO with indicator encoding in a large number of datasets, indicating that performance might have further improved with higher values.
Both KNN and SVM rely on numerical distances in feature space and tend to perform poorly in the presence of high dimensionality.
Thus, we expected that target encoding should work well here as it transforms categories into a single, smooth numerical feature.
This was backed by our benchmark results.
For some datasets (\emph{medical\_charges}, \emph{road-safety-drivers-sex}, and \emph{Midwest\_survey}) KNN (without tuning of the optimal number of nearest neighbors) could compete with the more sophisticated ML algorithms when combined with target encoding, but performed poorly with other encoders.
Although consistent with the big picture, SVM results have to be considered with care, as some experimental conditions resulted in unexpected computational errors.
In RF, a widely used strategy to deal with categorical features is to order levels by average target statistics for a given level.
For a small number of levels, this approach has been reported superior to indicator and integer encoding (Wright and König, 2019), while we observed poor performance for datasets with a larger number of levels.
When looking for a simple default encoding, indicator encoding in combination with collapsing small levels seems a robust alternative, although the glmm encoders performed better.
We found that one-hot encoding usually gave a slightly better performance than dummy encoding, which has also been observed by Chiquet et al. (2016) and Tutz and Gertheiss (2016) (p.254).
Our results suggest that one-hot encoding is the better standard compared to dummy encoding when applying nonlinear regularized models like RF, GB or SVM with (high cardinal) categorical features.
We provide a more detailed comparison for indicator encoding in the Supplementary Material.

\hypertarget{runtime-analysis}{%
\subsection{Runtime Analysis}\label{runtime-analysis}}

\begin{table}

\caption{\label{tab:time-ranks}Proportional Increase in Runtime Compared to One-Hot Encoding}
\centering
\begin{threeparttable}
\begin{tabular}[t]{l|l|l|l|l|l}
\hline
Encoding & LASSO & RF & GB & KNN & SVM\\
\hline
integer & $0.4_{0}^{1.4}$ & $0.59_{0}^{1.1}$ & $0.67_{0}^{0.9}$ & $0.85_{0.4}^{1.5}$ & $1.01_{0.6}^{1.6}$\\
\hline
frequency & $0.34_{0}^{2.1}$ & $0.54_{0}^{1.7}$ & $0.54_{0}^{1.5}$ & $0.79_{0.2}^{1.1}$ & $1.11_{0.5}^{1.7}$\\
\hline
dummy & $1.13_{0.8}^{2.8}$ & $0.91_{0.5}^{1.7}$ & $0.97_{0.4}^{5.7}$ & $1.05_{0.6}^{1.6}$ & $1.02_{0.7}^{1.5}$\\
\hline
one-hot & $1_{1}^{1}$ & $1_{1}^{1}$ & $1_{1}^{1}$ & $1_{1}^{1}$ & $1_{1}^{1}$\\
\hline
hash & $0.87_{0.2}^{2.6}$ & $1.03_{0.4}^{3.3}$ & $0.98_{0.4}^{7.3}$ & $0.93_{0.5}^{2.2}$ & $1.16_{0.5}^{2}$\\
\hline
leaf & $0.46_{0}^{1.7}$ & $0.67_{0}^{2.3}$ & $0.85_{0}^{9.1}$ & $0.97_{0.5}^{2.2}$ & $1.01_{0.6}^{1.9}$\\
\hline
impact & $0.5_{0.1}^{1.9}$ & $0.6_{0.1}^{1.5}$ & $0.82_{0}^{120.5}$ & $1.11_{0.3}^{24}$ & $1.06_{0.7}^{1.8}$\\
\hline
glmm-noCV & $0.58_{0}^{2.8}$ & $0.57_{0.1}^{3.8}$ & $1.65_{0}^{12}$ & $1.09_{0.3}^{10.3}$ & $1.1_{0.7}^{2.1}$\\
\hline
glmm-5CV & $0.83_{0.1}^{2.2}$ & $0.66_{0.1}^{15.3}$ & $7.21_{0}^{69.3}$ & $2.77_{0.7}^{50.1}$ & $1.41_{0.6}^{4.6}$\\
\hline
glmm-10CV & $1.07_{0.1}^{4.2}$ & $0.96_{0.1}^{28.4}$ & $12.71_{0}^{127.6}$ & $5.37_{0.8}^{97.9}$ & $1.55_{0.5}^{4.6}$\\
\hline
none &  & $0.2_{0}^{1.4}$ &  &  & \\
\hline
remove & $0.4_{0}^{1.7}$ & $0.48_{0}^{1.1}$ & $0.58_{0}^{1.3}$ & $0.82_{0.2}^{1.2}$ & $1.09_{0.6}^{2}$\\
\hline
\end{tabular}
\begin{tablenotes}
\item \textit{Note: } 
\item Median $_{min}^{max}$ across datasets of the proportional increase in runtime from 5-CV, when comparing the respective encoder with one-hot encoding. Only the best HCT conditions are reported.
\end{tablenotes}
\end{threeparttable}
\end{table}

To determine whether traditional encodings are preferred when facing limited computational resources, we further analyzed runtimes of the whole analysis pipeline for different encoders and ML algorithms.
Aggregated results are shown in Table \ref{tab:time-ranks}.
To enable a meaningful comparison, we report runtime as the fraction of a full pipeline's strategy compared to the one-hot encoding condition and then further aggregate across datasets using the median.
Again, we only report the \textbf{best} HCT setting.
Absolute runtimes are hard to interpret and aggregate because runtime distributions across datasets are heavily skewed as proportionally larger runtimes are observed for big datasets.
While a pipeline's runtime can be dominated by the encoder for small datasets, the training of the consecutive ML algorithm is dominating for large datasets, which might render differences during encoding irrelevant.
Therefore we aim to report what is important in practice, the time differences for training the \textbf{full pipeline}.
The results clearly show that regularized target encoding does not consistently yield slower runtimes compared to simple strategies like indicator encoding. Supposedly, the more efficient representations produced by target encoding lead to faster runtimes of subsequent ML algorithms.
This suggests that a possible runtime vs.~predictive performance trade-off might also be in favour of target encoding, especially for large datasets where a high number of indicator variables increases computational load.
We did observe a substantial increase in runtimes when using regularized glmm with GB (with little tuning), where runtimes are relatively short in comparison to the time required to fit categorical encoders.
In other settings (LASSO, RF), the glmm encoders have been observed to be even faster than indicator encoding.

\hypertarget{analysing-high-cardinality-thresholds}{%
\subsection{Analysing High Cardinality Thresholds}\label{analysing-high-cardinality-thresholds}}

\label{sec:hct_study}

Until now, we ignored the HCT parameter by reporting only the condition with the best performance.
An interesting question is whether target encoding is only useful for features with a very high number of levels or also for fewer levels, where most practitioners would routinely use indicator encoding.
We tested this using HCT thresholds of \(10\), \(25\) and \(125\), but the results are not easy to interpret.
In general, the optimal threshold seemed to strongly depend on the dataset, but we also observed some weak patterns:
LASSO improved for larger HCT, indicating that its internal regularization can efficiently deal with the sparseness induced by indicator encoding.
In comparison, other methods generally yielded better performance if features above the very low HCT of \(10\) were encoded using one of the target encoding strategies.
Differences in regularization for the target based encoders seemed not to prefer different HCT values.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

In our benchmark, we compared encoding strategies for high cardinality categorical features on a variety of regression, binary, and multiclass classification datasets with different ML algorithms.
Regularized target encoding was superior across most datasets and ML algorithms.
Although the performance of other encoding strategies was comparable in some conditions, target encoding was never outperformed.
In general, our results suggest that regularized target encoding based on glmms with 5-fold CV (glmm-5CV) works well for all kinds of algorithms and should be a reasonable default strategy.
It sometimes leads to slightly longer runtimes (in comparison to indicator encoding), but especially for larger datasets this is often offset by the more efficient representation produced by target encoding.
The glmm encoder has a clear advantage over target encoding with a smoothing hyperparameter (Micci-Barreca, 2001), as costly tuning the whole ML pipeline with different smoothing values is not required.

What constitutes a ``high'' cardinality problem is a difficult question that might not only depend on the number of levels but also on other characteristics of the dataset and its features.
Supposedly, the number of datasets in our study was too small to discover consistent patterns between encoder performance and dataset characteristics.
Target encoding features with only 10 levels seemed to be effective in a substantive number of conditions, but for other datasets, higher HCT values performed better.
Thus, some form of hyperparameter tuning seems necessary to decide the level threshold for target encoding at this point, as no suitable defaults seem to be available.
Note that we compared different HCT values but then used the same encoding strategy for all affected features alongside one-hot encoding for the remaining ones.
A further improvement could be to decide whether to use target encoding on a feature by feature basis.
ML pipelines could introduce a categorical hyperparameter for each feature that represents which encoding is used.
To make this complicated meta optimization problem feasible, only a small number of encoders can be included.
Our study can help to decide which encoders could be safely omitted from consideration.

\hypertarget{limitations}{%
\subsection{Limitations}\label{limitations}}

Several design decisions were necessary to make answering our research questions computationally feasible:
We used minimal tuning for our ML algorithms, as we were not interested in comparing their performance against each other.
This probably led to suboptimal performance for the GB, KNN, and SVM learners.
When interpreting our results, we assume that the encoder rankings are comparable when more extensive tuning is used.
This is plausible because the meta rankings between algorithms were highly stable.
We only used 5-fold-CV without repetitions to estimate predictive performance, but due to a small variance between folds, we could confidently detect performance differences of encoders for many datasets.
Because we are interested in model-agnostic methods that can be combined with any supervised ML algorithm, we did not investigate recent model-specific strategies (Guo and Berkhahn, 2016; Prokhorenkova et al., 2018).
We did not include deep neural networks in our benchmark because they are still rarely applied in the statistical learning community and more dominant in computer science.
We do not differentiate between nominal and ordinal categorical features:
Most publicly available datasets simply do not contain any high cardinal ordinal features.
For many applications, ordinal feature information is not available as metadata which makes it less relevant for some applications like automatic ML.
Thus, our benchmark does not include specific ordinal encoding methods like Helmert and Polynomial contrasts (Chambers and Hastie, 1992) or ordinal penalization methods (Tutz and Gertheiss, 2016).
We only investigate traditional categorical variables and do not extend our analysis to multi-categorical variables or text-strings (Cerda et al., 2018; Cerda and Varoquaux, 2020), where other encoding techniques can harvest additional information.
We only investigate univariate encodings, i.e.~we always encode each variable separately.
Levels with comparable main effects can not be distinguished based on the transformed feature, which prevents the consecutive ML algorithm from learning interactions for specific levels.
Methods that jointly encode features and therefore leverage correlational structures between features are an interesting avenue for future research:
Recent approaches use target-based encoding strategies to learn a vector-valued representation of a level, similar to neural network embeddings (Guo and Berkhahn, 2016; Rodríguez et al., 2018).
Those should be compared to classical approaches from the optimal scaling literature (e.g. De Leeuw et al., 1976; Young et al., 1976) or the subsequent \emph{aspect} framework (Mair and Leeuw, 2010), which are unfortunately unfamiliar to many ML researchers.

\hypertarget{conclusion}{%
\subsection{Conclusion}\label{conclusion}}

This benchmark study compared the predictive performance of a variety of strategies to encode categorical features with a high number of unordered levels for different supervised ML algorithms.
Regularized versions of target encoding, which uses predictions of the target variable as numeric feature values, performed better than traditional strategies like integer or indicator encoding.
Most effective, with a consistently superior performance across ML algorithms and datasets, was a target encoder which combines simple generalized linear mixed models with cross-validation and does not require hyperparameter tuning.
GLMMs are a major workhorse in applied statistics but not well understood and often neglected by the ML community.
Refining target encoders that use statistical models for more efficient regularization and studying their theoretical properties could be a valuable research topic for the computational statistics community.

\hypertarget{declarations}{%
\subsection{Declarations}\label{declarations}}

\hypertarget{funding}{%
\subsubsection{Funding}\label{funding}}

This work has been funded by the German Federal Ministry of Education and Research (BMBF) under Grant No.~01IS18036A and by the Bavarian Ministry of Economic Affairs, Regional Development and Energy through the Center for Analytics -- Data -- Applications (ADA-Center) within the framework of „BAYERN DIGITAL II`` (20-3410-2-9-8).

\hypertarget{conflicts-of-interest}{%
\subsubsection{Conflicts of Interest}\label{conflicts-of-interest}}

The authors declare no conflict of interest.

\hypertarget{code-availability}{%
\subsubsection{Code Availability}\label{code-availability}}

All analysis code is publicly available at \url{https://github.com/slds-lmu/paper_2021_categorical_feature_encodings}.

\hypertarget{availability-of-data}{%
\subsubsection{Availability of Data}\label{availability-of-data}}

All benchmark datasets are publicly available at \url{https://www.openml.org/}.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\raggedright

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-bates_2018}{}}%
Bates, D., 2020. Computational methods for mixed models. Vignette for lme4.

\leavevmode\vadjust pre{\hypertarget{ref-bates_2015}{}}%
Bates, D., Mächler, M., Bolker, B., Walker, S., 2015. Fitting linear mixed-effects models using {lme4}. Journal of Statistical Software 67, 1--48. doi:\href{https://doi.org/10.18637/jss.v067.i01}{10.18637/jss.v067.i01}

\leavevmode\vadjust pre{\hypertarget{ref-binder_2018}{}}%
Binder, M., 2018. mlrCPO: Composable preprocessing operators and pipelines for machine learning.

\leavevmode\vadjust pre{\hypertarget{ref-bischl_2016}{}}%
Bischl, B., Lang, M., Kotthoff, L., Schiffner, J., Richter, J., Studerus, E., Casalicchio, G., Jones, Z.M., 2016. {mlr}: Machine learning in r. Journal of Machine Learning Research 17, 1--5.

\leavevmode\vadjust pre{\hypertarget{ref-bommert}{}}%
Bommert, A., Sun, X., Bischl, B., Rahnenführer, J., Lang, M., 2020. Benchmark for filter methods for feature selection in high-dimensional classification data. Computational Statistics \& Data Analysis 143. doi:\href{https://doi.org/10.1016/j.csda.2019.106839}{10.1016/j.csda.2019.106839}

\leavevmode\vadjust pre{\hypertarget{ref-boulesteix_2017}{}}%
Boulesteix, A.-L., Binder, H., Abrahamowicz, M., Sauerbrei, W., others, 2017. On the necessity and design of studies comparing statistical methods. Biometrical journal. Biometrische Zeitschrift 60, 216--218. doi:\href{https://doi.org/10.1002/bimj.201700129}{10.1002/bimj.201700129}

\leavevmode\vadjust pre{\hypertarget{ref-brown2012}{}}%
Brown, G., Pocock, A., Ming-Jie, Z., Luján, M., 2012. Conditional likelihood maximisation: A unifying framework for information theoretic feature selection. Journal of Machine Learning Research 13, 27--66.

\leavevmode\vadjust pre{\hypertarget{ref-cerda_2020}{}}%
Cerda, P., Varoquaux, G., 2020. Encoding high-cardinality string categorical variables. IEEE Transactions on Knowledge and Data Engineering 1--1. doi:\href{https://doi.org/10.1109/TKDE.2020.2992529}{10.1109/TKDE.2020.2992529}

\leavevmode\vadjust pre{\hypertarget{ref-cerda_2018}{}}%
Cerda, P., Varoquaux, G., Kégl, B., 2018. Similarity encoding for learning with dirty categorical variables. Machine Learning 107, 1477--1494. doi:\href{https://doi.org/10.1007/s10994-018-5724-2}{10.1007/s10994-018-5724-2}

\leavevmode\vadjust pre{\hypertarget{ref-chambers_1992}{}}%
Chambers, J., Hastie, T., 1992. Statistical models. Chapter 2 of statistical models in s. Wadsworth \& Brooks/Cole.

\leavevmode\vadjust pre{\hypertarget{ref-chen_2018}{}}%
Chen, T., He, T., Benesty, M., Khotilovich, V., Tang, Y., Cho, H., Chen, K., Mitchell, R., Cano, I., Zhou, T., Li, M., Xie, J., Lin, M., Geng, Y., Li, Y., 2018. Xgboost: Extreme gradient boosting.

\leavevmode\vadjust pre{\hypertarget{ref-chiquet_2016}{}}%
Chiquet, J., Grandvalet, Y., Rigaill, G., 2016. On coding effects in regularized categorical regression. Statistical Modelling 16, 228--237. doi:\href{https://doi.org/10.1177/1471082X16644998}{10.1177/1471082X16644998}

\leavevmode\vadjust pre{\hypertarget{ref-coors_2018}{}}%
Coors, S., 2018. Automatic gradient boosting (Master's thesis). LMU Munich.

\leavevmode\vadjust pre{\hypertarget{ref-deleeuw_1976}{}}%
De Leeuw, J., Young, F.W., Takane, Y., 1976. Additive structure in qualitative data: An alternating least squares method with optimal scaling features. Psychometrika 41, 471--503.

\leavevmode\vadjust pre{\hypertarget{ref-dehghani2021benchmark}{}}%
Dehghani, M., Tay, Y., Gritsenko, A.A., Zhao, Z., Houlsby, N., Diaz, F., Metzler, D., Vinyals, O., 2021. The benchmark lottery.

\leavevmode\vadjust pre{\hypertarget{ref-delgado}{}}%
Fernández-Delgado, M., Cernadas, E., Barro, S., Amorim, D., 2014. Do we need hundreds of classifiers to solve real world classification problems? Journal of Machine Learning Research 15, 3133--3181.

\leavevmode\vadjust pre{\hypertarget{ref-ferri_2009}{}}%
Ferri, C., Hernández-Orallo, J., Modroiu, R., 2009. An experimental comparison of performance measures for classification. Pattern Recognition Letters 30, 27--38. doi:\href{https://doi.org/10.1016/j.patrec.2008.08.010}{10.1016/j.patrec.2008.08.010}

\leavevmode\vadjust pre{\hypertarget{ref-feurer_2015}{}}%
Feurer, M., Klein, A., Eggensperger, K., Springenberg, J., Blum, M., Hutter, F., 2015. Efficient and robust automated machine learning, in: Cortes, C., Lawrence, N.D., Lee, D.D., Sugiyama, M., Garnett, R. (Eds.), Advances in Neural Information Processing Systems 28. Curran Associates, Inc., pp. 2962--2970.

\leavevmode\vadjust pre{\hypertarget{ref-friedman_2010}{}}%
Friedman, J., Hastie, T., Tibshirani, R., 2010. Regularization paths for generalized linear models via coordinate descent. Journal of Statistical Software 33, 1--22. doi:\href{https://doi.org/10.18637/jss.v033.i01}{10.18637/jss.v033.i01}

\leavevmode\vadjust pre{\hypertarget{ref-gelman_2006}{}}%
Gelman, A., Hill, J., 2006. Data analysis using regression and multilevel/hierarchical models. Cambridge university press.

\leavevmode\vadjust pre{\hypertarget{ref-grabczewski_2003}{}}%
Grąbczewski, K., Jankowski, N., 2003. Transformations of symbolic data for continuous data oriented models, in: Kaynak, O., Alpaydin, E., Oja, E., Xu, L. (Eds.), Artificial Neural Networks and Neural Information Processing --- ICANN/ICONIP 2003. Springer Berlin Heidelberg, Berlin, Heidelberg, pp. 359--366.

\leavevmode\vadjust pre{\hypertarget{ref-guo_2016}{}}%
Guo, C., Berkhahn, F., 2016. Entity embeddings of categorical variables. arXiv preprint arXiv:1604.06737.

\leavevmode\vadjust pre{\hypertarget{ref-hancock}{}}%
Hancock, J.T., Khoshgoftaar, T.M., 2020. Survey on categorical data for neural networks. Journal of Big Data 7, 1--41. doi:\href{https://doi.org/10.1186/s40537-020-00305-w}{10.1186/s40537-020-00305-w}

\leavevmode\vadjust pre{\hypertarget{ref-hand_1997}{}}%
Hand, D.J., Henley, W.E., 1997. Statistical classification methods in consumer credit scoring: A review. Journal of the Royal Statistical Society: Series A (Statistics in Society) 160, 523--541. doi:\href{https://doi.org/10.1111/j.1467-985X.1997.00078.x}{10.1111/j.1467-985X.1997.00078.x}

\leavevmode\vadjust pre{\hypertarget{ref-hornik_2007}{}}%
Hornik, K., Meyer, D., 2007. Deriving consensus rankings from benchmarking experiments, in: Advances in Data Analysis. Springer, pp. 163--170. doi:\href{https://doi.org/10.1007/978-3-540-70981-7_19}{10.1007/978-3-540-70981-7\_19}

\leavevmode\vadjust pre{\hypertarget{ref-hothorn_2005}{}}%
Hothorn, T., Leisch, F., Zeileis, A., Hornik, K., 2005. The design and analysis of benchmark experiments. Journal of Computational and Graphical Statistics 14, 675--699. doi:\href{https://doi.org/10.1198/106186005X59630}{10.1198/106186005X59630}

\leavevmode\vadjust pre{\hypertarget{ref-kuhn_2019}{}}%
Kuhn, M., Johnson, K., 2019. Feature engineering and selection: A practical approach for predictive models. Chapman; Hall/CRC.

\leavevmode\vadjust pre{\hypertarget{ref-lang_2017}{}}%
Lang, M., Bischl, B., Surmann, D., 2017. Batchtools: Tools for r to work on batch systems. The Journal of Open Source Software 2. doi:\href{https://doi.org/10.21105/joss.00135}{10.21105/joss.00135}

\leavevmode\vadjust pre{\hypertarget{ref-mair_2010}{}}%
Mair, P., Leeuw, J. de, 2010. A general framework for multivariate analysis with optimal scaling: The r package aspect. Journal of Statistical Software 32, 1--23. doi:\href{https://doi.org/10.18637/jss.v032.i09}{10.18637/jss.v032.i09}

\leavevmode\vadjust pre{\hypertarget{ref-meyer_2018}{}}%
Meyer, D., Hornik, K., 2018. Relations: Data structures and algorithms for relations.

\leavevmode\vadjust pre{\hypertarget{ref-micci_barreca_2001}{}}%
Micci-Barreca, D., 2001. A preprocessing scheme for high-cardinality categorical attributes in classification and prediction problems. SIGKDD Explor. Newsl. 3, 27--32. doi:\href{https://doi.org/10.1145/507533.507538}{10.1145/507533.507538}

\leavevmode\vadjust pre{\hypertarget{ref-nadeau_2003}{}}%
Nadeau, C., Bengio, Y., 2003. Inference for the generalization error. Machine Learning 52, 239--281. doi:\href{https://doi.org/10.1023/A:1024068626366}{10.1023/A:1024068626366}

\leavevmode\vadjust pre{\hypertarget{ref-niesl2021overoptimism}{}}%
Nie{\ss}l, C., Herrmann, M., Wiedemann, C., Casalicchio, G., Boulesteix, A.-L., 2021. Over-optimism in benchmark studies and the multiplicity of design and analysis options when interpreting their results.

\leavevmode\vadjust pre{\hypertarget{ref-probst_2019}{}}%
Probst, P., Wright, M.N., Boulesteix, A.-L., 2019. Hyperparameters and tuning strategies for random forest. Wiley Interdisciplinary Reviews: Data Mining and Knowledge Discovery 0, e1301. doi:\href{https://doi.org/10.1002/widm.1301}{10.1002/widm.1301}

\leavevmode\vadjust pre{\hypertarget{ref-prokhorenkova_2018}{}}%
Prokhorenkova, L., Gusev, G., Vorobev, A., Dorogush, A.V., Gulin, A., 2018. CatBoost: Unbiased boosting with categorical features, in: Bengio, S., Wallach, H., Larochelle, H., Grauman, K., Cesa-Bianchi, N., Garnett, R. (Eds.), Advances in Neural Information Processing Systems 31. Curran Associates, Inc., pp. 6638--6648.

\leavevmode\vadjust pre{\hypertarget{ref-prokopev_2018}{}}%
Prokopev, V., 2018. Mean (likelihood) encodings: A comprehensive study. Kaggle Forums.

\leavevmode\vadjust pre{\hypertarget{ref-r_2021}{}}%
R Core Team, 2021. R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria.

\leavevmode\vadjust pre{\hypertarget{ref-rodriguez_2018}{}}%
Rodríguez, P., Bautista, M.A., Gonzàlez, J., Escalera, S., 2018. Beyond one-hot encoding: Lower dimensional target embedding. Image and Vision Computing 75, 21--31. doi:\url{https://doi.org/10.1016/j.imavis.2018.04.004}

\leavevmode\vadjust pre{\hypertarget{ref-schliep_2016}{}}%
Schliep, K., Hechenbichler, K., 2016. Kknn: Weighted k-nearest neighbors.

\leavevmode\vadjust pre{\hypertarget{ref-seca_2021}{}}%
Seca, D., Mendes-Moreira, J., 2021. Benchmark of encoders of nominal features for regression, in: Rocha, Á., Adeli, H., Dzemyda, G., Moreira, F., Ramalho Correia, A.M. (Eds.), Trends and Applications in Information Systems and Technologies. Springer International Publishing, Cham, pp. 146--155.

\leavevmode\vadjust pre{\hypertarget{ref-steinwart_2017}{}}%
Steinwart, I., Thomann, P., 2017. {liquidSVM}: A fast and versatile SVM package. {ArXiv e-prints 1702.06899}.

\leavevmode\vadjust pre{\hypertarget{ref-therneau_2018}{}}%
Therneau, T., Atkinson, B., 2018. Rpart: Recursive partitioning and regression trees.

\leavevmode\vadjust pre{\hypertarget{ref-thomas_2018}{}}%
Thomas, J., Coors, S., Bischl, B., 2018. Automatic gradient boosting.

\leavevmode\vadjust pre{\hypertarget{ref-thornton_2013}{}}%
Thornton, C., Hutter, F., Hoos, H.H., Leyton-Brown, K., 2013. Auto-WEKA: Combined selection and hyperparameter optimization of classification algorithms, in: Proceedings of the 19th ACM SIGKDD International Conference on Knowledge Discovery and Data Mining, KDD '13. ACM, New York, NY, USA, pp. 847--855. doi:\href{https://doi.org/10.1145/2487575.2487629}{10.1145/2487575.2487629}

\leavevmode\vadjust pre{\hypertarget{ref-tutz_2016}{}}%
Tutz, G., Gertheiss, J., 2016. Rejoinder: Regularized regression for categorical data. Statistical Modelling 16, 249--260. doi:\href{https://doi.org/10.1177/1471082X16652780}{10.1177/1471082X16652780}

\leavevmode\vadjust pre{\hypertarget{ref-vanschoren_2013}{}}%
Vanschoren, J., N. van Rijn, J., Bischl, B., Torgo, L., 2013. {OpenML}: Networked science in machine learning. SIGKDD Explorations 15, 49--60. doi:\href{https://doi.org/10.1145/2641190.2641198}{10.1145/2641190.2641198}

\leavevmode\vadjust pre{\hypertarget{ref-weinberger_2009}{}}%
Weinberger, K.Q., Dasgupta, A., Langford, J., Smola, A.J., Attenberg, J., 2009. Feature hashing for large scale multitask learning, in: ICML.

\leavevmode\vadjust pre{\hypertarget{ref-wright_2019}{}}%
Wright, M.N., König, I.R., 2019. Splitting on categorical predictors in random forests. PeerJ 7. doi:\href{https://doi.org/10.7717/peerj.6339}{10.7717/peerj.6339}

\leavevmode\vadjust pre{\hypertarget{ref-wright_2017}{}}%
Wright, M.N., Ziegler, A., 2017. {ranger}: A fast implementation of random forests for high dimensional data in {C++} and {R}. Journal of Statistical Software 77, 1--17. doi:\href{https://doi.org/10.18637/jss.v077.i01}{10.18637/jss.v077.i01}

\leavevmode\vadjust pre{\hypertarget{ref-young_1976}{}}%
Young, F.W., De Leeuw, J., Takane, Y., 1976. Regression with qualitative and quantitative variables: An alternating least squares method with optimal scaling features. Psychometrika 41, 505--529. doi:\url{https://doi.org/10.1007/BF02296972}

\end{CSLReferences}

\bibliographystyle{spbasic}

\end{document}

### 2022-02-23 publication/manuscript.Rmd
fix typos and references

@@
 #bibstyle: spphys
-csl: elsevier-harvard.csl
+#csl: svjour3.cls
 
@@
 The encoded value for each level is based on the spherical conditional mode estimates.
-In regression, the conditional modes are similar to the mean target value for each level, weighted by the relative observed frequency of that level in the training set [@gelman_2006, @bates_2018].
+In regression, the conditional modes are similar to the mean target value for each level, weighted by the relative observed frequency of that level in the training set [@gelman_2006; @bates_2018].
 The estimate of the fixed intercept can be used during the prediction phase to encode new feature levels not observed in the training set.
@@
 In the remove control condition, features with more than HCT levels in the training set are removed from the feature set.\footnote{Based on the number of categorical features and levels per feature, some HCT settings were removed from the benchmark for some combinations of dataset $\times$ encoder.
-This ensured that encoders always affect at least one feature and the remove condition always removes at least one feature.
-If settings where aan encoder would lead to identical encoding strategies for all features of a dataset, we only kept the condition with the smallest HCT value.}
+This ensured that encoders always affect at least one feature and that the remove condition always removes at least one feature.
+If settings where an encoder would lead to identical encoding strategies for all features of a dataset, we only kept the condition with the smallest HCT value.}
 
@@
 For some datasets, the remove condition performed very similar to the other encodings (e.g. *ames-housing*, *porto-seguro*), suggesting that categorical features were not informative.
-Performance in some CV folds was below the FL learner for *flight-delay-usa-dec-2017* ($RMSE_{FL} =$ `r res[lrn.id == "FL" & Name == "flight-delay-usa-dec-2017", rmse.test.mean]`) and *nyc-taxi-green-dec-2016* ($RMSE_{FL} =$ `r res[lrn.id == "FL" & Name == "nyc-taxi-green-dec-2016", rmse.test.mean]`).
-<!-- den satz können wir nicht rauslassen, sonst gibt es zum FL keine Information im ganzen Paper -->
+Performance in some CV folds was below the FL learner for *flight-delay-usa-dec-2017* ($RMSE_{FL} =$ `r round(res[lrn.id == "FL" & Name == "flight-delay-usa-dec-2017", rmse.test.mean], 2)`) and *nyc-taxi-green-dec-2016* ($RMSE_{FL} =$ `r round(res[lrn.id == "FL" & Name == "nyc-taxi-green-dec-2016", rmse.test.mean], 2)`).
 
@@
 For a small number of levels, this approach has been reported superior to indicator and integer encoding [@wright_2019], while we observed poor performance for datasets with a larger number of levels.
+
 When looking for a simple default encoding, indicator encoding in combination with collapsing small levels seems a robust alternative, although the glmm encoders performed better.

### 2022-03-04 publication/manuscript.tex
update arxiv preprint

@@
         Florian Pargent \orcidlink{0000-0002-2388-553X} \at
-     Department of Psychology, Psychological Methods and Assessment, LMU Munich, Leopoldstra{\ss}e 13, 80802 Munich, Germany \\
+     Department of Psychology, Psychological Methods and Assessment, LMU Munich, Leopoldstraße 13, 80802 Munich, Germany \\
      \email{\href{mailto:florian.pargent@psy.lmu.de}{\nolinkurl{florian.pargent@psy.lmu.de}}}  %  \\
@@
         Florian Pfisterer \orcidlink{0000-0001-8867-762X} \at
-     Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstra{\ss}e 33, 80539 Munich, Germany \\
+     Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstraße 33, 80539 Munich, Germany \\
      %  \\
@@
         Janek Thomas \orcidlink{0000-0003-4511-6245} \at
-     Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstra{\ss}e 33, 80539 Munich, Germany \\
+     Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstraße 33, 80539 Munich, Germany \\
      %  \\
@@
         Bernd Bischl \orcidlink{0000-0001-6002-6980} \at
-     Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstra{\ss}e 33, 80539 Munich, Germany \\
+     Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstraße 33, 80539 Munich, Germany \\
      %  \\
@@
 
-\date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
+\date{\textbf{This preprint has since been published in Computational Statistics:} \newline Pargent, F., Pfisterer, F., Thomas, J., \& Bischl, B.
+Regularized target encoding outperforms traditional methods in supervised machine learning with high cardinality features.
+\textit{Computational Statistics} (2022).
+https://doi.org/10.1007/s00180-022-01207-6 \newline \textbf{Link to copy of record:} https://link.springer.com/article/10.1007/s00180-022-01207-6}
 
@@
 Despite this inefficiency, simpler strategies are often favored in practice because other methods are either not known, implementations are missing or because of a lack of trust due to missing validation studies.
-Although domain knowledge can sometimes be used to reduce the number of theoretically relevant levels, finding strategies that work well on a large variety of problems is highly important for many applications as well as in automated ML (Feurer et al., 2015; Thomas et al., 2018; Thornton et al., 2013).
+Although domain knowledge can sometimes be used to reduce the number of theoretically relevant levels, finding strategies that work well on a large variety of problems is highly important for many applications as well as in automated ML (Thomas, Coors, and Bischl 2018; Thornton et al.
+2013; Feurer et al.
+2015).
 Optimally, strategies should be model-agnostic because benchmarking encoding methods together with ML algorithms from different classes is often necessary for applications.
 While a variety of strategies exist, there are very few benchmarks that can be used to decide which technique is expected to yield good predictive performance.
-Furthermore, there has recently been increasing attention on scientific benchmark studies that compare different methods to provide a clearer picture in light of a large number of methods available to practitioners (Bommert et al., 2020; Fernández-Delgado et al., 2014), as they can provide at least partial answers to such questions.
+Furthermore, there has recently been increasing attention on scientific benchmark studies that compare different methods to provide a clearer picture in light of a large number of methods available to practitioners (Fernández-Delgado et al.
+2014; Bommert et al.
+2020), as they can provide at least partial answers to such questions.
 The goal of this study is to provide an overview of existing approaches for encoding categorical predictor variables and to study their effect on a model's predictive performance.
-Following calls in the computational statistics community for neutral benchmark studies (Boulesteix et al., 2017), which do not introduce a new method, thus reducing the risk of cherry picking methods (Dehghani et al., 2021) and reporting over-optimistic performance (Nie{\ss}l et al., 2021), we present a carefully designed experimental setting to discern the effect of encoding strategies and their interaction with different ML algorithms.
+Following calls in the computational statistics community for neutral benchmark studies (Boulesteix et al.
+2017), which do not introduce a new method, thus reducing the risk of cherry picking methods (Dehghani et al.
+2021) and reporting over-optimistic performance (Nießl et al., n.d.), we present a carefully designed experimental setting to discern the effect of encoding strategies and their interaction with different ML algorithms.
 
@@
 
-We broadly categorize feature encoding techniques into \emph{target-agnostic} methods and \emph{target-based} methods (Micci-Barreca, 2001).
+We broadly categorize feature encoding techniques into \emph{target-agnostic} methods and \emph{target-based} methods (Micci-Barreca 2001).
 Figure \ref{fig:overview} contains a taxonomy of our considered encoding methods.
 \emph{Target-agnostic} methods do not rely on any information about the target variable and can therefore also be used in unsupervised settings.
-Simple strategies from this domain e.g.~one-hot or dummy encoding are widely used -- in the scientific literature (Hancock and Khoshgoftaar, 2020; Kuhn and Johnson, 2019) but also on Kaggle\footnote{\url{https://www.kaggle.com}} to embed variables for classical ML algorithms as well as (deep) neural networks.
+Simple strategies from this domain e.g.~one-hot or dummy encoding are widely used -- in the scientific literature (Kuhn and Johnson 2019; Hancock and Khoshgoftaar 2020) but also on Kaggle\footnote{\url{https://www.kaggle.com}} to embed variables for classical ML algorithms as well as (deep) neural networks.
 Such indicator methods map each level of a categorical variable to a set of dichotomous features encoding the presence or absence of a particular level.
@@
 When indicator encoding leads to an unreasonable number of features, levels are often mapped to integer values with random order (integer encoding).
-Alternatively, the ``hashing trick'' (Weinberger et al., 2009) can be used to randomly collapse feature levels into a smaller number of indicator variables (Kuhn and Johnson, 2019), or levels can be encoded by using the observed frequency of a given level in the dataset (frequency encoding).
+Alternatively, the ``hashing trick'' (Weinberger et al.
+2009) can be used to randomly collapse feature levels into a smaller number of indicator variables (Kuhn and Johnson 2019), or levels can be encoded by using the observed frequency of a given level in the dataset (frequency encoding).
 
@@
 \emph{Target-based} methods try to incorporate information about the target values associated with a given level.
-Early strategies aimed to reduce the number of levels by methods like hierarchical clustering or decision trees based on statistics of the target variable, although this has been rarely described in the scientific literature (for a brief mention, see Micci-Barreca, 2001).
+Early strategies aimed to reduce the number of levels by methods like hierarchical clustering or decision trees based on statistics of the target variable, although this has been rarely described in the scientific literature (for a brief mention, see Micci-Barreca 2001).
 The basic idea of more advanced methods called target, impact, mean, or likelihood encoding is to use the training set to make a simple prediction of the target for each level of the categorical feature, and to use the prediction as the numerical feature value \(\hat{x}_l\) for the respective level.
@@
 ML models would place a high priority on such an encoded feature during training but would perform badly on test data.
-To avoid this, practitioners often use regularized target encoding with a smoothing parameter that shrinks those effects towards the global mean (Micci-Barreca, 2001).
-An alternative strategy is to combine target encoding with cross-validation (CV) techniques (Prokhorenkova et al., 2018).
+To avoid this, practitioners often use regularized target encoding with a smoothing parameter that shrinks those effects towards the global mean (Micci-Barreca 2001).
+An alternative strategy is to combine target encoding with cross-validation (CV) techniques (Prokhorenkova et al. 2018).
 
@@
 Those works did not yield conclusive results due to narrower scopes or not considering high cardinality variables.
-One benchmark (\(6\) datasets) on encoding high cardinality features (cardinality between \(103\) and \(9095\)) in combination with gradient boosting has been published on the Kaggle forums (Prokopev, 2018).
+One benchmark (\(6\) datasets) on encoding high cardinality features (cardinality between \(103\) and \(9095\)) in combination with gradient boosting has been published on the Kaggle forums (Prokopev 2018).
 Different versions of target encoding are compared with indicator, integer, and frequency encoding.
@@
 Interestingly, frequency encoding did perform well in many cases.
-Coors (2018) performed a benchmark (\(12\) datasets) on encoding high cardinality features (maximum number of levels per dataset between 10 and 25847) when developing the automatic gradient boosting (autoxgboost) library (Thomas et al., 2018).
+Coors (2018) performed a benchmark (\(12\) datasets) on encoding high cardinality features (maximum number of levels per dataset between 10 and 25847) when developing the automatic gradient boosting (autoxgboost) library (Thomas, Coors, and Bischl 2018).
 They compared different variants of target encoding with integer and indicator encoding.
@@
 As those studies only consider gradient boosting and a limited amount of datasets, it is unclear whether results generalize to other datasets and ML algorithms.
-A recent study (\(15\) datasets) found good performance of target encoding, but they only investigated categorical features in regression settings (Seca and Mendes-Moreira, 2021).
-Several other publications studied encoding text data based on similarity (Cerda et al., 2018; Cerda and Varoquaux, 2020) and employed indicator or target encoding as baselines.
+A recent study (\(15\) datasets) found good performance of target encoding, but they only investigated categorical features in regression settings (Seca and Mendes-Moreira 2021).
+Several other publications studied encoding text data based on similarity (Patricio Cerda, Varoquaux, and Kégl 2018; P.
+Cerda and Varoquaux 2020) and employed indicator or target encoding as baselines.
 Another line of work studies variable encodings employed within specific ML models.
@@
 While the CatBoost method performed best, hold-out came second and target encoding without CV performed worst.
-An overview of encoding techniques tailored towards neural networks (e.g.~the widely adopted \emph{entity embeddings} by Guo and Berkhahn, 2016) is provided in Hancock and Khoshgoftaar (2020).
+An overview of encoding techniques tailored towards neural networks (e.g.~the widely adopted \emph{entity embeddings} by Guo and Berkhahn 2016) is provided in Hancock and Khoshgoftaar (2020).
 They present a survey of indicator- and embedding-based methods but no benchmark study.
@@
 
-Hash Encoding can be used to compute indicator variables based on a hash function (Weinberger et al., 2009).
+Hash Encoding can be used to compute indicator variables based on a hash function (Weinberger et al. 2009).
 The basic idea is to transform each feature level \(l\) into an integer \(hash(l) \in \mathbb{N}\), based on its label.
-This integer is then transformed into an indicator representation, with \(1\) in indicator column number \((hash(l) \mod hash.size) + 1\) and \(0\) in all remaining columns (Kuhn and Johnson, 2019).
+This integer is then transformed into an indicator representation, with \(1\) in indicator column number \((hash(l) \mod hash.size) + 1\) and \(0\) in all remaining columns (Kuhn and Johnson 2019).
 Some levels are hashed to the same indicator representation.
@@
 In that way, leaf encoding combines feature levels with similar target values.
-We use the rpart package in R (Therneau and Atkinson, 2018) which grows CARTs with categorical feature support that can be pruned based on internal performance estimates from 10-fold CV.
+We use the rpart package in R (Therneau and Atkinson 2018) which grows CARTs with categorical feature support that can be pruned based on internal performance estimates from 10-fold CV.
 Thus, our leaf encoder automatically uses an ``optimal'' number of new levels.
@@
 We choose a small \(\epsilon = 0.0001\) as we want to compare simple target encoding with the regularized encoder introduced next.
-Weights of evidence encoding from the credit scoring classification literature (Hand and Henley, 1997) is almost identical to impact encoding, but without regularization or centering.
+Weights of evidence encoding from the credit scoring classification literature (Hand and Henley 1997) is almost identical to impact encoding, but without regularization or centering.
 
@@
 
-Smoothed target encoding (Micci-Barreca, 2001) can be interpreted as a simple (generalized) linear mixed model (glmm) in which the target is predicted by a random intercept for each feature level in addition to a fixed global intercept.
+Smoothed target encoding (Micci-Barreca 2001) can be interpreted as a simple (generalized) linear mixed model (glmm) in which the target is predicted by a random intercept for each feature level in addition to a fixed global intercept.
 This connection is described in Kuhn and Johnson (2019).
@@
 The encoded value for each level is based on the spherical conditional mode estimates.
-In regression, the conditional modes are similar to the mean target value for each level, weighted by the relative observed frequency of that level in the training set Bates (2020).
+In regression, the conditional modes are similar to the mean target value for each level, weighted by the relative observed frequency of that level in the training set (Gelman and Hill 2006; Bates 2020).
 The estimate of the fixed intercept can be used during the prediction phase to encode new feature levels not observed in the training set.
@@
 We study this method in three different settings: without CV (noCV), with \(5-\) (5CV) and with \(10-\) fold CV (10CV).
-In our study, we use the lmer (regression) and glmer (classification) functions from the lme4 package in R (Bates et al., 2015) as an efficient way to fit glmms.
+In our study, we use the lmer (regression) and glmer (classification) functions from the lme4 package in R (Bates et al. 2015) as an efficient way to fit glmms.
 
@@
 For each dataset, we also consider a RF without encoding (\textbf{none} condition), to compare the use of encoding methods with a natural categorical splitting approach.
-The ranger (Wright and Ziegler, 2017) implementation provides efficient categorical feature support by ordering levels once before starting the tree growing algorithm (Wright and König, 2019).
+The ranger (Wright and Ziegler 2017) implementation provides efficient categorical feature support by ordering levels once before starting the tree growing algorithm (Wright and König 2019).
 In the \textbf{remove} high cardinality features control condition, we omit features with a high number of levels above some threshold and use one-hot encoding (without collapsing rare levels) for the remaining features.
@@
 We specifically investigate datasets that contain categorical variables with a large number of levels, including many well-known datasets
-from previous studies (Cerda et al., 2018; Coors, 2018; Kuhn and Johnson, 2019; Prokhorenkova et al., 2018).
-All datasets can be downloaded from the \texttt{OpenML} platform (Vanschoren et al., 2013) based on the name or the displayed OmlId.
+from previous studies (Patricio Cerda, Varoquaux, and Kégl 2018; Prokhorenkova et al. 2018; Coors 2018; Kuhn and Johnson 2019).
+All datasets can be downloaded from the \texttt{OpenML} platform (Vanschoren et al. 2013) based on the name or the displayed OmlId.
 The datasets include 8 regression, 10 binary classification, and 6 multiclass classification problems (between 3 and 12 classes).
@@
 We kept tuning their hyperparameters to a minimum because we were interested in the effect of the encoding techniques instead of a comparison of ML algorithms.
-LASSOs were fitted with \texttt{glmnet} (Friedman et al., 2010), internally tuning the regularization using 5-fold CV.
-RFs with \(500\) trees were trained using \texttt{ranger} (Wright and Ziegler, 2017) without tuning, because RFs can be expected to give reasonable results with default settings (Probst et al., 2019).
-GB models were trained using \texttt{xgboost} (Chen et al., 2018), setting the learning rate to \(0.01\) and determining the number of iterations using early stopping on a 20\% holdout set.
-KNN was taken from package \texttt{kknn} (Schliep and Hechenbichler, 2016) standardizing features and using a constant \(k=15\) for the number of nearest neighbors, together with an information gain filter (Brown et al., 2012) to limit the number of features to \(25\).
-SVMs with radial basis function kernel were trained with \texttt{liquidSVM} (Steinwart and Thomann, 2017).
+LASSOs were fitted with \texttt{glmnet} (Friedman, Hastie, and Tibshirani 2010), internally tuning the regularization using 5-fold CV.
+RFs with \(500\) trees were trained using \texttt{ranger} (Wright and Ziegler 2017) without tuning, because RFs can be expected to give reasonable results with default settings (Probst, Wright, and Boulesteix 2019).
+GB models were trained using \texttt{xgboost} (Chen et al. 2018), setting the learning rate to \(0.01\) and determining the number of iterations using early stopping on a 20\% holdout set.
+KNN was taken from package \texttt{kknn} (Schliep and Hechenbichler 2016) standardizing features and using a constant \(k=15\) for the number of nearest neighbors, together with an information gain filter (Brown et al.
+2012) to limit the number of features to \(25\).
+SVMs with radial basis function kernel were trained with \texttt{liquidSVM} (Steinwart and Thomann 2017).
 The bandwidth and regularization parameters were internally tuned using 5-fold CV. We used a one-vs-all approach for multiclass-settings.
@@
 
-We perform all analyses in the open-source statistical software R (R Core Team, 2021).
-To enable a fair and reliable comparison in our study, we implemented all encoding methods on top of the \texttt{mlrCPO} package (Binder, 2018).
-The pre-processing, as well as the final ML algorithm described in \ref{sec:pipeline}, were trained and resampled using the \texttt{mlr} framework (Bischl et al., 2016) together with the \texttt{batchtools} package (Lang et al., 2017) to scale the benchmark analysis to HPC compute infrastructure.
+We perform all analyses in the open-source statistical software R (R Core Team 2021).
+To enable a fair and reliable comparison in our study, we implemented all encoding methods on top of the \texttt{mlrCPO} package (Binder 2018).
+The pre-processing, as well as the final ML algorithm described in \ref{sec:pipeline}, were trained and resampled using the \texttt{mlr} framework (Bischl et al.
+2016) together with the \texttt{batchtools} package (Lang, Bischl, and Surmann 2017) to scale the benchmark analysis to HPC compute infrastructure.
 All materials for this study, including reproducible code for this manuscript and result objects can be downloaded from our \textbf{online repository} \footnote{\url{https://github.com/slds-lmu/paper_2021_categorical_feature_encodings}}.
@@
 Throughout our experiments, we use 5-fold CV to obtain estimates of predictive performance.
-Depending on the target variable, we report root mean squared error (RMSE) for regression, area under the curve (AUC) for binary classification, and its extension AUNU (Ferri et al., 2009) for multiclass problems.
+Depending on the target variable, we report root mean squared error (RMSE) for regression, area under the curve (AUC) for binary classification, and its extension AUNU (Ferri, Hernández-Orallo, and Modroiu 2009) for multiclass problems.
 
@@
 For some datasets, the remove condition performed very similar to the other encodings (e.g.~\emph{ames-housing}, \emph{porto-seguro}), suggesting that categorical features were not informative.
-Performance in some CV folds was below the FL learner for \emph{flight-delay-usa-dec-2017} (\(RMSE_{FL} =\) 48.8052624) and \emph{nyc-taxi-green-dec-2016} (\(RMSE_{FL} =\) 2.2152442).
+Performance in some CV folds was below the FL learner for \emph{flight-delay-usa-dec-2017} (\(RMSE_{FL} =\) 48.81) and \emph{nyc-taxi-green-dec-2016} (\(RMSE_{FL} =\) 2.22).
 
@@
 
-To further analyze our results, we used statistical inference methods inspired by the benchmark community in computational statistics (e.g., Hothorn et al., 2005).
+To further analyze our results, we used statistical inference methods inspired by the benchmark community in computational statistics (e.g., Hothorn et al. 2005).
 First, we present meta rankings for each ML algorithm in Figure \ref{fig:rank-res}:
-We defined an encoder relation within each dataset, based on corrected resample t-tests (Nadeau and Bengio, 2003).
+We defined an encoder relation within each dataset, based on corrected resample t-tests (Nadeau and Bengio 2003).
 An encoding was defined to beat another encoding if the one-sided p-value of the t-test was \(< .05\).
@@
 
-where \(d\) is the symmetric difference distance and \(R_b\) is the relation for dataset \(b\) (Hornik and Meyer, 2007; Meyer and Hornik, 2018).
+where \(d\) is the symmetric difference distance and \(R_b\) is the relation for dataset \(b\) (Hornik and Meyer 2007; Meyer and Hornik 2018).
 The symmetric difference between two relations is the number of cases one encoding beats another encoding in one relation but not in the other one.
@@
 Performance often did not improve with glmm-10CV, suggesting that 5 folds might be a good regularization default in practice.
-In line with earlier research (Micci-Barreca, 2001; Prokhorenkova et al., 2018), target encoding with regularization (glmm) performed better or equally well in comparison with the unregularized impact encoder.
+In line with earlier research (Micci-Barreca 2001; Prokhorenkova et al.
+2018), target encoding with regularization (glmm) performed better or equally well in comparison with the unregularized impact encoder.
 When glmm performed better, impact encoding not only performed worse than glmm with CV but was sometimes also inferior compared to other encoders.
@@
 Surprisingly, integer encoding did not perform well with GB and target-based encoders (especially the glmm encoder) seem to be preferable.
-For LASSO, previous studies have suggested that indicator encoding works well, even with a very high number of levels (Cerda et al., 2018).
+For LASSO, previous studies have suggested that indicator encoding works well, even with a very high number of levels (Patricio Cerda, Varoquaux, and Kégl 2018).
 Although the glmm encoders ranked first in our benchmark for the LASSO, it was the only algorithm where the indicator encoders achieved the next best ranking.
@@
 In RF, a widely used strategy to deal with categorical features is to order levels by average target statistics for a given level.
-For a small number of levels, this approach has been reported superior to indicator and integer encoding (Wright and König, 2019), while we observed poor performance for datasets with a larger number of levels.
+For a small number of levels, this approach has been reported superior to indicator and integer encoding (Wright and König 2019), while we observed poor performance for datasets with a larger number of levels.
+
 When looking for a simple default encoding, indicator encoding in combination with collapsing small levels seems a robust alternative, although the glmm encoders performed better.
-We found that one-hot encoding usually gave a slightly better performance than dummy encoding, which has also been observed by Chiquet et al. (2016) and Tutz and Gertheiss (2016) (p.254).
+We found that one-hot encoding usually gave a slightly better performance than dummy encoding, which has also been observed by Chiquet, Grandvalet, and Rigaill (2016) and Tutz and Gertheiss (2016) (p.254).
 Our results suggest that one-hot encoding is the better standard compared to dummy encoding when applying nonlinear regularized models like RF, GB or SVM with (high cardinal) categorical features.
@@
 It sometimes leads to slightly longer runtimes (in comparison to indicator encoding), but especially for larger datasets this is often offset by the more efficient representation produced by target encoding.
-The glmm encoder has a clear advantage over target encoding with a smoothing hyperparameter (Micci-Barreca, 2001), as costly tuning the whole ML pipeline with different smoothing values is not required.
+The glmm encoder has a clear advantage over target encoding with a smoothing hyperparameter (Micci-Barreca 2001), as costly tuning the whole ML pipeline with different smoothing values is not required.
 
@@
 We only used 5-fold-CV without repetitions to estimate predictive performance, but due to a small variance between folds, we could confidently detect performance differences of encoders for many datasets.
-Because we are interested in model-agnostic methods that can be combined with any supervised ML algorithm, we did not investigate recent model-specific strategies (Guo and Berkhahn, 2016; Prokhorenkova et al., 2018).
+Because we are interested in model-agnostic methods that can be combined with any supervised ML algorithm, we did not investigate recent model-specific strategies (Guo and Berkhahn 2016; Prokhorenkova et al.
+2018).
 We did not include deep neural networks in our benchmark because they are still rarely applied in the statistical learning community and more dominant in computer science.
@@
 For many applications, ordinal feature information is not available as metadata which makes it less relevant for some applications like automatic ML.
-Thus, our benchmark does not include specific ordinal encoding methods like Helmert and Polynomial contrasts (Chambers and Hastie, 1992) or ordinal penalization methods (Tutz and Gertheiss, 2016).
-We only investigate traditional categorical variables and do not extend our analysis to multi-categorical variables or text-strings (Cerda et al., 2018; Cerda and Varoquaux, 2020), where other encoding techniques can harvest additional information.
+Thus, our benchmark does not include specific ordinal encoding methods like Helmert and Polynomial contrasts (Chambers and Hastie 1992) or ordinal penalization methods (Tutz and Gertheiss 2016).
+We only investigate traditional categorical variables and do not extend our analysis to multi-categorical variables or text-strings (Patricio Cerda, Varoquaux, and Kégl 2018; P.
+Cerda and Varoquaux 2020), where other encoding techniques can harvest additional information.
 We only investigate univariate encodings, i.e.~we always encode each variable separately.
@@
 Methods that jointly encode features and therefore leverage correlational structures between features are an interesting avenue for future research:
-Recent approaches use target-based encoding strategies to learn a vector-valued representation of a level, similar to neural network embeddings (Guo and Berkhahn, 2016; Rodríguez et al., 2018).
+Recent approaches use target-based encoding strategies to learn a vector-valued representation of a level, similar to neural network embeddings (Guo and Berkhahn 2016; Rodríguez et al. 2018).
 Those should be compared to classical approaches from the optimal scaling literature (e.g.
-De Leeuw et al., 1976; Young et al., 1976) or the subsequent \emph{aspect} framework (Mair and Leeuw, 2010), which are unfortunately unfamiliar to many ML researchers.
+De Leeuw, Young, and Takane 1976; Young, De Leeuw, and Takane 1976) or the subsequent \emph{aspect} framework (Mair and Leeuw 2010), which are unfortunately unfamiliar to many ML researchers.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bates_2018}{}}%
-Bates, D., 2020. Computational methods for mixed models. Vignette for lme4.
+Bates, Douglas. 2020. {``Computational Methods for Mixed Models.''} \emph{Vignette for Lme4}. \url{https://cran.r-project.org/web/packages/lme4/vignettes/Theory.pdf}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-bates_2015}{}}%
-Bates, D., Mächler, M., Bolker, B., Walker, S., 2015.
-Fitting linear mixed-effects models using {lme4}.
-Journal of Statistical Software 67, 1--48.
-doi:\href{https://doi.org/10.18637/jss.v067.i01}{10.18637/jss.v067.i01}
+Bates, Douglas, Martin Mächler, Ben Bolker, and Steve Walker.
+2015.
+{``Fitting Linear Mixed-Effects Models Using {lme4}.''} \emph{Journal of Statistical Software} 67 (1): 1--48.
+\url{https://doi.org/10.18637/jss.v067.i01}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-binder_2018}{}}%
-Binder, M., 2018. mlrCPO: Composable preprocessing operators and pipelines for machine learning.
+Binder, Martin. 2018. \emph{mlrCPO: Composable Preprocessing Operators and Pipelines for Machine Learning}. \url{https://github.com/mlr-org/mlrCPO}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-bischl_2016}{}}%
-Bischl, B., Lang, M., Kotthoff, L., Schiffner, J., Richter, J., Studerus, E., Casalicchio, G., Jones, Z.M., 2016. {mlr}: Machine learning in r. Journal of Machine Learning Research 17, 1--5.
+Bischl, Bernd, Michel Lang, Lars Kotthoff, Julia Schiffner, Jakob Richter, Erich Studerus, Giuseppe Casalicchio, and Zachary M.
+Jones.
+2016.
+{``{mlr}: Machine Learning in r.''} \emph{Journal of Machine Learning Research} 17 (170): 1--5.
+\url{http://jmlr.org/papers/v17/15-066.html}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-bommert}{}}%
-Bommert, A., Sun, X., Bischl, B., Rahnenführer, J., Lang, M., 2020.
-Benchmark for filter methods for feature selection in high-dimensional classification data.
-Computational Statistics \& Data Analysis 143.
-doi:\href{https://doi.org/10.1016/j.csda.2019.106839}{10.1016/j.csda.2019.106839}
+Bommert, Andrea, Xudong Sun, Bernd Bischl, Jörg Rahnenführer, and Michel Lang.
+2020.
+{``Benchmark for Filter Methods for Feature Selection in High-Dimensional Classification Data.''} \emph{Computational Statistics \& Data Analysis} 143 (C).
+\url{https://doi.org/10.1016/j.csda.2019.106839}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-boulesteix_2017}{}}%
-Boulesteix, A.-L., Binder, H., Abrahamowicz, M., Sauerbrei, W., others, 2017.
-On the necessity and design of studies comparing statistical methods.
-Biometrical journal.
-Biometrische Zeitschrift 60, 216--218.
-doi:\href{https://doi.org/10.1002/bimj.201700129}{10.1002/bimj.201700129}
+Boulesteix, Anne-Laure, Harald Binder, Michal Abrahamowicz, Willi Sauerbrei, et al.
+2017.
+{``On the Necessity and Design of Studies Comparing Statistical Methods.''} \emph{Biometrical Journal.
+Biometrische Zeitschrift} 60 (1): 216--18.
+\url{https://doi.org/10.1002/bimj.201700129}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-brown2012}{}}%
-Brown, G., Pocock, A., Ming-Jie, Z., Luján, M., 2012.
-Conditional likelihood maximisation: A unifying framework for information theoretic feature selection.
-Journal of Machine Learning Research 13, 27--66.
+Brown, Gavin, Adam Pocock, Zhao Ming-Jie, and Mikel Luján.
+2012.
+{``Conditional Likelihood Maximisation: A Unifying Framework for Information Theoretic Feature Selection.''} \emph{Journal of Machine Learning Research} 13 (January): 27--66.
+
+\leavevmode\vadjust pre{\hypertarget{ref-cerda_2018}{}}%
+Cerda, Patricio, Gaël Varoquaux, and Balázs Kégl.
+2018.
+{``Similarity Encoding for Learning with Dirty Categorical Variables.''} \emph{Machine Learning} 107 (8): 1477--94.
+\url{https://doi.org/10.1007/s10994-018-5724-2}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-cerda_2020}{}}%
-Cerda, P., Varoquaux, G., 2020.
-Encoding high-cardinality string categorical variables.
-IEEE Transactions on Knowledge and Data Engineering 1--1.
-doi:\href{https://doi.org/10.1109/TKDE.2020.2992529}{10.1109/TKDE.2020.2992529}
-
-\leavevmode\vadjust pre{\hypertarget{ref-cerda_2018}{}}%
-Cerda, P., Varoquaux, G., Kégl, B., 2018.
-Similarity encoding for learning with dirty categorical variables.
-Machine Learning 107, 1477--1494.
-doi:\href{https://doi.org/10.1007/s10994-018-5724-2}{10.1007/s10994-018-5724-2}
+Cerda, P., and G.
+Varoquaux.
+2020.
+{``Encoding High-Cardinality String Categorical Variables.''} \emph{IEEE Transactions on Knowledge and Data Engineering}, 1--1.
+\url{https://doi.org/10.1109/TKDE.2020.2992529}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-chambers_1992}{}}%
-Chambers, J., Hastie, T., 1992. Statistical models. Chapter 2 of statistical models in s. Wadsworth \& Brooks/Cole.
+Chambers, JM, and TJ Hastie. 1992. {``Statistical Models. Chapter 2 of Statistical Models in s (1st Ed.).''} \emph{Routledge}. https://doi.org/\url{https://doi.org/10.1201/9780203738535}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-chen_2018}{}}%
-Chen, T., He, T., Benesty, M., Khotilovich, V., Tang, Y., Cho, H., Chen, K., Mitchell, R., Cano, I., Zhou, T., Li, M., Xie, J., Lin, M., Geng, Y., Li, Y., 2018. Xgboost: Extreme gradient boosting.
+Chen, Tianqi, Tong He, Michael Benesty, Vadim Khotilovich, Yuan Tang, Hyunsu Cho, Kailong Chen, et al.
+2018.
+\emph{Xgboost: Extreme Gradient Boosting}.
+\url{https://CRAN.R-project.org/package=xgboost}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-chiquet_2016}{}}%
-Chiquet, J., Grandvalet, Y., Rigaill, G., 2016.
-On coding effects in regularized categorical regression.
-Statistical Modelling 16, 228--237.
-doi:\href{https://doi.org/10.1177/1471082X16644998}{10.1177/1471082X16644998}
+Chiquet, Julien, Yves Grandvalet, and Guillem Rigaill.
+2016.
+{``On Coding Effects in Regularized Categorical Regression.''} \emph{Statistical Modelling} 16 (3): 228--37.
+\url{https://doi.org/10.1177/1471082X16644998}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-coors_2018}{}}%
-Coors, S., 2018. Automatic gradient boosting (Master's thesis). LMU Munich.
+Coors, Stefan. 2018. {``Automatic Gradient Boosting.''} Master's thesis, LMU Munich. \url{https://epub.ub.uni-muenchen.de/59108/1/MA_Coors.pdf}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-deleeuw_1976}{}}%
-De Leeuw, J., Young, F.W., Takane, Y., 1976. Additive structure in qualitative data: An alternating least squares method with optimal scaling features. Psychometrika 41, 471--503.
+De Leeuw, Jan, Forrest W Young, and Yoshio Takane.
+1976.
+{``Additive Structure in Qualitative Data: An Alternating Least Squares Method with Optimal Scaling Features.''} \emph{Psychometrika} 41 (4): 471--503.
 
 \leavevmode\vadjust pre{\hypertarget{ref-dehghani2021benchmark}{}}%
-Dehghani, M., Tay, Y., Gritsenko, A.A., Zhao, Z., Houlsby, N., Diaz, F., Metzler, D., Vinyals, O., 2021. The benchmark lottery.
+Dehghani, Mostafa, Yi Tay, Alexey A. Gritsenko, Zhe Zhao, Neil Houlsby, Fernando Diaz, Donald Metzler, and Oriol Vinyals. 2021. {``The Benchmark Lottery.''} \url{https://arxiv.org/abs/2107.07002}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-delgado}{}}%
-Fernández-Delgado, M., Cernadas, E., Barro, S., Amorim, D., 2014. Do we need hundreds of classifiers to solve real world classification problems? Journal of Machine Learning Research 15, 3133--3181.
+Fernández-Delgado, Manuel, Eva Cernadas, Senén Barro, and Dinani Amorim.
+2014.
+{``Do We Need Hundreds of Classifiers to Solve Real World Classification Problems?''} \emph{Journal of Machine Learning Research} 15 (90): 3133--81.
+\url{http://jmlr.org/papers/v15/delgado14a.html}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-ferri_2009}{}}%
-Ferri, C., Hernández-Orallo, J., Modroiu, R., 2009.
-An experimental comparison of performance measures for classification.
-Pattern Recognition Letters 30, 27--38.
-doi:\href{https://doi.org/10.1016/j.patrec.2008.08.010}{10.1016/j.patrec.2008.08.010}
+Ferri, C., J.
+Hernández-Orallo, and R.
+Modroiu.
+2009.
+{``An Experimental Comparison of Performance Measures for Classification.''} \emph{Pattern Recognition Letters} 30 (1): 27--38.
+\url{https://doi.org/10.1016/j.patrec.2008.08.010}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-feurer_2015}{}}%
-Feurer, M., Klein, A., Eggensperger, K., Springenberg, J., Blum, M., Hutter, F., 2015.
-Efficient and robust automated machine learning, in: Cortes, C., Lawrence, N.D., Lee, D.D., Sugiyama, M., Garnett, R.
-(Eds.), Advances in Neural Information Processing Systems 28.
-Curran Associates, Inc., pp.
-2962--2970.
+Feurer, Matthias, Aaron Klein, Katharina Eggensperger, Jost Springenberg, Manuel Blum, and Frank Hutter.
+2015.
+{``Efficient and Robust Automated Machine Learning.''} In \emph{Advances in Neural Information Processing Systems 28}, edited by C.
+Cortes, N.
+D.
+Lawrence, D.
+D.
+Lee, M.
+Sugiyama, and R.
+Garnett, 2962--70.
+Curran Associates, Inc.
+\url{http://papers.nips.cc/paper/5872-efficient-and-robust-automated-machine-learning.pdf}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-friedman_2010}{}}%
-Friedman, J., Hastie, T., Tibshirani, R., 2010.
-Regularization paths for generalized linear models via coordinate descent.
-Journal of Statistical Software 33, 1--22.
-doi:\href{https://doi.org/10.18637/jss.v033.i01}{10.18637/jss.v033.i01}
+Friedman, Jerome, Trevor Hastie, and Robert Tibshirani.
+2010.
+{``Regularization Paths for Generalized Linear Models via Coordinate Descent.''} \emph{Journal of Statistical Software} 33 (1): 1--22.
+\url{https://doi.org/10.18637/jss.v033.i01}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-gelman_2006}{}}%
-Gelman, A., Hill, J., 2006. Data analysis using regression and multilevel/hierarchical models. Cambridge university press.
+Gelman, Andrew, and Jennifer Hill. 2006. \emph{Data Analysis Using Regression and Multilevel/Hierarchical Models}. Cambridge university press.
 
 \leavevmode\vadjust pre{\hypertarget{ref-grabczewski_2003}{}}%
-Grąbczewski, K., Jankowski, N., 2003.
-Transformations of symbolic data for continuous data oriented models, in: Kaynak, O., Alpaydin, E., Oja, E., Xu, L.
-(Eds.), Artificial Neural Networks and Neural Information Processing --- ICANN/ICONIP 2003.
-Springer Berlin Heidelberg, Berlin, Heidelberg, pp.
-359--366.
+Grąbczewski, Krzysztof, and Norbert Jankowski.
+2003.
+{``Transformations of Symbolic Data for Continuous Data Oriented Models.''} In \emph{Artificial Neural Networks and Neural Information Processing --- ICANN/ICONIP 2003}, edited by Okyay Kaynak, Ethem Alpaydin, Erkki Oja, and Lei Xu, 359--66.
+Berlin, Heidelberg: Springer Berlin Heidelberg.
 
 \leavevmode\vadjust pre{\hypertarget{ref-guo_2016}{}}%
-Guo, C., Berkhahn, F., 2016. Entity embeddings of categorical variables. arXiv preprint arXiv:1604.06737.
+Guo, Cheng, and Felix Berkhahn. 2016. {``Entity Embeddings of Categorical Variables.''} \emph{arXiv Preprint arXiv:1604.06737}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-hancock}{}}%
-Hancock, J.T., Khoshgoftaar, T.M., 2020. Survey on categorical data for neural networks. Journal of Big Data 7, 1--41. doi:\href{https://doi.org/10.1186/s40537-020-00305-w}{10.1186/s40537-020-00305-w}
+Hancock, John T, and Taghi M Khoshgoftaar. 2020. {``Survey on Categorical Data for Neural Networks.''} \emph{Journal of Big Data} 7: 1--41. \url{https://doi.org/10.1186/s40537-020-00305-w}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-hand_1997}{}}%
-Hand, D.J., Henley, W.E., 1997.
-Statistical classification methods in consumer credit scoring: A review.
-Journal of the Royal Statistical Society: Series A (Statistics in Society) 160, 523--541.
-doi:\href{https://doi.org/10.1111/j.1467-985X.1997.00078.x}{10.1111/j.1467-985X.1997.00078.x}
+Hand, D.
+J., and W.
+E.
+Henley.
+1997.
+{``Statistical Classification Methods in Consumer Credit Scoring: A Review.''} \emph{Journal of the Royal Statistical Society: Series A (Statistics in Society)} 160 (3): 523--41.
+\url{https://doi.org/10.1111/j.1467-985X.1997.00078.x}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-hornik_2007}{}}%
-Hornik, K., Meyer, D., 2007.
-Deriving consensus rankings from benchmarking experiments, in: Advances in Data Analysis.
-Springer, pp.
-163--170.
-doi:\href{https://doi.org/10.1007/978-3-540-70981-7_19}{10.1007/978-3-540-70981-7\_19}
+Hornik, Kurt, and David Meyer.
+2007.
+{``Deriving Consensus Rankings from Benchmarking Experiments.''} In \emph{Advances in Data Analysis}, 163--70.
+Springer.
+\url{https://doi.org/10.1007/978-3-540-70981-7_19}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-hothorn_2005}{}}%
-Hothorn, T., Leisch, F., Zeileis, A., Hornik, K., 2005.
-The design and analysis of benchmark experiments.
-Journal of Computational and Graphical Statistics 14, 675--699.
-doi:\href{https://doi.org/10.1198/106186005X59630}{10.1198/106186005X59630}
+Hothorn, Torsten, Friedrich Leisch, Achim Zeileis, and Kurt Hornik.
+2005.
+{``The Design and Analysis of Benchmark Experiments.''} \emph{Journal of Computational and Graphical Statistics} 14 (3): 675--99.
+\url{https://doi.org/10.1198/106186005X59630}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-kuhn_2019}{}}%
-Kuhn, M., Johnson, K., 2019. Feature engineering and selection: A practical approach for predictive models. Chapman; Hall/CRC.
+Kuhn, Max, and Kjell Johnson. 2019. \emph{Feature Engineering and Selection: A Practical Approach for Predictive Models}. Chapman; Hall/CRC. \url{http://www.feat.engineering/index.html}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-lang_2017}{}}%
-Lang, M., Bischl, B., Surmann, D., 2017. Batchtools: Tools for r to work on batch systems. The Journal of Open Source Software 2. doi:\href{https://doi.org/10.21105/joss.00135}{10.21105/joss.00135}
+Lang, Michel, Bernd Bischl, and Dirk Surmann.
+2017.
+{``Batchtools: Tools for r to Work on Batch Systems.''} \emph{The Journal of Open Source Software} 2 (10).
+\url{https://doi.org/10.21105/joss.00135}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-mair_2010}{}}%
-Mair, P., Leeuw, J.
-de, 2010.
-A general framework for multivariate analysis with optimal scaling: The r package aspect.
-Journal of Statistical Software 32, 1--23.
-doi:\href{https://doi.org/10.18637/jss.v032.i09}{10.18637/jss.v032.i09}
+Mair, Patrick, and Jan de Leeuw.
+2010.
+{``A General Framework for Multivariate Analysis with Optimal Scaling: The r Package Aspect.''} \emph{Journal of Statistical Software} 32 (1): 1--23.
+\url{https://doi.org/10.18637/jss.v032.i09}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-meyer_2018}{}}%
-Meyer, D., Hornik, K., 2018. Relations: Data structures and algorithms for relations.
+Meyer, David, and Kurt Hornik. 2018. \emph{Relations: Data Structures and Algorithms for Relations}. \url{https://CRAN.R-project.org/package=relations}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-micci_barreca_2001}{}}%
-Micci-Barreca, D., 2001.
-A preprocessing scheme for high-cardinality categorical attributes in classification and prediction problems.
-SIGKDD Explor.
-Newsl.
-3, 27--32.
-doi:\href{https://doi.org/10.1145/507533.507538}{10.1145/507533.507538}
+Micci-Barreca, Daniele.
+2001.
+{``A Preprocessing Scheme for High-Cardinality Categorical Attributes in Classification and Prediction Problems.''} \emph{SIGKDD Explor.
+Newsl.} 3 (1): 27--32.
+\url{https://doi.org/10.1145/507533.507538}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-nadeau_2003}{}}%
-Nadeau, C., Bengio, Y., 2003. Inference for the generalization error. Machine Learning 52, 239--281. doi:\href{https://doi.org/10.1023/A:1024068626366}{10.1023/A:1024068626366}
+Nadeau, Claude, and Yoshua Bengio. 2003. {``Inference for the Generalization Error.''} \emph{Machine Learning} 52 (3): 239--81. \url{https://doi.org/10.1023/A:1024068626366}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-niesl2021overoptimism}{}}%
-Nie{\ss}l, C., Herrmann, M., Wiedemann, C., Casalicchio, G., Boulesteix, A.-L., 2021.
-Over-optimism in benchmark studies and the multiplicity of design and analysis options when interpreting their results.
+Nießl, Christina, Moritz Herrmann, Chiara Wiedemann, Giuseppe Casalicchio, and Anne-Laure Boulesteix.
+n.d.
+{``Over-Optimism in Benchmark Studies and the Multiplicity of Design and Analysis Options When Interpreting Their Results.''} \emph{WIREs Data Mining and Knowledge Discovery} n/a (n/a): e1441.
+https://doi.org/\url{https://doi.org/10.1002/widm.1441}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-probst_2019}{}}%
-Probst, P., Wright, M.N., Boulesteix, A.-L., 2019.
-Hyperparameters and tuning strategies for random forest.
-Wiley Interdisciplinary Reviews: Data Mining and Knowledge Discovery 0, e1301.
-doi:\href{https://doi.org/10.1002/widm.1301}{10.1002/widm.1301}
+Probst, Philipp, Marvin N.
+Wright, and Anne-Laure Boulesteix.
+2019.
+{``Hyperparameters and Tuning Strategies for Random Forest.''} \emph{Wiley Interdisciplinary Reviews: Data Mining and Knowledge Discovery} 0 (0): e1301.
+\url{https://doi.org/10.1002/widm.1301}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-prokhorenkova_2018}{}}%
-Prokhorenkova, L., Gusev, G., Vorobev, A., Dorogush, A.V., Gulin, A., 2018.
-CatBoost: Unbiased boosting with categorical features, in: Bengio, S., Wallach, H., Larochelle, H., Grauman, K., Cesa-Bianchi, N., Garnett, R.
-(Eds.), Advances in Neural Information Processing Systems 31.
-Curran Associates, Inc., pp.
-6638--6648.
+Prokhorenkova, Liudmila, Gleb Gusev, Aleksandr Vorobev, Anna Veronika Dorogush, and Andrey Gulin.
+2018.
+{``CatBoost: Unbiased Boosting with Categorical Features.''} In \emph{Advances in Neural Information Processing Systems 31}, edited by S.
+Bengio, H.
+Wallach, H.
+Larochelle, K.
+Grauman, N.
+Cesa-Bianchi, and R.
+Garnett, 6638--48.
+Curran Associates, Inc.
+\url{http://papers.nips.cc/paper/7898-catboost-unbiased-boosting-with-categorical-features.pdf}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-prokopev_2018}{}}%
-Prokopev, V., 2018. Mean (likelihood) encodings: A comprehensive study. Kaggle Forums.
+Prokopev, Viacheslav.
+2018.
+{``Mean (Likelihood) Encodings: A Comprehensive Study.''} \emph{Kaggle Forums}.
+Kaggle.
+\url{https://www.kaggle.com/vprokopev/mean-likelihood-encodings-a-comprehensive-study}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-r_2021}{}}%
-R Core Team, 2021. R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria.
+R Core Team. 2021. \emph{R: A Language and Environment for Statistical Computing}. Vienna, Austria: R Foundation for Statistical Computing. \url{https://www.R-project.org/}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-rodriguez_2018}{}}%
-Rodríguez, P., Bautista, M.A., Gonzàlez, J., Escalera, S., 2018.
-Beyond one-hot encoding: Lower dimensional target embedding.
-Image and Vision Computing 75, 21--31.
-doi:\url{https://doi.org/10.1016/j.imavis.2018.04.004}
+Rodríguez, Pau, Miguel A.
+Bautista, Jordi Gonzàlez, and Sergio Escalera.
+2018.
+{``Beyond One-Hot Encoding: Lower Dimensional Target Embedding.''} \emph{Image and Vision Computing} 75: 21--31.
+https://doi.org/\url{https://doi.org/10.1016/j.imavis.2018.04.004}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-schliep_2016}{}}%
-Schliep, K., Hechenbichler, K., 2016. Kknn: Weighted k-nearest neighbors.
+Schliep, Klaus, and Klaus Hechenbichler. 2016. \emph{Kknn: Weighted k-Nearest Neighbors}. \url{https://CRAN.R-project.org/package=kknn}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-seca_2021}{}}%
-Seca, D., Mendes-Moreira, J., 2021.
-Benchmark of encoders of nominal features for regression, in: Rocha, Á., Adeli, H., Dzemyda, G., Moreira, F., Ramalho Correia, A.M.
-(Eds.), Trends and Applications in Information Systems and Technologies.
-Springer International Publishing, Cham, pp.
-146--155.
+Seca, Diogo, and João Mendes-Moreira.
+2021.
+{``Benchmark of Encoders of Nominal Features for Regression.''} In \emph{Trends and Applications in Information Systems and Technologies}, edited by Álvaro Rocha, Hojjat Adeli, Gintautas Dzemyda, Fernando Moreira, and Ana Maria Ramalho Correia, 146--55.
+Cham: Springer International Publishing.
 
 \leavevmode\vadjust pre{\hypertarget{ref-steinwart_2017}{}}%
-Steinwart, I., Thomann, P., 2017. {liquidSVM}: A fast and versatile SVM package. {ArXiv e-prints 1702.06899}.
+Steinwart, Ingo, and Philipp Thomann. 2017. {``{liquidSVM}: A Fast and Versatile SVM Package.''} \emph{{ArXiv e-Prints 1702.06899}}, February. \url{http://www.isa.uni-stuttgart.de/software}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-therneau_2018}{}}%
-Therneau, T., Atkinson, B., 2018. Rpart: Recursive partitioning and regression trees.
+Therneau, Terry, and Beth Atkinson. 2018. \emph{Rpart: Recursive Partitioning and Regression Trees}. \url{https://CRAN.R-project.org/package=rpart}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-thomas_2018}{}}%
-Thomas, J., Coors, S., Bischl, B., 2018. Automatic gradient boosting.
+Thomas, Janek, Stefan Coors, and Bernd Bischl. 2018. \emph{Automatic Gradient Boosting}. \url{http://arxiv.org/abs/1807.03873v2}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-thornton_2013}{}}%
-Thornton, C., Hutter, F., Hoos, H.H., Leyton-Brown, K., 2013.
-Auto-WEKA: Combined selection and hyperparameter optimization of classification algorithms, in: Proceedings of the 19th ACM SIGKDD International Conference on Knowledge Discovery and Data Mining, KDD '13.
-ACM, New York, NY, USA, pp.
-847--855.
-doi:\href{https://doi.org/10.1145/2487575.2487629}{10.1145/2487575.2487629}
+Thornton, Chris, Frank Hutter, Holger H.
+Hoos, and Kevin Leyton-Brown.
+2013.
+{``Auto-WEKA: Combined Selection and Hyperparameter Optimization of Classification Algorithms.''} In \emph{Proceedings of the 19th ACM SIGKDD International Conference on Knowledge Discovery and Data Mining}, 847--55.
+KDD '13.
+New York, NY, USA: ACM.
+\url{https://doi.org/10.1145/2487575.2487629}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-tutz_2016}{}}%
-Tutz, G., Gertheiss, J., 2016. Rejoinder: Regularized regression for categorical data. Statistical Modelling 16, 249--260. doi:\href{https://doi.org/10.1177/1471082X16652780}{10.1177/1471082X16652780}
+Tutz, Gerhard, and Jan Gertheiss. 2016. {``Rejoinder: Regularized Regression for Categorical Data.''} \emph{Statistical Modelling} 16 (3): 249--60. \url{https://doi.org/10.1177/1471082X16652780}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-vanschoren_2013}{}}%
-Vanschoren, J., N.
-van Rijn, J., Bischl, B., Torgo, L., 2013.
-{OpenML}: Networked science in machine learning.
-SIGKDD Explorations 15, 49--60.
-doi:\href{https://doi.org/10.1145/2641190.2641198}{10.1145/2641190.2641198}
+Vanschoren, Joaquin, Jan N.
+van Rijn, Bernd Bischl, and Luis Torgo.
+2013.
+{``{OpenML}: Networked Science in Machine Learning.''} \emph{SIGKDD Explorations} 15: 49--60.
+\url{https://doi.org/10.1145/2641190.2641198}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-weinberger_2009}{}}%
-Weinberger, K.Q., Dasgupta, A., Langford, J., Smola, A.J., Attenberg, J., 2009. Feature hashing for large scale multitask learning, in: ICML.
+Weinberger, Kilian, Anirban Dasgupta, John Langford, Alex Smola, and Josh Attenberg.
+2009.
+{``Feature Hashing for Large Scale Multitask Learning.''} In \emph{Proceedings of the 26th Annual International Conference on Machine Learning}, 1113--20.
+ICML '09.
+New York, NY, USA: Association for Computing Machinery.
+\url{https://doi.org/10.1145/1553374.1553516}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-wright_2019}{}}%
-Wright, M.N., König, I.R., 2019. Splitting on categorical predictors in random forests. PeerJ 7. doi:\href{https://doi.org/10.7717/peerj.6339}{10.7717/peerj.6339}
+Wright, Marvin N., and Inke R. König. 2019. {``Splitting on Categorical Predictors in Random Forests.''} \emph{PeerJ} 7. \url{https://doi.org/10.7717/peerj.6339}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-wright_2017}{}}%
-Wright, M.N., Ziegler, A., 2017.
-{ranger}: A fast implementation of random forests for high dimensional data in {C++} and {R}.
-Journal of Statistical Software 77, 1--17.
-doi:\href{https://doi.org/10.18637/jss.v077.i01}{10.18637/jss.v077.i01}
+Wright, Marvin N., and Andreas Ziegler.
+2017.
+{``{ranger}: A Fast Implementation of Random Forests for High Dimensional Data in {C++} and {R}.''} \emph{Journal of Statistical Software} 77 (1): 1--17.
+\url{https://doi.org/10.18637/jss.v077.i01}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-young_1976}{}}%
-Young, F.W., De Leeuw, J., Takane, Y., 1976.
-Regression with qualitative and quantitative variables: An alternating least squares method with optimal scaling features.
-Psychometrika 41, 505--529.
-doi:\url{https://doi.org/10.1007/BF02296972}
+Young, Forrest W, Jan De Leeuw, and Yoshio Takane.
+1976.
+{``Regression with Qualitative and Quantitative Variables: An Alternating Least Squares Method with Optimal Scaling Features.''} \emph{Psychometrika} 41 (4): 505--29.
+https://doi.org/\url{https://doi.org/10.1007/BF02296972}.
 
@@
 \bibliographystyle{spbasic}
+\bibliography{literature.bib}
 
